---
title: "Rate per 100,000 People in State Prisons from Each Census Tract, 2020"
subtitle: "Visuals for Equity Tracker (tract data)"
author: "Kristin Mitchell"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
    word_document:
---
  
```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
#output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'a-regional-health-collaboration',
                   ind_dir = 'a03-imprisonment',
                   map = 'a03-imprisonment-map-data',
                   chart = 'a03-imprisonment-chart-data',
                   map_name = 'a03-imprisonment-map.html',
                   column_chart_name = 'a03-imprisonment-column',
                   line_chart_name = 'a03-imprisonment-line')

# set most recent year
year <- "2020"
# years <- c(2012, 2017, 2022)

# set past version year - this will set organize archived files in Y drive
past_year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time. 
\
\
**If the indicator is available through a tract-level data set.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-tract-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
The rate of incarcerated individuals in Washington State prisons per 100,000 residents by census tract, based on each incarcerated person’s last known address. Understanding where incarcerated individuals come from can help us understand which communities are disproportionately impacted so better policies can be created to change negative outcomes. 

## 1. Map of most recent data
To map data in this form, there should be a value corresponding to each census tract. Depending on the year or source of the data, this could be about 700 rows for data at the 2010 census tract resolution, or about 900 rows for data at the 2020 census tract resolution.

The map scale has been adjusted to the natural log of the rate to make the map colors meaningful.

### Create Visual
```{r map .rda}

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r tract data map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# set up palettes
min_value_legend<- min(log10(1+data_tract$estimate), na.rm=TRUE)
max_value_legend<-max(round(log10(data_tract$estimate),0), na.rm=TRUE)

# set up palettes
psrc_purple_plus<-c("#FFFFFF", "#FAE6FA","#D8BFD8","#4a0048","#4a0048")

# tract_data_plus_zero<-append(data_tract$estimate_prelog,0)
# set up palettes
# psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
#                                       domain = data_tract$estimate)



psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min_value_legend, max_value_legend, NA)) # this is the value field

# acs_data_plus_zero<-append(data_tract$estimate,0)
# 
# psrc_palette <- leaflet::colorNumeric(palette = psrc_colors$purples_inc,
#                                       domain = acs_data_plus_zero)

# set the variable
var_name <- "Imprisonment*"
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>", data_tract$geoid20, # the code in <> makes the geoid value italic 
  paste0("Tract value: ", prettyNum((round(data_tract$estimate, digits=1)), big.mark = ","), "/100,000"),
  paste0("Region value: ", prettyNum((round(data_tract$reg_estimate, digits=1)), big.mark = ","), "/100,000"),
  paste0("County value: ", prettyNum((round(data_tract$cnty_estimate, digits=1))), "/100,000", 
        paste0("(",data_tract$county_name," County)"))) %>% 
  lapply(htmltools::HTML)

# if the label required a prefix, for example a $, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$Census.Tract, # the code in <> makes the geoid value italic
#   paste("$",
#         prettyNum((round(data_tract$estimate, digits=-2)),
#                   big.mark = ","))) %>%
#   lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$Census.Tract,
#   paste(prettyNum((round(data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels")%>%
  addPolygons(data=data_tract,
              fillColor =psrc_palette(log10(1+data_tract$estimate)),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              # label = round(data_tract$estimate, digits=1),
              label = labels) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min_value_legend, max_value_legend, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       bins=4,
                       decreasing = TRUE, labFormat = labelFormat(
                         transform = function(x) round(10^x, 0)))%>%
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```
Source(s): Prison Policy Initiative, Number of People in Prison in 2020 from Each Washington Census Tract; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles

* Note: color ramp uses log scale

```{r generate tract map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\a03-imprisonment\webpage-html-outputs

# interactive webpage output process
htmlwidgets::saveWidget(tract_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\

#### Data call outs
```{r map calculations - tract, include=FALSE, eval=FALSE}
# simplify data set
data <- data_tract %>%
  st_drop_geometry()

# calculate call out information
calc <- data %>%
  mutate(prison_rate=(sum(estimate*total_pop20, na.rm =TRUE))/(sum(total_pop20, na.rm=TRUE)))


x<-rep(data$estimate,times=data$total_pop20)
df1<-data.frame(x)
summary(df1$x)
median(df1$x, na.rm=TRUE)
hist(df1$x)

#   Min. 1st Qu.  Median  Mean   3rd Qu.  Max. 
#  0.000  43.66   94.25  180.12  200.80  8006.63 

sort <- data %>% 
  dplyr::arrange(desc(estimate))

# 2 highest: 53053061400 (downtown Tacoma that includes the jail) Pierce (8006.6262), 53033008500 (Yesler Terrace are of Seattle that includes Harborview) King (5127.6504), 53053060200 (Port of Tacoma area) Pierce (4624.0450), 53061040700 (downtown Everett which includes the Snohomish County jail) Snohomish (3608.0442)

sort <- data %>% 
  dplyr::arrange(estimate)

# 3 lowest: there are 99 census tracts with a rate of zero

by_county <- data %>% 
  dplyr::group_by(County.Name) %>% 
  dplyr::summarise(prison_rate=(sum(estimate*total_pop20, na.rm =TRUE))/(sum(total_pop20, na.rm=TRUE))) %>% 
  dplyr::arrange(desc(prison_rate))
# King (139.1627), Snohomish (158.3497), Kitsap (231.0399), Pierce (285.3764)

# State average: state prison pop = 15094, state pop with prisoners added back into pop = 7527559, rate per 100,000 = 200.5165287, natural log of rate = 5.300896681

```

1. 180: The region's average imprisonment rate
2. 200: Washington State's average imprisonment rate
3. 276: Number of census tracts in the region above the region's average imprisonment rate
\

#### Insights & Analysis

* The four census tracts with the highest average imprisonment rates all include large facilities such as jails or Harborview Medical Center, which may indicate that these places were used as last known addresses for some incarcerated individuals
* The census tracts with the highest imprisonment rate are Downtown Tacoma, Pierce County (8006); Yesler Terrace area of Seattle, King County (5127); Port of Tacoma area, Pierce County (4624); Downtown Everett, Snohomish County (3608)
* Pierce County has the highest average imprisonment rate (285), followed by Kitsap (231), Snohomish (158), and King (139)
* There are 99 census tracts with an imprisonment rate of zero in the region. For Pierce County, 3% of census tracts (6) have a rate of zero for imprisonment, followed by Kitsap with 5% (3), Snohomish with 6% (10), and then King county with 16% (80).
\
\

## 2. Facet of most recent data
### Create Visual
<style type="text/css">
  .main-container {
    max-width: 100% !important;
    margin: auto;
  }
</style>
  
```{r load rda for chart, include=FALSE}
# load visualization functions for charts
# source('C:/Coding/equity-tracker/data-visualization/equity-tracker-chart-functions.R') # this needs to be adjusted based on your local GitHub directory - it is required if you wish to run chunks and view the visuals (without first knitting)
# source('equity-tracker-chart-functions.R')
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - tract}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county_ord"
x = "quintile_ord"
y = "estimate"
facet = "equity_group_ord"
fill = "quintile_ord"
y_min = 0 
y_max = 800
dec = 1
esttype = "number"
color = "oranges"
num_colors = 5
color_rev = FALSE
title = "Imprisonment by Community"
subtitle = "Rate per 100,000 residents"
#source = paste("Prison Policy Initiative, Number of People in Prison in 2020 from Each Washington Census Tract;", "U.S. Census Bureau, 2016-2020 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002", sep="\n")

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color, num_colors = num_colors,
                                            color_rev = color_rev)

```

```{r output final column chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\a03-imprisonment\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
  `r title`
</p>
  <p style="font-family: Poppins; font-size:12pt; font-weight:normal">
  `r subtitle`
</p>
```{r column chart output - tract, echo=FALSE}

if (output_type == "html") {
  
  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
 
</p> 
  
```{r column facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         equity_group=="poc_quintile")

# region low quintile poc households: 107.0394
# region high quintile poc households: 230.6096
# 230.6096-107.0394=123.5702

filter <- data_clean_column %>% 
  filter(quintile=="High",
         equity_group=="income_quintile")

# region high quintile low income households: 383.3097
# king high quintile low income households: 266.8788
# kitsap high quintile low income households: 492.9817
# pierce high quintile low income households: 713.1442
# snohomish high quintile low income households: 321.0439
# 713.1442-383.3097=329.8345

filter <- data_clean_column %>% 
  filter(equity_group=="lep_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for lep: 214.62451-108.30399=106.3205

filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for income: 383.30974-48.66483=334.6449
# king dif bt high/low quintiles for income: 266.87878-35.73872=231.1401
# kitsap dif bt high/low quintiles for income: 492.98171-117.62698=375.3547
# pierce dif bt high/low quintiles for income: 713.14421-74.01117=639.133
# snohomish dif bt high/low quintiles for income: 321.04386-69.00852=252.0353

filter <- data_clean_column %>% 
  filter(equity_group=="disability_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for disability: 326.40311-61.42979=264.9733
# king dif bt high/low quintiles for disability: 221.33108-58.01088=163.3202
# kitsap dif bt high/low quintiles for disability: 311.52873-125.09387=186.4349
# pierce dif bt high/low quintiles for disability: 451.45180-101.29526=350.1565
# snohomish dif bt high/low quintiles for disability: 317.52730-77.37100=240.1563

filter <- data_clean_column %>% 
  filter(county!="Region",
         equity_group=="poc_quintile",
         quintile=="Low"|
           quintile=="High")
# king dif bt high/low quintiles for income: 238.82189-57.27276=181.5491
# kitsap dif bt high/low quintiles for income: 324.01960-99.66457=224.355
# pierce dif bt high/low quintiles for income: 417.13152-117.40980=299.7217
# snohomish dif bt high/low quintiles for income: 123.56042-117.04332=6.5171

# Information about incarceration is connected to poorer outcomes: https://www.irp.wisc.edu/resource/connections-among-poverty-incarceration-and-inequality/#_edn8

```
\

#### Data Call Outs
1. 8x: As a region, communities with high concentrations of households with lower income have almost eight times higher imprisonment rate
2. 5x higher: As a region, communities with high concentrations of people with disabilities have five times higher imprisonment rate
3. 2x: Pierce County has almost twice the imprisonment rate as the region for communities with high concentrations of households with people of color or lower income


#### Insights & Analysis

* Communities with high concentrations of households with lower income or people with disabilities are most disproportionately impacted by imprisonment
* Pierce County has the highest imprisonment rate (713) for communities with high concentration of households with lower income, followed by Kitsap (493), Snohomish (321), and King (267); the average rate for high concentration of households with lower income across the Region is 383
* The highest imprisonment rate for communities with high concentration of people with disabilities is Pierce County (451), followed by Kitsap (312), King (221), Snohomish (318); the average rate for high concentration of people with disabilities across the Region is 326
and the rate for the Region is 326
\

This dataset first became available in 2020 due to a change in Washington State law requiring the Department of Corrections to report last known addresses for people incarcerated in state prisons during the 2020 Decennial Census. At this time, 2020 is the only year available and it is unclear if this data will be available again after the 2030 Decennial Census.
\
\

## 3. Facet of trend data
### Create Visual
<style type="text/css">
  .main-container {
    max-width: 100% !important;
    margin: auto;
  }
</style>
  
```{r line facet - tract, include=FALSE}
### Placeholder language for future use/updating when multiple years are available 

# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "quintile_ord" # different from variables for facet column
y_min = 0
y_max = 800
color = "blues"
title = "Imprisonment Trend by Community"
subtitle = "Rate per 100,000 residents, in 5-year spans between 2010 and 2020"
#source = paste("Prison Policy Initiative, Number of People in Prison in 2020 from Each Washington Census Tract;", "U.S. Census Bureau, 2006-2010, 2011-2015, 2016-2020 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002", sep="\n")


line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\a03-imprisonment\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
  `r title`
</p>
  <p style="font-family: Poppins; font-size:12pt; font-weight:normal">
  `r subtitle`
</p>
```{r line chart - tract, echo=FALSE}
if (output_type == "html") {
  
  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
 
</p> 
  
```{r line facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2010" |
           data_year=="2020")

# 2020 region low quintile poc households: 4.035320
# 2020 region high quintile poc households: 4.615563
# 4.615563-4.035320=0.580243

filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2020")

# 2020 region low quintile income: 2.882521
# 2020 region high quintile income: 5.352679
# 5.352679-2.882521=2.470158

filter <- data_clean %>% 
  filter(county=="Snohomish",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High")

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2020")

# 2020 King dif low/high income: 5.083316-2.306517=2.776799
# 2020 Kitsap dif low/high income: 5.906103-4.232797=1.673306
# 2020 Pierce dif low/high income: 5.911497-3.856045=2.055452
# 2020 Snohomish dif low/high income: 5.306512-3.715537=1.590975

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="disability_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2020")

# 2020 King dif low/high disability: 4.912824-2.659749=2.253075
# 2020 Kitsap dif low/high disability: 5.468296-3.715005=1.753291
# 2020 Pierce dif low/high disability: 5.547184-3.812307=1.734877
# 2020 Snohomish dif low/high disability: 5.137279-3.679690=1.457589

```



\
\

# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. *Always run this chunk of code. If this is the first time you are generating visuals for an indicator, comment out the 'update' part of the 'y.drive.folder.update' variable because this is the first version that is being created.*
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update' #comment out if this is the first time you are generating visuals
                                   )

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r to archive, include=FALSE, eval=FALSE}
# add previous year folder to archive folder - to retain and organize all archived files
dir.create(file.path(file_names$base_dir,
                     file_names$theme_dir,
                     file_names$ind_dir,'archive', past_year))

# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive', past_year)

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall). *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage. *Always run this chunk of code.*
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>