---
title: "Voter Participation" 
subtitle: "Data Gen: Exploring, Cleaning, Transforming (PUMS/OSPI data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts
library(stringr) #for string wrapping

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(RMySQL)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Voter Participation**

# Download data 
## Primary data 
The data set referenced in this script is generated from the [Washington Secretary of State elections](https://www.sos.wa.gov/elections#data_research) voter turnout data [(PUMS)]. These data sets provide data at the precinct-level, which is then translated to census geographies and related to the distribution/concentraion of the six equity demographic groups of interest. 
```{r, include=FALSE}
# PUMS/OSPI data set
# pums_ospi_elmer <- get_table(schema="equity",
#                              tbl_name="v_tracker_indicators")

### Get 2020 presidential election votes by precinct ---------------------
# From: https://www.sos.wa.gov/elections/research/data-and-statistics.aspx
election = read_excel("Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data/2020_precinct_results.xlsx",
                      sheet = "Sheet1")

voters_base <- election %>% 
  dplyr::filter(RaceName == "Turnout",
                Candidate %in% c("Registered Voters", "Ballots Cast"),
                County %in% c("King","Kitsap","Snohomish","Pierce")) %>% 
  dplyr::select(Candidate,PrecinctCode,County,Votes) %>% 
  pivot_wider(names_from = Candidate, values_from = Votes) %>% 
  rename(registered_voters = 'Registered Voters', ballots_cast = 'Ballots Cast')

voters <- voters_base %>%
  mutate(Precinct = factor(PrecinctCode), 
         County = factor(County),
         CPrecinct = str_c(County, Precinct, sep = "_"))

### Get population and number of residential units by parcel ---------------------
mydb = dbConnect(MySQL(), 
                 user='psrcurbansim', 
                 password='psrc_urbansim', 
                 dbname='2018_parcel_baseyear_rtp', 
                 host='aws-modelmysql')

# dbListTables(mydb)

# Total population for each parcel - used to calculate percent of people in precinct who voted
# Join households and buildings to get population per parcel
building_pop = dbGetQuery(mydb, "select building_id, sum(persons) as total_ppl 
                                from households
                                group by building_id")

parcel_data = dbGetQuery(mydb, "select * 
                                from buildings")

pop_by_parcel = parcel_data %>% 
  dplyr::select(building_id, parcel_id) %>% 
  left_join(building_pop,  by = "building_id") %>% 
  group_by(parcel_id) %>% 
  summarise(pop_sum = sum(total_ppl, na.rm = TRUE))

# Total residential units per parcel - used to calculate weight of clipped precinct in tract
resunits_by_parcel = parcel_data %>%
  group_by(parcel_id) %>%
  summarise(resunit_sum = sum(residential_units, na.rm = TRUE))

# Get parcel centroids (X,Y coordinates)
parcel_centroid = dbGetQuery(mydb, "select * from parcels")

# Select only relevant columns and set projection to NAD83 Washington State Plane North (CRS 2285) -> WGS84 (match other spatial layers)
parcel_centroid <- parcel_centroid %>%
  dplyr::select(parcel_id, census_tract_id, x_coord_sp, y_coord_sp) %>% 
  st_as_sf(coords = c("x_coord_sp", "y_coord_sp")) %>% 
  st_set_crs(2285) %>% 
  st_transform(2285) #WGS84: 4326

# plot(parcel_centroid)

# Attach parcel population/residential unit information to parcel centroids
parcel_centroid <- parcel_centroid %>%
  left_join(pop_by_parcel, by = "parcel_id") %>% 
  left_join(resunits_by_parcel, by = "parcel_id") %>%
  # remove NAs 
  filter(!is.na(resunit_sum))

# write_rds(parcel_centroid, "./components/parcel_centroid_pop_resunits.rds")

# rm(mydb, building_pop, parcel_data, pop_by_parcel, resunits_by_parcel)

### Get precinct shapefile for intersections ---------------------
# From: https://www.sos.wa.gov/elections/research/precinct-shapefiles.aspx
# precinct <- read_sf("Y:/VISION 2050/Data/Displacement/Displacement Index 2021/data/15-Voter Turnout/statewide_precincts_2020general/Statewide_Precincts_2020General.shp")
precinct_data <- read_sf("Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data/statewide_precincts_2020general/Statewide_Precincts_2020General.shp")

# Retain only PSRC counties, select only relevant columns, set CRS 2285 projection
precinct <- precinct_data %>%
  filter(CountyName %in% c("King", "Kitsap", "Pierce", "Snohomish")) %>%
  mutate(County = factor(County),
         Precinct = factor(PrecName),
         PrecinctCode = factor(PrecCode),
         CPrecinct = str_c(County, Precinct, sep = "_")) %>%
  dplyr::select(County, PrecinctCode, Precinct, CPrecinct, St_Code, geometry) %>%
  st_transform(2285) #2285 4326

# Attach election results to precinct spatial features
precinct_elec <- precinct %>%
  left_join(voters %>% dplyr::select(PrecinctCode, registered_voters, ballots_cast), 
            by = c("St_Code" = "PrecinctCode")) %>% 
  mutate(ballots_cast = replace_na(ballots_cast, 0))

# rm(precinct, voters)


### Get tract shapefile for intersections ---------------------
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
# tracts10.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)
# tracts10.lyr<-st_read(tracts10.url)

# tract <- read_sf("Y:/VISION 2050/Data/Displacement/Displacement_Risk_Script/gis/tract2010_nowater.shp")
# Set correct projection
tract <- tracts20.lyr %>%
  mutate(GEOID = factor(geoid20)) %>%
  st_transform(2285) # transform to match the coordinates of the precinct centroids (need to change back later)

### Spatial intersections

# PRECINCT/PARCEL: Intersect parcel centroids with precinct spatial feature - yields number of people and residential units in each precinct
precinct_parcel = st_join(precinct_elec, parcel_centroid)
# 25 precincts have no parcels or residential units, so the parcel columns are NA. Make these zero to indicate absence of population/residential units
precinct_parcel$pop_sum <- replace_na(precinct_parcel$pop_sum, 0)
precinct_parcel$resunit_sum <- replace_na(precinct_parcel$resunit_sum, 0)

# Sum up population and total residential by precinct
precinct_pop = precinct_parcel %>%
  # Do not need spatial features to calculate population/total res. units by parcel, and runs faster making geometry NULL
  st_set_geometry(NULL) %>% 
  group_by(St_Code) %>% 
  summarise(precinct_pop = sum(pop_sum),
            precinct_resunit = sum(resunit_sum))

# rm(precinct_parcel)

# Attach population/resunit totals to precinct spatial features/election info
# Then can calculate "turnout", or number of votes out of all people living in the precinct (not just eligible voters)
# Note adding 1 to precinct_pop to avoid zero divide
precinct_pop <- precinct_elec %>% 
  left_join(precinct_pop, by = "St_Code" ) %>% 
  mutate(turnout = ballots_cast/(precinct_pop + 1),
         # 64 precincts have more ballots cast than population. Cap turnout at 1 (100%)
         turnout = ifelse(turnout > 1, 1, turnout)) 

# rm(precinct_elec)
# write_rds(precinct_pop, "./components/precinct_turnout_cap100.rds")

# PRECINCT/TRACT: Intersect precinct and tract spatial features to yield clipped precincts
precinct_tract <- st_intersection(precinct_pop, tract) %>% 
  # Add ids to clipped precincts to group on later
  mutate(id = row_number())

# rm(precinct_pop)

# PARCELS/CLIPPED PRECINCTS: Intersect parcel centroids with clipped precincts - yields number of people and residential units in each clipped precinct
precinct_tract_parcel = st_join(precinct_tract, parcel_centroid)

# rm(parcel_centroid)

# summary(precinct_tract_parcel$pop_sum) #5109 NAs
# summary(precinct_tract_parcel$resunit_sum) #5109 NAs
# 5109 clipped precincts have no parcels, so the parcel columns are NA. Change total population and residential units to 0
precinct_tract_parcel$pop_sum <- precinct_tract_parcel$pop_sum %>% replace_na(0)
precinct_tract_parcel$resunit_sum <- precinct_tract_parcel$resunit_sum %>% replace_na(0)

# Count all residential units in clipped precinct
precinct_tract_population = precinct_tract_parcel %>% 
  st_set_geometry(NULL) %>% 
  group_by(id) %>% 
  summarise(precinct_resunit_int = sum(resunit_sum)) 

# rm(precinct_tract_parcel)

# Attach total residential units to clipped precinct spatial features and data
precinct_tract_pop <- precinct_tract %>% 
  left_join(precinct_tract_population, by = "id" )

# rm(precinct_tract_population)

# WEIGHTS: Calculate total residential units in each tract by grouping clipped precincts by tract
tractresunit <- precinct_tract_pop %>% 
  st_set_geometry(NULL) %>% 
  group_by(GEOID) %>% 
  summarise(tractresunit = sum(precinct_resunit_int))

cntyresunit <- precinct_tract_pop %>% 
  st_set_geometry(NULL) %>% 
  group_by(county_name) %>% 
  summarise(cntyresunit = sum(precinct_resunit_int))

regresunit <- precinct_tract_pop %>% 
  st_set_geometry(NULL) %>% 
  # group_by(county_name) %>% 
  summarise(regresunit = sum(precinct_resunit_int))

# Attach total residential units by tract to clipped precincts, calculate weights as (# resunits clipped precinct)/(# resunits in tract + 1)
# One tract has no residential units, so add 1 to tractresunit
precinct_tract_wt <- precinct_tract_pop %>%
  left_join(tractresunit, by = "GEOID") %>%
  mutate(weight = precinct_resunit_int/(tractresunit + 1),
         weighted_vote = weight * turnout)

precinct_cnty_wt <- precinct_tract_pop %>%
  left_join(cntyresunit, by = "county_name") %>%
  mutate(weight = precinct_resunit_int/cntyresunit,
         weighted_vote = weight * turnout)

precinct_reg_wt <- precinct_tract_pop %>%
  # left_join(regresunit, by = "county_name") %>%
  mutate(weight = precinct_resunit_int/regresunit$regresunit,
         weighted_vote = weight * turnout)

# FINAL VALUES: Calculate turnout score by tract as weighted sum of turnout in clipped precincts
# tabular data
votes_tract <- precinct_tract_wt %>%
  st_set_geometry(NULL) %>%
  group_by(GEOID) %>%
  summarise(votes = sum(weighted_vote, na.rm = TRUE) * 100)

votes_cnty <- precinct_cnty_wt %>% 
  st_set_geometry(NULL) %>% 
  group_by(county_name) %>% 
  summarise(votes_cnty = sum(weighted_vote, na.rm = TRUE) *100)

votes_reg <- precinct_reg_wt %>% 
  st_set_geometry(NULL) %>%
  summarise(votes_reg = sum(weighted_vote, na.rm = TRUE) *100)

# Join tabular data to tract (need common fields), transform to data frame
tract_votes_df <- tract %>%
  left_join(votes_tract, by = ("GEOID")) %>% 
  left_join(votes_cnty, by = ("county_name")) %>% 
  mutate(votes_reg = votes_reg$votes_reg)%>%
  st_set_geometry(NULL) #keep in tabular form (remove geometry)

# votes_tract_viz <- precinct_tract %>%
#   st_set_geometry(NULL) %>%
#   group_by(GEOID) %>%
#   summarise(votes = sum(weighted_vote, na.rm = TRUE),
#             precincts = toString(St_Code),
#             clipped_resunits = toString(precinct_resunit_int),
#             turnouts = (round(turnout,2)) * 100)

# rm(precinct_tract, tractresunit)
```
\
This data set was compiled from PUMS data.
\
Looking at the fields in the data set:
```{r}
# explore the data
head(pums_ospi_elmer)
# view the equity demographic groups
unique(pums_ospi_elmer$focus_type)
# view the different indicators available within the data set
unique(pums_ospi_elmer$indicator_type)
# view the different data years available within the data set
unique(pums_ospi_elmer$data_year)
```

```{r indicator specific information, include=FALSE}
# set indicator information
indicator_measure <- 'median_household_income'
indicator_title <- 'Median household income'
```

## 1. Explore data
*In this section we make sure that the data set has the correct dimensions - year, geography, equity focus group, etc.*  
```{r}
# clean data set
data_full <- pums_ospi_elmer %>% 
  # distinct(indicator_fact_id, .keep_all = TRUE) %>% #this may be necessary depending on indicator
  dplyr::filter(indicator_type==indicator_measure) %>% 
  dplyr::filter(focus_type!="Total") %>% 
  filter(focus_attribute!="Total") %>%
  # filter(indicator_attribute!="Total") %>%
  dplyr::mutate(data_year_yr=format(data_year,format="%Y"))
```

### Data fields
#### *consistent base data*

* There should be 5 geographies - the 4 counties and the Region. 
* There should be 6 equity focus categories - POC, income, disability, youth, older adult, and LEP
    + 2 sub-groups in each of the 6 equity focus group (e.g. people of color, non-people of color)
```{r}
# geographies
num_county <- length(unique(data_full$county)) #5
unique(data_full$county)

# equity focus groups
num_group <- length(unique(data_full$focus_type)) #6
unique(data_full$focus_type)
```
\
\

#### *indicator-specific data*
These fields will vary by indicator:

* Type of metric - this will determine how the data are visualized *(est ="percent" or "currency" or "number")*
* Number of years (5-year span) - this can vary depending on data availability
* Number of indicator-specific categories - this can vary depending on the indicator of interest, ranging from N/A (*median income*) to multiple levels (*crowding, housing cost burden*)
```{r}
# metric
unique(data_full$fact_type) #median

# years
num_yr <- length(unique(data_full$data_year)) #3
unique(data_full$data_year)

# indicator attributes
num_indatt <- length(unique(data_full$indicator_attribute)) #1
unique(data_full$indicator_attribute) #N/A

# calculate expected number of rows
num_row <- num_yr*num_county*num_group*2*num_indatt #180
```

There are **`r num_county`** geographies and **`r num_group`** equity focus groups (each with **2** subgroups). There are **`r num_yr`** years in the data set and the indicator specific field has **`r num_indatt`** attribute(s), which means there should be a total of **`r num_row`** rows.
```{r}
# count number of rows
nrow(data_full) #180 in the full data set

data_full_nona <- data_full %>% 
  dplyr::filter(!is.na(fact_value))

nrow(data_full_nona) #170 - there are some NAs in the data
```
<span style="color: #00A7A0">There are some missing data.</span> 
\
\

#### *checking for missing data*
##### Year / geography
```{r}
num_yr_geo <- num_group*2*num_indatt #12
```
If we look at the data by year and geography, there should be **`r num_yr_geo`** entries per year/geography.
```{r, include=FALSE, eval=FALSE}
table(data_full_nona$data_year,
      data_full_nona$county)
```
<span style="color: #00A7A0">
2011 is missing some data for all geographies.
</span> 
\
\

##### Year / equity focus group
```{r}
num_yr_grp <- num_county*2*num_indatt #10
```
If we look at the data by year and focus group, there should be **`r num_yr_grp`** entries per year/focus group.
```{r}
table(data_full_nona$data_year,
      data_full_nona$focus_type)
```
<span style="color: #00A7A0">
The disability category is missing all data in 2011.
</span> 
\
\

##### Year / equity focus sub-group
```{r}
num_yr_subgrp <- num_county*num_indatt #5
```
If we look at the data by year and focus sub-group, there should be **`r num_yr_subgrp`** entries per year/focus sub-group.
```{r}
table(data_full_nona$data_year,
      data_full_nona$focus_attribute)
```
\
\

##### Year / indicator attribute
```{r}
num_yr_ind <- num_county*num_group*2 #60
```
If we look at the data by year and indicator attribute, there should be **`r num_yr_ind`** entries per year/indicator attribute.
```{r}
table(data_full_nona$data_year,
      data_full_nona$indicator_attribute)
```
\
\

### Numeric data
To check for *0*s and *NULL*s
```{r}
# checking for NULLs
summary(data_full$fact_value)
```
<span style="color: #00A7A0">
There are nulls.
</span> 
\
\

To look at distribution of all data - not the most useful visual, but provides a sense of the range of values at a high level in one plot. 
\
```{r}
qplot(data_year_yr, fact_value, 
      colour = county, shape = focus_type, 
      data = data_full, 
      main = indicator_title) + theme(axis.title.x = element_blank(),
                                               axis.title.y = element_blank())
```
\
This table includes a lot of information about the data set and helps to show the different levels of each field. It provides another way to check if data are available for all counties and all years, or where there may be gaps in the data set. 
```{r}
data_refined <- data_full %>% 
  dplyr::select(data_year_yr, county, vulnerability, focus_type, focus_attribute,
                indicator_type, indicator_attribute, fact_value)

sumtable(data_refined,
         add.median=TRUE,
         group="county",
         out="return")
```

```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

focus_type_order <- c("People of Color", 
                      "Households with Lower Income", 
                      "People with a Disability", 
                      "Households with Limited English Proficiency", 
                      "Households with Youth <18", 
                      "Households with Older Adults 65+")

focus_attribute_order <- c("People of color",
                           "White\nnon-Hispanic",
                           "Households with\nlower income",
                           "With a\ndisability",
                           "Without a\ndisability",
                           "Limited English\nproficiency",
                           "English\nproficient",
                           "Households with\nyouth",
                           "Households with\nolder adults",
                           "Other households")

# transforming data labels ----
data_clean <- data_full %>% 
  mutate(county = factor(county, levels=county_order)) %>%
  mutate(focus_type_ord = case_when(
    focus_type=="POC_cat"~"People of Color",
    focus_type=="Disability_cat"~"People with a Disability",
    focus_type=="LEP_cat"~"Households with Limited English Proficiency",
    focus_type=="Income_cat"~"Households with Lower Income",
    focus_type=="Youth_cat"~"Households with Youth <18",
    focus_type=="Older_cat"~"Households with Older Adults 65+")) %>%
  mutate(focus_type_ord = factor(focus_type_ord, levels = focus_type_order)) %>%
  mutate(focus_attribute_ord = case_when(
    focus_attribute== "POC"~ "People of color",
    focus_attribute== "Non-POC"~ "White non-Hispanic",
    focus_attribute== "Low Income"~ "Households with lower income",
    focus_attribute== "Non-Low Income"~ "Other households",
    focus_attribute== "With disability"~ "With a disability",
    focus_attribute== "Without disability"~ "Without a disability",
    focus_attribute== "Limited English proficiency"~ "Limited English proficiency",
    focus_attribute== "English proficient"~ "English proficient",
    focus_attribute== "Household with youth"~ "Households with youth",
    focus_attribute== "Household without youth"~ "Other households",
    focus_attribute== "Household with older adult"~ "Households with older adults",
    focus_attribute== "Household without older adult"~ "Other households")) %>% 
  mutate(focus_attribute_ord = str_wrap(focus_attribute_ord, width=16)) %>%
  mutate(focus_attribute_ord = factor(focus_attribute_ord, levels = focus_attribute_order)) 

# echarts wants axis to be factors/characters so convert year to a character
data_clean$data_year <- as.character(data_clean$data_year)

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county, focus_type_ord, focus_attribute_ord, data_year)
```

```{r}
# indicator specific re-ordering - median income is N/A so this step is unnecessary
# data_clean$indicator_attribute <- factor(data_clean$indicator_attribute,
#                                     levels = c("Less than a Bachelor's degree",
#                                                "Bachelor's degree or higher"))
```
\

### Data labels, shares
These charts were generated to ensure the labels across years are consistent/make sense. There had been an issue with misassigned labels because tidycensus::pums_variables, i.e. the only digital data dictionary available to associate labels with codes, exists only from 2017 forward. Most variables have had consistent codes, but in cases where the codes have shifted over time, using the 2017 lookup winds up mischaracterizing categories.
\
\
These charts also help to confirm that the shares add up to 100% - only relevant when *indicator_attribute* has more than one category. The *indicator_attribute* for median household income is NA.
\
\
*The colors of the charts may not be consistent between the years depending on missing data.* 
\
```{r}
# the years below reflect the years that are available from unique(data_year)
just_region_11 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2011) %>% 
  arrange(focus_attribute_ord)
just_region_16 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2016)%>% 
  arrange(focus_attribute_ord)
just_region_21 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2021)%>% 
  arrange(focus_attribute_ord)

# check est type - make sure it corresponds to your indicator (percent, currency, or number)
facet_chart <- static_facet_column_chart(t=just_region_11,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "currency",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2011)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")

facet_chart

facet_chart <- static_facet_column_chart(t=just_region_16,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "currency",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2016)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")
facet_chart

facet_chart <- static_facet_column_chart(t=just_region_21,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "currency",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2021)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")

facet_chart
```

## 2. Visually explore data
### Scatter plots
*In this section we start to explore the data visually - distribution by the different dimensions within the data set. These plots are helpful to check for outliers and get a higher level understanding of the data in one visual, before slicing the data by geography and equity focus group in the following sections.* 
\
\
The following code will need to be adjusted to fit the fields specific to the data indicator. For educational attainment, we focus on those with a Bachelor's degree or higher. The following code establishes the data frame that the rest of the analysis uses. If there are fewer than 2 *indicator attributes*, this section can be skipped/commented out, but the code will need to be adjusted throughout. 
\
\

#### By *indicator_attribute*
This section isn't relevant for this specific indicator because there aren't unique indicator attributes. 

#### By Year
```{r}
# separating by year ----
data_clean11 <- data_clean %>% 
  filter(data_year==2011)
data_clean16 <- data_clean %>% 
  filter(data_year==2016)
data_clean21 <- data_clean %>% 
  filter(data_year==2021)
```

```{r}
# plot data, adjust y limits based on data set
qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean11, 
      main = paste0(indicator_title, ": 2011")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 150000)

qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean16,
      main = paste0(indicator_title, ": 2016")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 150000)

qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean21, 
      main = paste0(indicator_title, ": 2021")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 150000)
```

# **CHART .rda**
```{r, include=FALSE}
head(data_clean) # make sure this is the correct data set

# make any additional edits 
data_clean <- data_clean %>% 
  mutate(fact_value_notrounded=fact_value) %>% #create column to retain full median income values
  mutate(fact_value=round(fact_value, digits=-2)) #round median income value to hundreds and set retain consistent name - if you are not working with $ you will need to change this to the correct number of digits you would like to round to (for example, if using decimals, you may want to set digits=2 or if you using whole numbers you may want to set digits=0)

data_clean #check to make sure the rounding worked for your indicator - if not will need to go back to generate data set in chunk around line 245
```
Save final data set (.rda) for charts
```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'h-public-services'
ind_dir <- 'h02-voter-participation'
file_name <- 'h02-voter-participation-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```


## Create data set for map .rda 
The spatial data was created above when it was first downloaded and transformed. Very little additional processing is needed in this section, other than changing the variable names to match existing 

### Transform data
```{r}
# Join tabular data to tract spatial data
tract_votes <- tract %>%
  left_join(votes_tract, by = ("GEOID")) %>% 
  left_join(votes_cnty, by = ("county_name")) %>% 
  mutate(votes_reg = votes_reg$votes_reg)%>%
  st_transform(4326)

# consistent field names
data_tract <- tract_votes %>% 
  dplyr::rename(estimate=votes,
                reg_estimate=votes_reg,
                cnty_estimate=votes_cnty)
```

### Explore map/acs data
```{r map legend settings, include=FALSE}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - this code helps to set up the legend so that it is arranged high-low with correct color order
addLegend_decreasing <- function (map, position = c("topright", "bottomright", "bottomleft", 
			    "topleft"), pal, values, na.label = "NA", bins = 7, colors, 
		  opacity = 0.5, labels = NULL, labFormat = labelFormat(), 
		  title = NULL, className = "info legend", layerId = NULL, 
		  group = NULL, data = getMapData(map), decreasing = FALSE) {
	position <- match.arg(position)
	type <- "unknown"
	na.color <- NULL
	extra <- NULL
	if (!missing(pal)) {
		if (!missing(colors)) 
			stop("You must provide either 'pal' or 'colors' (not both)")
		if (missing(title) && inherits(values, "formula")) 
			title <- deparse(values[[2]])
		values <- evalFormula(values, data)
		type <- attr(pal, "colorType", exact = TRUE)
		args <- attr(pal, "colorArgs", exact = TRUE)
		na.color <- args$na.color
		if (!is.null(na.color) && col2rgb(na.color, alpha = TRUE)[[4]] == 
		    0) {
			na.color <- NULL
		}
		if (type != "numeric" && !missing(bins)) 
			warning("'bins' is ignored because the palette type is not numeric")
		if (type == "numeric") {
			cuts <- if (length(bins) == 1) 
				pretty(values, bins)
			else bins	
			
			if (length(bins) > 2) 
				if (!all(abs(diff(bins, differences = 2)) <= 
				         sqrt(.Machine$double.eps))) 
					stop("The vector of breaks 'bins' must be equally spaced")
			n <- length(cuts)
			r <- range(values, na.rm = TRUE)
			cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
			n <- length(cuts)
			p <- (cuts - r[1])/(r[2] - r[1])
			extra <- list(p_1 = p[1], p_n = p[n])
			p <- c("", paste0(100 * p, "%"), "")
			if (decreasing == TRUE){
				colors <- pal(rev(c(r[1], cuts, r[2])))
				labels <- rev(labFormat(type = "numeric", cuts))
			}else{
				colors <- pal(c(r[1], cuts, r[2]))
				labels <- rev(labFormat(type = "numeric", cuts))
			}
			colors <- paste(colors, p, sep = " ", collapse = ", ")
			
		}
		else if (type == "bin") {
			cuts <- args$bins
			n <- length(cuts)
			mids <- (cuts[-1] + cuts[-n])/2
			if (decreasing == TRUE){
				colors <- pal(rev(mids))
				labels <- rev(labFormat(type = "bin", cuts))
			}else{
				colors <- pal(mids)
				labels <- labFormat(type = "bin", cuts)
			}
			
		}
		else if (type == "quantile") {
			p <- args$probs
			n <- length(p)
			cuts <- quantile(values, probs = p, na.rm = TRUE)
			mids <- quantile(values, probs = (p[-1] + p[-n])/2, 
				 na.rm = TRUE)
			if (decreasing == TRUE){
				colors <- pal(rev(mids))
				labels <- rev(labFormat(type = "quantile", cuts, p))
			}else{
				colors <- pal(mids)
				labels <- labFormat(type = "quantile", cuts, p)
			}
		}
		else if (type == "factor") {
			v <- sort(unique(na.omit(values)))
			colors <- pal(v)
			labels <- labFormat(type = "factor", v)
			if (decreasing == TRUE){
				colors <- pal(rev(v))
				labels <- rev(labFormat(type = "factor", v))
			}else{
				colors <- pal(v)
				labels <- labFormat(type = "factor", v)
			}
		}
		else stop("Palette function not supported")
		if (!any(is.na(values))) 
			na.color <- NULL
	}
	else {
		if (length(colors) != length(labels)) 
			stop("'colors' and 'labels' must be of the same length")
	}
	legend <- list(colors = I(unname(colors)), labels = I(unname(labels)), 
	               na_color = na.color, na_label = na.label, opacity = opacity, 
	               position = position, type = type, title = title, extra = extra, 
	               layerId = layerId, className = className, group = group)
	invokeMethod(map, data, "addLegend", legend)
}
```

```{r explore map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Voter Participation"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

# **MAP .rda**
Save final data set (.rda) for map
```{r}
# This is the data set used to create the map of the most recent data. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk.

# set folder structure
file_name <- 'h02-voter-participation-map-data'

# save final data set as .rda
save(data_tract, file = file.path(base_dir,
                                      theme_dir,
                                      ind_dir, "rda-data",
                                      paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>