---
title: "2a.) Voter Turnout: 2020"
subtitle: "Exploring weighting methods to determine sensitivity"
author: "Mary Richards"
date: today
date-format: long
format: 
  html:
    toc: true
    toc-location: right-body
    toc-expand: 2
    # toc-depth: 6
    toc-title: "Contents"
execute:
  # echo: false
  warning: false
  error: false
---

This documents the process for exploring voter turnout based on data from the WA Secretary of State. In this exercise, we are using precinct-tract geography splits to calculate voter turnout by tract geographies.  
\
The final output is voter turnout (%) by census tract. Ultimately, this measurement will be used in the Equity Tracker and Displacement Risk Mapping projects.
```{r setup, include=FALSE}
# Read in Libraries
library(magrittr)
library(stringr)
library(data.table)
library(tidycensus)
library(tidyverse)
library(openxlsx)
library(psrcelmer)
library(psrccensus)

library(sf)
library(leaflet)
library(leafem) #home button
library(psrcplot)
library(here)

library(readxl)

# reference supplemental script with supporting settings/functions
# source(here::here('GitHub/equity-tracker/data-visualization/equity-tracker-supplemental-script.R'))
source('C:/Users/mrichards/Documents/GitHub/equity-tracker/data-visualization/equity-tracker-supplemental-script.R')
```

::: panel-tabset
# Data 
## 1.) Elmer geography splits: precinct-tract
These data tables include the distribution of the total population based on OFM between voting precincts (2024 and 2020, from SOS) and census tracts (2020). The 2024 and 2020 geography split numbers are weighted based on the 2024 and 2020 5y ACS tract estimates for adult 18+ population, respectively.
```{r}
sql20 <- "select *
    from general.get_any_geography_splits(
      'voter_precinct_2020', --@data_geog_type
      'tract20',              --@planning_geog_type
      2020,                   --@ofm_estimate_year
      2024,                   --@ofm_vintage
      2018)                   --@parcel_year"

df20_adult <- get_query(sql=sql20, db_name = "Elmer")
```

## 2.) Secretary of State: turnout data
These data sets (2012, 2016, 2020, 2024) include the number of registered voters and the number of ballots cast by voting precinct. The voting precinct geographies change between election years. 
```{r}
voting_url <-"https://www.sos.wa.gov/sites/default/files/"

turnout_zip_url24 <- paste0(voting_url, "2025-01/2024Gen_Precinct_Results_GIS-Ready.zip")
turnout_zip_url20 <- paste0(voting_url, "2024-02/2020Gen_Precinct_Results_GIS-Ready.zip") 


psrc_counties <- data.frame("CountyName"=c("King","Kitsap","Pierce","Snohomish"),
                            "FIPS"=c("033","035","053","061"),
                            "CountyCode" = c("KI", "KI", "PI", "SN"))

# 2024 and 2020 saved locally in case of issue with the SOS website
rawdata_folder <- 'Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data'
vt24 <- read.csv(file.path(rawdata_folder, '2024Gen_Precinct_Turnout_GIS-Ready.csv'))
vt20 <- read.xlsx(file.path(rawdata_folder, '2020Gen_Precinct_Turnout_GIS-Ready.xlsx'))


# Clean turnout data sets - filter to PSRC region, rename columns
vt20_df <- vt20 %>% setDT() %>%                                             
  .[County %in% psrc_counties$CountyName] %>%                   # Filter to PSRC counties
  .[, .(county = County, precinctcode = PrecCode, precinctname = PrecName, 
        reg_voters = G20TREGVOT, ballots_cast = G20TBALCST)]

# precinct spatial data
# https://www.sos.wa.gov/elections/data-research/reports-data-and-statistics/precinct-shapefiles
folder <- "Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data"


precinct_20 <- st_read(file.path(folder, "Statewide_Precincts_2020General/Statewide_Precincts_2020General.shp"))
```

## 3.) ACS and Decennial census population data
These values will be used for the denominator - dividing the ballots cast by the total population 18+ or weighting the total population (OFM SAEP) by the ratio of adult 18+/total population estimates.
```{r}
# # ACS 5y for 2012, 2016, and 2024
# acs_year <- c(2024, 2016, 2012) #changed from 2023 to 2024 (Census released 5y estimates, 2/2026)
# 
# base_acs_data <- get_acs_recs(geography ='tract',
#                               table.names = 'S0101', #subject table code
#                               years = c(as.numeric(acs_year)),
#                               acs.type = 'acs5')
# 
# # calculate the proportion of 18+ adults/total population to check if there are some tracts with more/fewer adults
# age_18over_total_24 <- base_acs_data %>%
#   filter(year==2024,
#          variable=="S0101_C01_001" | #Total population
#            variable=="S0101_C01_026") %>% #2024: Total population!!SELECTED AGE CATEGORIES!!18 years and over
#            select(GEOID, variable, estimate, year) %>%
#   pivot_wider(names_from = c(variable),
#               values_from = estimate) %>%
#   reframe(GEOID,
#           year_census=year,
#           pop_18=S0101_C01_026,
#           pop_tot=S0101_C01_001,
#           pop18_total=S0101_C01_026/S0101_C01_001)
# 
# # the variables changed between 2016 and 2023, so need to recalculate
# age_18over_total_16_12 <- base_acs_data %>%
#   filter(year==2016 |
#            year==2012,
#          variable=="S0101_C01_001" | #Total population
#            variable=="S0101_C01_025") %>% #2016: Total!!Estimate!!SELECTED AGE CATEGORIES!!18 years and over
#   select(GEOID, variable, estimate, year) %>%
#   pivot_wider(names_from = c(variable),
#               values_from = estimate) %>%
#   reframe(GEOID,
#           year_census=year,
#           # pop_18=S0101_C01_025, #these are mislabeled and are actually percentages, not estimate
#           # pop_tot=S0101_C01_001, #these are mislabeled and are actually percentages, not estimates
#           pop18_total=S0101_C01_025*0.01)
# 
# # separate out years
# age_18over_total_16 <- age_18over_total_16_12 %>%
#   filter(year_census==2016)
# 
# age_18over_total_12 <- age_18over_total_16_12 %>%
#   filter(year_census==2012)

## decennial census for 2020
base_dec_data <- get_decennial_recs(geography='tract',
                                    table_codes = "DP1",
                                    sumfile = "dp",
                                    years=2020)

# filter variables based on: http://aws-linux/mediawiki/index.php/Decennial_Variables_2020#DP1
age_18over_total_20 <- base_dec_data %>% 
  filter(variable=="DP1_0001C" | #Count:SEX AND AGE:Total population
           variable=="DP1_0021C") %>% #Count:SEX AND AGE:Total population:Selected Age Categories:18 years and over
  select(GEOID, variable, value, year) %>% 
  pivot_wider(names_from = c(variable),
              values_from = value) %>% 
  reframe(GEOID,
          year_census=year,
          pop_18=DP1_0021C,
          pop_tot=DP1_0001C,
          pop18_total=DP1_0021C/DP1_0001C)
```

## 4.) PSRC Region spatial file
This data set includes the census tracts within the region and provides a way to visually explore the voter turnout values. 
```{r, echo=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
arc_service <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services"

tracts20.url <- file.path(arc_service, "Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
tracts10.url <- file.path(arc_service, "Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")

tracts20.lyr <- st_read(tracts20.url)
tracts10.lyr <- st_read(tracts10.url)
```

## 5.) OFM small area estimates (2020) population data
These values will be used as the denominator for the fourth approach - dividing the ballots cast by the total population (OFM - Small Area Demographic Estimates, 2020-2024), weighted by the proportion of adults 18+/total population from 2020 decennial census estimates.
```{r}
# OFM Small Estimates Program: https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/small-area-estimates-program
# folder_loc <- here::here("data_development/voter_turnout")
folder_loc <- "C:/Users/mrichards/Documents/GitHub/equity-tracker/data_development/voter_turnout"


ofm_raw <- read_excel(file.path(folder_loc,
                                "saep_tract20.xlsx"),
                      sheet = "Total Population",
                      skip=10) #first rows includes metadata

# clean data set
ofm_psrc_20 <- ofm_raw %>% 
  select(`County Name`, `Census Tract Code Complete`, 
         # `Estimated Total Population 2012`,
         # `Estimated Total Population 2016`,
         `OFM Adjusted Total Population 2020`#,
         # `Estimated Total Population 2024`
         ) %>% 
  rename("County_Name"=`County Name`,
         "GEOID20"=`Census Tract Code Complete`,
         # "population_2012"=`Estimated Total Population 2012`,
         # "population_2016"=`Estimated Total Population 2016`,
         "population_2020"=`OFM Adjusted Total Population 2020`#,
         # "population_2024"=`Estimated Total Population 2024`
         ) %>% 
  filter(County_Name=='King' |
           County_Name=='Kitsap' |
           County_Name=='Pierce' |
           County_Name=='Snohomish')
```

# Methods
**Voter turnout = ballots cast / eligible population** 
\
The main challenge is that the raw voting data is available at voting precincts, which are unique geographies that do not align with census geographies. This requires translation between the two geographies. 
\
\

1. For the 2020 election data, the numerator is calculated by weighting the ballots_cast per precinct-tract pair by the percent_total_population. These values are then summed up to the tract level. The denominator is calculated by using OFM Small Area Estimates Program (SAEP) 2020 total population estimates (TRACT), weighted by the 2020 decennial census adult 18+ population/total population ratio (TRACT). These data will be calculated based on 2020 census geographies.
2. Ava's method of calculating 2020 voter participation (Y:/VISION 2050/Data/Displacement/Displacement Index 2021/scripts/15-Voter Turnout/DataGen_15_CivicEngagement.R) and the output data set (.../Displacement Index 2021/data/15_CivicEngagement.csv)
3. We will test weighting sensitivity by changing the denominator - using the total population instead of weighting the total population by the adult 18+ population/total population ratio. 
\
\
Calculations:

1. (ballots_cast*percent_total_population) / (OFM 2020 total_population*dec. 2020 census ratio of 18+/total_population)
2.
3. (ballots_cast*percent_total_population) / OFM 2020 total_population



*DATA CAVEATS:* 

* For this exercise, it is assumed that the eligible population most closely aligns with the adult 18+ population. Although this may include individuals who are not able to legally vote (non-citizens), the alternative is using the number of registered voters (available at voting precincts), which is a subset of the adult 18+ population who are already engaged enough to have registered. While using the general 18+ population as a denominator may result in slightly lower voter turnout values (based on our working definition), using registered voters as the denominator would results in inflated voter turnout values.
* Some voters may be registered in one location, but reside in another. This could cause mismatches in the way that they are accounted for in the numerator (ballots cast) and the denominator (population). 
* There is a distinction between residential addresses (which must be in Washington State and where you are considered registered for voting purposes) and mailing addresses (which can be anywhere). Therefore, the number of registered voters refers to the number of voters who have their residential address in that precinct. This is based on correspondence with staff at the Washington Secretary of State office (8/13/2025).
* According to correspondences with staff at the Washington Secretary of State office (10/9/2025), data from smaller precincts are occasionally “masked” into a larger precinct in order to protect the privacy of voters and maintain confidentiality. It is also possible for last minuet ballots to be cast (such as by UOCAVA voters, or Military and Overseas Voters, defined by https://www.eac.gov/uocava) that must be accorded to a precinct. As a result, the voter turnout results calculated at the tract level are estimates and may reflect some margin of error. 


# 1) 2020 - 'new' methodology 
The denominator is weighted by the adult population 18+/total population ratio.
\
\
This voter participation data set was generated in the 3-voter-turnout-12-16-20-24.qmd file. It includes presidential election years (2012, 2016, 2020, and 2024). For this exercise we will be focusing on the 2020 election data so that it can be compared with the Displacement Risk Index 2021 update's 2020 election voter participation data set. 
```{r}
# download data
rawdata_folder <- 'Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data'

all_election <- read.csv(paste0(rawdata_folder,"/voter_participation.csv"))

new_adult_pop <- all_election %>% 
  select(planning_geog, pop18_turnout, election_year) %>% 
  rename(GEOID=planning_geog,
         votes=pop18_turnout,
         year=election_year) %>% 
  mutate(votes=round(votes*100, digits=2),
         year=as.character(year)) %>% 
  filter(year=="2020")

new_adult_pop$method <- "new_adult_pop"
```

## visualize data
### summary stats
```{r}
summary(new_adult_pop$votes)
```

### density plot
```{r}
new_adult_pop %>% ggplot(aes(x = votes, fill = year)) + geom_density(alpha=.2) + 
  coord_cartesian(xlim = c(0, 100)) + 
  coord_cartesian(ylim = c(0, 0.035))
```

### map
2020 census tracts
```{r}
tracts20.lyr$GEOID <- as.numeric(tracts20.lyr$geoid20)
tracts_new_adult_population <- tracts20.lyr %>% left_join(new_adult_pop, by = "GEOID")

bins <- c(0, 40, 50, 60, 70, 80, 90, 100, 102)
pal <- colorBin("YlOrRd", domain = tracts_new_adult_population$votes, bins = bins)

map_new_method_adultpop18 <- leaflet(tracts20.lyr) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = tracts_new_adult_population,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_new_adult_population$votes),
              fillOpacity = 0.7,
              popup = paste(
                "Vote percentage: ", round(tracts_new_adult_population$votes, 2))) %>% 
  addLegend(pal = pal, values = tracts_new_adult_population$votes, opacity = 0.7, title = "Turnout, Tracts",
            position = "bottomright")

map_new_method_adultpop18
```

# 2) 2020 - Ava's methodology
This voter participation data set was generated for the Displacement Risk Index 2021 update. Although it is data for the 2020 election year, it is mapped to 2010 census tracts. 
```{r}
# download data
old_2020 <- read_excel("Y:/VISION 2050/Data/Displacement/Displacement Index 2021/data/FinalIndicators.xlsx", 
                  sheet = 1, skip = 1)
old_2020 <- old_2020[,c(1,27)]
names(old_2020) <- c("GEOID", "votes")
old_2020$votes <- as.numeric(old_2020$votes)
old_2020$year <- "2020"
old_2020$method <- "old_Ava's"
```

## visualize data
### summary stats
```{r}
summary(old_2020$votes)
```

### density plot
```{r}
old_2020 %>% ggplot(aes(x = votes, fill = year)) + geom_density(alpha=.2) + 
  coord_cartesian(xlim = c(0, 100)) + 
  coord_cartesian(ylim = c(0, 0.035))
```

### map
2010 census tracts
```{r}
tracts10.lyr$GEOID <- as.numeric(tracts10.lyr$geoid10)
tracts_old_method <- tracts10.lyr %>% left_join(old_2020, by = "GEOID")

bins <- c(0, 40, 50, 60, 70, 80, 90, 100, 102)
pal <- colorBin("YlOrRd", domain = tracts_old_method$votes, bins = bins)

map_old_method <- leaflet(tracts_old_method) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = tracts_old_method,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_old_method$votes),
              fillOpacity = 0.7,
              popup = paste(
                "Vote percentage: ", round(tracts_old_method$votes, 2))) %>% 
  addLegend(pal = pal, values = tracts_old_method$votes, opacity = 0.7, title = "Turnout, Tracts",
            position = "bottomright")

map_old_method
```

# 3) 2020 - 'new' methodology adjusted
The denominator is the total population.
\
\
This voter participation data set will vary from the data set used in the first method because the denominator will not be weighted by the adult population (additional steps using the ratio of the ACS 5y 18+ population and total population numbers by tract to weight the OFM total population by tract), we will simply be dividing the ballots cast by the total population. 

## generate data set
### join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
str(df20_adult) #5,912 - voting precincts x census tracts
str(vt20_df) #4,196 - ballots cast info
sum(vt20_df$ballots_cast) #2,288,765 total number of ballots cast

join_df20 <- df20_adult %>% 
  full_join(vt20_df, 
            by = c('data_geog'='precinctcode')) #5941 records

# simplify data - remove unnecessary fields
join_df20_simp <- join_df20 %>% 
  select(-c(ofm_estimate_year,
            ofm_vintage, 
            parcel_year,
            percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units,
            precinctname,
            reg_voters))

# review data
summary(join_df20_simp$ballots_cast) #61 NAs

# calculate weighted ballots cast
df20_calc <- join_df20_simp %>%
  mutate(ballots_cast_wt = ballots_cast*percent_of_total_pop)

# check totals
sum(df20_calc$ballots_cast_wt, na.rm=TRUE) #2,288,762 #total number of ballots cast after weighting
```

### calculate voter turnout by total population
(ballots cast/total population)
\
```{r}
str(df20_calc) #5,941 - weighted ballots cast by voting precincts with associated census tracts
sum(df20_calc$ballots_cast_wt, na.rm=TRUE) #2,288,762
summary(df20_calc$ballots_cast_wt) #90 NAs
str(ofm_psrc_20) #923 - population 2020
sum(ofm_psrc_20$population_2020) #4,293,636 - total population 

# need to aggregate weighted ballots up to census tracts
df20_calc_tract <- df20_calc %>% 
  mutate(ballots_cast_wt= case_when(is.na(ballots_cast_wt)~0,
                                    TRUE~ballots_cast_wt)) %>% # needed to convert NAs to 0 so that total ballots cast would reflect correct number
  group_by(planning_geog) %>% 
  summarise(ballots_by_tract = sum(ballots_cast_wt), .groups = 'drop')

sum(df20_calc_tract$ballots_by_tract) #2,288,762

# join to ballots cast data by census tract to OFM population 
df20_calc_ofm_pop <- df20_calc_tract %>% 
  left_join(ofm_psrc_20, by = c('planning_geog'='GEOID20'))

str(df20_calc_ofm_pop)
  
# calculate voter turnout by total population
df20_tract_voter <- df20_calc_ofm_pop %>% 
  mutate(voter_turnout_tot_pop = ballots_by_tract/population_2020) 

# check
summary(df20_tract_voter$voter_turnout_tot_pop) # 0.87896 max

# rename to reflect method, transform to match other data sets
new_total_population <- df20_tract_voter %>% 
  rename(GEOID = planning_geog,
         votes = voter_turnout_tot_pop) %>% 
  mutate(year = "2020",
         method = "new_total_population",
         votes = votes*100) %>% 
  select(GEOID, votes, year, method) %>% 
  mutate(GEOID = as.numeric(GEOID))
```

## visualize data
### summary stats
```{r}
summary(new_total_population$votes)
```

### density plot
```{r}
new_total_population %>% ggplot(aes(x = votes, fill = year)) + geom_density(alpha=.2) + 
  coord_cartesian(xlim = c(0, 100)) + 
  coord_cartesian(ylim = c(0, 0.035))
```

### map
2020 census tracts
```{r}
tracts20.lyr$GEOID <- as.numeric(tracts20.lyr$geoid20)
tracts_new_total_population <- tracts20.lyr %>% left_join(new_total_population, 
                                                          by = ("geoid20"="GEOID"))

bins <- c(0, 40, 50, 60, 70, 80, 90, 100, 102)
pal <- colorBin("YlOrRd", domain = new_total_population$votes, bins = bins)

map_new_total_population <- leaflet(tracts_new_total_population) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = tracts_new_total_population,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_new_total_population$votes),
              fillOpacity = 0.7,
              popup = paste(
                "Vote percentage: ", round(tracts_new_total_population$votes, 2))) %>% 
  addLegend(pal = pal, values = tracts_new_total_population$votes, opacity = 0.7, title = "Turnout, Tracts",
            position = "bottomright")

map_new_total_population
```


# Combine data 
```{r}
all <- rbind(new_adult_pop, #weighted by pop 18+/total population
             old_2020, 
             new_total_population #weighted by total population
            )

unique(all$method)
```

## visualize data
### summary stats
```{r}
summary(new_adult_pop$votes)
summary(old_2020$votes)
summary(new_total_population$votes)
```


### density plot
```{r}
all %>% ggplot(aes(x = votes, fill = method)) + geom_density(alpha=.2) + 
  coord_cartesian(xlim = c(0, 100)) + 
  coord_cartesian(ylim = c(0, 0.035))
```


### map
```{r}
# preparing adult population data for mapping
age_18over_total_20_map <- age_18over_total_20 %>% 
  mutate(pop18_total = pop18_total*100,
         GEOID = as.numeric(GEOID))
tract_age18over <- tracts20.lyr %>% left_join(age_18over_total_20_map, 
                                              by = ("geoid20"="GEOID"))
# exploring adult population data
summary(age_18over_total_20$pop18_total)
test <- age_18over_total_20_map %>% 
  arrange(desc(pop18_total))

# preparing precinct data
precinct_20_psrc <- precinct_20 %>% 
  filter(CountyName == "King" |
           CountyName == "Kitsap" | 
           CountyName == "Pierce" |
           CountyName == "Snohomish")
```

```{r}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8

# set up palettes
bins <- c(0, 40, 50, 60, 70, 80, 90, 100, 102)
pal <- colorBin("YlOrRd", domain = all$votes, bins = bins) # palette based on combined data
pal_adult <- colorBin("YlOrRd", domain = age_18over_total_20_map$pop18_total, bins = bins) # palette based on adult pop

# map settings
vote_tract_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addLayersControl(#baseGroups = c("Proportion of Adults 18+",
                                  # "Voting Precincts 2020",
                                  # "New Method Adult Population", #basegroup=radio buttons
                                  # "Old Method",
                                  # "New Method Total Population"),
                   overlayGroups = c("Proportion of Adults 18+",
                                  "Voting Precincts 2020",
                                  "New Method Adult Population", #overlaygroup=check boxes
                                  "Old Method",
                                  "New Method Total Population"),
                   options = layersControlOptions(collapsed = TRUE)) %>%
  addPolygons(data=tract_age18over,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(age_18over_total_20_map$pop18_total),
              fillOpacity = 0.7,
              group = "Proportion of Adults 18+",
              label = ~paste0("% adults: ", round(pop18_total,2))) %>%
  addPolygons(data=precinct_20_psrc,
              stroke = T,
              opacity = 0.7,
              weight = 1,
              color = "blue",
              group = "Voting Precincts 2020",
              label = ~paste0("precinct name: ", PrecName)) %>% 
  addPolygons(data=tracts_new_adult_population,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_new_adult_population$votes),
              fillOpacity = 0.7,
              group = "New Method Adult Population",
              label = ~paste0("% participation: ", round(votes, 2))) %>%
  addPolygons(data=tracts_old_method,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_old_method$votes),
              fillOpacity = 0.7,
              group = "Old Method",
              label = ~paste0("% participation: ", round(votes, 2))) %>% 
  addPolygons(data=tracts_new_total_population,
              stroke = T,
              opacity = 1,
              weight = 0.4,
              color = "gray",
              fillColor = ~pal(tracts_new_total_population$votes),
              fillOpacity = 0.7,
              group = "New Method Total Population",
              label = ~paste0("% participation: ", round(votes, 2))) %>%

  # legends
  addLegend(pal = pal, 
            values = all$votes, 
            opacity = 0.7, 
            title = "Voter Turnout, Tracts",
            position = "bottomright") %>% 
  addLegend(pal = pal_adult, 
            values = all$votes, 
            opacity = 0.7, 
            title = "Adult Population (18+)",
            group = "Proportion of Adults 18+",
            position = "bottomright") %>% 
  
  #default hide overlap groups
  hideGroup("Old Method") %>% 
  hideGroup("New Method Total Population") %>%
  hideGroup("New Method Adult Population") %>%
  
  showGroup("Proportion of Adults 18+") %>% 
  showGroup("Voting Precincts 2020") %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon="fa-globe", title="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8); }")))

# fix the legend NA placement
css_fix <- "div.info.legend.leaflet-control br {clear: both;}" # CSS to correct spacing
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
vote_tract_map %<>% htmlwidgets::prependContent(html_fix)                   # Insert into leaflet HTML code


# print map
vote_tract_map
```

```{r}
# # save HTML map
# library(htmlwidgets) #save visuals as html
# 
# htmlwidgets::saveWidget(vote_tract_map,
#                         file='T:/60day-TEMP/Mary/Other/forBrian/voting_tract_map.html')
```


# Exploring sensitivities
Create test data set with Bainbridge precincts and tracts
```{r}
# identifying the precincts that overlap with the southern border of census tract >100% voter participation
precinct_20_bi <- precinct_20 %>% 
  filter(PrecName=="ISLAND CENTER 320" | #precinct code = KP00000320 
           PrecName=="NEW BROOKLYN 322") #precinct code = KP00000322 

# isolating the geographies of interest
geography_split_bi <- df20_adult %>% 
  filter(data_geog == "KP00000320" |
           data_geog == "KP00000322" &
           planning_geog=="53035090800") # census tract >100%

# merging geographies with voting data and population data (ACS and OFM)
str(vt20)
str(age_18over_total_20)
str(ofm_psrc_20)

voting_data_bi <- geography_split_bi %>% 
  left_join(vt20, by = c("data_geog"="PrecCode")) %>% 
  left_join(age_18over_total_20, by = c("planning_geog"="GEOID")) %>% 
  left_join(ofm_psrc_20, by = c('planning_geog'="GEOID20")) %>% 
  # simplify data frame
  select(-c(data_geog_type, planning_geog_type,
            percent_of_group_quarters_pop, percent_of_household_pop,
            percent_of_housing_units, percent_of_occupied_housing_units,
            PrecName, G20TREGVOT)) %>% 
  rename(ballots= G20TBALCST)
```


## Testing weighting

a. trial #0: sum precinct-tract > tract: (ballots_cast x percent_total_population)/ tract: (total_population x ratio 18+/totalpop) - 'new method'
b. trial #1: sum precinct-tract > tract: (ballots_cast x percent_total_population)/ tract: total_population
c. trial #2: sum precinct-tract > tract: (ballots_cast x adj_percent_total_population)/ tract: total_population

     trial_2_wt_denom: sum precinct-tract > tract: (ballots_cast x adj_percent_total_population)/ tract: (total_population x ratio 18+/totalpop)
     
d. trial #3: sum precinct-tract > tract: (ballots_cast x adj2_percent_total_population)/ tract: total_population

     trial_3_wt_denom: sum precinct-tract > tract: (ballots_cast x adj_percent_total_population)/ tract: (total_population x ratio 18+/totalpop)

[make the adjusted percent_total_population shares smaller - not more than 10% - based on the results]

### a. trial #0
```{r}
trial_0 <- voting_data_bi %>% 
  mutate(split_ballots = ballots*percent_of_total_pop) %>% 
  group_by(planning_geog, pop_18, pop18_total, population_2020) %>% #group by tract
  summarise(ballots_sum = sum(split_ballots), .groups = "drop") %>% #sum split ballots cast
  mutate(denom_pop_weighted = population_2020*pop18_total,
         trial_0_wt_denom = ballots_sum/denom_pop_weighted)

trial_0
```

### b. trial #1
```{r}
trial_1 <- voting_data_bi %>% 
  mutate(split_ballots = ballots*percent_of_total_pop) %>% 
  group_by(planning_geog, pop_18, pop18_total, population_2020) %>% #group by tract
  summarise(ballots_sum = sum(split_ballots), .groups = "drop") %>% #sum split ballots cast
  mutate(denom_pop_weighted = population_2020*pop18_total,
         trial_1 = ballots_sum/population_2020)

trial_1
```

### c. trial #2
make the 'percent_of_total_pop' values smaller by 10%
```{r}
trial_2 <- voting_data_bi %>% 
  #create adjusted 'percent_of_total_pop' values
  mutate(percent_of_total_pop=round(percent_of_total_pop,7),
         adj1_percent_of_total_pop = case_when(percent_of_total_pop==0.5350032~0.435,
                                               percent_of_total_pop==0.4974429~0.397,
                                               percent_of_total_pop==0.4649968~0.465)) %>% 
  mutate(split_ballots = ballots*adj1_percent_of_total_pop) %>% 
  group_by(planning_geog, pop_18, pop18_total, population_2020) %>% #group by tract
  summarise(ballots_sum = sum(split_ballots), .groups = "drop") %>% #sum split ballots cast
  mutate(denom_pop_weighted = population_2020*pop18_total,
         trial_2 = ballots_sum/population_2020,
         trial_2_wt_denom = ballots_sum/denom_pop_weighted)

trial_2
```

### c. trial #3
make the 'percent_of_total_pop' values smaller by 5%
```{r}
trial_3 <- voting_data_bi %>% 
  #create adjusted 'percent_of_total_pop' values
  mutate(percent_of_total_pop=round(percent_of_total_pop,7),
         adj1_percent_of_total_pop = case_when(percent_of_total_pop==0.5350032~0.485,
                                               percent_of_total_pop==0.4974429~0.447,
                                               percent_of_total_pop==0.4649968~0.415)) %>% 
  mutate(split_ballots = ballots*adj1_percent_of_total_pop) %>% 
  group_by(planning_geog, pop_18, pop18_total, population_2020) %>% #group by tract
  summarise(ballots_sum = sum(split_ballots), .groups = "drop") %>% #sum split ballots cast
  mutate(denom_pop_weighted = population_2020*pop18_total,
         trial_3 = ballots_sum/population_2020,
         trial_3_wt_denom = ballots_sum/denom_pop_weighted)

trial_3
```

### Conclusion
This exercise in testing the sensitivity of the methods, by adjusting the percent_of_total_population (the value used to split the ballots across precinct-tract pairs and aggregate to tracts), reveals that the maximum possible change is about 4% (with the 10% adjustment) or 2% (5% adjustment). This is based on the results of trial #0 and trials #2 and #3, when using the denominator total population (OFM) that's weighted by the ACS ratio of 18+ adults/total population. These values are comparing 'trial_0_wt_denom' with 'trial_3_wt_denom' and 'trial_3_wt_denom'.

When calculating the tract's total voter participation, there would be more precinct-tract pairs (7) than the 2 tested in this exploration, therefore the overall effect of these 2 precincts would be smaller than indicated here. This implies that the results using the 'new' method (weighting the denominator's total population by the ratio of 18+/total population) does shift the distribution to the right (closer to 100%), when compared to Ava's method in the 2021 update, but it doesn't fully account for the much higher participation rates.

Despite these higher estimates of participation rates, the 'new' method, which uses the total population multiplied by the ratio of adults 18+/total population in the denominator, is more intuitive because only adults 18+ are eligible to vote (even if some adults cannot legally vote or do not vote). Therefore, we will continue to calculate the numerator (ballots_cast) using 'percent_total_population' since there is no closer measure available when splitting precincts that overlap with more than one tract (other 4 options: group quarters, household population, housing units, occupied housing units) and use the adult population value in the denominator to determine participation rate.

# 'New' method process - reference 
This includes the steps used to generate the results used in Method #1 (from 3-voter-turnout-12-16-20-24.qmd file).
\
II) 2020 (OFM SAEP 2020; 2020 decennial census)
\
In 2020 census geographies

## join precinct-tract to geographies to turnout data
```{r}
nrow(df20_adult) #precinct-tract records: 5912
nrow(vt20_df) #voting data by unique precinct: 4196
vt20_df_0 <- vt20_df %>% 
  filter(reg_voters<1) # there are 30 precincts without data
# data_geog = precinct
# planning_geog = tract
length(unique(df20_adult$data_geog)) #unique precincts: 4169
length(unique(df20_adult$planning_geog)) #unique tracts: 919
```
There is a difference in the number of records - in the voting data, there are 4,196 records (precincts) while the geography split data only has 4,169 unique precincts. The difference between these two data sets is 27 precincts. There are 30 precincts in the voting data without registered voters.

## calculate total numbers for reference 
```{r}
sum(vt20_df$reg_voters) #2,706,778
sum(vt20_df$ballots_cast) #2,288,765
```
In the 2020 voting data (by precinct), there were 2,706,778 registered voters and 2,288,765 ballots cast.  

## join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
join_df20 <- df20_adult %>% 
  full_join(vt20_df, 
            by = c('data_geog'='precinctcode')) #5941 records


# simplify data - remove unnecessary fields
join_df20_simp <- join_df20 %>% 
  select(-c(percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units))

# review data
summary(join_df20_simp$reg_voters) #61 NAs
summary(join_df20_simp$ballots_cast) #61 NAs

# calculate weighted registered voters and ballots cast 
df20_calc <- join_df20_simp %>% 
  mutate(reg_voters_wt = reg_voters*percent_of_total_pop,
         ballots_cast_wt = ballots_cast*percent_of_total_pop)

# check totals
sum(df20_calc$reg_voters_wt, na.rm=TRUE) #2706774
sum(df20_calc$ballots_cast_wt, na.rm=TRUE) #2288762
```

## join tract level OFM estimates to decennial census adult 18+ ratio
```{r}
tract_population <- ofm_psrc_20 %>% 
  left_join(age_18over_total_20, by= c("GEOID20"="GEOID"))

# weighting the OFM population estimates by decennial census 18+/total proportion, 
tract_pop_18plus_20 <- tract_population %>% 
  mutate(population_18plus_20=round(as.numeric(population_2020)*pop18_total,0))
```

## calculate voter turnout by registered voters 
(ballots cast/registered voters)
\
This is not the final calculation because it is based on those who are already engaged to an extent, not the general population who are capable of voting.
```{r}
df20_calc_tract <- df20_calc %>% 
  mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
                                   TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
         ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
                                     TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog_type, planning_geog_type, 
           ofm_estimate_year, ofm_vintage, parcel_year,
           planning_geog) %>% # grouping on tract ids, retaining other fields for reference
  summarise(#reg_voters = sum(reg_voters_wt), # need to round?
            # ballots_cast = sum(ballots_cast_wt), # need to round?
            reg_voters = round(sum(reg_voters_wt), digits = 0),
            ballots_cast = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop') %>% 
  mutate(regvote_turnout = ballots_cast/reg_voters)

# calculate total numbers to check
sum(df20_calc_tract$reg_voters, na.rm=T) #2706766 when rounded
sum(df20_calc_tract$ballots_cast, na.rm=T) #2288758 when rounded
```

## calculate voter turnout by population 18+ 
(ballots cast/pop 18+)
\
This is closer to the final calculation because it takes into account all people who are capable of voting. This value may include non-citizens, or adults 18+ who are not legally allowed to vote, but it is more comprehensive than just considering those who are registered. The denominator will be based on OFM population estimates (2024), weighted by the ratio of adults 18+/total population from ACS (2023). 
```{r}
# join to precinct/voter turnout data to tract 18+ population data (OFM/ACS) to calculate population 18+ turnout
df20_calc_tract_adults_ofm <- df20_calc_tract %>% 
  left_join(tract_pop_18plus_20, by = c('planning_geog'='GEOID20')) 
  
# calculate voter turnout by population 18+
df20_tract_adult_ofm <- df20_calc_tract_adults_ofm %>% 
  mutate(pop18_turnout = ballots_cast/population_18plus_20,
         registered_pop18 = reg_voters/population_18plus_20) %>% 
  rename(total_pop_ofm=population_2020)

# check 
summary(df20_tract_adult_ofm$pop18_turnout) #1.01164 max
summary(df20_tract_adult_ofm$registered_pop18) #1.09840 max
```
There are tracts with higher than 100% voter turnout and higher than 100% registered voters. 

### Checking tracts with over 100%
#### where # of ballots cast is greater than the number of adults 18+
```{r}
df20_tract_adult_ofm_simp <- df20_tract_adult_ofm %>% 
  select(planning_geog, ballots_cast, population_18plus_20, pop18_turnout)

df20_tract_above1_adultpop_ofm <- df20_tract_adult_ofm_simp %>% 
  filter(pop18_turnout>1)

df20_tract_above1_adultpop_ofm
```
There are 2 tracts over 100% voter turnout.

##### map of tracts by voter turnout
```{r}
# join full data to tract spatial file
df20_tract_adult_ofm_geo <- merge(tracts20.lyr, df20_tract_adult_ofm,
                              by.x="geoid20",
                              by.y="planning_geog", 
                              all.x=TRUE)

df20_tract_adult_ofm_geo <- df20_tract_adult_ofm_geo %>% 
  mutate(pop18_turnout=pop18_turnout*100)

# filter to tracts >100%
df20_tract_adult_ofm_geo_100plus <- df20_tract_adult_ofm_geo %>% 
  filter(pop18_turnout>100)
```

```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = df20_tract_adult_ofm_geo$pop18_turnout)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   overlayGroups = c("tracts >100%", "voting precincts (2020)"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=df20_tract_adult_ofm_geo,
              fillColor = psrc_palette(df20_tract_adult_ofm_geo$pop18_turnout),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(df20_tract_adult_ofm_geo$pop18_turnout, digits=0),"%")) %>%
  # addPolygons(data=precinct_20, #if wanted to look at precint boundaries
  #             color = "gray",
  #             stroke=TRUE, 
  #             weight = 3,
  #             smoothFactor = 0.5,
  #             fillOpacity = 0,
  #             group = "voting precincts (2020)",
  #             label = precinct_20$St_Code) %>%
  addPolygons(data=df20_tract_adult_ofm_geo_100plus,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "tracts >100%",
              label = paste(round(df20_tract_adult_ofm_geo_100plus$pop18_turnout, digits=0),"%")) %>% 
   

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = df20_tract_adult_ofm_geo$pop18_turnout,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

##### top 10 tracts
Looking at the ten tracts with the highest voter turnout
```{r}
topten <- df20_tract_adult_ofm %>% 
  slice_max(order_by = pop18_turnout, n = 10) %>% 
  select(planning_geog, pop18_turnout)

topten
```

#### where # of registered voters is greater than the number of adults 18+
```{r}
df20_tract_adult_ofm_simp_reg <- df20_tract_adult_ofm %>% 
  select(planning_geog, reg_voters, population_18plus_20, registered_pop18)

df20_tract_above1_adultpop_reg <- df20_tract_adult_ofm_simp_reg %>% 
  filter(registered_pop18>1)

df20_tract_above1_adultpop_reg
```
There are 58 tracts over 100% where the number of registered voters is greater than the number of adults 18+.