---
title: "Homeownership Rates"
subtitle: "Visuals for Equity Tracker (PUMS data)"
author: "Meg Grzybowski, updated by Grant Gibson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'e-housing',
                   ind_dir = 'e03-homeownership-rates',
                   map = 'e03-homeownership-rates-map-data',
                   chart = 'e03-homeownership-rates-chart-data',
                   map_name = 'e03-homeownership-rates-map.html',
                   column_chart_name = 'e03-homeownership-rates-column',
                   line_chart_name = 'e03-homeownership-rates-line')

# set most recent year
year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Homeownership rates are directly related to income, but can also be related to indicators regardless of income. Discrimination and discriminatory policies of the past have created generational wealth gaps and disparities amongst communities, leading to a need for intentional and active systemic changes.

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# move the decimal point for reg_estimate and cnty_estimate
# acs_data_tract <- acs_data_tract %>%
#    mutate(reg_estimate = (reg_estimate*100)) %>%
#    mutate(cnty_estimate = (cnty_estimate*100))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 98.81, so rounding to 100 makes sense
max_value <- 100 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
acs_data_plus_zero<-append(acs_data_tract$estimate,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Homeownership <br> Rates" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", 
        prettyNum((round(acs_data_tract$estimate, digits=0))), " %"), 
  paste("Tract data reliability: ", acs_data_tract$reliability),
  paste("Region value: ",
        prettyNum((round(acs_data_tract$reg_estimate, digits=0))), " %"), 
  paste("County value: ",
        prettyNum((round(acs_data_tract$cnty_estimate, digits=0))), " %"), 
  paste0("(",acs_data_tract$county_name," County)")) %>%
  lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", acs_data_tract$geoid20,
#   paste(prettyNum((round(acs_data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)


# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))
# 53061050600 (Woodway and Edmonds Waterfront), 53061052112 (Maltby), and 53061041607 (east of Silver Lake) in Snohomish County, 53053070314 (Lake Tapps) in Pierce County, and 53033032328 (Sammamish) in King County
# Snohomish and King County are top 5 highest home-owning counties

# national average
# acs table: S2201, https://data.census.gov/table/ACSST5Y2022.S2201?q=S2201

sort <- data %>% 
  dplyr::arrange(estimate)
# 53033032331 (Downtown Redmond), 53033005305 (Seattle UW), 53053061601 (Tacoma)
# 53033030501 (Auburn, around 167)
```

1. 61%: The region's average rate of homeownership
2. 21%: Regionally, households with limited English proficiency have homeownership rates 21 percentage points lower than households proficient in English
3. 13%: The regional homeownership gap between households with lower income and other households decreased 13% between 2012 and 2022

\

#### Insights & Analysis

* Kitsap County has the highest homeownership rates (70%), followed by Snohomish (69%), Pierce (65%), and King (56%)
* The five census tracts tied with the highest rates of homeownership (98-99%) are in Snohomish County (Woodway and Edmonds Waterfront, Maltby, east of Silver Lake), Pierce County (Lake Tapps), and King County (Sammamish)
* Snohomish County is the only county that does not have any census tracts with 0% homeownership, while there are 4 tracts in King, 2 in Kitsap, and 2 in Pierce. The census tract with the lowest homeownership rate in Snohomish (14%) is in Everett, east of Paine Field.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = 1
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Homeownership Rates"
subtitle = "Percent of households living in owner-occupied housing units"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")

# region low income households: 0.34	
# region non-low income households: 0.67
# 67-34= 33%

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat",
         county=="Region")
# region lep households: 0.41	
# region non-lep households: 0.62
# 62-41= 23%

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat",
         vulnerability=="high vulnerability")

# region lep households: 41%	
# kitsap lep households: 60%
# 21%

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Youth_cat")

filter <- data_clean_column %>% 
  filter(focus_type=="POC_cat")

# king 16 pt diff
# kitsap 16 pt diff
# pierce 20 pt diff
# snohomish 10 pt diff
# region 16 pt diff

filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# kitsap dis: 84171.34
# kitsap non-dis: 104284.10
# 104284.10-84171.34=20112.76

# king dis: 81508.16
# king non-dis: 134637.12
# 134637.12-81508.16=53128.96

filter <- data_clean_column %>% 
  filter(focus_type=="Income_cat",
         vulnerability=="high vulnerability") %>% 
  dplyr::arrange(desc(fact_value))
```
\

#### Insights & Analysis

* The regional homeownership rate for households with lower income (below 200% federal poverty level) were highest in Kitsap County (44%) followed by Snohomish (39%), Pierce (38%), and King (29%)
* The largest difference in homeownership rates between people of color and white non-Hispanic people is in Pierce County (20 percentage points), while the smallest difference is in Snohomish County (10 percentage points)
* King County is the only county where limited English proficient households have lower homeownership rates (39%) than the region’s homeownership rate for households with limited English proficiency (41%)


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "Homeownership Rates"
subtitle = "Percent of households living in owner-occupied housing units"
source = "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Region",
         focus_type=="Income_cat")
# 2022 low income: 34
# 2022 non-low income: 67
# 67-34=33
# 2012 low income: 35
# 2012 non-low income: 73
# 73-35=38
# ((33-38)/(38))*100: -13% change


filter <- data_clean %>%
  filter(county=="Region",
         focus_type=="POC_cat")
# 2021 poc: 102481.29
# 2021 non-poc: 118053.44
# 118053.44-102481.29=15572.15

filter <- data_clean %>%
  filter(county=="Region",
         focus_type=="Older_cat")
# 2022 older: 76
# 2022 non-older: 55
# 76-55 = 21
# 2012 older: 78
# 2012 non-older: 61
# 78-61 = 17
# ((21-17)/(17))*100: 23.5% change

filter <- data_clean %>%
  filter(county != "Region",
         focus_type=="Disability_cat")
# 2012 king disability-no disability: 58-65=-7
# 2022 king disability-no disability: 54-57=-3
# 2012 kitsap disability-no disability: 64-69=-5
# 2022 kitsap disability-no disability: 68-71=-3
# 2012 pierce disability-no disability: 65-64=1
# 2022 pierce disability-no disability: 63-64=-1
# 2012 snohomish disability-no disability: 60-69=-9
# 2022 snohomish disability-no disability: 64-70=-6

filter <- data_clean %>%
  filter(focus_type=="POC_cat")
# 2012 king:69-49=20
# 2022 king:62-46=16
# ((16-20)/(20))*100: -20% change
# 2012 kitsap:70-53=17
# 2022 kitsap:73-57=16
# ((16-17)/(17))*100: -6% change
# 2012 pierce:68-48=20
# 2022 pierce:70-50=20
# ((20-20)/(20))*100: 0% change
# 2012 snohomish:68-55=13
# 2022 snohomish:71-61=10
# ((10-13)/(13))*100: -23% change

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="Income_cat",
         vulnerability=="low vulnerability",
         data_year=="2011" |
           data_year=="2021")
# king non-low income:150511.89-98506.82=52005.07
# Snohomish non-low income:130900.06-97445.42=33454.64
```
\

#### Insights & Analysis

The difference in homeownership rates for people of color and white non-Hispanic people over time has shifted the greatest in King and Snohomish counties (-4 percentage points), while decreasing slightly in Kitsap County (-1 percentage point), and remaining unchanged in Pierce County
* Between 2012 and 2022, the lowest homeownership rates for people of color have consistently been in King and Pierce counties (with a slight increase in Pierce in 2022), while the highest rates have been in Snohomish County
* The difference in homeownership rates of people with a disability compared to those living without has narrowed between 2012 and 2022 in all counties except Pierce, where there has been almost no difference


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
Once the data and planning reviews are complete, this step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>