---
title: "Median Household Income"
subtitle: "Visuals for Equity Tracker (PUMS data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'f-economy',
                   ind_dir = 'f01-median-income',
                   map = 'f01-median-income-map-data',
                   chart = 'f01-median-income-chart-data',
                   map_name = 'f01-median-income-map.html',
                   column_chart_name = 'f01-median-income-column',
                   line_chart_name = 'f01-median-income-line')

# set most recent year
year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
[Please include the following for a general, layperson audience: 1-2 sentences explaining what your indicator is/measures, what it says about people's life outcomes; 1 sentence describing why it is useful/relevant in the indicator's theme]

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# move the decimal point for reg_estimate and cnty_estimate
# acs_data_tract <- acs_data_tract %>%
#    mutate(reg_estimate = (reg_estimate*100)) %>%
#    mutate(cnty_estimate = (cnty_estimate*100))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 250001, so rounding to 260000 makes sense
max_value <- 260000 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
acs_data_plus_zero<-append(acs_data_tract$estimate,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Median Income" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ","$", # if doesn't require prefix, this can be removed
        prettyNum((round(acs_data_tract$estimate, digits=-2)),
                  big.mark = ",")),
  paste("Tract data reliability: ", acs_data_tract$reliability),
  paste("Region value: ","$",
        prettyNum((round(acs_data_tract$reg_estimate, digits=-2)),
                  big.mark = ",")),
  paste("County value: ","$",
        prettyNum((round(acs_data_tract$cnty_estimate, digits=-2)),
                  big.mark = ","), paste0("(",acs_data_tract$county_name," County)"))) %>%
  lapply(htmltools::HTML)

# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = acs_data_plus_zero,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(prefix = "$"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>%
  select(county_name, geoid20, estimate, cnty_estimate, reg_estimate) %>% 
  dplyr::arrange(desc(estimate))

filter <- sort %>% 
  filter(estimate > 250000)
# Highest median income is 250001 (9 tracts in King County)	- 53033004101 (Laurelhurst), 53033023902 (southwest Bellevue along the lakeshore), 53033024100 (Clyde Hill/Hunts Point/Yarrow Point), 53033024500 (Mercer), 53033024601 (Mercer), 53033024602 (Mercer), 53033024905 (Hilltop neighborhood in Bellevue), 53033032215 (east Sammamish), 53033032321 (Woodinville and Cottage Lake)

sort <- data %>% 
  select(county_name, geoid20, estimate, cnty_estimate, reg_estimate) %>% 
  dplyr::arrange(estimate)
# 2 census tracts below $30k in King: 53033005304 (19375) and 53033005305 (23201) - UW
# 5 census tracts $30-40k in King (4) and Pierce (1): 53033030501 (30327) - Auburn, 53053061601 (32740) - Tacoma, 53033005303 (35875) - UW, 53033009100 (36857) - Chinatown/ID, 53033005307 (37989)  

# national average
# acs table: B19013, https://data.census.gov/table?q=B19013 
```

1. \$106,100: Regional median household income
2. \$49,900: Households with limited English proficiency in the region have median income that is $52,300 lower than households proficient in English
3. 18%: Regionally, people of color have median household income $12,800 lower than white non-Hispanic people, a gap which has decreased 18% in the last ten years

\

#### Insights & Analysis

* King County has the highest median household income (\$116,300), followed by Snohomish (\$104,100), Kitsap (\$93,700), and Pierce (\$91,500).

* Highest median household income at the census tract level is \$250,000.

* Nine census tracts tied with the highest median household income ($250,000) are all in King County. Six are around Lake Washington: one tract in southwest Bellevue along the lakeshore, one in Clyde Hill/Hunts Point/Yarrow Point, three on the southern half of Mercer Island, and one in Seattle’s Laurelhurst neighborhood. The remaining three are in the Hilltop neighborhood in Bellevue, east Sammamish, and between Woodinville and Cottage Lake. 

* Lowest median household income at the census tract level is $19,400.

* Seven census tracts with median household incomes below \$40,000 are in Seattle's University District (\$19,400 and \$23,200), Auburn along Route 167 (\$30,300), Downtown Tacoma (\$32,700), around Seattle's University of Washington campus (\$35,900 and \$38,000), and in Seattle's Chinatown and International District (\$36,900).

* The national median household income is \$77,700.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = 180000
dec = 0
esttype = "currency"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Household Income by Community"
subtitle = "Median values adjusted to 2022 dollars"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# Region: 0.21 vs. 0.06
# with disability has higher coverage across all counties/region

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat")
# King lep households: .22	
# King non-lep households: .07

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")
# region low income households: .32	
# region non-low income households: .04
# .32-.04=.28

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="LEP_cat")
# region non-lep:0.09
# region lep: 0.21
# 0.21/0.09=1.857143

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="POC_cat")
# region non-poc:0.07
# region poc: 0.13
# 0.13/0.07=2.3333
```
\

#### Insights & Analysis

* Regional median household income for those living below 200% of federal poverty level (i.e., households with lower income) is \$23,200.

* Regional median household income for those with youth is $38,700 higher than those without.

* Snohomish County has the smallest difference in median household income by race and ethnicity, where white non-Hispanic households earn \$1,600 more than households of color, while Pierce County has the largest difference (\$18,100) with white non-Hispanic households earning more than households of color.

* For households with lower income, the median household income is lowest in King County (\$22,100), followed by Snohomish (\$23,800), Kitsap (\$24,000), and Pierce (\$25,300).

* The largest difference in median household income by disability status is in King County, where people with a disability earn \$50,000 less than those without, while the smallest difference is in Kitsap County (\$16,500).

* For households with limited English proficiency, Snohomish County is the only county where the median income (\$64,000) is higher than the region’s median (\$56,900).


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "Household Income Trend by Community"
subtitle = "Median values adjusted to 2022 dollars, in 5-year spans between 2012 and 2022"
source = "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# king 2012: 0.17-0.07=0.10
# king 2022: 0.12-0.06=0.06
# kitsap 2012: 0.19-0.10=0.09
# kitsap 2022:0.14-0.08=0.06
# pierce 2012: 0.19-0.10=0.09
# pierce 2022:0.17-0.10=0.07
# snohomish 2012: 0.16-0.10=0.06
# snohomish 2022: 0.11-0.08=0.03

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="LEP_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# king 2022-2012: 0.22-0.24=-0.02
# kitsap 2022-2012: 0.13-0.18=-0.05
# pierce 2022-2012: 0.23-0.21=0.02
# snohomish 2022-2012: 0.18-0.26=-0.08

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Disability_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# King 2022-2012: 0.21-0.22=-0.01
# Kitsap 2022-2012: 0.20-0.20=0
# Pierce 2022-2012: 0.22-0.22=0
# Snohomish: 0.19-0.23=-0.04

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# King 2012 poc-non poc: 0.17-0.07=0.10
# King 2022: 0.12-0.06=0.06
# King: ((0.06-0.10)/(0.10))*100: -40% change
# Kitsap 2012: 0.19-0.10=0.09
# Kitsap 2022: 0.14-0.08=0.06
# Kitsap: ((0.06-0.09)/(0.09))*100: -33% change
# Pierce 2012:0.19-0.10=0.09
# Pierce 2022: 0.17-0.10=0.07
# Pierce: ((0.07-0.09)/(0.09))*100: -22% change
# Snohomish 2012: 0.16-0.10=0.06
# Snohomish 2022: 0.11-0.08=0.03 
# Snohomish: ((0.03-0.06)/(0.06))*100: -50% change


filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="Income_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# 2012: 0.35
# 2022: 0.32


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year=="2016" |
           data_year=="2021")


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="Income_cat",
         vulnerability=="low vulnerability",
         data_year=="2016" |
           data_year=="2021")

```
\

#### Insights & Analysis

* In 2022, households in the region with lower income made $103,200 less than other households, a 30% larger gap than in 2012.

* The regional median household income for those with older adults is $38,500 lower than those without in 2022, a 33% larger gap than in 2012.

* The differences in median household income between people of color and white non-Hispanic people from 2012 to 2022 has changed across counties: decreasing in King County (-34%) and Snohomish County (-84%), and increasing in Kitsap County (+1%) and Pierce County (+27%).

* The median household income for households with lower income stayed relatively static over time, while the median household income for other households increased between 2012 and 2022, especially in King County (increased by \$28,500) and Snohomish County (increased by \$20,600).

* The difference in median household income for those with a disability compared to those without between 2012 and 2022 decreased in Kitsap County (-21%) slightly increased in Pierce and Snohomish counties (+7% and +13%, respectively), and increased to a larger extent in King County (+36%), where the disparity increased by \$13,200.


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
Once the data and planning reviews are complete, this step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>