---
title: "Flood Risk by Census Tract, 2022-2052"
subtitle: "Visuals for Equity Tracker (tract data)"
author: "Kristin Mitchell"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
    word_document:

---
  
```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
#output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'c-climate-change',
                   ind_dir = 'c01-flood-risk',
                   map = 'c01-flood-risk-map-data',
                   chart = 'c01-flood-risk-chart-data',
                   map_name = 'c01-flood-risk-map.html',
                   column_chart_name = 'c01-flood-risk-column',
                   line_chart_name = 'c01-flood-risk-line')

# set most recent year
year <- "2021"
# years <- c(2012, 2017, 2022)

# set past version year - this will set organize archived files in Y drive
past_year <- "2021"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time. 
\
\
**If the indicator is available through a tract-level data set.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-tract-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Flood risk scores are from the [First Street Foundation Flood Model](https://firststreet.org/risk-factor/flood-factor/). Understanding where flood risk occurs and may occur in the future with a changing climate can help communities better prepare for flood risk. 

First Street Foundation determines the flood risk score by combining hydraulic and hydrology models along with climate change projections. This creates a risk score range from 1 to 10. More information on the [methodology can be found here](https://firststreet.org/methodology/flood/).

## 1. Map of most recent data
To map data in this form, there should be a value corresponding to each census tract. Depending on the year or source of the data, this could be about 700 rows for data at the 2010 census tract resolution, or about 900 rows for data at the 2020 census tract resolution.

### Create Visual
```{r map .rda}

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r tract data map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(data_tract$estimate) # max is 9.13, so rounding to 10 makes sense
max_value <- 10 # this value will be the max value in the legend

# set up palettes
# psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
#                                       domain = data_tract$estimate)

psrc_purple_plus<-c("#FFFFFF","#FFFFFF","#F6CEFC", psrc_colors$purples_inc)
tract_data_plus_zero<-append(data_tract$estimate,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                    domain = c(min(tract_data_plus_zero,na.rm = TRUE),max_value, NA))# this is the value field

# acs_data_plus_zero<-append(data_tract$estimate,0)
 
# psrc_palette <- leaflet::colorNumeric(palette = psrc_colors$purples_inc,
#                                       domain = acs_data_plus_zero)

# set the variable
var_name <- "Flood Risk <br> Score"
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>", data_tract$geoid20, # the code in <> makes the geoid value italic 
  paste("Tract value: ", prettyNum((round(data_tract$estimate, digits=1))), " score"),
  paste("Region value: ", prettyNum((round(data_tract$reg_estimate, digits=1))), " score"),
  paste("County value: ", prettyNum((round(data_tract$cnty_estimate, digits=1))), 
        paste0("(",data_tract$county_name," County)"))) %>% 
  lapply(htmltools::HTML)

# if the label required a prefix, for example a $, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$Census.Tract, # the code in <> makes the geoid value italic
#   paste("$",
#         prettyNum((round(data_tract$estimate, digits=-2)),
#                   big.mark = ","))) %>%
#   lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$Census.Tract,
#   paste(prettyNum((round(data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(tract_data_plus_zero),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              # label = round(data_tract$estimate, digits=1),
              label = labels) %>%
  
  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(tract_data_plus_zero,na.rm = TRUE),max_value, NA),
                       bins = 5,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```
Sources: First Street Foundation Flood Model, 2022; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles


```{r generate tract map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\c-climate-change\c01flood-risk\webpage-html-outputs

# set folder structure
map_name <- 'c01-flood-risk-map.html'

# interactive webpage output process
htmlwidgets::saveWidget(tract_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\

#### Data call outs
```{r map calculations - tract, include=FALSE, eval=FALSE}
# simplify data set
data <- data_tract %>%
  st_drop_geometry()

# calculate call out information
calc <- data %>%
  mutate(average=(sum(estimate*prop_w_score, na.rm =TRUE))/(sum(prop_w_score, na.rm=TRUE)))


x<-rep(data$estimate,times=data$prop_w_score)
df1<-data.frame(x)
summary(df1$x)
median(df1$x, na.rm=TRUE)
hist(df1$x)

# 1  Min. 1st Qu.  Median  Mean   3rd Qu.  Max. 
#   1.000   1.175   1.350   1.599   1.610   9.132 

sort <- data %>% 
  dplyr::arrange(desc(estimate))

# 2 highest: 53053073407 (Puyallup along the Puyallup River) Pierce (9.132), 53033029205 (west Kent just south of Green River NRA) King (7.636), 53033029206 (west Kent including Green River NRA) King (7.539), 53053073408 (Puyallup along the Puyallup River) Pierce (7.135)

sort <- data %>% 
  dplyr::arrange(estimate)

# 3 lowest: there are 25 census tracts with a score of 1.000

by_county <- data %>% 
  dplyr::group_by(County.Name) %>% 
  dplyr::summarise(average=(sum(estimate*prop_w_score, na.rm =TRUE))/(sum(prop_w_score, na.rm=TRUE))) %>% 
  dplyr::arrange(desc(average))

# King (1.504131), Snohomish (1.382133), Kitsap (1.434871), Pierce (1.825793)

```

1. 1.6: The region's average flood risk score
2. Top 20 census tracts: out of the 20 census tracts with the highest flood risk scores, 10 are in King County and 10 are in Pierce County
3. 2%: the region's population that live in the top 20 census tracts with the highest flood risk scores
4. 5: number of rivers that exit into the Puget Sound in the Puget Sound region - the Duwamish, Nisqually, Puyallup, Snohomish, and Stillaguamish rivers
\

#### Insights & Analysis

* The four census tracts with the highest average flood risk score are all adjacent to either the Puyallup or Green rivers
* The census tracts with the highest flood risk score are along the Puyallup river in Puyallup, Pierce County (9.132); Kent just south of the Green River Natural Resources Area (NRA), King County (7.636); Kent including the Green River NRA, King County (7.539); again along the Puyallup River in Puyallup, Pierce County (7.135)
* Pierce County has the highest average flood risk score (1.9), followed by King (1.5), Kitsap (1.5), and Snohomish (1.5)

\
\

## 2. Facet of most recent data
### Create Visual
<style type="text/css">
  .main-container {
    max-width: 100% !important;
    margin: auto;
  }
</style>
  
```{r load rda for chart, include=FALSE}
# load visualization functions for charts
# source('C:/Coding/equity-tracker/data-visualization/equity-tracker-chart-functions.R') # this needs to be adjusted based on your local GitHub directory - it is required if you wish to run chunks and view the visuals (without first knitting)
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - tract}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county_ord"
x = "quintile_ord"
y = "estimate"
facet = "equity_group_ord"
fill = "quintile_ord"
y_min = 0 
y_max = 3
dec = 1
esttype = "number"
color = "oranges"
num_colors = 5
color_rev = FALSE
title = "Flood Risk by Community"
subtitle = "Projected likelihood of flooding in next 30 years, score out of 10"
#source = paste("First Street Foundation Flood Model, 2022;", "U.S. Census Bureau, 2017-2021 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002", sep="\n")

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color, num_colors = num_colors,
                                            color_rev = color_rev)

```

```{r output final column chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\c-climate-change\c01-flood-risk\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\c-climate-change\static-images

# set folder structure
file_name <- 'c01-flood-risk-column'

# column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
  `r title`
</p>
  <p style="font-family: Poppins; font-size:12pt; font-weight:normal">
  `r subtitle`
</p>
```{r column chart output - tract, echo=FALSE}

if (output_type == "html") {
  
  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
 
</p> 
  
```{r column facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         equity_group=="poc_quintile")

# region low quintile poc households: 1.771193
# region high quintile poc households: 1.557723
# 1.557723-1.771193=-0.21347

filter <- data_clean_column %>% 
  filter(quintile=="High",
         equity_group=="income_quintile")

# region high quintile low income households: 
# king high quintile low income households: 
# kitsap high quintile low income households: 
# pierce high quintile low income households: 
# snohomish high quintile low income households: 
#  - =

filter <- data_clean_column %>% 
  filter(equity_group=="lep_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for lep: 

filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for income: 
# king dif bt high/low quintiles for income: 
# kitsap dif bt high/low quintiles for income: 
# pierce dif bt high/low quintiles for income: 
# snohomish dif bt high/low quintiles for income: 

filter <- data_clean_column %>% 
  filter(equity_group=="disability_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for disability: 
# king dif bt high/low quintiles for disability: 
# kitsap dif bt high/low quintiles for disability: 
# pierce dif bt high/low quintiles for disability: 
# snohomish dif bt high/low quintiles for disability: 

filter <- data_clean_column %>% 
  filter(county!="Region",
         equity_group=="poc_quintile",
         quintile=="Low"|
           quintile=="High")
# king dif bt high/low quintiles for income: 
# kitsap dif bt high/low quintiles for income: 
# pierce dif bt high/low quintiles for income: 
# snohomish dif bt high/low quintiles for income: 

```
\

#### Data Call Outs
1. 30% higher: As a region, communities with high concentrations of households with lower income have about a 30% higher average flood risk score
2. 30% higher: As a region, communities with high concentrations of people with disabilities have about a 30% higher average flood risk score
2. 15% lower: As a region, communities with high concentrations of people of color have about a 15% lower average flood risk score


#### Insights & Analysis

* In King County, communities with high concentration of households with people with disabilities have a flood risk score almost 40% higher than communities with a low concentration
* In Pierce County, communities with high concentration of households with people with disabilities have a flood risk score almost 20% higher than communities with a low concentration
* In Snohomish County, communities with high concentration of people of color have a flood risk score 40% lower than communities with a low concentration
\
\

## 3. Facet of trend data
### Create Visual
<style type="text/css">
  .main-container {
    max-width: 100% !important;
    margin: auto;
  }
</style>
  
```{r line facet - tract, include=FALSE}
### Placeholder language for future use/updating when multiple years are available 

# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "quintile_ord" # different from variables for facet column
y_min = 0
y_max = 3
color = "blues"
title = "Flood Risk Trend by Community"
subtitle = "Projected likelihood of flooding in next 30 years, score out of 10, in 5-year spans between 2010 and 2020"
#source = paste("First Street Foundation Flood Model, 2022;", "U.S. Census Bureau, 2006-2010, 2011-2015, 2016-2020 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002", sep="\n")

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\c-climate-change\c01-flood-risk\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\c-climate-change\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
  `r title`
</p>
  <p style="font-family: Poppins; font-size:12pt; font-weight:normal">
  `r subtitle`
</p>
```{r line chart - tract, echo=FALSE}
if (output_type == "html") {
  
  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
  
</p> 
  
```{r line facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High",
           data_year=="2022")

# 2020 region low quintile poc households: 1.771193
# 2020 region high quintile poc households: 1.557723
# 1.557723-1.771193=-0.21347

filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2022")

# 2020 region low quintile income: 2.882521
# 2020 region high quintile income: 5.352679
# 5.352679-2.882521=2.470158

filter <- data_clean %>% 
  filter(county=="Snohomish",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High")

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2020")

# 2020 King dif low/high income: 5.083316-2.306517=2.776799
# 2020 Kitsap dif low/high income: 5.906103-4.232797=1.673306
# 2020 Pierce dif low/high income: 5.911497-3.856045=2.055452
# 2020 Snohomish dif low/high income: 5.306512-3.715537=1.590975

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="disability_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2020")

# 2020 King dif low/high disability: 4.912824-2.659749=2.253075
# 2020 Kitsap dif low/high disability: 5.468296-3.715005=1.753291
# 2020 Pierce dif low/high disability: 5.547184-3.812307=1.734877
# 2020 Snohomish dif low/high disability: 5.137279-3.679690=1.457589

```



\
\

# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. *Always run this chunk of code. If this is the first time you are generating visuals for an indicator, comment out the 'update' part of the 'y.drive.folder.update' variable because this is the first version that is being created.*
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update' #comment out if this is the first time you are generating visuals
                                   )

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r to archive, include=FALSE, eval=FALSE}
# add previous year folder to archive folder - to retain and organize all archived files
dir.create(file.path(file_names$base_dir,
                     file_names$theme_dir,
                     file_names$ind_dir,'archive', past_year))

# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive', past_year)

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall). *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage. *Always run this chunk of code.*
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>