---
title: "Safety Data: Serious Injuries and Fatalities"
subtite: "Data Gen: Exploring, Cleaning, Transforming (tract data)"
author: "Kristin Mitchell"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Crash Data: Serious Injuries and Fatalities**

# Download Data
Data comes from Washington Traffic Safety Commission. The data includes number of serious injuries and number of fatalities by census tract from 2010 to 2022. PSRC processes the data including pulling the census tract population using Census Bureau Five Year Estimates for the matching data year.

```{r, include=FALSE}
folder_path <- "Y:/Equity Indicators/tracker-webpage-content/g-transportation/g01-safety/raw-data"
c <- read.csv(file.path(folder_path,"serious_fatal_data_2010_to_2021_raw.csv")) %>%
  rename(data_year = year) 

c$geoid <- as.character(c$geoid)

fat_inj_2010 <- c %>% filter(data_year == 2010)
fat_inj_2011 <- c %>% filter(data_year == 2011)
fat_inj_2012 <- c %>% filter(data_year == 2012)
fat_inj_2013 <- c %>% filter(data_year == 2013)
fat_inj_2014 <- c %>% filter(data_year == 2014) 
fat_inj_2015 <- c %>% filter(data_year == 2015)
fat_inj_2016 <- c %>% filter(data_year == 2016)
fat_inj_2017 <- c %>% filter(data_year == 2017)
fat_inj_2018 <- c %>% filter(data_year == 2018)
fat_inj_2019 <- c %>% filter(data_year == 2019)
fat_inj_2020 <- c %>% filter(data_year == 2020)
fat_inj_2021 <- c %>% filter(data_year == 2021)

```
\
## Additional data

### equity quintile data and 2010-20 crosswalk file 

* low, low medium, etc.) at the tract-level for each equity focus group for the years that correspond to the serious injury and fatality data (2010-2022)
```{r, include=FALSE}
# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
# equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares") %>%
#   filter(data_year > "2009", data_year < "2022")
# table(equity_tracts$data_year)
# 
# crosswalk_10_20 <- get_table(schema="census",
#                              tbl_name="v_geo_relationships_tracts")
# 
# e <- left_join(c, equity_tracts, by=c("geoid", "data_year")) %>%
#   mutate(data_year=as.character(data_year))

# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares")
table(equity_tracts$data_year)

crosswalk_10_20 <- get_table(schema="census",
                             tbl_name="v_geo_relationships_tracts")
```

### tract-level population data for 2021
```{r, include=FALSE}
# getting census population data by tract
tracts_10_21 <- get_acs_recs(geography ='tract', 
                                table.names = 'B01003',
                                years = c(2010:2021),
                                acs.type = 'acs5')

tract.10 <- tracts_10_21 %>% 
  filter(year==2010)

tract.11 <- tracts_10_21 %>% 
  filter(year==2011)

tract.12 <- tracts_10_21 %>% 
  filter(year==2012)

tract.13 <- tracts_10_21 %>% 
  filter(year==2013)

tract.14 <- tracts_10_21 %>% 
  filter(year==2014)

tract.15 <- tracts_10_21 %>% 
  filter(year==2015)

tract.16 <- tracts_10_21 %>% 
  filter(year==2016)

tract.17 <- tracts_10_21 %>% 
  filter(year==2017)

tract.18 <- tracts_10_21 %>% 
  filter(year==2018)

tract.19 <- tracts_10_21 %>% 
  filter(year==2019)

tract.20 <- tracts_10_21 %>% 
  filter(year==2020)

tract.21 <- tracts_10_21 %>% 
  filter(year==2021)

```

### tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
tracts10.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)
tracts10.lyr<-st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

# Explore data
The safety data are available in 2020 geographies.
```{r, include=FALSE}
# check number of rows in data sets
# nrow(fat_inj_2011) #
nrow(fat_inj_2021) #923
```
The 2020 data set that could be used without overlapping years include 1,465 records each. This closely matches the number of records in the 2010 census tract shapefile for WA state, published by the [US Census Bureau, Geography Division](https://www.census.gov/cgi-bin/geo/shapefiles/index.php) (1,458; in comparison to the 1,784 tracts in the 2020 shapefile).

```{r, include=FALSE}
# to check the field names and data class
str(fat_inj_2021)

# add column with the last year of the 5-year span
# filter to PSRC counties
fat_inj_2010 <- fat_inj_2010 %>% 
  mutate(year=2010) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2011 <- fat_inj_2011 %>% 
  mutate(year=2011) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2012 <- fat_inj_2012 %>% 
  mutate(year=2012) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2013 <- fat_inj_2013 %>% 
  mutate(year=2013) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2014 <- fat_inj_2014 %>% 
  mutate(year=2014) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2015 <- fat_inj_2015 %>% 
  mutate(year=2015) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2016 <- fat_inj_2016 %>% 
  mutate(year=2016) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2017 <- fat_inj_2017 %>% 
  mutate(year=2017) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2018 <- fat_inj_2018 %>% 
  mutate(year=2018) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2019 <- fat_inj_2019 %>% 
  mutate(year=2019) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2020 <- fat_inj_2020 %>% 
  mutate(year=2020) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
fat_inj_2021 <- fat_inj_2021 %>% 
  mutate(year=2021) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
```

## Most Recent Data
### Map
#### Join recent data to spatial file
```{r, include=FALSE}
# join geographic data to safety data 
data_tract <- merge(tracts20.lyr, fat_inj_2021,
                    by.x="geoid20",
                    by.y="geoid", 
                    all.x=TRUE)

tract_acs_21data <- get_acs_recs(geography ='tract', table.names = 'B01003', years = c(2021), acs.type = 'acs5')
  
tract_acs_21data <- tract_acs_21data %>%
  rename(total_pop21=estimate) %>%
  rename(geoid20=GEOID)
 
tract_acs_21data <- tract_acs_21data %>%
  dplyr::select(geoid20, total_pop21)

data_tract <- left_join(data_tract, tract_acs_21data, 
                        by="geoid20")

data_tract <- data_tract %>% 
  dplyr::mutate(estimate=as.numeric(combo_rate)) # rename value variable for consistency
```

```{r}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - to get legend high-low with correct color order
addLegend_decreasing <-
  function (map,
            position = c("topright", "bottomright", "bottomleft",
                         "topleft"),
            pal,
            values,
            na.label = "NA",
            bins = 7,
            colors,
            opacity = 0.5,
            labels = NULL,
            labFormat = labelFormat(),
            title = NULL,
            className = "info legend",
            layerId = NULL,
            group = NULL,
            data = getMapData(map),
            decreasing = FALSE) {
    position <- match.arg(position)
    type <- "unknown"
    na.color <- NULL
    extra <- NULL
    if (!missing(pal)) {
      if (!missing(colors))
        stop("You must provide either 'pal' or 'colors' (not both)")
      if (missing(title) && inherits(values, "formula"))
        title <- deparse(values[[2]])
      values <- evalFormula(values, data)
      type <- attr(pal, "colorType", exact = TRUE)
      args <- attr(pal, "colorArgs", exact = TRUE)
      na.color <- args$na.color
      if (!is.null(na.color) &&
          col2rgb(na.color, alpha = TRUE)[[4]] ==
          0) {
        na.color <- NULL
      }
      if (type != "numeric" && !missing(bins))
        warning("'bins' is ignored because the palette type is not numeric")
      if (type == "numeric") {
        cuts <- if (length(bins) == 1)
          pretty(values, bins)
        else
          bins
        
        if (length(bins) > 2)
          if (!all(abs(diff(bins, differences = 2)) <=
                   sqrt(.Machine$double.eps)))
            stop("The vector of breaks 'bins' must be equally spaced")
        n <- length(cuts)
        r <- range(values, na.rm = TRUE)
        cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
        n <- length(cuts)
        p <- (cuts - r[1]) / (r[2] - r[1])
        extra <- list(p_1 = p[1], p_n = p[n])
        p <- c("", paste0(100 * p, "%"), "")
        if (decreasing == TRUE) {
          colors <- pal(rev(c(r[1], cuts, r[2])))
          labels <- rev(labFormat(type = "numeric", cuts))
        } else{
          colors <- pal(c(r[1], cuts, r[2]))
          labels <- rev(labFormat(type = "numeric", cuts))
        }
        colors <- paste(colors, p, sep = " ", collapse = ", ")
        
      }
      else if (type == "bin") {
        cuts <- args$bins
        n <- length(cuts)
        mids <- (cuts[-1] + cuts[-n]) / 2
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "bin", cuts))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "bin", cuts)
        }
        
      }
      else if (type == "quantile") {
        p <- args$probs
        n <- length(p)
        cuts <- quantile(values, probs = p, na.rm = TRUE)
        mids <- quantile(values, probs = (p[-1] + p[-n]) / 2,
                         na.rm = TRUE)
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "quantile", cuts, p))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "quantile", cuts, p)
        }
      }
      else if (type == "factor") {
        v <- sort(unique(na.omit(values)))
        colors <- pal(v)
        labels <- labFormat(type = "factor", v)
        if (decreasing == TRUE) {
          colors <- pal(rev(v))
          labels <- rev(labFormat(type = "factor", v))
        } else{
          colors <- pal(v)
          labels <- labFormat(type = "factor", v)
        }
      }
      else
        stop("Palette function not supported")
      if (!any(is.na(values)))
        na.color <- NULL
    }
    else {
      if (length(colors) != length(labels))
        stop("'colors' and 'labels' must be of the same length")
    }
    legend <-
      list(
        colors = I(unname(colors)),
        labels = I(unname(labels)),
        na_color = na.color,
        na_label = na.label,
        opacity = opacity,
        position = position,
        type = type,
        title = title,
        extra = extra,
        layerId = layerId,
        className = className,
        group = group
      )
    invokeMethod(map, data, "addLegend", legend)
  }

```

```{r tract data map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Serious Injuries and Fatalities - 2021"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Missing Data
None

#### Isolate NA or missing data
```{r}
# join NA data to the shapefile to map
data_tract_na <- data_tract %>% 
  filter(is.na(estimate))

nrow(data_tract_na) #0 tracts
```

```{r missing tract map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Serious Injuries and Fatalities - 2021"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = var_name,
                   overlayGroups = "missing data",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%
  addPolygons(data=data_tract_na,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "missing data") %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Descriptive Statistics
The following descriptive statistics reflect the census tracts that have values - the r na_rows_num census tracts with NAs have been removed. These are not weighted by population numbers. 
```{r}
data_tract_nona <- data_tract %>% 
  filter(!is.na(estimate))

summary(data_tract_nona$estimate)
```

**When weighted by tract populations.** 
```{r, include=FALSE}
# data_tract_nona_weight <- data_tract_nona %>% 
#   mutate(tot_age = estimate*total_pop10,
#          reg_tot = sum(tot_age),
#          tot_pop = sum(total_pop10),
#          est_weight_avg = reg_tot/tot_pop)

distrib <- rep(data_tract_nona$estimate, 1)

avg.value <- data_tract_nona$estimate[1] #20.1 - pulling first value
med.value <- median(distrib) #28.90
```
*Average:* `r avg.value`
\
*Median:* `r med.value`
\
\

#### Histogram
A histogram is a visual representation of the distribution of a dataset...The y-axis shows how frequently the values on the x-axis occur in the data, while the bars group ranges of values or continuous categories on the x-axis [(source)](https://www.datacamp.com/tutorial/make-histogram-basic-r).  
```{r}
hist(distrib)
```
\

#### Boxplot
A boxplot helps to visualize a quantitative variable by displaying five common location summary (minimum, median, first and third quartiles and maximum) and any observation that was classified as a suspected outlier using the interquartile range (IQR) criterion [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
boxplot(distrib, horizontal = TRUE)
```
\

#### Outliers
The IQR criterion means that all observations above or below the first and third quartile respectively, and IQR is the difference between the third and first quartile) are considered as potential outliers by R [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
outliers <- boxplot.stats(distrib)$out
outlier_rle <- rle(outliers)
outlier_df <- data.frame(unclass(outlier_rle)) %>% 
  rename(number_of_census_tracts = lengths,
         indicator = values) %>% 
  arrange(indicator)

outlier_df
```
\

###### Identify census tracts with outliers
```{r include=FALSE}
outlier_values <- outlier_df[,2]

data_tract_nona_outliers <- data_tract_nona %>% 
  dplyr::filter(estimate %in% outlier_values)

num_outliers <- nrow(data_tract_nona_outliers) #7
```
There are `r num_outliers` tracts that have outliers. 
\
\

###### Outliers by Displacement Risk score
This check uses the displacement risk data, which is mapped to 2010 census tracts. This will help determine if there is a connection between displacement risk and `r indicator_measure`.
```{r displacement risk data, inlcude=FALSE}
# Connecting to Portal for displacement tract geographies ----
displacement.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Displacement_Risk_Data/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

displacement.lyr <-st_read(displacement.url)
# head(displacement.lyr)
# plot(displacement.lyr$geometry)
# class(displacement.lyr$geoid10)

displacement_simp <- displacement.lyr %>%
  dplyr::select(geoid10,
                risk_score) %>% 
  as_tibble() %>% 
  dplyr::select(-geometry)
```

```{r, include=FALSE}
data_tract_nona_outliers %<>% 
  dplyr::mutate(geoid=trimws(geoid20))

# join outlier tract datas to spatial dispalcement tract data to isolate tracts with outlier data
data_tract_risk_outliers <- merge(x=displacement_simp, 
                                      y=data_tract_nona_outliers,
                                      by.x="geoid10",
                                      by.y="geoid",
                                      all.y=TRUE)

head(data_tract_risk_outliers)
# class(data_tract_risk_outliers)
# plot(data_tract_risk_outliers$geometry)
nrow(data_tract_risk_outliers) #103
pop_outliers <- as.numeric(sum(data_tract_risk_outliers$total_pop21, na.rm = T)) #33152
pop_outliers_num <- label_comma()(pop_outliers)
```
There are `r pop_outliers_num` people (2020 Decennial Census) in these outlier tracts.
\
\
```{r}
# compare the two measures
plot(data_tract_risk_outliers$risk_score, 
     data_tract_risk_outliers$estimate)
```

```{r}
# this will be different for each indicator - the code and the text included below...
# based on the plot in the last chunk, this code is meant to isolate outlier(s) - in this case it filters for the outliers with high risk scores and fatality and serious injury rate

# outlier #1
data_tract20_outliers_1 <- data_tract_risk_outliers %>% 
  filter(risk_score > 40)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_1

# outlier #2
data_tract20_outliers_2 <- data_tract_risk_outliers %>% 
  filter(risk_score > 150)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_2

# To locate the census tracts on the map, its easy to use the Data Portal (https://psrc-psregcncl.hub.arcgis.com/datasets/census-tracts-2020/explore?location=47.506129%2C-121.980700%2C9.23) - just filter on the geoid20 field. 
```


# **MAP .rda**
Save final data set (.rda) for map
```{r, include=FALSE}
# This is the data set used to create the map of the most recent data. For this data set, there is one additional transformation that is needed before saving it for final visualization in the 'vis' script. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk. 

head(tract.21) # to determine field names
tract.21.edit <- tract.21 %>% 
   dplyr::mutate(pop21=estimate) # the rate is already 'estimate'

# merge 2020 acs population data to crosswalk file
pop21_crosswalk <- merge(tract.21.edit, crosswalk_10_20, 
                         by.x="GEOID",
                         by.y="geoid20", 
                         all.x=TRUE)

# calculate population by adding 2020 tracts that were split from 2010 tracts
pop21 <- pop21_crosswalk %>% 
  dplyr::group_by(geoid10) %>% 
  dplyr::summarise(total_pop21=sum(pop21)) %>%
  as_tibble()

```

```{r, region and county numbers for map labels}
# calculating data by county 
county_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  # dplyr::select(county_name, estimate, total_pop21) %>% #to simplify table to necessary fields
  dplyr::group_by(county_name) %>% 
  dplyr::mutate(cnty_estimate=(sum(estimate*total_pop21, na.rm =TRUE))/(sum(total_pop21, na.rm=TRUE))) %>% #calculate weighted value by county
  dplyr::distinct(county_name, cnty_estimate)

# calculating data by region
reg_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  dplyr::select(estimate, total_pop21) %>% #to simplify table to necessary fields
  dplyr::mutate(reg_estimate=(sum(estimate*total_pop21, na.rm =TRUE))/(sum(total_pop21, na.rm=TRUE))) %>% #calculate weighted value for region
  dplyr::distinct(reg_estimate)

# add region data to county data
county_reg_data <- county_tract_data %>% 
  mutate(reg_estimate=reg_tract_data$reg_estimate)

# merge tract, county, and region data into one data set 
data_tract <- merge(data_tract, county_reg_data,
                    by.x="county_name",
                    by.y="county_name",
                    all.x=TRUE)
```

```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'g-transportation'
ind_dir <- 'g01-safety'
file_name <- 'g01-safety-map-data'

# save final data set as .rda
save(data_tract, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

# Explore data by equity quintile
## Join primary data to equity quintiles and acs population

```{r}
# 2021 data -----
equity_tracts_21 <- equity_tracts %>% 
  dplyr::filter(data_year==2021)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2021, equity_tracts_21,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2021 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract21 <- merge(tract.21, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2020 data -----
equity_tracts_20 <- equity_tracts %>% 
  dplyr::filter(data_year==2021)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2020, equity_tracts_20,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2021 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract20 <- merge(tract.20, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2019 data -----
equity_tracts_19 <- equity_tracts %>% 
  dplyr::filter(data_year==2019)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2019,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2019, equity_tracts_19,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2019 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract19 <- merge(tract.19, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2018 data -----
equity_tracts_18 <- equity_tracts %>% 
  dplyr::filter(data_year==2018)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2018,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2018, equity_tracts_18,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2018 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract18 <- merge(tract.18, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2017 data -----
equity_tracts_17 <- equity_tracts %>% 
  dplyr::filter(data_year==2017)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2017,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2017, equity_tracts_17,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2017 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract17 <- merge(tract.17, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2016 data -----
equity_tracts_16 <- equity_tracts %>% 
  dplyr::filter(data_year==2016)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2016,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2016, equity_tracts_16,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2016 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract16 <- merge(tract.16, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2015 data -----
equity_tracts_15 <- equity_tracts %>% 
  dplyr::filter(data_year==2015)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2015,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2015, equity_tracts_15,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2015 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract15 <- merge(tract.15, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2014 data -----
equity_tracts_14 <- equity_tracts %>% 
  dplyr::filter(data_year==2014)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2014,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2014, equity_tracts_14,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2014 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract14 <- merge(tract.14, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2013 data -----
equity_tracts_13 <- equity_tracts %>% 
  dplyr::filter(data_year==2013)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2013,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2013, equity_tracts_13,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2013 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract13 <- merge(tract.13, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2012 data -----
equity_tracts_12 <- equity_tracts %>% 
  dplyr::filter(data_year==2012)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2012,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2012, equity_tracts_12,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2012 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract12 <- merge(tract.12, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2011 data -----
equity_tracts_11 <- equity_tracts %>% 
  dplyr::filter(data_year==2011)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2011,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2011, equity_tracts_11,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2011 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract11 <- merge(tract.11, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# 2010 data -----
equity_tracts_10 <- equity_tracts %>% 
  dplyr::filter(data_year==2010)

# join WTN data (in 2020 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, fat_inj_2010,
                        by.x="geoid10",
                        by.y="geoid",
                        all.x=TRUE)

# join data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(fat_inj_2010, equity_tracts_10,
                            by.x="geoid",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2010 acs population data to safety data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract10 <- merge(tract.10, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid", 
                             all.x=TRUE)

# check data set
nrow(data_equity_tract21)
nrow(data_equity_tract11)

```
This data set should include about twice the number of census tracts in the region - census tracts are listed twice - once as part of the region and another as part of their corresponding county.

## Transform/pivot
The data sets need to be cleaned and pivoted to a longer table so that there is only one field for the 6 different equity categories, instead of one field for each.
```{r, include=FALSE}
# clean/simplify data sets
data_equity_quintiles21 <- data_equity_tract21 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles20 <- data_equity_tract20 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles19 <- data_equity_tract19 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles18 <- data_equity_tract18 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles17 <- data_equity_tract17 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles16 <- data_equity_tract16 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles15 <- data_equity_tract15 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles14 <- data_equity_tract14 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles13 <- data_equity_tract13 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles12 <- data_equity_tract12 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles11 <- data_equity_tract11 %>%
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>%
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

data_equity_quintiles10 <- data_equity_tract10 %>%
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>%
  dplyr::mutate(combo_rate_num=as.numeric(combo_rate))

# pivot data set so that there is one column with quintile designation
data_equity_pivot21<- data_equity_quintiles21 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot20<- data_equity_quintiles20 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot19<- data_equity_quintiles19 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot18<- data_equity_quintiles18 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot17<- data_equity_quintiles17 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot16<- data_equity_quintiles16 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot15<- data_equity_quintiles15 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot14<- data_equity_quintiles14 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot13<- data_equity_quintiles13 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot12<- data_equity_quintiles12 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot11<- data_equity_quintiles11 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot10<- data_equity_quintiles10 %>%
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")
```

```{r}
# check data sets
nrow(data_equity_pivot11)
nrow(data_equity_pivot21)
```

These data sets should include records for:

* 2 geographies (region and whichever county the tract belongs to) x # census tracts (~919 for 2020)
* 6 equity focus groups (disability and lep are not available in for 2010, lep not avail - but still show up as NA)


## Calculations
### Calculate by region
The safety data are available at the census tract level, but we want to calculate the regional average, using the census bureau's tract-level population estimates (2021).
```{r}
data_equity_pivot21_calc <- data_equity_pivot21 %>% 
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>% 
  as_tibble() %>%
  mutate(data_year=2021) 

data_equity_pivot20_calc <- data_equity_pivot20 %>% 
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>% 
  as_tibble() %>%
  mutate(data_year=2020) 

data_equity_pivot19_calc <- data_equity_pivot19 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2019)

data_equity_pivot18_calc <- data_equity_pivot18 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2018)

data_equity_pivot17_calc <- data_equity_pivot17 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2017)

data_equity_pivot16_calc <- data_equity_pivot16 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2016)

data_equity_pivot15_calc <- data_equity_pivot15 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2015)

data_equity_pivot14_calc <- data_equity_pivot14 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2014)

data_equity_pivot13_calc <- data_equity_pivot13 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2013)

data_equity_pivot12_calc <- data_equity_pivot12 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2012)

data_equity_pivot11_calc <- data_equity_pivot11 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2011)

data_equity_pivot10_calc <- data_equity_pivot10 %>%
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile, county) %>%
  summarise(wt_combo_rate=(sum(combo_rate_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>%
  mutate(County.Name="Region") %>%
  as_tibble() %>%
  mutate(data_year=2010)

```

```{r}
# check data sets
nrow(data_equity_pivot11_calc)
nrow(data_equity_pivot21_calc)
```
These data sets should include records for:

* 1 geography: region
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015)
     * 2010: only 4 equity focus groups
     * 2015: only 5 equity focus groups
* 5 equity quintiles 


### Calculate by equity quintile
The rates are available at the census tract level but we want to calculate the weighted values by the 6 equity/5 quintile groups, using the census bureau's tract-level population estimates (2020). 

```{r}

# calculate 2020 region population
population_test <- data_equity_pivot21 %>%
  distinct(GEOID, .keep_all = TRUE) %>%
  dplyr::summarise(pop=formatC((sum(estimate, na.rm=TRUE)), format="d", big.mark=",")) #4,197,443

sum(tracts20.lyr$total_pop21) #4,294,373

# calculations to get weighted average for safety percentage per equity group/quintile
data_equity_wt21 <- data_equity_pivot21 %>%
  filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
  dplyr::mutate(tot_risk_tract=estimate*combo_rate_num) %>%
  dplyr::group_by(county, data_year, equity_group, quintile) %>%
  dplyr::summarise(tracts=length(GEOID),
                   tot_combo_rate=(sum(tot_risk_tract, na.rm = TRUE)),
                   tot_pop=(sum(estimate, na.rm = TRUE))) %>%
  dplyr::mutate(wt_combo_rate=tot_combo_rate/tot_pop) 

# check population
data_equity_wt21_test <- data_equity_wt21 %>%
  dplyr::filter(equity_group=="income_quintile")
  sum(data_equity_wt21_test$tot_pop) #4,197,443

# calculations to get weighted average for safety percentage per equity group/quintile
# data_equity_wt11 <- data_equity_pivot11 %>%
#   filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
#   dplyr::mutate(tot_risk_tract=estimate*Change_num) %>%
#   dplyr::group_by(county, data_year, equity_group, quintile) %>%
#   dplyr::summarise(tracts=length(GEOID),
#                    tot_Change=(sum(tot_risk_tract, na.rm = TRUE)),
#                    tot_pop=(sum(estimate, na.rm = TRUE))) %>%
#   dplyr::mutate(wt_Change=tot_Change/tot_pop) 

# check population
# data_equity_wt11_test <- data_equity_wt11 %>%
#   dplyr::filter(equity_group=="income_quintile")
#   sum(data_equity_wt11_test$tot_pop) #4,197,443

# check data set
# nrow(data_equity_wt11)
nrow(data_equity_wt21)

# region_rate <- data_equity_pivot20 %>%
#   dplyr::select(-"county", -"equity_group", -"quintile") %>%
#   distinct() %>%
#   mutate(rate_pop=(estimate*Log_rate)) %>%
#   mutate(rate_pop=(replace_na(rate_pop, 0))) %>%
#   group_by(year.x) %>%
#   summarise(region_rate=(sum(rate_pop)/sum(estimate))) %>%
#   as_tibble() %>%
#   dplyr::select("region_rate") %>%
#   pull()

```
The regional population in 2021: `r population_test`
\
\
This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 6 equity focus groups
* 5 equity quintiles 

## Combine years into one data set
```{r}
data_tract <- as.data.frame(rbind(data_equity_pivot10_calc, data_equity_pivot11_calc, data_equity_pivot12_calc, data_equity_pivot13_calc, data_equity_pivot14_calc, data_equity_pivot15_calc, data_equity_pivot16_calc, data_equity_pivot17_calc, data_equity_pivot18_calc, data_equity_pivot19_calc, data_equity_pivot20_calc,
        data_equity_pivot21_calc)) %>%
   dplyr::select(-County.Name) %>%
   filter(!is.na(quintile))


# data_tract <- data_equity_pivot21_calc %>% 
#   filter(!is.na(quintile))

# check data set
str(data_tract)
```
This data set should include records for: 

* 5 geographies: 4 PSRC counties + region
* 3 years
* 6 equity focus groups
    * 2010: (subtracting) 2 missing focus groups(disability, lep) * 5 equity quintiles * 5 geographies
    * 2015: (subtracting) 1 missing focus group (lep) * 5 equity quintiles * 5 geographies
* 5 equity quintiles

### Address "Low - Low Medium" quintile group
This code was added because of Kitsap/LEP data - there were too many tracts with 0, so we combined the bottom two quintiles into low - low medium. This change requires some additional processing so that all of the tracts (bottom 40%) in this combined quintile category are averaged and this average value is assigned to separate 'low' and 'low medium' quintile categories. 
```{r}
# nrow(data_tract) #1594
#unique(data_tract$quintile) #6 quintiles because of additional "Low - Low Medium"

low_quintile <- data_tract %>% 
  filter(quintile=="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>% 
  mutate(quintile="Low") %>% 
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications
lowmed_quintile <- data_tract %>% 
  filter(quintile=="Low - Low Medium") %>%
  rename(quintile_original=quintile) %>%
  mutate(quintile="Low Medium") %>% 
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications

low_lowmed <- rbind(low_quintile,
                    lowmed_quintile)
# nrow(low_lowmed) #4

# gather counties that have low - medium categorization
low_lowmed_counties <- unique(low_lowmed$county) #Kitsap, Pierce
low_lowmed_years <- unique(low_lowmed$data_year) #2020
low_lowmed_equity <- unique(low_lowmed$equity_group) #lep_quintile

other_data <- data_tract %>% 
  filter(quintile!="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>% 
  mutate(quintile=quintile_original)
# nrow(other_data) #396

data_tract <- rbind(low_lowmed, #4
                    other_data) #396

nrow(data_tract) #400 - should be fewer because of duplicates (396)- low and low medium present in data set because of aggregation/calculations in previous steps - need to be removed and replaced with averaged values
# table(data_tract$quintile, data_tract$county) #should not have extra in low and low medium quintile categories
# unique(data_tract$quintile) #5 quintiles

# remove rows where it is one of the "Low - Low Medium" quintiles AND the original quintile is "Low" or "Low Medium" - only want to keep the rows that are "Low"/"Low Medium" and they share a value
data_tract <- data_tract %>% 
  filter(!(county %in% low_lowmed_counties & 
           data_year %in% low_lowmed_years &
           equity_group %in% low_lowmed_equity &
           (quintile_original=="Low" | 
             quintile_original=="Low Medium")))

# nrow(data_tract) #400
# table(data_tract$quintile, data_tract$county) #should not have extra in low and low medium quintile categories
```

## Rename and factor
```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

quintile_order <- c("Low", 
                    "Low\nMedium", 
                    "Medium", 
                    "Medium\nHigh", 
                    "High")

equity_group_order <- c("People of Color", 
                        "Households with Lower Income", 
                        "People with a Disability", 
                        "Households with Limited English Proficiency", 
                        "Households with Youth <18", 
                        "Households with Older Adults 65+")

# transforming data labels ----
data_clean <- data_tract %>% 
  mutate(county_ord = factor(county, levels=county_order)) %>%
  mutate(equity_group_ord = case_when(
    equity_group=="poc_quintile"~"People of Color",
    equity_group=="disability_quintile"~"People with a Disability",
    equity_group=="lep_quintile"~"Households with Limited English Proficiency",
    equity_group=="income_quintile"~"Households with Lower Income",
    equity_group=="youth_quintile"~"Households with Youth <18",
    equity_group=="older_quintile"~"Households with Older Adults 65+")) %>%
  mutate(equity_group_ord = factor(equity_group_ord, levels = equity_group_order)) %>% 
  mutate(quintile_ord = str_wrap(quintile, width=7)) %>%
  mutate(quintile_ord = factor(quintile_ord, levels = quintile_order))

data_clean$data_year <- as.character(data_clean$data_year)

# Minor edits to the data
data_clean <- data_clean %>%
  dplyr::mutate(estimate=as.numeric(wt_combo_rate)) #rename value variable for consistency

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county_ord, equity_group_ord, quintile_ord, data_year)
```


# **CHART .rda**
Save final data set for charts (.rda)
```{r}
# set folder structure
file_name <- 'g01-safety-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>