---
title: "Overcrowding" 
subtitle: "Data Gen: Exploring, Cleaning, Transforming (PUMS/OSPI data)"
author: "Meg Grzybowski"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts
library(stringr) #for string wrapping

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Overcrowding**

# Download data 
## Primary data 
The data set referenced in this script is generated from the Census Bureau’s American Community Survey (ACS) Public Use Microdata Sample [(PUMS)](https://www.census.gov/programs-surveys/acs/microdata/documentation.html). These data sets provide data at the **person- or household-level**, with the ability to look at the different indicators by the six equity demographic groups of interest. 
```{r, include=FALSE}
# PUMS/OSPI data set
pums_ospi_elmer <- get_table(schema="equity", tbl_name="v_tracker_indicators")
```
\
This data set was compiled from PUMS data.
\
Looking at the fields in the data set:
```{r}
# explore the data
head(pums_ospi_elmer)
# view the equity demographic groups
unique(pums_ospi_elmer$focus_type)
# view the different indicators available within the data set
unique(pums_ospi_elmer$indicator_type)
# view the different data years available within the data set
unique(pums_ospi_elmer$data_year)
```

```{r indicator specific information, include=FALSE}
# set indicator information
indicator_measure <- 'crowding'
indicator_title <- 'Overcrowding (>1.5 person/bedroom)'
```

## 1. Explore data
*In this section we make sure that the data set has the correct dimensions - year, geography, equity focus group, etc.*  
```{r}
# clean data set
data_full <- pums_ospi_elmer %>% 
  distinct(indicator_fact_id, .keep_all = TRUE) %>% #this may be necessary depending on indicator
  dplyr::filter(indicator_type==indicator_measure) %>% 
  dplyr::filter(focus_type!="Total") %>% 
  filter(focus_attribute!="Total") %>%
  filter(indicator_attribute!="Total") %>%
  dplyr::mutate(data_year_yr=format(data_year,format="%Y"))
```

### Data fields
#### *consistent base data*

* There should be 5 geographies - the 4 counties and the Region. 
* There should be 6 equity focus categories - POC, income, disability, youth, older adult, and LEP
    + 2 sub-groups in each of the 6 equity focus group (e.g. people of color, non-people of color)
```{r}
# geographies
num_county <- length(unique(data_full$county)) #5
unique(data_full$county)

# equity focus groups
num_group <- length(unique(data_full$focus_type)) #6
unique(data_full$focus_type)
```
\
\

#### *indicator-specific data*
These fields will vary by indicator:

* Type of metric - this will determine how the data are visualized *(est ="percent" or "currency" or "number")*
* Number of years (5-year span) - this can vary depending on data availability
* Number of indicator-specific categories - this can vary depending on the indicator of interest, ranging from N/A (*median income*) to multiple levels (*crowding, housing cost burden*)
```{r}
# metric
unique(data_full$fact_type) #share

# years
num_yr <- length(unique(data_full$data_year)) #3
unique(data_full$data_year)

# indicator attributes
num_indatt <- length(unique(data_full$indicator_attribute)) #1
unique(data_full$indicator_attribute) #N/A

# calculate expected number of rows
num_row <- num_yr*num_county*num_group*2*num_indatt #540
```

There are **`r num_county`** geographies and **`r num_group`** equity focus groups (each with **2** subgroups). There are **`r num_yr`** years in the data set and the indicator specific field has **`r num_indatt`** attribute(s), which means there should be a total of **`r num_row`** rows.
```{r}
# count number of rows
nrow(data_full) #510
```
<span style="color: #00A7A0">There are some missing data.</span> 
\
\

#### *checking for missing data*
##### Year / geography
```{r}
num_yr_geo <- num_group*2*num_indatt #36
```
If we look at the data by year and geography, there should be **`r num_yr_geo`** entries per year/geography.
```{r, include=FALSE, eval=FALSE}
table(data_full$data_year,
      data_full$county)
```
<span style="color: #00A7A0">
2011 is missing some data for all geographies.
</span> 
\
\

##### Year / equity focus group
```{r}
num_yr_grp <- num_county*2*num_indatt #30
```
If we look at the data by year and focus group, there should be **`r num_yr_grp`** entries per year/focus group.
```{r}
table(data_full$data_year,
      data_full$focus_type)
```
<span style="color: #00A7A0">
The disability category is missing all data in 2011.
</span> 
\
\

##### Year / equity focus sub-group
```{r}
num_yr_subgrp <- num_county*num_indatt #5
```
If we look at the data by year and focus sub-group, there should be **`r num_yr_subgrp`** entries per year/focus sub-group.
```{r}
table(data_full$data_year,
      data_full$focus_attribute)
```
\
\

##### Year / indicator attribute
```{r}
num_yr_ind <- num_county*num_group*2 #60
```
If we look at the data by year and indicator attribute, there should be **`r num_yr_ind`** entries per year/indicator attribute.
```{r}
table(data_full$data_year,
      data_full$indicator_attribute)
```
\
\

### Numeric data
To check for *0*s and *NULL*s
```{r}
# checking for NULLs
summary(data_full$fact_value)
```
<span style="color: #00A7A0">
There are no 0s or nulls.
</span> 
\
\

To look at distribution of all data - not the most useful visual, but provides a sense of the range of values at a high level in one plot. 
\
```{r}
qplot(data_year_yr, fact_value, 
      colour = county, shape = focus_type, 
      data = data_full, 
      main = indicator_title) + theme(axis.title.x = element_blank(),
                                               axis.title.y = element_blank())
```
\
This table includes a lot of information about the data set and helps to show the different levels of each field. It provides another way to check if data are available for all counties and all years, or where there may be gaps in the data set. 
```{r}
data_refined <- data_full %>% 
  dplyr::select(data_year_yr, county, vulnerability, focus_type, focus_attribute,
                indicator_type, indicator_attribute, fact_value)

sumtable(data_refined,
         add.median=TRUE,
         group="county",
         out="return")
```

```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

focus_type_order <- c("People of Color", 
                      "Households with Lower Income", 
                      "People with a Disability", 
                      "People with Limited English Proficiency", 
                      "Households with Youth <18", 
                      "Households with Older Adults 65+")

focus_attribute_order <- c("People of color",
                           "White\nnon-Hispanic",
                           "Households with\nlower income",
                           "With a\ndisability",
                           "Without a\ndisability",
                           "Limited English\nproficiency",
                           "English\nproficient",
                           "Households with\nyouth",
                           "Households with\nolder adults",
                           "Other households")

# transforming data labels ----
data_clean <- data_full %>% 
  mutate(county = factor(county, levels=county_order)) %>%
  mutate(focus_type_ord = case_when(
    focus_type=="POC_cat"~"People of Color",
    focus_type=="Disability_cat"~"People with a Disability",
    focus_type=="LEP_cat"~"People with Limited English Proficiency",
    focus_type=="Income_cat"~"Households with Lower Income",
    focus_type=="Youth_cat"~"Households with Youth <18",
    focus_type=="Older_cat"~"Households with Older Adults 65+")) %>%
  mutate(focus_type_ord = factor(focus_type_ord, levels = focus_type_order)) %>%
  mutate(focus_attribute_ord = case_when(
    focus_attribute== "POC"~ "People of color",
    focus_attribute== "Non-POC"~ "White non-Hispanic",
    focus_attribute== "Low Income"~ "Households with lower income",
    focus_attribute== "Non-Low Income"~ "Other households",
    focus_attribute== "With disability"~ "With a disability",
    focus_attribute== "Without disability"~ "Without a disability",
    focus_attribute== "Limited English proficiency"~ "Limited English proficiency",
    focus_attribute== "English proficient"~ "English proficient",
    focus_attribute== "Household with youth"~ "Households with youth",
    focus_attribute== "Household without youth"~ "Other households",
    focus_attribute== "Household with older adult"~ "Households with older adults",
    focus_attribute== "Household without older adult"~ "Other households")) %>% 
  mutate(focus_attribute_ord = str_wrap(focus_attribute_ord, width=16)) %>%
  mutate(focus_attribute_ord = factor(focus_attribute_ord, levels = focus_attribute_order)) 

# echarts wants axis to be factors/characters so convert year to a character
data_clean$data_year <- as.character(data_clean$data_year)

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county, focus_type_ord, focus_attribute_ord, data_year)
```

```{r}
# indicator specific re-ordering - median income is N/A so this step is unnecessary
# data_clean$indicator_attribute <- factor(data_clean$indicator_attribute,
#                                     levels = c("Less than a Bachelor's degree",
#                                                "Bachelor's degree or higher"))
```
\

### Data labels, shares
These charts were generated to ensure the labels across years are consistent/make sense. There had been an issue with misassigned labels because tidycensus::pums_variables, i.e. the only digital data dictionary available to associate labels with codes, exists only from 2017 forward. Most variables have had consistent codes, but in cases where the codes have shifted over time, using the 2017 lookup winds up mischaracterizing categories.
\
\
These charts also help to confirm that the shares add up to 100% - only relevant when *indicator_attribute* has more than one category. The *indicator_attribute* for median household income is NA.
\
\
*The colors of the charts may not be consistent between the years depending on missing data.* 
\
```{r}
# the years below reflect the years that are available from unique(data_year)
just_region_11 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2011) %>% 
  arrange(focus_attribute_ord)
just_region_16 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2016)%>% 
  arrange(focus_attribute_ord)
just_region_21 <- data_clean %>% 
  filter(county=="Region") %>% 
  filter(data_year==2021)%>% 
  arrange(focus_attribute_ord)

# check est type - make sure it corresponds to your indicator (percent, currency, or number)
facet_chart <- static_facet_column_chart(t=just_region_11,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "percent",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2011)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")

facet_chart

facet_chart <- static_facet_column_chart(t=just_region_16,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "percent",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2016)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")
facet_chart

facet_chart <- static_facet_column_chart(t=just_region_21,
                                         x="data_year", y="fact_value",
                                         fill="vulnerability", facet="focus_type_ord",
                                         est = "percent",
                                         scales= "fixed",
                                         color = "psrc_pairs", 
                                         ncol=3,
                                         title = paste0(indicator_title, ": Region (2021)"),
                                         subtitle = "by equity focus group",
                                         source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")

facet_chart
```

## 2. Visually explore data
### Scatter plots
*In this section we start to explore the data visually - distribution by the different dimensions within the data set. These plots are helpful to check for outliers and get a higher level understanding of the data in one visual, before slicing the data by geography and equity focus group in the following sections.* 
\
\
The following code will need to be adjusted to fit the fields specific to the data indicator. For educational attainment, we focus on those with a Bachelor's degree or higher. The following code establishes the data frame that the rest of the analysis uses. If there are fewer than 2 *indicator attributes*, this section can be skipped/commented out, but the code will need to be adjusted throughout. 
\
\

#### By *indicator_attribute*
This section isn't relevant for this specific indicator because there aren't unique indicator attributes. 

#### By Year
```{r}
# separating by year ----
data_clean11 <- data_clean %>% 
  filter(data_year==2011)
data_clean16 <- data_clean %>% 
  filter(data_year==2016)
data_clean21 <- data_clean %>% 
  filter(data_year==2021)
```

```{r}
# plot data, adjust y limits based on data set
qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean11, 
      main = paste0(indicator_title, ": 2011")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 1)

qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean16,
      main = paste0(indicator_title, ": 2016")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 1)

qplot(focus_type_ord, fact_value, 
      colour = county, shape = vulnerability, 
      data = data_clean21, 
      main = paste0(indicator_title, ": 2021")) + theme(axis.title.x = element_blank(),
                                                        axis.title.y = element_blank()) + ylim(0, 1)
```

# **CHART .rda**
```{r, include=FALSE}
head(data_clean) # make sure this is the correct data set

# make any additional edits 
data_clean <- data_clean %>% 
  mutate(fact_value_notrounded=fact_value) %>% #create column to retain full median income values
  mutate(fact_value=round(fact_value, digits=2)) #round median income value to hundreds - retain consistent name
```
Save final data set (.rda) for charts
```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'e-housing'
ind_dir <- 'e04-overcrowding'
file_name <- 'e04-overcrowding-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```


## Create data set for map .rda 
### Additional data
#### Tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
tracts10.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)
tracts10.lyr<-st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

#### Corresponding ACS table
```{r acs data, include=FALSE}
# Finding the corresponding ACS table - this works if you know the correct concept label. If not, another option would be to visit https://data.census.gov/table and search for the right subject table and skip to the next step ----

# getting data by tract
acs_data <- get_acs_recs(geography ='tract', 
                         table.names = 'B25014', #subject table code
                         years = c(2021),
                         acs.type = 'acs5')

# filter to the variables of interest
acs_data_filter <- acs_data %>%
  # dplyr::filter(type %in% row_names) %>%
  dplyr::filter(variable != 'B25014_001') %>% #remove total households
  mutate(per_bedroom = factor(
    case_when(variable == "B25014_002" ~ "Total",
              variable == "B25014_008" ~ "Total",
              grepl("0.50 or less|0.51 to 1.00|1.01 to 1.50", label) ~ "Less than or equal to 1.5 persons per bedroom",
              grepl("1.51 to 2.00|2.01 or more", label) ~ "Overcrowding",
              TRUE ~ as.character(label)))) %>%
  mutate(type = factor(
    case_when(grepl("B25014_002|B25014_003|B25014_004|B25014_005|B25014_006|B25014_007", variable) ~ "Owner",
              grepl("B25014_008|B25014_009|B25014_010|B25014_011|B25014_012|B25014_013", variable) ~ "Renter",
              TRUE ~ as.character(variable)))) #%>%
  # dplyr::select(GEOID, variable, per_bedroom, type, estimate, moe) #%>% #simplify table
  # dplyr::filter(type != 'Owner')    #this can be commented out for inclusion of both Owners and Renters

# ----------------
# for separate owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom, type) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom, type), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding_rent=(estimate_Overcrowding_Renter/estimate_Total_Renter),
         estimate_perc_overcrowding_own=(estimate_Overcrowding_Owner/estimate_Total_Owner))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_rent=tidycensus::moe_sum(estimate = estimate_Overcrowding_Renter,
                                             moe = moe_Overcrowding_Renter),
                # moe_own=tidycensus:moe_prop(num=estimate_Overcrowding_Owner,
                #                           denom=estimate_Total_units,
                #                           moe_num=moe_Overcrowding_Owner,
                #                           moe_denom=moe_Total_units),
                moe_own=tidycensus::moe_sum(estimate = estimate_Overcrowding_Owner,
                                            moe = moe_Overcrowding_Owner))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding_Renter),
                !is.na(estimate_Overcrowding_Owner))

# calculate reliability values
acs_data_reliability_rent <- reliability_calcs(acs_data_final,
                                               estimate='estimate_Overcrowding_Renter',
                                               moe='moe_rent')
acs_data_reliability_own <- reliability_calcs(acs_data_final,
                                              estimate='estimate_Overcrowding_Owner',
                                              moe='moe_own')

# rename and simplify
acs_data <- acs_data_reliability_rent %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding_rent') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)

# ----------------
# for combined owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding=(estimate_Overcrowding/estimate_Total))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_overcrowding=tidycensus::moe_sum(estimate = estimate_Overcrowding,
                                             moe = moe_Overcrowding))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding))

# calculate reliability values
acs_data_reliability <- reliability_calcs(acs_data_final,
                                          estimate='estimate_Overcrowding',
                                          moe='moe_Overcrowding')

# rename and simplify
acs_data <- acs_data_reliability %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)
```

```{r acs data - place level, include=FALSE}
# Finding the corresponding ACS table - this works if you know the correct concept label. If not, another option would be to visit https://data.census.gov/table and search for the right subject table and skip to the next step ----

# getting data by tract
acs_data <- get_acs_recs(geography ='place', 
                         table.names = 'B25014', #subject table code
                         years = c(2021),
                         acs.type = 'acs5')

# filter to the variables of interest
acs_data_filter <- acs_data %>%
  # dplyr::filter(type %in% row_names) %>%
  dplyr::filter(variable != 'B25014_001') %>% #remove total households
  mutate(per_bedroom = factor(
    case_when(variable == "B25014_002" ~ "Total",
              variable == "B25014_008" ~ "Total",
              grepl("0.50 or less|0.51 to 1.00|1.01 to 1.50", label) ~ "Less than or equal to 1.5 persons per bedroom",
              grepl("1.51 to 2.00|2.01 or more", label) ~ "Overcrowding",
              TRUE ~ as.character(label)))) %>%
  mutate(type = factor(
    case_when(grepl("B25014_002|B25014_003|B25014_004|B25014_005|B25014_006|B25014_007", variable) ~ "Owner",
              grepl("B25014_008|B25014_009|B25014_010|B25014_011|B25014_012|B25014_013", variable) ~ "Renter",
              TRUE ~ as.character(variable)))) #%>%
  # dplyr::select(GEOID, variable, per_bedroom, type, estimate, moe) #%>% #simplify table
  # dplyr::filter(type != 'Owner')    #this can be commented out for inclusion of both Owners and Renters

# ----------------
# for separate owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom, type) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom, type), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding_rent=(estimate_Overcrowding_Renter/estimate_Total_Renter),
         estimate_perc_overcrowding_own=(estimate_Overcrowding_Owner/estimate_Total_Owner))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_rent=tidycensus::moe_sum(estimate = estimate_Overcrowding_Renter,
                                             moe = moe_Overcrowding_Renter),
                # moe_own=tidycensus:moe_prop(num=estimate_Overcrowding_Owner,
                #                           denom=estimate_Total_units,
                #                           moe_num=moe_Overcrowding_Owner,
                #                           moe_denom=moe_Total_units),
                moe_own=tidycensus::moe_sum(estimate = estimate_Overcrowding_Owner,
                                            moe = moe_Overcrowding_Owner))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding_Renter),
                !is.na(estimate_Overcrowding_Owner))

# calculate reliability values
acs_data_reliability_rent <- reliability_calcs(acs_data_final,
                                               estimate='estimate_Overcrowding_Renter',
                                               moe='moe_rent')
acs_data_reliability_own <- reliability_calcs(acs_data_final,
                                              estimate='estimate_Overcrowding_Owner',
                                              moe='moe_own')

# rename and simplify
acs_data <- acs_data_reliability_rent %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding_rent') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)

# ----------------
# for combined owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding=(estimate_Overcrowding/estimate_Total))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_overcrowding=tidycensus::moe_sum(estimate = estimate_Overcrowding,
                                             moe = moe_Overcrowding))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding))

# calculate reliability values
acs_data_reliability <- reliability_calcs(acs_data_final,
                                          estimate='estimate_Overcrowding',
                                          moe='moe_Overcrowding')

# rename and simplify
acs_data <- acs_data_reliability %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)
```

```{r acs data - county level, include=FALSE}
# Finding the corresponding ACS table - this works if you know the correct concept label. If not, another option would be to visit https://data.census.gov/table and search for the right subject table and skip to the next step ----

# getting data by tract
acs_data <- get_acs_recs(geography ='county', 
                         table.names = 'B25014', #subject table code
                         years = c(2021),
                         acs.type = 'acs5')

# filter to the variables of interest
acs_data_filter <- acs_data %>%
  # dplyr::filter(type %in% row_names) %>%
  dplyr::filter(variable != 'B25014_001') %>% #remove total households
  mutate(per_bedroom = factor(
    case_when(variable == "B25014_002" ~ "Total",
              variable == "B25014_008" ~ "Total",
              grepl("0.50 or less|0.51 to 1.00|1.01 to 1.50", label) ~ "Less than or equal to 1.5 persons per bedroom",
              grepl("1.51 to 2.00|2.01 or more", label) ~ "Overcrowding",
              TRUE ~ as.character(label)))) %>%
  mutate(type = factor(
    case_when(grepl("B25014_002|B25014_003|B25014_004|B25014_005|B25014_006|B25014_007", variable) ~ "Owner",
              grepl("B25014_008|B25014_009|B25014_010|B25014_011|B25014_012|B25014_013", variable) ~ "Renter",
              TRUE ~ as.character(variable)))) #%>%
  # dplyr::select(GEOID, variable, per_bedroom, type, estimate, moe) #%>% #simplify table
  # dplyr::filter(type != 'Owner')    #this can be commented out for inclusion of both Owners and Renters

# ----------------
# for separate owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom, type) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom, type), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding_rent=(estimate_Overcrowding_Renter/estimate_Total_Renter),
         estimate_perc_overcrowding_own=(estimate_Overcrowding_Owner/estimate_Total_Owner))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_rent=tidycensus::moe_sum(estimate = estimate_Overcrowding_Renter,
                                             moe = moe_Overcrowding_Renter),
                # moe_own=tidycensus:moe_prop(num=estimate_Overcrowding_Owner,
                #                           denom=estimate_Total_units,
                #                           moe_num=moe_Overcrowding_Owner,
                #                           moe_denom=moe_Total_units),
                moe_own=tidycensus::moe_sum(estimate = estimate_Overcrowding_Owner,
                                            moe = moe_Overcrowding_Owner))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding_Renter),
                !is.na(estimate_Overcrowding_Owner))

# calculate reliability values
acs_data_reliability_rent <- reliability_calcs(acs_data_final,
                                               estimate='estimate_Overcrowding_Renter',
                                               moe='moe_rent')
acs_data_reliability_own <- reliability_calcs(acs_data_final,
                                              estimate='estimate_Overcrowding_Owner',
                                              moe='moe_own')

# rename and simplify
acs_data <- acs_data_reliability_rent %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding_rent') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)

# ----------------
# for combined owners and renters
# use the moe_sum function from tidycensus
acs_data_group <- acs_data_filter %>% 
  dplyr::group_by(GEOID, per_bedroom) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), 
                   moe=tidycensus::moe_sum(estimate=estimate, moe=moe)) %>% 
  dplyr::filter(per_bedroom != "Less than or equal to 1.5 persons per bedroom")

# pivot data to calculate percentage by census tract
acs_data_pivot<- acs_data_group %>%   
  pivot_wider(names_from = c(per_bedroom), values_from = c(estimate,moe)) #add 'type' if want to distinguish b/t Owners and Renters

# calculating percentage/share of population per tract instead of integer
acs_data_calc <- acs_data_pivot %>%
  mutate(estimate_perc_overcrowding=(estimate_Overcrowding/estimate_Total))

# calculate moes
acs_data_moes <- acs_data_calc %>%
  rowwise() %>% 
  dplyr::mutate(moe_overcrowding=tidycensus::moe_sum(estimate = estimate_Overcrowding,
                                             moe = moe_Overcrowding))
acs_data_final<- acs_data_moes %>% 
  dplyr::filter(!is.na(estimate_Overcrowding))

# calculate reliability values
acs_data_reliability <- reliability_calcs(acs_data_final,
                                          estimate='estimate_Overcrowding',
                                          moe='moe_Overcrowding')

# rename and simplify
acs_data <- acs_data_reliability %>% #just interested in renters for now
  rename(estimate_dec='estimate_perc_overcrowding') %>%
  mutate(estimate=estimate_dec*100) %>% 
  dplyr::select(GEOID,estimate,reliability)
```

### Joining data
```{r}
# merge to 2020 census tract spatial file
acs_data_tract_merge <- merge(tracts20.lyr, acs_data,
                              by.x="geoid20",
                              by.y="GEOID",
                              all.x=TRUE)

# getting data by county and region from acs 
county_reg_acs_data <- get_acs_recs(geography ='county', 
                                table.names = 'B25014', #subject table code
                                years = c(2021),
                                acs.type = 'acs5')

county_reg_data_rename <- county_reg_acs_data %>% 
  dplyr::filter(variable != 'B25014_001') %>% #remove total households
  mutate(per_bedroom = factor(
    case_when(variable == "B25014_002" ~ "Total",
              variable == "B25014_008" ~ "Total",
              grepl("0.50 or less|0.51 to 1.00|1.01 to 1.50", label) ~ "Less than or equal to 1.5 persons per bedroom",
              grepl("1.51 to 2.00|2.01 or more", label) ~ "Overcrowding",
              TRUE ~ as.character(label)))) %>%
  mutate(type = factor(
    case_when(grepl("B25014_002|B25014_003|B25014_004|B25014_005|B25014_006|B25014_007", variable) ~ "Owner",
              grepl("B25014_008|B25014_009|B25014_010|B25014_011|B25014_012|B25014_013", variable) ~ "Renter",
              TRUE ~ as.character(variable)))) %>%
  dplyr::select(GEOID, name, variable, per_bedroom, type, estimate, moe) #%>% #simplify table
# dplyr::filter(type != 'Owner')    #this can be commented out for inclusion of both Owners and Renters

county_reg_data_aggregate <- county_reg_data_rename %>% 
  dplyr::group_by(GEOID, name, per_bedroom) %>% #add 'type' if want to distinguish b/t Owners and Renters
  dplyr::summarise(estimate=sum(estimate), .groups='drop')

# pivot data to calculate percentage by census tract
county_reg_data_group <- county_reg_data_aggregate %>% 
  pivot_wider(names_from = per_bedroom, values_from = estimate) %>%
  dplyr::mutate(overcrowding_perc = ((Overcrowding/Total)*100),
                county_name=str_split(name, " ", simplify = TRUE)[,1]) #for join to tract level data

county_data <- county_reg_data_group %>% 
  dplyr::filter(name != "Region") %>% 
  dplyr::rename(cnty_estimate=overcrowding_perc) %>% #to distinguish between tract and county values
  dplyr::select(county_name, cnty_estimate) #simplify table

# getting data by region
region_data <- county_reg_data_group %>% 
  dplyr::filter(name == "Region") %>% 
  dplyr::rename(reg_estimate=overcrowding_perc) %>% #to distinguish between tract and county values
  dplyr::select(county_name, reg_estimate) #simplify table

# add region median to county data
county_reg_data <- county_data %>% 
  mutate(reg_estimate=region_data$reg_estimate)

# merge tract, county, and region data into one data set 
acs_data_tract <- merge(acs_data_tract_merge, county_reg_data,
                        by.x="county_name",
                        by.y="county_name",
                        all.x=TRUE)
```

### Explore map/acs data
```{r map legend settings, include=FALSE}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - this code helps to set up the legend so that it is arranged high-low with correct color order
addLegend_decreasing <- function (map, position = c("topright", "bottomright", "bottomleft", 
			    "topleft"), pal, values, na.label = "NA", bins = 9, colors, 
		  opacity = 0.5, labels = NULL, labFormat = labelFormat(), 
		  title = NULL, className = "info legend", layerId = NULL, 
		  group = NULL, data = getMapData(map), decreasing = FALSE) {
	position <- match.arg(position)
	type <- "unknown"
	na.color <- NULL
	extra <- NULL
	if (!missing(pal)) {
		if (!missing(colors)) 
			stop("You must provide either 'pal' or 'colors' (not both)")
		if (missing(title) && inherits(values, "formula")) 
			title <- deparse(values[[2]])
		values <- evalFormula(values, data)
		type <- attr(pal, "colorType", exact = TRUE)
		args <- attr(pal, "colorArgs", exact = TRUE)
		na.color <- args$na.color
		if (!is.null(na.color) && col2rgb(na.color, alpha = TRUE)[[4]] == 
		    0) {
			na.color <- NULL
		}
		if (type != "numeric" && !missing(bins)) 
			warning("'bins' is ignored because the palette type is not numeric")
		if (type == "numeric") {
			cuts <- if (length(bins) == 1) 
				pretty(values, bins)
			else bins	
			
			if (length(bins) > 2) 
				if (!all(abs(diff(bins, differences = 2)) <= 
				         sqrt(.Machine$double.eps))) 
					stop("The vector of breaks 'bins' must be equally spaced")
			n <- length(cuts)
			r <- range(values, na.rm = TRUE)
			cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
			n <- length(cuts)
			p <- (cuts - r[1])/(r[2] - r[1])
			extra <- list(p_1 = p[1], p_n = p[n])
			p <- c("", paste0(100 * p, "%"), "")
			if (decreasing == TRUE){
				colors <- pal(rev(c(r[1], cuts, r[2])))
				labels <- rev(labFormat(type = "numeric", cuts))
			}else{
				colors <- pal(c(r[1], cuts, r[2]))
				labels <- rev(labFormat(type = "numeric", cuts))
			}
			colors <- paste(colors, p, sep = " ", collapse = ", ")
			
		}
		else if (type == "bin") {
			cuts <- args$bins
			n <- length(cuts)
			mids <- (cuts[-1] + cuts[-n])/2
			if (decreasing == TRUE){
				colors <- pal(rev(mids))
				labels <- rev(labFormat(type = "bin", cuts))
			}else{
				colors <- pal(mids)
				labels <- labFormat(type = "bin", cuts)
			}
			
		}
		else if (type == "quantile") {
			p <- args$probs
			n <- length(p)
			cuts <- quantile(values, probs = p, na.rm = TRUE)
			mids <- quantile(values, probs = (p[-1] + p[-n])/2, 
				 na.rm = TRUE)
			if (decreasing == TRUE){
				colors <- pal(rev(mids))
				labels <- rev(labFormat(type = "quantile", cuts, p))
			}else{
				colors <- pal(mids)
				labels <- labFormat(type = "quantile", cuts, p)
			}
		}
		else if (type == "factor") {
			v <- sort(unique(na.omit(values)))
			colors <- pal(v)
			labels <- labFormat(type = "factor", v)
			if (decreasing == TRUE){
				colors <- pal(rev(v))
				labels <- rev(labFormat(type = "factor", v))
			}else{
				colors <- pal(v)
				labels <- labFormat(type = "factor", v)
			}
		}
		else stop("Palette function not supported")
		if (!any(is.na(values))) 
			na.color <- NULL
	}
	else {
		if (length(colors) != length(labels)) 
			stop("'colors' and 'labels' must be of the same length")
	}
	legend <- list(colors = I(unname(colors)), labels = I(unname(labels)), 
	               na_color = na.color, na_label = na.label, opacity = opacity, 
	               position = position, type = type, title = title, extra = extra, 
	               layerId = layerId, className = className, group = group)
	invokeMethod(map, data, "addLegend", legend)
}
```

```{r test map, include=FALSE,eval=FALSE}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
# psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
#                                       domain = acs_data_tract$estimate)
psrc_palette <- leaflet::colorFactor(palette=psrc_colors$purples_inc,
                                      domain = acs_data_tract$reliability)

# set the variable
var_name <- "Overcrowding (% of hh's with >1.5 persons)"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$reliability),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(acs_data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = acs_data_tract$reliability,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

```{r explore map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = acs_data_tract$estimate)

# set the variable
var_name <- "Overcrowding (% of hh's with >1.5 persons)"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(acs_data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = acs_data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

# **MAP .rda**
Save final data set (.rda) for map
```{r}
# This is the data set used to create the map of the most recent data. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk.

# set folder structure
file_name <- 'e04-overcrowding-map-data'

# save final data set as .rda
save(acs_data_tract, file = file.path(base_dir,
                                      theme_dir,
                                      ind_dir, "rda-data",
                                      paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>