---
title: "Renter Overcrowding"
subtitle: "Visuals for Equity Tracker (PUMS data)"
author: "Meg Grzybowski and Mary Richards, updated by Kristin Mitchell"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    word_document:
    html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'e-housing',
                   ind_dir = 'e04-overcrowding',
                   map = 'e04-overcrowding-map-data',
                   chart = 'e04-overcrowding-chart-data',
                   map_name = 'e04-overcrowding-map.html',
                   column_chart_name = 'e04-overcrowding-column',
                   line_chart_name = 'e04-overcrowding-line')
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Overcrowding is an indicator that people are living in homes that may be too small for their household needs or family size. For this measure, we are focusing on renter households and dividing the number of people in the household by the number of rooms in their home. We are defining overcrowding as households with more than 1.5 person(s) per room. Although overcrowding can occur in owner-occupied homes, renters are often at a disadvantage in terms of housing stability [this explanation needs more work about why we are focusing on renters].

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# move the decimal point for reg_estimate and cnty_estimate
# acs_data_tract <- acs_data_tract %>%
#    mutate(reg_estimate = (reg_estimate*100)) %>%
#    mutate(cnty_estimate = (cnty_estimate*100))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 250001, so rounding to 260000 makes sense
max_value <- 8 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
acs_data_plus_zero<-append(acs_data_tract$estimate,0)

# psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
#                                       domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                       domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Renter <br> Overcrowding" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "County: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body>",
  paste(acs_data_tract$county_name, " County"), # the code in <> makes the geoid value bold/italic
  # paste("Tract value: ", 
  #       prettyNum((round(acs_data_tract$estimate, digits=2))), " %"), 
  paste("County value: ",
        prettyNum((round(acs_data_tract$estimate, digits=1))), " %"),
  paste("County data reliability: ", acs_data_tract$reliability),
  paste("Region value: ",
        prettyNum((round(acs_data_tract$reg_estimate, digits=1))), " %")
  # , 
  # paste0("(",acs_data_tract$county_name," County)")
  ) %>%
  lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", acs_data_tract$geoid20,
#   paste(prettyNum((round(acs_data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)


# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              #  psrc_palette(acs_data_plus_zero),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       # values = acs_data_tract$estimate,
                       values = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))
# Highest percentage of SNAP participation is 53053940006 (east Tacoma where two low income housing complexes are - 49.4%), 53053061601 (downtown Tacoma - 44.1%), 53033029004 (southeast Des Moines - 39.3%), 53033029604 (southeast Kent - 35.8%), 53053063301 (east Tacoma just south of 53053940006 - 35.0%)

# national average
# acs table: S2201, https://data.census.gov/table/ACSST5Y2022.S2201?q=S2201

sort <- data %>% 
  dplyr::arrange(estimate)

filter <- data %>% 
  filter(estimate==0)
# There are 38 census tracts with a 0% participation. 
table(filter$county_name)
# 30 tracts in King County, 2 in Kitsap, 2 in Pierce, and 4 in Snohomish 
```

1. 6.3%: The regional average of overcrowded renter households
2. 2.5X: Among renter households in the region, the overcrowding rate for people of color is 2.5 times higher than white non-Hispanic people
3. 10%: The regional share of overcrowding for renter households with lower income is 10%, an 11% increase between 2012 and 2022

\

#### Insights & Analysis

* King and Snohomish counties have the highest rate of overcrowding among renters (6.8%), followed by Pierce (5.3%), and Kitsap (3.5%).
* The higher rate of overcrowding for renters in King and Snohomish counties is likely related to having lower access to affordable rent.
* Compared to King and Snohomish counties, Kitsap County’s overcrowding rate among renters is about half as high, while Pierce is almost four-fifths as high.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year=="2022") #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = .2
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Renter Overcrowding by Community"
subtitle = "Percent of households with more than 1 person per room"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 


```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# Region: 0.21 vs. 0.06
# with disability has higher coverage across all counties/region

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat")
# King lep households: .22	
# King non-lep households: .07

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")
# region low income households: .32	
# region non-low income households: .04
# .32-.04=.28

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="LEP_cat")
# region non-lep:0.09
# region lep: 0.21
# 0.21/0.09=1.857143

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="POC_cat")
# region non-poc:0.07
# region poc: 0.13
# 0.13/0.07=2.3333
```
\

#### Insights & Analysis

* A greater share of renters of color experience overcrowding compared to white non-Hispanic renters. The difference is highest in King County (eight percentage points), followed by Snohomish County (seven percentage points) while Kitsap and Pierce counties have much smaller differences (two and three percentage points, respectively).

* The share of the region’s renter households with lower income experiencing overcrowding is 10%, twice as high as other renter households - King County has the highest rate of overcrowding among renter households with lower income (11%), followed by Snohomish (10%), Pierce (9%), and Kitsap (5%).

* The regional share of renter households with limited English proficiency experiencing overcrowding is 15%, over two and a half times higher than renter households proficient in English (6%).

* Renter households with limited English proficiency have the highest rates of overcrowding across almost all of demographic groups, regardless of county, with the exception of renter households with youth in King, Kitsap, or Snohomish counties.


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "Renter Overcrowding Trend by Community"
subtitle = "Percent of households with more than 1 person per room, in 5-year spans between 2012 and 2022"
source = "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 


```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# king 2012: 0.17-0.07=0.10
# king 2022: 0.12-0.06=0.06
# kitsap 2012: 0.19-0.10=0.09
# kitsap 2022:0.14-0.08=0.06
# pierce 2012: 0.19-0.10=0.09
# pierce 2022:0.17-0.10=0.07
# snohomish 2012: 0.16-0.10=0.06
# snohomish 2022: 0.11-0.08=0.03

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="LEP_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# king 2022-2012: 0.22-0.24=-0.02
# kitsap 2022-2012: 0.13-0.18=-0.05
# pierce 2022-2012: 0.23-0.21=0.02
# snohomish 2022-2012: 0.18-0.26=-0.08

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Disability_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# King 2022-2012: 0.21-0.22=-0.01
# Kitsap 2022-2012: 0.20-0.20=0
# Pierce 2022-2012: 0.22-0.22=0
# Snohomish: 0.19-0.23=-0.04

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# King 2012 poc-non poc: 0.17-0.07=0.10
# King 2022: 0.12-0.06=0.06
# King: ((0.06-0.10)/(0.10))*100: -40% change
# Kitsap 2012: 0.19-0.10=0.09
# Kitsap 2022: 0.14-0.08=0.06
# Kitsap: ((0.06-0.09)/(0.09))*100: -33% change
# Pierce 2012:0.19-0.10=0.09
# Pierce 2022: 0.17-0.10=0.07
# Pierce: ((0.07-0.09)/(0.09))*100: -22% change
# Snohomish 2012: 0.16-0.10=0.06
# Snohomish 2022: 0.11-0.08=0.03 
# Snohomish: ((0.03-0.06)/(0.06))*100: -50% change


filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="Income_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# 2012: 0.35
# 2022: 0.32


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year=="2016" |
           data_year=="2021")


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="Income_cat",
         vulnerability=="low vulnerability",
         data_year=="2016" |
           data_year=="2021")

```
\

#### Insights & Analysis

* Regionally, the share of overcrowded renter households of color remained the same (10%) between 2012 and 2022, with a slight increase in 2017, while the overcrowding for white, non-Hispanic renter households increased from 2% to 4%.

* Renters of color have consistently experienced higher rates of overcrowding compared to white non-Hispanic renter households; however, between 2012 and 2022, the differences between these two groups decreased across most of the region: decreasing in Kitsap (-60%), Pierce (-50%), and King (-20%) counties while staying the same in Snohomish County.

* Overcrowding among renter households with lower income in the region slightly increased between 2012 and 2022 (going from 9% to 10%).

* Between 2012 and 2022, the share of overcrowded renter households with lower income increased 50% in Pierce County (going from 6% to 9%) and 11% in Snohomish County (going from 9% to 10%), while the shares remained the same in King and Kitsap counties (11% and 5%, respectively).

* Overcrowding experienced by renters with a disability stayed the same in King County (6%) between 2012 to 2022, while it increased two percentage points in Kitsap County (going from 3% to 5%), increased one percentage point in Pierce County (going from 5% to 6%), and increased two percentage points in Snohomish County (going from 6% to 8%).

* A higher share of renter households with limited English proficiency live in overcrowded housing than those proficient in English. Between 2012 and 2022, the difference between the two groups decreased in King (-25%) and Snohomish (-9%) counties and increased in Pierce (+29%) and Kitsap (+100%) counties. 

* The share of overcrowding for renter households with youth increased between 2012 and 2022 the most in Snohomish County (five percentage points), followed by an increase of two percentage points in King, Kitsap, and Pierce counties. 


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>