---
title: "Access to Affordable Rent"
subtitle: 'Visuals for Equity Tracker (PUMS data)'
author: "Mary Richards updated by Christy Lam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  word_document:
  # html_document:
  #   keep_md: yes
  #   df_print: paged
  #   toc: yes
  #   toc_depth: 6
  #   toc_float: yes

  
    
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
output_type <- "word"
# output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'e-housing',
                   ind_dir = 'e01-affordable-rent-access',
                   map = 'e01-affordable-rent-access-map-data',
                   chart = 'e01-affordable-rent-access-data',
                   map_name = 'e01-affordable-rent-access-map.html',
                   column_chart_name = 'e01-affordable-rent-access-column',
                   line_chart_name = 'e01-affordable-rent-access-line')

# set most recent year
year <- "2022"

# set past version year - this will set organize archived files in Y drive
past_year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Access to affordable rent allows people to choose to live in neighborhoods with the communities, services, and amenities that can support their households to thrive. Increasing access to housing provides greater opportunity for all people to choose safe and stable homes to meets their needs.

For this measure, we are focused on renter households and define a census tract as affordable (i.e., having access to affordable rent) if its median gross rent is equal to or less than 30% of the median household income for renters within the region or for a specific group. Across the region or a county, the share of affordable census tracts is calculated to represent the proportion of neighborhoods that is accessible to a typical renter household.

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
year <- "2022" # most recent year

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.


```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
psrc_palette <- leaflet::colorFactor(palette = psrc_colors$purples_inc,
                                     domain = acs_data_tract$affordability)

# setting the legend/label variables
var_name <- "Affordability"

labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", acs_data_tract$affordability),
  paste("Tract data reliability: ", acs_data_tract$reliability),
  paste("Regional monthly median income: ","$",
        prettyNum((round(acs_data_tract$reg_med_income_monthly, digits=-1)),
                  big.mark = ",")),
  paste("Regional affordability threshold*: ","$",
        prettyNum((round(acs_data_tract$income_30perc, digits=-1)),
                  big.mark = ",")),
  paste("Tract Median Gross Rent: ","$",
        prettyNum((round(acs_data_tract$estimate, digits=-1)),
                  big.mark = ","))) %>%
  lapply(htmltools::HTML)

# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$affordability),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = acs_data_tract$affordability,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```

Source(s): U.S. Census Bureau, 2018-2022 American Community Survey (ACS) 5-Year Estimates, Table B25064; U.S. Census Bureau, 2018-2022 ACS 5-Year Public Use Microdata Sample; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles

```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# table(data$affordability) 
# table(data$county_name, data$affordability)

data |> 
  group_by(affordability) |> 
  tally() |> 
  pivot_wider(names_from = affordability, values_from = n) |>
  mutate(denom = (Affordable + `Not affordable`)) |> 
  mutate(share_aff = Affordable / denom) # NA not included in denominator

# 2022 dollars = 383/(383 + 520) = 0.4241417
# 2011 Region = 347/(347+550) = 0.386845

data |> 
  group_by(county_name, affordability) |> 
  tally() |> 
  pivot_wider(names_from = affordability, values_from = n) |>
  mutate(denom = (Affordable + `Not affordable`)) |> 
  mutate(share_aff = Affordable / denom) # NA not included in denominator

# 2022 King: 150/(150+337) = .3080082
# 2022 Kitsap: 37/(37+21) = .6379310
# 2022 Pierce: 124/(124+68) = .6458333
# 2022 Snohomish: 72/(72+94) = .4337349

# 2021 King: 123/(123+362)= 0.2536082
# 2021 Kitsap: 41/(41+18)= 0.6949153
# 2021 Pierce: 114/(114+76)= 0.6
# 2021 Snohomish: 69/(69+94)= 0.4233129

sort <- data %>% 
  dplyr::arrange(desc(estimate))

highest_mgr <- acs_data_tract |>
  arrange(desc(estimate)) |>
  head(10)

highest_mgr_map <- create_map(highest_mgr, psrc_colors)
# highest_mgr_map
# Ten census tracts tied with the highest median gross rent ($3,500) are all in King County: 
# 1 is in Seattle (Montlake, south of UW)
# 2 are in Mercer Island, 
# 2 are in Clyde Hill/Hunts Point/Medina/Yarrow Point, 
# 4 are in Bellevue
# 1 in Auburn (West Hill)

# The highest median gross rent in Pierce, Kitsap, and Snohomish County

highest_mgr_cnty <- acs_data_tract |>
  group_by(county_name) |> 
  slice_max(order_by = estimate, n = 2) |> 
  filter(county_name != "King") |> 
  arrange(county_name, desc(estimate))

highest_mgr_map_kit <- create_map(highest_mgr_cnty |> filter(county_name == 'Kitsap'), psrc_colors)
highest_mgr_map_prc <- create_map(highest_mgr_cnty|> filter(county_name == 'Pierce'), psrc_colors)
highest_mgr_map_sno <- create_map(highest_mgr_cnty|> filter(county_name == 'Snohomish'), psrc_colors)

# highest_mgr_map_kit
# highest_mgr_map_prc
# highest_mgr_map_sno
```

1. 42%: Share of tracts in the region where rent is affordable based on the regional median household income
2. 1%: Regionally, 1% of census tracts are affordable to renter households with limited English proficiency, 46X less than the what is affordable to English-proficient renter households (46%)
3. 2.5X: The share of tracts with affordable rent decreased 45% for renters of color between 2012 and 2022, about 2.5 times more than the decrease for white non-Hispanic renters (38%)

\

#### Insights & Analysis

* Regional affordability threshold for monthly rent, based on 30% of the regional median household income for renters (\$71,91000/year), is \$1,800.
* Access to affordable rent is highest in Pierce (65% of tracts), followed by Kitsap County (64%), Snohomish (43%), and King (31%).
* Ten census tracts tied with the highest median gross rent ($3,500) are all in King County: one is in Seattle (Montlake neighborhood), two are in southern Mercer Island, two are in Clyde Hill/Hunts Point/Medina/Yarrow Point, four are in Bellevue, one is in Auburn (West Hill).
* The highest median gross rent in Kitsap, Pierce, and Snohomish County are in: the southeastern portion of Bainbridge Island (\$2,720), Tehaleh community south of Bonney Lake (\$3,090), and the Edmonds/southern Mukilteo area (\$3,280), respectively. 
* Relatively affordable areas are in south Seattle, south King County, Tacoma/central Pierce County, along the I-5 corridor in Snohomish County and in communities farther out from the region’s urban centers.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# # load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year == year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df <- data_clean_column
geo <- "county"
x <- "focus_attribute_ord"
y <- "fact_value"
facet <- "focus_type_ord"
fill <- "focus_attribute_ord"
y_min <- 0
y_max <- 1
dec <- 0
esttype <- "percent"
color <- "oranges"
num_colors <- 2
color_rev <- TRUE
title <- "Access to Affordable Rent by Community"
subtitle  <- "Percent of census tracts that are affordable"
source <- "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)

# column_chart
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information

## Older Cat ----
o_df <- data_clean_column %>% 
  filter(focus_type == "Older_cat")

o_df_wide <- o_df |> 
  select(county, focus_attribute_ord, fact_value_notrounded) |> 
  pivot_wider(id_cols = county,
              names_from = focus_attribute_ord,
              values_from = fact_value_notrounded) |> 
  mutate(diff = abs(`Households with\nolder adults` - `Other households`))

# see scripts:
# insights-analysis-config.R
# insights-analysis-map.R
# insights-analysis-charts.R

```
\

#### Insights & Analysis

* Renters with a disability have access to affordable rent in 3% of the region’s census tracts. 
* There is a 53 percentage point difference in the regional share of affordable tracts between renter households with older adults (1%) and other renter households (54%).
* Renters of color in King County have access to affordable rent in 32% of the census tracts, while white non-Hispanic renters may find 57% of the tracts to be affordable, the largest difference among the counties.
* While the renter affordability gap between people of color and white non-Hispanic people is notable, it is likely even bigger for some racial subgroups (more information about the way in which housing opportunities are different among racial subgroups is available in the Regional Housing Strategy: 2023 Monitoring Report.
* For households living below 200% of poverty level, there is no access to affordable rent.
* The disparity in access to affordable rent between households with older adults and other households is greatest in King County (60%), followed by Kitsap (48%), Pierce (42%), and Snohomish (37%).

\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df <- data_clean
x <- "data_year" # different from variables for facet column
fill <- "focus_attribute_ord" # different from variables for facet column
color <- "blues"
title <- "Access to Affordable Rent Trend by Community"
subtitle <- "Percent of census tracts that are affordable, in 5-year spans between 2012 and 2022"
source <- "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')

# line_chart
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information

### Biggest disparities ----

# is the gap between two lines (low vulnerability vs high vulnerability) lessening or widening?
disparity_df <- data_clean |> 
  select(data_year, county, vulnerability, focus_type, focus_attribute_ord, fact_value_notrounded) 

disparity_df_wide <- disparity_df |> 
  filter(data_year != middle_year) |> 
  pivot_wider(id_cols = c(data_year, county, focus_type),
              names_from = vulnerability,
              values_from = fact_value_notrounded) |> 
  relocate(`high vulnerability`, .after = `low vulnerability`) |> 
  mutate(disparity = `low vulnerability` - `high vulnerability`) |> 
  group_by(county, focus_type) |> 
  mutate(disparity_btwn_years = disparity-lag(disparity, 1)) |>  # numerator
  mutate(disparity_share = (disparity_btwn_years/lag(disparity, 1)))

disparity_top <- disparity_df_wide |> 
  filter(focus_type != "Youth_cat") |> 
  group_by(data_year, county) |> 
  slice_min(disparity_share, n = 10) |> 
  arrange(data_year, county, disparity_share)

### Deltas ----

delta_df <- data_clean |> 
  select(data_year, county, vulnerability, focus_type, focus_attribute_ord, fact_value_notrounded) |> 
  filter(data_year != middle_year) 

delta_years <- unique(delta_df$data_year)
names(delta_years) <- c("start", "end")

delta_df_wide <- delta_df |> 
  pivot_wider(id_cols = c(county, focus_type, vulnerability),
              names_from = data_year,
              values_from = fact_value_notrounded
              ) |> 
  rename(delta_years) |> 
  mutate(delta = end - start) |> 
  mutate(delta_share = delta/start)

### HH with Lower Income ----

li_dis_df <- disparity_df_wide |> 
  filter(focus_type == "Income_cat", county == "Region")

li_df <- data_clean |> 
  select(data_year, county, vulnerability, focus_type, focus_attribute_ord, aff_tracts, tot_tracts, fact_value_notrounded) |> 
  filter(data_year != middle_year, focus_type == "Income_cat", county == "Region")

# Other households, dropped 12 percentage points. Absolute number/difference for affordable tracts is 28. How does that 
# explain the drop when the denominator (tot tracts) is different across the two years?

### Disability ----

disab_df <- disparity_df_wide |> 
  filter(focus_type == "Disability_cat")

# The disparity in access to affordable rent between those with and without a disability decreased 
# over time for most counties, with the greatest difference in Snohomish (41 percentage points), 
# followed by Kitsap (32 percentage points), and King (14 percentage points). 
# The disparity for Pierce grew by 5 percentage points.

### LEP ----

lep_df <- disparity_df_wide |> 
  filter(focus_type == "LEP_cat")

# The disparity in access to affordable rent between households 
# with limited English proficiency and English proficient households decreased in Snohomish (40 percentage pts), 
# followed by  King (23 percentage points), and Pierce (17 percentage points), while disparity greatly increased in Kitsap. 
# Access to affordable rent in Kitsap dropped nearly 70 percentage points.

lep_delta_df <- delta_df_wide |> 
  filter(focus_type == "LEP_cat")

# The regional share of affordable tracts for renter households with limited English proficiency 
# decreased 80% between 2012 and 2022, a 2x larger decrease than for English proficient renter households (40%).


### Older Adults ----

older_df <- disparity_df_wide |> 
  filter(focus_type == "Older_cat")

### POC ----

delta_df_wide |> 
  filter(focus_type == "POC_cat", county == 'Region')

# 2.5X: The share of tracts with affordable rent decreased 45% for renters of color between 2012 and 2022, 
# about 2.5 times more than the decrease for white non-Hispanic renters (38%)

# see scripts:
# insights-analysis-config.R
# insights-analysis-map.R
# insights-analysis-charts.R
```
\

#### Insights & Analysis

* The difference in regional access to affordable rent between renter households with lower income and other renter households decreased 12 percentage points because the number of affordable tracts for other renter households decreased by 28. 
* The disparity in access to affordable rent between those with and without a disability decreased over time for most counties, with the greatest difference in Snohomish (41%), followed by Kitsap (33%) and King (14%). The disparity in Pierce grew by 5%.
* The regional share of affordable tracts for renter households with limited English proficiency decreased 86% between 2012 and 2022, a 2.2x larger decrease than for English proficient renter households (-39%).
* The disparity in access to affordable rent between households with limited English proficiency and English proficient households decreased in Snohomish (40 percentage points), followed by King (23 percentage points), and Pierce (18 percentage points), while the disparity greatly increased in Kitsap. 
* Snohomish is the only county where access to affordable rent for households with older adults stayed the same over time, compared to the other counties, where access decreased over time: Pierce (17 percentage points), King (6 percentage points), and Kitsap (4 percentage points).
\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. *Always run this chunk of code. If this is the first time you are generating visuals for an indicator, comment out the 'update' part of the 'y.drive.folder.update' variable because this is the first version that is being created.*
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(Rmd|R)$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r to archive, include=FALSE, eval=FALSE}
# add previous year folder to archive folder - to retain and organize all archived files
dir.create(file.path(file_names$base_dir,
                     file_names$theme_dir,
                     file_names$ind_dir,'archive', past_year))

# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive', past_year)

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall). *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage. *Always run this chunk of code.*
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>