---
title: "Housing Cost Burden"
subtitle: "Visuals for Equity Tracker (PUMS data)"
author: "Meg Grzybowski, Mary Richards, updated by Grant Gibson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
output_type <- "word"
# output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'e-housing',
                   ind_dir = 'e02-housing-cost-burden',
                   map = 'e02-housing-cost-burden-map-data',
                   chart = 'e02-housing-cost-burden-chart-data',
                   map_name = 'e02-housing-cost-burden-map.html',
                   column_chart_name = 'e02-housing-cost-burden-column',
                   line_chart_name = 'e02-housing-cost-burden-line')
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Housing cost burden, in this case, refers to renters that are paying more than 30% of their income on rent, rendering them "burdened" by the cost of housing. As rent increases and income remains unchanged, or as affordable housing becomes less available, people are over-extending themselves in order to afford housing. This contributes to the housing crisis and also to people being unable to access opportunity in the same way that affluent, or low-vulnerability, communities can.

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# move the decimal point for reg_estimate and cnty_estimate
# acs_data_tract <- acs_data_tract %>%
#    mutate(reg_estimate = (reg_estimate*100)) %>%
#    mutate(cnty_estimate = (cnty_estimate*100))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 100
max_value <- 100 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
acs_data_plus_zero<-append(acs_data_tract$estimate,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Renter Cost <br> Burden" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", 
        prettyNum((round(acs_data_tract$estimate, digits=0))), " %"), 
  paste("Tract data reliability: ", acs_data_tract$reliability),
  paste("Region value: ",
        prettyNum((round(acs_data_tract$reg_estimate, digits=0))), " %"), 
  paste("County value: ",
        prettyNum((round(acs_data_tract$cnty_estimate, digits=0))), " %"), 
  paste0("(",acs_data_tract$county_name," County)")) %>%
  lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", acs_data_tract$geoid20,
#   paste(prettyNum((round(acs_data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)


# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_plus_zero),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

sort <- data %>% 
  dplyr::arrange(desc(estimate))
# 53033031709 (King)
# 53033032102 (King), 
# 53033032328 (King), 
# 53061050403 (Snohomish), 
# 53061050600 (Snohomish),
# 53061053605 (Snohomish), 

sort <- data %>% 
  dplyr::arrange(estimate)
# 53033005304 (Seattle UW), 53033005305 (Seattle UW), 53053061601 (Tacoma)
# 53033030501 (Auburn, around 167)

filter <- data %>% 
  dplyr::filter(reliability=="estimate is 0, cannot compute")

filter <- data %>% 
  dplyr::filter(estimate>50) #409/919:
table(filter$county_name)
# King: 201/409: 49%
# Kitsap: 28/409: 7% 
# Pierce: 97/409: 24%
# Snohomish: 83/409: 20%
table(data$county_name)
# King: 494/919: 54%
# Kitsap: 59/919: 6%
# Pierce: 193/919: 21%
# Snohomish: 173/919: 19%

filter_tracts_ok <- data %>% 
  dplyr::filter(reliability=="good" |
                  reliability=="fair" |
                  reliability=="use with caution")

filter <- data %>% 
  dplyr::filter(estimate==50)

# calculate population density
# get population for 2022
tract_acs_data <- psrccensus::get_acs_recs(geography ='tract', 
                                           table.names = 'B01003', 
                                           years = c(2022), 
                                           acs.type = 'acs5')
tract_pop <- tract_acs_data %>% 
  dplyr::select(GEOID, estimate) %>% 
  dplyr::rename(pop_estimate=estimate)

# join population data to 
data_2022pop <- merge(data,
                      tract_pop,
                      by.x="geoid20",
                      by.y="GEOID",
                      all.x=TRUE)

# calculation
calc <- data_2022pop %>% 
  mutate(pop_acres=pop_estimate/land_acres) %>% 
  mutate(area_sqmi=aland20*3.861E-07) %>% 
  mutate(pop_sqmi=pop_estimate/area_sqmi)

# 50+ 
calc_50plus <- calc %>% 
  dplyr::filter(estimate>=50) %>% 
  dplyr::arrange(estimate)

calc_50minus <- calc %>% 
  dplyr::filter(estimate<50)

median(calc_50plus$pop_acres) #6.965862
median(calc_50minus$pop_acres) #6.452537
median(calc_50plus$pop_sqmi) #4501.258
median(calc_50minus$pop_sqmi) #4167.978

# quintiles
quartile_burden <- calc %>% 
  dplyr::mutate(quartile = ntile(estimate, 4)) %>% 
  dplyr::group_by(quartile) %>% 
  dplyr::summarise(median_dens_acres=median(pop_acres),
                   median_dens_sqmi=median(pop_sqmi),
                   mean_dens_acres=mean(pop_acres),
                   mean_dens_sqmi=mean(pop_sqmi))
```

1. 48%: Regional share of renter households who are cost burdened
2. 3/5: Almost three-fifths of households with limited English proficiency and households with older adults are rent cost burdened
3. 23%: The gap in the regional share of renter cost burden between people with and without disabilities decreased by 23% between 2012 and 2022

\

#### Insights & Analysis

* Pierce and Snohomish Counties have the highest percentage of cost burdened households (52%), followed by Kitsap (51%) and King (46%).
* In almost half (45%) of the region’s census tracts, cost burdened households make up more than half of the renters.
* Half of these majority cost burdened renter household tracts are in King County (49%), followed by Pierce (24%), Snohomish (20%), and Kitsap (7%), which roughly reflect the distribution of tracts across the region (King 54%, Pierce 21%, Snohomish 19%, and Kitsap 6%) and point to this being a region-wide problem.
* The median population density of the tracts where more than half of the renter households are cost burdened is higher (7 people/acre) than that of the tracts where fewer than half of households are rent-burdened (6 people/acre).

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year=="2022") #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = 1
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Renter Cost Burden by Community"
subtitle = "Percent of renters paying more than 30% of income on housing"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")

# low-income rent burden regional: 80%
# region non-low income households: 30%
# 2.7% more likely

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat")

#         county=="Region")
# 50% versus 60%

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat",
         vulnerability=="high vulnerability")

# same across the region

filter <- data_clean_column %>% 
  filter(focus_type=="Youth_cat")

# region youth households: 141389.8	
# region non-youth households: 102097.0
# 141389.8-102097.0=39292.8

filter <- data_clean_column %>% 
  filter(focus_type=="POC_cat")
# snohomish poc: 108886.4
# snohomish non-poc: 112974.4
# 112974.4-108886.4=4088

# king poc: 111529.77
# king non-poc: 130977.15
# 130977.15-111529.77=19447.38

filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# kitsap dis: 84171.34
# kitsap non-dis: 104284.10
# 104284.10-84171.34=20112.76

# king dis: 81508.16
# king non-dis: 134637.12
# 134637.12-81508.16=53128.96

filter <- data_clean_column %>% 
  filter(focus_type=="Income_cat",
         vulnerability=="high vulnerability") %>% 
  dplyr::arrange(desc(fact_value))
```
\

#### Insights & Analysis

* People of color are more likely to be rent cost burdened than their white non-Hispanic counterparts, with the largest percentage point gap in Snohomish County (7%), followed by Pierce (3%), King (2%), and Kitsap (1%).
* Renter households with lower income in the region are 2.4 times more likely to be cost burdened than other renter households.
* Regionally, 57% of people with disabilities are rent cost burdened, 13% higher than those without disabilities.
* Renter households with limited English proficiency are most cost burdened in Snohomish County (65%) and least in Kitsap and Pierce Counties (53%).
* Snohomish is the only county where the share of renter households with youth (18 years or younger) that are cost burdened (50%) is lower than the share for those without (52%).


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "Renter Cost Burden Trend by Community"
subtitle = "Percent of renters paying more than 30% of income on housing"
source = "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Region",
         focus_type=="Income_cat")

filter <- data_clean %>%
  filter(county=="Region",
         focus_type=="POC_cat",
         indicator_attribute != "Less than 30 percent")
# 2021 poc: 102481.29
# 2021 non-poc: 118053.44
# 118053.44-102481.29=15572.15

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Youth_cat",
         indicator_attribute != "Less than 30 percent",
         data_year==2012 |
           data_year==2022)
# king 2012 youth-no youth: 0.52-0.44=0.08
# king 2022 youth-no youth: 0.48-0.44=0.04
# kitsap 2012 youth-no youth: 0.52-0.50=0.02
# kitsap 2022 youth-no youth: 0.50-0.48=0.02
# pierce 2012 youth-no youth: 0.54-0.48=0.06
# pierce 2022 youth-no youth: 0.51-0.48=0.03
# snohomish 2012 youth-no youth: 0.53-0.45=0.08
# snohomish 2022 youth-no youth: 0.50-0.52=0.02

filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="Disability_cat",
         data_year==2012 |
           data_year==2022)
# 2012 disability-no disability: 0.60-0.44=0.16
# 2022 disability-no disability: 0.57-0.44=0.13
# ((0.16-0.13)/(0.13))*100: 23.07692% change

filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="LEP_cat",
         data_year==2012 |
           data_year==2022)
# 2012 lep-no lep: 0.59-0.47=0.12
# 2022 lep-no lep: 0.57-0.46=0.11

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year=="2011" |
           data_year=="2021")
# 2011 king:82987.29-66233.98=16753.31
# 2021 king:130977.15-111529.7=19447.45
# ((19447.45-16753.31)/(16753.31))*100: 16.08124% change
# 2011 kitsap:101045.37-78865.29=22180.08
# 2021 kitsap:103185.85-85187.57=17998.28
# ((17998.28-22180.08)/(22180.08))*100: -18.85385% change
# 2011 pierce:87650.57-77922.34=9728.23
# 2021 pierce:103124.26-84547.06=18577.2
# ((18577.2-9728.23)/(9728.23))*100: 90.96177% change
# 2011 snohomish:80736.67-64959.63=15777.04
# 2021 snohomish:112974.44-108886.37=4088.07
# ((4088.07-15777.04)/(15777.04))*100: -74.08849% change

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Older_cat",
         vulnerability=="high vulnerability",
         data_year=="2012" |
           data_year=="2022",
         indicator_attribute != "Less than 30 percent")

```
\

#### Insights & Analysis

* Regionally, a little less than half of renters of color are rent cost burdened in 2022 (48%), a decrease of three percentage points from 2012 (51%).
* The regional share of rent cost burdened households with lower income has remained unchanged between 2012 and 2022, while the share of rent cost burdened households with higher income has increased by 6 percentage points during the same period.
* The gap in the regional share of renter cost burden between households with limited English proficiency and households proficient in English narrowed by one percentage point between 2012 (12 percentage points) and 2022 (11 percentage points).
* Kitsap is the only county where the difference between renter households with youth and those without has remained constant between 2012 and 2022, while the difference has decreased in Snohomish (6 percentage points), King (4 percentage points), and Pierce (3 percentage points).
* The share of cost burdened renter households with older adults has decreased in Pierce County (4 percentage points) and King County (3 percentage points) between 2012 and 2022, while it has increased in Kitsap (1 percentage point) and Snohomish (3 percentage points).


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>