---
title: "SNAP" 
subtitle: "Data Gen: Exploring, Cleaning, Transforming (PUMS/OSPI data)"
author: "Kristin Mitchell, updated by Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

# Organize workspace
```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# install.packages("devtools")
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)

library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)

library(vtable) #summary stats
library(forcats) #for factor re-leveling

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(ggspatial)
library(lubridate) #year formatting
library(readxl)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(here)
```

```{r sources}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - this code helps to set up the map legend so that it is arranged high-low with correct color order

# reference supplemental script with supporting settings/functions
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))
```

```{r variables setup}

file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'a-regional-health-collaboration',
                   ind_dir = 'a05-snap',
                   chart = 'a05-snap-chart-data',
                   map = 'a05-snap-map-data')
```

# Download data 
## Primary data 
The data set referenced in this script is generated from the Census Bureau’s American Community Survey (ACS) Public Use Microdata Sample [(PUMS)](https://www.census.gov/programs-surveys/acs/microdata/documentation.html). These data sets provide data at the **person- or household-level**, with the ability to look at the different indicators by the six equity demographic groups of interest. 
```{r, include=FALSE}
# pums_ospi_year <- "2021"
# years <- c("2011", "2016", "2021") # these years are no longer available/reproducible

pums_ospi_year <- "2022"
years <- c("2012", "2017", "2022")

# PUMS/OSPI data set
pums_ospi_elmer <- get_table(schema = "equity", tbl_name = "v_tracker_indicators")
```
\
This data set was compiled from PUMS data.
\
Looking at the fields in the data set:
```{r}
# explore the data
head(pums_ospi_elmer)

# view the equity demographic groups, different indicators, and data years
pums_ospi_cols <- c('focus_type', 'indicator_type', 'data_year') 
walk(pums_ospi_cols, ~print(unique(pums_ospi_elmer[[.x]])))
```

```{r indicator specific information, include=FALSE}
# set indicator information
indicator_measure <- 'SNAP'
indicator_title <- 'SNAP'
```


## 1. Explore data
*In this section we make sure that the data set has the correct dimensions - year, geography, equity focus group, etc.*  
```{r}
# clean data set
data_full <- pums_ospi_elmer %>% 
  # distinct(indicator_fact_id, .keep_all = TRUE) %>% #this may be necessary depending on indicator
  dplyr::filter(indicator_type==indicator_measure) %>% 
  dplyr::filter(focus_type!="Total") %>% 
  filter(focus_attribute!="Total") %>%
  filter(indicator_attribute!="Total") %>%
  dplyr::mutate(data_year_yr=format(data_year,format="%Y"))
```

### Data fields
#### *consistent base data*

* There should be 5 geographies - the 4 counties and the Region. 
* There should be 6 equity focus categories - POC, income, disability, youth, older adult, and LEP
    + 2 sub-groups in each of the 6 equity focus group (e.g. people of color, non-people of color)
```{r, include=FALSE}
# geographies
if(length(unique(data_full$county)) < 4){
  warning("There is a missing geography. Please check.")
} else if (length(unique(data_full$county)) < 5) {
  warning("There is a missing geography. Please check.")
} else {
  message("All geographies are included: King, Kitsap, Pierce, Snohomish, and Region.")
}

# equity focus groups
if(length(unique(data_full$focus_type)) < 5){
  warning("There are missing equity groups. Please check.")
} else if (length(unique(data_full$focus_type)) < 6) {
  warning("There is a missing equity group. Please check.")
} else {
  message("All 6 equity groups are included.")
}

# equity focus attributes
if(length(unique(data_full$focus_attribute)) < 11){
  warning("There are missing equity sub-groups. Please check.")
} else if (length(unique(data_full$focus_attribute)) < 12) {
  warning("There is a missing equity sub-groups. Please check.")
} else {
  message("All 12 equity sub-groups are included.")
}
```

#### *indicator-specific data*
These fields will vary by indicator:

* Type of metric - this will determine how the data are visualized *(est ="percent" or "currency" or "number")*
* Number of years (5-year span) - this can vary depending on data availability
* Number of indicator-specific categories - this can vary depending on the indicator of interest, ranging from N/A (*median income*) to multiple levels (*crowding, housing cost burden*)
```{r}
# metric
data_type <- unique(data_full$fact_type) #median
message("This indicator's data type/format: ", data_type)

# years
num_yr <- length(unique(data_full$data_year)) #3
range <- list(paste(as.character(unique(data_full$data_year)),collapse=", ",sep=""))
message("There are ", num_yr, " available years: ", range)

# indicator attributes
num_indatt <- length(unique(data_full$indicator_attribute)) #1
attributes_list <- list(paste(unique(data_full$indicator_attribute), collapse=", ", sep=""))
if(length(unique(data_full$indicator_attribute)) > 1){
  warning(paste("There are",num_indatt, "indicator attributes: ",
              attributes_list, ". Please choose 1."))
} else {
  message("There is only one indicator attribute: ", attributes_list,
              ". No need to filter.")
}
```

```{r}
# space to filter if necessary
data_full <- data_full %>%
  filter(indicator_attribute=="Yes")
```

```{r}
# calculate expected number of rows
num_row <- num_yr*5*6*2 #180
message("There should be data for each year (",num_yr,"), geography (5), equity focus group (6), and equity sub-group (2) = ",num_row ," rows")

if(nrow(data_full) < num_row){
  warning(paste("There are", nrow(data_full), "rows, which is fewer than expected. Please check."))
} else if (nrow(data_full )> num_row){
  warning(paste("There are", nrow(data_full), "rows, which is more than expected. Please check and filter as necessary."))
} else {
  message(paste("There are the expected number of rows in the dataset:", nrow(data_full)))
}
```

```{r check data elements}
# extra detail: if there are any missing elements, can run this code
check_data_fields <- function(column) {
  l <- length(unique(column))
  v <- unique(column)
  return(list(length = l, values = v))
}

data_fields <- c("num_county" = "county", # geographies
                 "num_group" = "focus_type", # equity focus groups
                 "metric" = "fact_type", 
                 "num_yr" = "data_year", 
                 "num_sub_group" = "focus_attribute") #equity sub-focus groups

data_fields_summary <- map(data_fields, ~check_data_fields(data_full[[.x]]))

data_fields_summary


# calculate expected number of rows
num_row <- data_fields_summary |> 
  map(~pluck(.x, "length")) |> # isolate nums
  discard_at("metric") |> # remove metric element
  reduce(`*`) * 2 # multiply all numbers and by 2 for subgroups

#180
```

#### *checking for missing data*
##### Year / geography
```{r}
# Function in equity-tracker-supplemental-script.R
num_yr_geo <- check_missing_data(vars = c("num_yr", "num_county"),
                                 multiply_by_subgroups = FALSE)
#15
```

```{r}
table(data_full$data_year,
      data_full$county)
```


##### Year / equity focus group
```{r}
num_yr_grp <- check_missing_data(vars = c("num_yr", "num_group"),
                                 multiply_by_subgroups = FALSE)
#18
```

```{r}
table(data_full$data_year,
      data_full$focus_type)
```


##### Year / equity focus sub-group
```{r}
num_yr_subgrp <- check_missing_data(vars = c("num_yr", "num_sub_group"),
                                    multiply_by_subgroups = FALSE)
#36
```

```{r}
table(data_full$data_year,
      data_full$focus_attribute)
```
\
\

### Numeric data
To check for *0*s and *NULL*s
```{r}
# checking for NULLs
summary(data_full$fact_value)
```
\
\

To look at distribution of all data - not the most useful visual, but provides a sense of the range of values at a high level in one plot. 
\
```{r}
qplot(data_year_yr, fact_value, 
      colour = county, shape = focus_type, 
      data = data_full, 
      main = indicator_title) + theme(axis.title.x = element_blank(),
                                               axis.title.y = element_blank())
```
\
This table includes a lot of information about the data set and helps to show the different levels of each field. It provides another way to check if data are available for all counties and all years, or where there may be gaps in the data set. 
```{r}
data_refined <- data_full |> 
  select(data_year_yr, county, vulnerability, focus_type, focus_attribute,
        indicator_type, indicator_attribute, fact_value)

sumtable(data_refined,
         add.median = TRUE,
         group = "county",
         out = "return")
```

```{r}
# wrap/order labels ----
# variable ordering and label function in equity-tracker-supplemental-script.R
data_clean <- transform_data_labels(data_full)
```

```{r}
# indicator specific re-ordering - this step is unnecessary for SNAP and likely for most indicators since we are focused on one aspect of the variable
# data_clean$indicator_attribute <- factor(data_clean$indicator_attribute,
#                                     levels = c("Less than a Bachelor's degree",
#                                                "Bachelor's degree or higher"))
```
\

### Data labels, shares
These charts were generated to ensure the labels across years are consistent/make sense. There had been an issue with misassigned labels because tidycensus::pums_variables, i.e. the only digital data dictionary available to associate labels with codes, exists only from 2017 forward. Most variables have had consistent codes, but in cases where the codes have shifted over time, using the 2017 lookup winds up mischaracterizing categories.
\
\
These charts also help to confirm that the shares add up to 100% - only relevant when *indicator_attribute* has more than one category. The *indicator_attribute* for median household income is NA.
\
\
*The colors of the charts may not be consistent between the years depending on missing data.* 
\
```{r}
check_region <- function(df, year) {
  df |> 
    filter(county == "Region") |> 
    filter(data_year == year) |> 
    arrange(focus_attribute_ord)
}

check_region_facet <- function(df, year) {
  static_facet_column_chart(t = df,
                            x = "data_year", 
                            y = "fact_value",
                            fill = "vulnerability", 
                            facet = "focus_type_ord",
                            est = "percent",
                            scales = "fixed",
                            color = "psrc_pairs", 
                            ncol = 3,
                            title = paste0(indicator_title, ": Region (", year, ")"),
                            subtitle = paste(str_to_title(str_replace_all(indicator_measure, "_", " ")), "by Equity Focus Group"),
                            source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")
}

# the years below reflect the years that are available from unique(data_year)
just_region_dfs <- map(years, ~check_region(df = data_clean, 
                                            year = .x)) |> 
  set_names(paste0('r', years))

# check est type - make sure it corresponds to your indicator (percent, currency, or number)
check_region_facets <- map2(just_region_dfs, 
                            years, 
                            ~check_region_facet(df = .x, year = .y))

walk(check_region_facets, ~print(.x))

```

## 2. Visually explore data
### Scatter plots
*In this section we start to explore the data visually - distribution by the different dimensions within the data set. These plots are helpful to check for outliers and get a higher level understanding of the data in one visual, before slicing the data by geography and equity focus group in the following sections.* 
\
\
The following code will need to be adjusted to fit the fields specific to the data indicator. For educational attainment, we focus on those with a Bachelor's degree or higher. The following code establishes the data frame that the rest of the analysis uses. If there are fewer than 2 *indicator attributes*, this section can be skipped/commented out, but the code will need to be adjusted throughout. 
\
\

#### By *indicator_attribute*
This section isn't relevant for this specific indicator because there aren't unique indicator attributes. 

#### By Year
```{r}
# separating by year ----

data_clean_dfs <- map(years, ~data_clean |> filter(data_year == .x))
```

```{r}
# plot data, adjust y limits based on data set
data_clean_plot <- function(df, year) {
  qplot(focus_type_ord, 
        fact_value, 
        colour = county, 
        shape = vulnerability, 
        data = df, 
        main = paste0(indicator_title, ": ", year)) + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()#,
          # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
          ) + 
    ylim(0, (max(data_clean$fact_value)*1.5))
}

data_clean_plots <- map2(data_clean_dfs, years, ~data_clean_plot(.x, .y))

walk(data_clean_plots, ~print(.x))
```


# **CHART .rda**
```{r, include=FALSE}
head(data_clean) # make sure this is the correct data set

# make any additional edits 
data_clean <- data_clean %>% 
  mutate(fact_value_notrounded=fact_value) %>% #create column to retain full median income values
  mutate(fact_value=round(fact_value, digits=2)) #round median income value to hundreds and set retain consistent name - if you are not working with $ you will need to change this to the correct number of digits you would like to round to (for example, if using decimals, you may want to set digits=2 or if you using whole numbers you may want to set digits=0)

data_clean #check to make sure the rounding worked for your indicator - if not will need to go back to generate data set in chunk around line 245
```
Save final data set (.rda) for charts
```{r}
# save final data set as .rda
save(data_clean, file = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir, 
                                  "update",
                                  "rda-data",
                                  paste0(file_names$chart,'.rda')))
```


## Create data set for map .rda 
### Additional data
#### Tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
arc_service <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services"

tracts20.url <- file.path(arc_service, "Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
tracts10.url <- file.path(arc_service, "Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")

tracts20.lyr <- st_read(tracts20.url)
tracts10.lyr <- st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

#### Corresponding ACS table
```{r acs data, include=FALSE}
# Finding the corresponding ACS table - this works if you know the correct concept label. If not, another option would be to visit https://data.census.gov/table and search for the right subject table and skip to the next step ----
# x <- tidycensus::load_variables(2021,"acs5") %>% 
#   dplyr::filter(grepl("median household income",
#                       concept, ignore.case=TRUE) & 
#                   (geography=="tract"))

# getting data by tract
base_acs_data <- get_acs_recs(geography ='tract', 
                              table.names = 'S2201', #subject table code
                              years = c(as.numeric(pums_ospi_year)),
                              acs.type = 'acs5')


#filter for percentage instead of total number
acs_data_per <- base_acs_data %>%
  dplyr::filter(variable == "S2201_C04_001")

# filter only fields of interest - census tract, value, and reliability (may need to be adjusted depending on indicator and if you need to run any calculations/transformations) - look at labels and filter on variable if necessary
acs_data <- acs_data_per %>% 
  dplyr::select(GEOID, estimate, reliability)

# you may need to do some additional data wrangling to get the acs data into the desired format - for example, aggregating education attainment to two categories - less than bachelor's and bachelors and higher or dividing number of people 25+ who have a bachelor's degree or higher by the total population 25+ to get the share
```

### Joining data
```{r}
# merge to 2020 census tract spatial file
acs_data_tract_merge <- merge(tracts20.lyr, acs_data,
                              by.x="geoid20",
                              by.y="GEOID",
                              all.x=TRUE)

# getting data by county from acs 
county_acs_data <- get_acs_recs(geography ='county', 
                                table.names = 'S2201', #subject table code
                                years = c(as.numeric(pums_ospi_year)),
                                acs.type = 'acs5')

#filter for percentage instead of total number
# county_acs_data_per <- county_acs_data %>%
#   dplyr::filter(variable == "S2701_C03_001")
# 
# county_data <- county_acs_data_per %>% 
#   dplyr::select(name,estimate) %>% #to simplify table
#   dplyr::rename(cnty_estimate=estimate) %>% #to distinguish between tract and county values
#   mutate(county_name=str_split(name, " ", simplify = TRUE)[,1]) #for join to tract level data

# getting data by region from PUMS
# pums_data <- get_psrc_pums(span = 5,
#                            dyear = 2021,
#                            level = "p",
#                            vars = c("PRIVCOV", "PUBCOV"))
# 
# reg_median <- psrc_pums_median(pums_data, stat_var = c("PRIVCOV", "PUBCOV"))
# 
# # add region median to county data
# county_reg_data <- county_data %>% 
#   mutate(reg_estimate=reg_median$POVPIP_median)
# 
# # merge tract, county, and region data into one data set 
# acs_data_tract <- merge(acs_data_tract_merge, county_reg_data,
#                         by.x="county_name",
#                         by.y="county_name",
#                         all.x=TRUE)


# county transformations
county_snap <- county_acs_data %>% 
  dplyr::filter(variable=="S2201_C03_001",
                name!="Region") %>% 
  dplyr::select(name, estimate) %>% 
  dplyr::rename(snap=estimate)
county_pop <- county_acs_data %>% 
  dplyr::filter(variable=="S2201_C01_001",
                name!="Region") %>% 
  dplyr::select(name, estimate) %>% 
  dplyr::rename(pop_snap=estimate)

county_merge <- merge(county_snap, county_pop,
                      by.x="name",
                      by.y="name",
                      all.x=TRUE)
county_data_simp <- county_merge %>% 
  dplyr::mutate(cnty_estimate=snap/pop_snap) %>% 
  dplyr::mutate(county_name=str_split(name, " ", simplify = TRUE)[,1]) %>% #for join to tract level data
  dplyr::select(county_name,cnty_estimate) #to simplify table

# region transformations
reg_snap <- county_acs_data %>% 
  dplyr::filter(variable=="S2201_C03_001",
                name=="Region") %>% 
  dplyr::select(name, estimate) %>% 
  dplyr::rename(snap=estimate)
reg_pop <- county_acs_data %>% 
  dplyr::filter(variable=="S2201_C01_001",
                name=="Region") %>% 
  dplyr::select(name, estimate) %>% 
  dplyr::rename(pop_snap=estimate)

reg_merge <- merge(reg_snap, reg_pop,
                      by.x="name",
                      by.y="name",
                      all.x=TRUE)
reg_data <- reg_merge %>% 
  dplyr::mutate(reg_estimate=snap/pop_snap) %>% 
  dplyr::select(name,reg_estimate)

# add region median to county data
county_reg_data <- county_data_simp %>% 
  mutate(reg_estimate=reg_data$reg_estimate)

# merge tract, county, and region data into one data set 
acs_data_tract <- merge(acs_data_tract_merge, county_reg_data,
                        by.x="county_name",
                        by.y="county_name",
                        all.x=TRUE)
```

### Explore map/acs data
```{r explore map}
# map settings and legend function in equity-tracker-supplemental-script.R

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = acs_data_tract$estimate)

# set the variable
var_name <- "SNAP"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(acs_data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette, #included in equity-tracker-supplemental-script.R
                       values = acs_data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

# **MAP .rda**
Save final data set (.rda) for map
```{r}
# This is the data set used to create the map of the most recent data. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk.

# save final data set as .rda
save(acs_data_tract, file = file.path(file_names$base_dir,
                                      file_names$theme_dir,
                                      file_names$ind_dir, 
                                      "update",
                                      "rda-data",
                                      paste0(file_names$map,'.rda')))
```

<a href="#top">Back to top of the page</a>