---
title: "SNAP"
subtitle: "Visuals for Equity Tracker (PUMS data)"
author: "Kristin Mitchell, updated by Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'a-regional-health-collaboration',
                   ind_dir = 'a05-snap',
                   map = 'a05-snap-map-data',
                   chart = 'a05-snap-chart-data',
                   map_name = 'a05-snap-map.html',
                   column_chart_name = 'a05-snap-column',
                   line_chart_name = 'a05-snap-line')
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
The Supplemental Nutrition Assistance Program (SNAP) provides food benefits for households with lower income. This helps families supplement their food budget and access healthy food options. It is useful to identify communities who have many participants in SNAP to help address food security and food access disparities.

The Census Bureau question asks if an individual or household member has received SNAP benefits in the past 12 months.

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# move the decimal point for reg_estimate and cnty_estimate
acs_data_tract <- acs_data_tract %>%
  # mutate(reg_estimate = formatC( round((reg_estimate*100), 1), format='f', digits=1)) %>%
  mutate(reg_estimate = sprintf('%.1f',reg_estimate*100)) %>% #keep zero after decimal, while round to nearest tenth
  mutate(cnty_estimate = sprintf('%.1f',cnty_estimate*100))

# acs_data_tract <- acs_data_tract %>%
#    mutate(reg_estimate = round((reg_estimate*100), 1)) %>%
#    mutate(cnty_estimate = round((cnty_estimate*100),1))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 49.4, so rounding to 50 makes sense
max_value <- 50 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
acs_data_plus_zero<-append(acs_data_tract$estimate,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "SNAP" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", 
        prettyNum((round(acs_data_tract$estimate, digits=2))), " %"), 
  paste("Tract data reliability: ", acs_data_tract$reliability),
  # paste("Region value: ",
  #       prettyNum((round(acs_data_tract$reg_estimate, digits=2))), " %"), 
  # paste("County value: ",
  #       prettyNum((round(acs_data_tract$cnty_estimate, digits=2))), " %"), 
  paste("Region value: ", acs_data_tract$reg_estimate, " %"), 
  paste("County value: ", acs_data_tract$cnty_estimate, " %"), 
  paste0("(",acs_data_tract$county_name," County)")) %>%
  lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", acs_data_tract$geoid20,
#   paste(prettyNum((round(acs_data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)


# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))
# Highest percentage of SNAP participation is 53053940006 (east Tacoma where two low income housing complexes are - 49.4%), 53053061601 (downtown Tacoma - 44.1%), 53033029004 (southeast Des Moines - 39.3%), 53033029604 (southeast Kent - 35.8%), 53053063301 (east Tacoma just south of 53053940006 - 35.0%)

# national average
# acs table: S2201, https://data.census.gov/table/ACSST5Y2022.S2201?q=S2201

sort <- data %>% 
  dplyr::arrange(estimate)

filter <- data %>% 
  filter(estimate==0)
# There are 38 census tracts with a 0% participation. 
table(filter$county_name)
# 30 tracts in King County, 2 in Kitsap, 2 in Pierce, and 4 in Snohomish 
```

1. 9.0%: The region’s average SNAP participation
2. 49.4%: The highest average SNAP participation at the census tract level in the region
3. 3.5x: People with a disability are more than 3.5 times more likely to participate in SNAP

\

#### Insights & Analysis

* King County has the lowest average for SNAP participation (8.0%), followed by Snohomish (8.7%), Kitsap (9.6%), and Pierce (11.6%).
* There are 38 census tracts with no SNAP participation with 30 tracts in King County, two in Kitsap, two in Pierce, and four in Snohomish.
* The four census tracts with the highest participation include one tract in the Eastside neighborhood of Tacoma (49.4%), Downtown Tacoma (44.1%), southeast Des Moines (39.3%), and southeast Kent (35.8%).
* The national average SNAP participation is 11.5%.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year=="2022") #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = 0.5
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "SNAP Participation by Community"
subtitle = "Percent of households participating in SNAP"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# Region: 0.21 vs. 0.06
# with disability has higher coverage across all counties/region

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat")
# King lep households: .22	
# King non-lep households: .07

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")
# region low income households: .32	
# region non-low income households: .04
# .32-.04=.28

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="LEP_cat")
# region non-lep:0.09
# region lep: 0.21
# 0.21/0.09=1.857143

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="POC_cat")
# region non-poc:0.07
# region poc: 0.13
# 0.13/0.07=2.3333
```
\

#### Insights & Analysis

* In King County, households with limited English proficiency are more than three times as likely to participate in SNAP.
* As a region, 32% of households with lower income participate in SNAP, leaving room for increased participation. 
* SNAP uses an income threshold of 200% of the federal poverty level for qualification, the same definition used here to identify households with lower income.
* 4% of non-lower income households in the region reported participating in SNAP, which is likely a data anomaly due to how the Census Bureau asks about household income.
* Households with limited English proficiency are over twice as likely to participate in SNAP.
* People of color are almost twice as likely to participate in SNAP.


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "SNAP Participation Trend by Community"
subtitle = "Percent of households participating in SNAP"
source = "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# king 2012: 0.17-0.07=0.10
# king 2022: 0.12-0.06=0.06
# kitsap 2012: 0.19-0.10=0.09
# kitsap 2022:0.14-0.08=0.06
# pierce 2012: 0.19-0.10=0.09
# pierce 2022:0.17-0.10=0.07
# snohomish 2012: 0.16-0.10=0.06
# snohomish 2022: 0.11-0.08=0.03

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="LEP_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# king 2022-2012: 0.22-0.24=-0.02
# kitsap 2022-2012: 0.13-0.18=-0.05
# pierce 2022-2012: 0.23-0.21=0.02
# snohomish 2022-2012: 0.18-0.26=-0.08

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Disability_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# King 2022-2012: 0.21-0.22=-0.01
# Kitsap 2022-2012: 0.20-0.20=0
# Pierce 2022-2012: 0.22-0.22=0
# Snohomish: 0.19-0.23=-0.04

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# King 2012 poc-non poc: 0.17-0.07=0.10
# King 2022: 0.12-0.06=0.06
# King: ((0.06-0.10)/(0.10))*100: -40% change
# Kitsap 2012: 0.19-0.10=0.09
# Kitsap 2022: 0.14-0.08=0.06
# Kitsap: ((0.06-0.09)/(0.09))*100: -33% change
# Pierce 2012:0.19-0.10=0.09
# Pierce 2022: 0.17-0.10=0.07
# Pierce: ((0.07-0.09)/(0.09))*100: -22% change
# Snohomish 2012: 0.16-0.10=0.06
# Snohomish 2022: 0.11-0.08=0.03 
# Snohomish: ((0.03-0.06)/(0.06))*100: -50% change


filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="Income_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# 2012: 0.35
# 2022: 0.32


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year=="2016" |
           data_year=="2021")


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="Income_cat",
         vulnerability=="low vulnerability",
         data_year=="2016" |
           data_year=="2021")

```
\

#### Insights & Analysis

* Regionally, people of color have consistently been almost twice as likely to participate in SNAP compared to the white non-Hispanic population; however, the disparity between these two groups decreased in all counties - Snohomish (-50%), King (-40%), Kitsap (-33%), Pierce (-22%).
* SNAP participation decreased in all counties for households with limited English proficiency between 2012 and 2022, except for Pierce County where participation for the same group increased from 21% in 2012 to 23% in 2022.
* SNAP participation among people with disabilities in the region either stayed the same (Kitsap and Pierce counties) or decreased (King and Snohomish counties) from 2012 to 2022.
* The share of participating lower income households reached 36% in 2017, the highest for any group in the past decade; however, between 2012 and 2022, there was a three percentage point decrease in SNAP participation for households with lower income across the region


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
Once the data and planning reviews are complete, this step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>