---
title: "Air Quality: PM2.5 Pollution"
subtitle: "Visuals for Equity Tracker (tract data)"
author: "Grant Gibson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'b-environment',
                   ind_dir = 'b03-air-quality-pm25',
                   map = 'b03-air-quality-pm25-map-data',
                   chart = 'b03-air-quality-pm25-chart-data',
                   map_name = 'b03-air-quality-pm25-map.html',
                   column_chart_name = 'b03-air-quality-pm25-column',
                   line_chart_name = 'b03-air-quality-pm25-line')

# set most recent year
year <- "2024"
years <- c(2024)

# set past version year - this will set organize archived files in Y drive
# past_year <- "2022"
```

# 3 visuals for webpage

This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.\
\
**If the indicator is available through a tract-level data set.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-tract-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.

## Indicator Explanation

Fine particles (PM2.5) are tiny, microscopic pieces of pollution that can easily enter your bloodstream and cause breathing and heart problems. Fine particle pollution is especially dangerous for children, the elderly and people with sensitive immune systems. In our region, main sources of fine particle pollution come from wood stoves and fireplaces, backyard burn piles, wildfires, and transportation.

The data shown below represent annual average measurements of PM2.5 particles by volume. Daily measurements from Purple Air sensors, excluding July 3-5 and days with wildfire smoke, are averaged and expanded to cover whole Census tracts. The units of measure are notated as micrograms per cubic meter ($\mu\text{g}/\text{m}^3$). For more information, see the Puget Sound Clean Air Agency's Air Quality Data webpage [https://pscleanair.gov/154/Air-Quality-Data].

## 1. Map of most recent data

### Create Visual

```{r map .rda}

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    #"update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
```

To map data in this form, there should be a value corresponding to each census tract. Depending on the year or source of the data, this could be about 700 rows for data at the 2010 census tract resolution, or about 900 rows for data at the 2020 census tract resolution.

```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# check data set
str(data_tract)
```

```{r tract data map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(data_tract$variable_value) # max is 6.54, so rounding to 7 makes sense
# PSCAA's guidance has max value at 9, then decreases by 1 for each bin
max_value <- 9 # this value will be the max value in the legend
min_value <- 3 # this value will be the min value in the legend

# set up palettes
psrc_purple_plus<-c("#F6CEFC", psrc_colors$purples_inc)
# tract_data_plus_zero<-append(data_tract$variable_value,0)
# 

# psrc_palette <- leaflet::colorNumeric(palette=psrc_purple_plus,
#                                       domain = data_tract$variable_value) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min_value, max_value, NA)) # this is the value field

# set the variable
var_name <- "Fine Particle <br> Pollution (PM2.5)<br style='line-height:0.7;'><span style='font-size: 0.7em;'>(micrograms/cubic meter)</span>" #if needed, includes subtitle with units

labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>", data_tract$geoid20, # the code in <> makes the geoid value italic 
  paste("Tract value: ", prettyNum((round(data_tract$variable_value, digits=1))), "ug/m<sup>3</sup>"),
  paste("Region value: ", prettyNum((round(data_tract$reg_estimate, digits=1))), "ug/m<sup>3</sup>"),
  paste("County value: ", prettyNum((round(data_tract$cnty_estimate, digits=1))), "ug/m<sup>3</sup>",
        paste0("(",data_tract$county_name," County)"))) %>% 
  lapply(htmltools::HTML)

# if the label required a prefix, for example a $, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20, # the code in <> makes the geoid value italic
#   paste("$",
#         prettyNum((round(data_tract$variable_value, digits=-2)),
#                   big.mark = ","))) %>%
#   lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20,
#   paste(prettyNum((round(data_tract$variable_value, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$variable_value), #tract_data_plus_zero
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              # label = round(data_tract$variable_value, digits=1),
              label = labels) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min_value, max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       bins = 6,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

Source(s): Puget Sound Clean Air Agency, 2024, Particulate Matter Pollution; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles

```{r generate tract map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(tract_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       #'update/webpage-html-outputs',
                                       'webpage-html-outputs',
                                       file_names$map_name))
```

\
\

#### Data call outs

```{r map calculations - tract, include=FALSE, eval=FALSE}
# simplify data set
data <- data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(variable_value))
# 3 highest: 53053073404 (Puyallup) Pierce (6.539873), 53053071307 (Uninc.) Pierce (6.519665), 53053071306 (Uninc.) Pierce (6.488128)

sort <- data %>% 
  dplyr::arrange(variable_value)
# 3 lowest: 53033032003 (Uninc.) King (3.479040), 53033032002 (Uninc.) King (3.553932), 53033031501 (Uninc.) King (3.617258)

# difference between highest and lowest

```

1.  4.9 $\mu\text{g}/\text{m}^3$: The average amount of fine particle pollution in the region in 2024
2.  9.0 $\mu\text{g}/\text{m}^3$: The upper limit of annual average amount of fine particle pollution considered to be "good" air quality, below which every tract in the region falls
3.  4.3 $\mu\text{g}/\text{m}^3$: The lowest countywide average (Kitsap County) in the region

\

#### Insights & Analysis

-   The tract with the lowest average concentration of fine particle pollution (3.5 $\mu\text{g}/\text{m}^3$) is in unincorporated King County. The tract with the highest average concentration of fine particle pollution (6.5 $\mu\text{g}/\text{m}^3$) is in Puyallup, Pierce County.
-   Higher annual averages of fine partilce pollution (6.0 $\mu\text{g}/\text{m}^3$ or greater) are found in communities around Kent, Federal Way, and Tacoma.

\
\

## 2. Facet of most recent data

### Create Visual

```{=html}
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>
```

```{r load rda for chart, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    #"update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - tract}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (y_min, y_max, dec, esttype, title, subtitle, source)
df = data_clean_column
geo = "county"
x = "quintile_ord"
y = "variable_value"
facet = "equity_group_ord"
fill = "quintile_ord"
y_min = 0
# y_max = 9 # will be based on indicator data set
y_max = 6
dec = 1
esttype = "number"
color = "oranges"
num_colors = 5
color_rev = FALSE
title = "Fine Particle Pollution (PM2.5)"
subtitle = "micrograms/cubic meter"
source = "Puget Sound Clean Air Agency, 2024; U.S. Census Bureau, 2019-2023 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\d01-life-expectancy\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         #'update/webpage-html-outputs'
                                         'webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  #'update/webpage-html-outputs',
                                  'webpage-html-outputs',
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">

`r title`

</p>

<p style="font-family: Poppins; font-size:12pt; font-weight:normal">

`r subtitle`

</p>

```{r column chart output - tract, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```

<p style="font-family: Poppins; font-size:10pt; font-style:italic">

`r source`

</p>

#### Data call outs

```{r column facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information

# data call out (#2) from above 
filter <- data_clean_column %>% 
  filter(county=="Region",
         equity_group=="poc_quintile")



filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High",
         county!="Region")



filter <- data_clean_column %>% 
  filter(equity_group=="lep_quintile",
         quintile=="Low"|
           quintile=="High",
         county!="Region")




filter <- data_clean_column %>% 
  filter(equity_group=="disability_quintile",
         quintile=="Low"|
           quintile=="High")


filter <- data_clean_column %>% 
  filter(county!="Region",
         equity_group=="youth_quintile",
         quintile=="Low"|
           quintile=="High")



filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High")


```

\

#### Insights & Analysis

-   In the region, communities with high concentrations of people of color and households with limited English proficiency have higher average amounts of fine particle pollution than those with low concentrations of the same.
-   King County communities with high concentrations of lower income households have greater average yearly amounts of fine particle pollution than communities with lower concentrations of the same.
-   Pierce County communities with high concentrations of people of color, households with limited English proficiency, and households with lower income have greater average yearly amounts of fine particle pollution than communities with lower concentrations of the same groups.

\
\

# Transfer files

## Copy files from Github \> Y drive/update folder

This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. *Always run this chunk of code. If this is the first time you are generating visuals for an indicator, comment out the 'update' part of the 'y.drive.folder.update' variable because this is the first version that is being created.*

```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir#,
                                   #'update' #comment out if this is the first time you are generating visuals
                                   )

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder \> Y drive/indicator/archive folder

This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*

```{r to archive, include=FALSE, eval=FALSE}
# add previous year folder to archive folder - to retain and organize all archived files
dir.create(file.path(file_names$base_dir,
                     file_names$theme_dir,
                     file_names$ind_dir,'archive', past_year))

# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive', past_year)

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder

This step will clean the indicator folder to make room for the new versions. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*

```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder \> Y drive/indicator folder

This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall). *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*

```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder

This step will help keep the folders organized and ready for the next update. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*

```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder \> website folder

This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage. *Always run this chunk of code.*

```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>
