---
title: "Life Expectancy"
subtite: "Data Gen: Exploring, Cleaning, Transforming (tract data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Life Expectancy at Birth**

# Download data 
## Primary data 
Data comes from [WTN](https://fortress.wa.gov/doh/wtn/WTNPortal#!q0=655). Separate CSVs at the census tract geography are available (manual download) for 5y time spans at 1y increments, starting in 2000, until 2020.
```{r, include=FALSE}
folder_path <- "Y:/Equity Indicators/tracker-webpage-content/a-regional-health-collaboration/a01-life-expectancy/raw-data"
life_expec_01_05 <- read.csv(file.path(folder_path,"Life_Expectancy_at_Birth_01_05.csv"))
life_expec_06_10 <- read.csv(file.path(folder_path,"Life_Expectancy_at_Birth_06_10.csv"))
life_expec_11_15 <- read.csv(file.path(folder_path,"Life_Expectancy_at_Birth_11_15.csv"))
life_expec_16_20 <- read.csv(file.path(folder_path,"Life_Expectancy_at_Birth_16_20.csv"))
```
\
Although there are 4 possible data sets (*2001-2005, 2006-2010, 2011-2015, 2016-2020*), we will be using the three more recent data sets to keep it to 2010 and onward. 


## Additional data

### equity quintile data and 2010-20 crosswalk file 

* low, low medium, etc.) at the tract-level for each equity focus group for the years that correspond to the life expectancy data (2010, 2015, 2020
* for the 2016-2020 life expectancy data set
```{r, include=FALSE}
# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares")
table(equity_tracts$data_year)

crosswalk_10_20 <- get_table(schema="census",
                             tbl_name="v_geo_relationships_tracts")
```

### tract-level population data
to weight life expectancy for 2010, 2015, 2020
```{r, include=FALSE}
# getting census population data by tract
tracts_10_15_20 <- get_acs_recs(geography ='tract', 
                                table.names = 'B01003',
                                years = c(2010, 2015, 2020),
                                acs.type = 'acs5')
tract.10 <- tracts_10_15_20 %>% 
  filter(year==2010)
tract.15 <- tracts_10_15_20 %>% 
  filter(year==2015)
tract.20 <- tracts_10_15_20 %>% 
  filter(year==2020)
```

### tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
tracts10.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)
tracts10.lyr<-st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

# Explore data
The 2010 and 2015 equity tracts are in 2010 geographies and include ~770 census tracts, each of which are assigned one of the 5 quintile categories for each of the 6 equity demographic groups. The 2010, 2015, an 2020 life expectancy data are available from WTN in 2010 geographies, so the 2020-year data will need to be translated to 2020 geographies to be able to use the 2020 population information.
```{r, include=FALSE}
# check number of rows in data sets
nrow(life_expec_01_05) #1465
nrow(life_expec_06_10) #1465
nrow(life_expec_11_15) #1465
nrow(life_expec_16_20) #1465
```
The four possible WTN data sets that could be used without overlapping years include 1,465 records each. This closely matches the number of records in the 2010 census tract shapefile for WA state, published by the [US Census Bureau, Geography Division](https://www.census.gov/cgi-bin/geo/shapefiles/index.php) (1,458; in comparison to the 1,784 tracts in the 2020 shapefile).

```{r, include=FALSE}
# to check the field names and data class
str(life_expec_16_20)

# add column with the last year of the 5-year span
# filter to PSRC counties
life_expec_01_05 <- life_expec_01_05 %>% 
  mutate(year=2005) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
life_expec_06_10 <- life_expec_06_10 %>% 
  mutate(year=2010) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
life_expec_11_15 <- life_expec_11_15 %>% 
  mutate(year=2015) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
life_expec_16_20 <- life_expec_16_20 %>% 
  mutate(year=2020) %>% 
  filter(County.Name=="King" |
           County.Name=="Kitsap" |
           County.Name=="Pierce" |
           County.Name=="Snohomish")
```

## Most Recent Data
### Map
#### Join recent data to spatial file
```{r, include=FALSE}
# join base 2010 geographic data to WTN health data - staying in 2010 geographies to not display false sense of accuracy
data_tract <- merge(tracts10.lyr, life_expec_16_20,
                    by.x="geoid10",
                    by.y="Census.Tract", 
                    all.x=TRUE)

data_tract <- data_tract %>% 
  dplyr::mutate(estimate=as.numeric(Life.Expectancy)) # rename value variable for consistency
```

```{r}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - to get legend high-low with correct color order
addLegend_decreasing <-
  function (map,
            position = c("topright", "bottomright", "bottomleft",
                         "topleft"),
            pal,
            values,
            na.label = "NA",
            bins = 7,
            colors,
            opacity = 0.5,
            labels = NULL,
            labFormat = labelFormat(),
            title = NULL,
            className = "info legend",
            layerId = NULL,
            group = NULL,
            data = getMapData(map),
            decreasing = FALSE) {
    position <- match.arg(position)
    type <- "unknown"
    na.color <- NULL
    extra <- NULL
    if (!missing(pal)) {
      if (!missing(colors))
        stop("You must provide either 'pal' or 'colors' (not both)")
      if (missing(title) && inherits(values, "formula"))
        title <- deparse(values[[2]])
      values <- evalFormula(values, data)
      type <- attr(pal, "colorType", exact = TRUE)
      args <- attr(pal, "colorArgs", exact = TRUE)
      na.color <- args$na.color
      if (!is.null(na.color) &&
          col2rgb(na.color, alpha = TRUE)[[4]] ==
          0) {
        na.color <- NULL
      }
      if (type != "numeric" && !missing(bins))
        warning("'bins' is ignored because the palette type is not numeric")
      if (type == "numeric") {
        cuts <- if (length(bins) == 1)
          pretty(values, bins)
        else
          bins
        
        if (length(bins) > 2)
          if (!all(abs(diff(bins, differences = 2)) <=
                   sqrt(.Machine$double.eps)))
            stop("The vector of breaks 'bins' must be equally spaced")
        n <- length(cuts)
        r <- range(values, na.rm = TRUE)
        cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
        n <- length(cuts)
        p <- (cuts - r[1]) / (r[2] - r[1])
        extra <- list(p_1 = p[1], p_n = p[n])
        p <- c("", paste0(100 * p, "%"), "")
        if (decreasing == TRUE) {
          colors <- pal(rev(c(r[1], cuts, r[2])))
          labels <- rev(labFormat(type = "numeric", cuts))
        } else{
          colors <- pal(c(r[1], cuts, r[2]))
          labels <- rev(labFormat(type = "numeric", cuts))
        }
        colors <- paste(colors, p, sep = " ", collapse = ", ")
        
      }
      else if (type == "bin") {
        cuts <- args$bins
        n <- length(cuts)
        mids <- (cuts[-1] + cuts[-n]) / 2
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "bin", cuts))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "bin", cuts)
        }
        
      }
      else if (type == "quantile") {
        p <- args$probs
        n <- length(p)
        cuts <- quantile(values, probs = p, na.rm = TRUE)
        mids <- quantile(values, probs = (p[-1] + p[-n]) / 2,
                         na.rm = TRUE)
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "quantile", cuts, p))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "quantile", cuts, p)
        }
      }
      else if (type == "factor") {
        v <- sort(unique(na.omit(values)))
        colors <- pal(v)
        labels <- labFormat(type = "factor", v)
        if (decreasing == TRUE) {
          colors <- pal(rev(v))
          labels <- rev(labFormat(type = "factor", v))
        } else{
          colors <- pal(v)
          labels <- labFormat(type = "factor", v)
        }
      }
      else
        stop("Palette function not supported")
      if (!any(is.na(values)))
        na.color <- NULL
    }
    else {
      if (length(colors) != length(labels))
        stop("'colors' and 'labels' must be of the same length")
    }
    legend <-
      list(
        colors = I(unname(colors)),
        labels = I(unname(labels)),
        na_color = na.color,
        na_label = na.label,
        opacity = opacity,
        position = position,
        type = type,
        title = title,
        extra = extra,
        layerId = layerId,
        className = className,
        group = group
      )
    invokeMethod(map, data, "addLegend", legend)
  }

```

```{r tract data map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Life Expectancy"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Missing Data
Life expectancy calculations can fluctuate considerably in smaller populations or populations experiencing low or no deaths for the year(s) being calculated. Because of these issues the Life Expectancy at birth calculation for the Census Tract geographies is suppressed for Census Tracts with a population (for the 5 years combined) of <5000 or a result with a Standard Error >2 or a record of <50 deaths for the time period. The Washington Department of Health, Center for Health Statistics estimates data gathered from death certificates to be 99% complete.

#### Isolate NA or missing data
```{r}
# join NA data to the shapefile to map
data_tract_na <- data_tract %>% 
  filter(is.na(estimate))

nrow(data_tract_na) #27 tracts
```

```{r missing tract map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Life Expectancy"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = var_name,
                   overlayGroups = "missing data",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%
  addPolygons(data=data_tract_na,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "missing data") %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Descriptive Statistics
The following descriptive statistics reflect the census tracts that have values - the `r na_rows_num` census tracts with NAs have been removed. These are not weighted by population numbers. 
```{r}
data_tract_nona <- data_tract %>% 
  filter(!is.na(estimate))

summary(data_tract_nona$estimate)
```

**When weighted by tract populations.** 
```{r, include=FALSE}
data_tract_nona_weight <- data_tract_nona %>% 
  mutate(tot_age = estimate*total_pop10,
         reg_tot = sum(tot_age),
         tot_pop = sum(total_pop10),
         est_weight_avg = reg_tot/tot_pop)

distrib <- rep(data_tract_nona$estimate, data_tract_nona$total_pop10)

avg.value <- data_tract_nona_weight$est_weight_avg[1] #80.78997
med.value <- median(distrib) #80.74
```
*Average:* `r avg.value`
\
*Median:* `r med.value`
\
\

#### Histogram
A histogram is a visual representation of the distribution of a dataset...The y-axis shows how frequently the values on the x-axis occur in the data, while the bars group ranges of values or continuous categories on the x-axis [(source)](https://www.datacamp.com/tutorial/make-histogram-basic-r).  
```{r}
hist(distrib)
```
\

#### Boxplot
A boxplot helps to visualize a quantitative variable by displaying five common location summary (minimum, median, first and third quartiles and maximum) and any observation that was classified as a suspected outlier using the interquartile range (IQR) criterion [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
boxplot(distrib, horizontal = TRUE)
```
\

#### Outliers
The IQR criterion means that all observations above or below the first and third quartile respectively, and IQR is the difference between the third and first quartile) are considered as potential outliers by R [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
outliers <- boxplot.stats(distrib)$out
outlier_rle <- rle(outliers)
outlier_df <- data.frame(unclass(outlier_rle)) %>% 
  rename(population = lengths,
         health_in = values) %>% 
  arrange(health_in)

outlier_df
```
\

##### *Check outliers*
The following steps are specific to the life expectancy dataset and may or may not be applicable depending on the data.

###### Identify census tracts with outliers
```{r include=FALSE}
outlier_values <- outlier_df[,2]

data_tract_nona_weight_outliers <- data_tract_nona_weight %>% 
  dplyr::filter(estimate %in% outlier_values)

num_outliers <- nrow(data_tract_nona_weight_outliers) #7
```
There are `r num_outliers` tracts that have outliers. 
\
\

###### Outliers by Displacement Risk score
This check uses the displacement risk data, which is mapped to 2010 census tracts. This will help determine if there is a connection between displacement risk and `r indicator_measure`.
```{r displacement risk data, inlcude=FALSE}
# Connecting to Portal for displacement tract geographies ----
displacement.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Displacement_Risk_Data/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

displacement.lyr <-st_read(displacement.url)
# head(displacement.lyr)
# plot(displacement.lyr$geometry)
# class(displacement.lyr$geoid10)

displacement_simp <- displacement.lyr %>%
  dplyr::select(geoid10,
                risk_score) %>% 
  as_tibble() %>% 
  dplyr::select(-geometry)
```

```{r, include=FALSE}
data_tract_nona_weight_outliers %<>% 
  dplyr::mutate(geoid=trimws(geoid10))

# join outlier tract datas to spatial dispalcement tract data to isolate tracts with outlier data
data_tract_risk_outliers <- merge(x=displacement_simp, 
                                      y=data_tract_nona_weight_outliers,
                                      by.x="geoid10",
                                      by.y="geoid",
                                      all.y=TRUE)
head(data_tract_risk_outliers)
# class(data_tract_risk_outliers)
# plot(data_tract_risk_outliers$geometry)
nrow(data_tract_risk_outliers) #7
pop_outliers <- as.numeric(sum(data_tract_risk_outliers$total_pop10, na.rm = T)) #33152
pop_outliers_num <- label_comma()(pop_outliers)
```
There are `r pop_outliers_num` people (2020 Decennial Census) in these outlier tracts.
\
\
```{r}
# compare the two measures
plot(data_tract_risk_outliers$risk_score, 
     data_tract_risk_outliers$estimate)
```

```{r}
# this will be different for each indicator - the code and the text included below...
# based on the plot in the last chunk, this code is meant to isolate outlier(s) - in this case it filters for the outliers with high risk scores and high life expectancy

# outlier #1
data_tract20_outliers_1 <- data_tract_risk_outliers %>% 
  filter(risk_score > 50)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_1

# outlier #2
data_tract20_outliers_2 <- data_tract_risk_outliers %>% 
  filter(risk_score < 17)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_2

# To locate the census tracts on the map, its easy to use the Data Portal (https://psrc-psregcncl.hub.arcgis.com/datasets/census-tracts-2020/explore?location=47.506129%2C-121.980700%2C9.23) - just filter on the geoid20 field. 
```


# **MAP .rda**
Save final data set (.rda) for map
```{r, include=FALSE}
# This is the data set used to create the map of the most recent data. For this data set, there is one additional transformation that is needed before saving it for final visualization in the 'vis' script. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk. 

# Because the measure (life expectancy) does not account for the population of the tract, it hasn't been normalized and needs to be weighted in order to aggregate to the region level. The map is going to visualize 2020 data, but in 2010 geographies. The original WTN data set was merged to the 2010 spatial file which has 2010 population numbers. To get 2020 population numbers, we need to aggregate population numbers in the 2020 tracts that were split from 2010 tracts. 

 
head(tract.20) # to determine field names
tract.20.edit <- tract.20 %>% 
  dplyr::mutate(pop20=estimate) # the life expectancy value is already 'estimate'

# merge 2020 acs population data to crosswalk file
pop20_crosswalk <- merge(tract.20.edit, crosswalk_10_20, 
                         by.x="GEOID",
                         by.y="geoid20", 
                         all.x=TRUE)

# calculate population by adding 2020 tracts that were split from 2010 tracts
pop20 <- pop20_crosswalk %>% 
  dplyr::group_by(geoid10) %>% 
  dplyr::summarise(total_pop20=sum(pop20))

# merge 2020 population by 2010 tracts to life expectancy data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_tract <- merge(data_tract, pop20, 
                         by.x="geoid10",
                         by.y="geoid10", 
                         all.x=TRUE)
```

```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'a-regional-health-collaboration'
ind_dir <- 'a01-life-expectancy'
file_name <- 'a01-life-expectancy-map-data'

# save final data set as .rda
save(data_tract, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

# Explore data by equity quintile
## Join primary data to equity quintiles and acs population
```{r}
# 2010 data -----
equity_tracts_10 <- equity_tracts %>% 
  dplyr::filter(data_year==2010)

# join WTN data to equity quintiles (on 2010 census tracts)
data_equity_tracts <- merge(life_expec_06_10, equity_tracts_10,
                            by.x="Census.Tract",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2010 acs population data to WTN data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract10 <- merge(tract.10, data_equity_tracts,
                             by.x="GEOID",
                             by.y="Census.Tract", 
                             all.x=TRUE)

# check data set
nrow(data_equity_tract10) #1548
```
This data set should include about twice the number of census tracts in the region - census tracts are listed twice - once as part of the region and another as part of their corresponding county.

```{r}
# 2015 data ----- 
equity_tracts_15 <- equity_tracts %>% 
  dplyr::filter(data_year==2015)

# join WTN data to equity quintiles (on 2010 census tracts)
data_equity_tracts <- merge(life_expec_11_15, equity_tracts_15,
                            by.x="Census.Tract",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2015 acs population data to WTN data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract15 <- merge(tract.15, data_equity_tracts,
                             by.x="GEOID",
                             by.y="Census.Tract", 
                             all.x=TRUE)

# check data set
nrow(data_equity_tract15)
```
This data set should include about twice the number of census tracts in the region - census tracts are listed twice - once as part of the region and another as part of their corresponding county.

```{r}
# 2020 data -----
equity_tracts_20 <- equity_tracts %>% 
  dplyr::filter(data_year==2020)

# join WTN data (in 2010 census tracts) to cross walk
data_crosswalk <- merge(crosswalk_10_20, life_expec_16_20,
                        by.x="geoid10",
                        by.y="Census.Tract",
                        all.x=TRUE)

# join WTN data to equity quintiles (on 2020 census tracts)
data_equity_tracts <- merge(data_crosswalk, equity_tracts_20,
                            by.x="geoid20",
                            by.y="geoid", 
                            all.x=TRUE)

# join 2020 acs population data to WTN data - this is for eventual weighting (which may not be necessary depending on the indicator)
data_equity_tract20 <- merge(tract.20, data_equity_tracts,
                             by.x="GEOID",
                             by.y="geoid20", 
                             all.x=TRUE)

# check data set
nrow(data_equity_tract20)
```
This data set should include about twice the number of census tracts in the region - census tracts are listed twice - once as part of the region and another as part of their corresponding county.

## Transform/pivot
The data sets need to be cleaned and pivoted to a longer table so that there is only one field for the 6 different equity categories, instead of one field for each.
```{r, include=FALSE}
# clean/simplify data sets
data_equity_quintiles10 <- data_equity_tract10 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(Life.Expectancy_num=as.numeric(Life.Expectancy))

data_equity_quintiles15 <- data_equity_tract15 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(Life.Expectancy_num=as.numeric(Life.Expectancy))

data_equity_quintiles20 <- data_equity_tract20 %>% 
  dplyr::mutate(data_year=format(year.x,format="%Y")) %>% 
  dplyr::mutate(Life.Expectancy_num=as.numeric(Life.Expectancy))

# pivot data set so that there is one column with quintile designation
data_equity_pivot10 <- data_equity_quintiles10 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot15 <- data_equity_quintiles15 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

data_equity_pivot20 <- data_equity_quintiles20 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")
```

```{r}
# check data sets
nrow(data_equity_pivot10)
nrow(data_equity_pivot15)
nrow(data_equity_pivot20)
```

These data sets should include records for:

* 2 geographies (region and whichever county the tract belongs to) x # census tracts (~773 for 2010/15, ~919 for 2020)
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015 - but still show up as NA)


## Calculations
### Calculate by region
The life expectancy values are available at the census tract level, but we want to calculate the regional average, using the census bureau's tract-level population estimates (2010, 2015, 2020).
```{r}
data_equity_pivot10_calc <- data_equity_pivot10 %>% 
  filter(!is.na(quintile)) %>% #missing lep and disability data
  dplyr::group_by(equity_group, quintile) %>%
  summarise(wt_life_expec=(sum(Life.Expectancy_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>% 
  mutate(County.Name="Region") %>% 
  mutate(data_year=2010)

data_equity_pivot15_calc <- data_equity_pivot15 %>% 
  filter(!is.na(quintile)) %>% #missing lep data
  dplyr::group_by(equity_group, quintile) %>%
  summarise(wt_life_expec=(sum(Life.Expectancy_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>% 
  mutate(County.Name="Region") %>% 
  mutate(data_year=2015)

data_equity_pivot20_calc <- data_equity_pivot20 %>% 
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile) %>%
  summarise(wt_life_expec=(sum(Life.Expectancy_num*estimate, na.rm =TRUE))/(sum(estimate, na.rm=TRUE))) %>% 
  mutate(County.Name="Region") %>% 
  mutate(data_year=2020)
```

```{r}
# check data sets
nrow(data_equity_pivot10_calc)
nrow(data_equity_pivot15_calc)
nrow(data_equity_pivot20_calc)
```
These data sets should include records for:

* 1 geography: region
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015)
     * 2010: only 4 equity focus groups
     * 2015: only 5 equity focus groups
* 5 equity quintiles 


### Calculate by equity quintile
The life expectancy values are available at the census tract level but we want to calculate the weighted values by the 6 equity/5 quintile groups, using the census bureau's tract-level population estimates (2010, 2015, 2020). 
```{r}
# calculate 2010 region population
population_test <- data_equity_pivot10 %>% 
  distinct(GEOID, .keep_all = TRUE) %>%
  dplyr::summarise(pop=formatC((sum(estimate, na.rm=TRUE)), format="d", big.mark=",")) #3,603,425

sum(tract.10$estimate) #3,603,425

# calculations to get weighted average life expectancy
data_equity_wt10 <- data_equity_pivot10 %>% 
  filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
  dplyr::mutate(tot_risk_tract=estimate*Life.Expectancy_num) %>% 
  dplyr::group_by(County.Name, data_year, equity_group, quintile) %>%
  dplyr::summarise(tracts=length(GEOID),
                   tot_age=(sum(tot_risk_tract, na.rm = TRUE)),
                   tot_pop=(sum(estimate, na.rm = TRUE))) %>% 
  dplyr::mutate(wt_life_expec=tot_age/tot_pop)

# # check population
# data_equity_wt10_test <- data_equity_wt10 %>% 
#   dplyr::filter(equity_group=="income_quintile")
# sum(data_equity_wt10_test$tot_pop) #3,603,425

# check data set
nrow(data_equity_wt10)
```
The regional population in 2010: `r population_test`
\
\
This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 4 equity focus groups (disability and lep are not available in for 2010)
* 5 equity quintiles 

```{r}
# calculate 2015 region population
population_test <- data_equity_pivot15 %>% 
  distinct(GEOID, .keep_all = TRUE) %>%
  dplyr::summarise(pop=formatC((sum(estimate, na.rm=TRUE)), format="d", big.mark=",")) #3,869,802

# calculations to get weighted average life expectancy
data_equity_wt15 <- data_equity_pivot15 %>% 
  filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
  dplyr::mutate(tot_risk_tract=estimate*Life.Expectancy_num) %>% 
  dplyr::group_by(County.Name, data_year, equity_group, quintile) %>%
  dplyr::summarise(tracts=length(GEOID),
                   tot_age=(sum(tot_risk_tract, na.rm = TRUE)),
                   tot_pop=(sum(estimate, na.rm = TRUE))) %>% 
  dplyr::mutate(wt_life_expec=tot_age/tot_pop)

# # check population
# data_equity_wt15_test <- data_equity_wt15 %>%
#   dplyr::filter(equity_group=="income_quintile")
# sum(data_equity_wt15_test$tot_pop) #3,869,802

# check data set
nrow(data_equity_wt15)
```
The regional population in 2015: `r population_test`
\
\
This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 5 equity focus groups (lep not available in 2015)
* 5 equity quintiles 

```{r}
# calculate 2020 region population
population_test <- data_equity_pivot20 %>% 
  distinct(GEOID, .keep_all = TRUE) %>%
  dplyr::summarise(pop=formatC((sum(estimate, na.rm=TRUE)), format="d", big.mark=",")) #4,197,443	

sum(tracts20.lyr$total_pop20) #4,294,373

# calculations to get weighted average life expectancy per equity group/quintile
data_equity_wt20 <- data_equity_pivot20 %>% 
  filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
  dplyr::mutate(tot_risk_tract=estimate*Life.Expectancy_num) %>% 
  dplyr::group_by(County.Name, data_year, equity_group, quintile) %>%
  dplyr::summarise(tracts=length(GEOID),
                   tot_age=(sum(tot_risk_tract, na.rm = TRUE)),
                   tot_pop=(sum(estimate, na.rm = TRUE))) %>% 
  dplyr::mutate(wt_life_expec=tot_age/tot_pop)

# # check population
# data_equity_wt20_test <- data_equity_wt20 %>%
#   dplyr::filter(equity_group=="income_quintile")
# sum(data_equity_wt20_test$tot_pop) #4,197,443

# check data set
nrow(data_equity_wt20)
```
The regional population in 2020: `r population_test`
\
\
This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 6 equity focus groups
* 5 equity quintiles 

## Combine years into one data set
```{r}
data_tract <- as.data.frame(rbind(data_equity_wt10,
                                  data_equity_wt15,
                                  data_equity_wt20)) %>% 
  dplyr::select(-tot_age, -tot_pop, -tracts) %>% 
  rbind(data_equity_pivot10_calc,
        data_equity_pivot15_calc,
        data_equity_pivot20_calc) %>% 
  filter(!is.na(quintile))

# check data set
str(data_tract)
```
This data set should include records for: 

* 5 geographies: 4 PSRC counties + region
* 3 years
* 6 equity focus groups
    * 2010: (subtracting) 2 missing focus groups(disability, lep) * 5 equity quintiles * 5 geographies
    * 2015: (subtracting) 1 missing focus group (lep) * 5 equity quintiles * 5 geographies
* 5 equity quintiles

## Rename and factor
```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

quintile_order <- c("Low", 
                    "Low\nMedium", 
                    "Medium", 
                    "Medium\nHigh", 
                    "High")

equity_group_order <- c("People of Color", 
                        "Households with Lower Income", 
                        "People with a Disability", 
                        "Households with Limited English Proficiency", 
                        "Households with Youth <18", 
                        "Households with Older Adults 65+")

# transforming data labels ----
data_clean <- data_tract %>% 
  mutate(county=County.Name) %>%  #rename value variable for consistency
  mutate(county_ord = factor(county, levels=county_order)) %>%
  mutate(equity_group_ord = case_when(
    equity_group=="poc_quintile"~"People of Color",
    equity_group=="disability_quintile"~"People with a Disability",
    equity_group=="lep_quintile"~"Households with Limited English Proficiency",
    equity_group=="income_quintile"~"Households with Lower Income",
    equity_group=="youth_quintile"~"Households with Youth <18",
    equity_group=="older_quintile"~"Households with Older Adults 65+")) %>%
  mutate(equity_group_ord = factor(equity_group_ord, levels = equity_group_order)) %>% 
  mutate(quintile_ord = str_wrap(quintile, width=7)) %>%
  mutate(quintile_ord = factor(quintile_ord, levels = quintile_order))

# Minor edits to the data
data_clean <- data_clean %>%
  dplyr::mutate(estimate=as.numeric(wt_life_expec)) #rename value variable for consistency

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county_ord, equity_group_ord, quintile_ord, data_year)
```


# **CHART .rda**
Save final data set for charts (.rda)
```{r}
# set folder structure
file_name <- 'a01-life-expectancy-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>