---
title: "1.) Exploring Voter Turnout: 2024"
subtitle: "Applying geography splits to translate precinct-level data to census tract geographies"
author: "Mary Richards"
date: today
date-format: long
format: 
  html:
    toc: true
    toc-location: right-body
    toc-expand: 2
    # toc-depth: 6
    toc-title: "Contents"
execute:
  # echo: false
  warning: false
  error: false
---

This documents the process for exploring voter turnout based on data from the WA Secretary of State. In this exercise, we are using precinct-tract geography splits to calculate voter turnout by tract geographies.  
\
The final output is voter turnout (%) by census tract. Ultimately, this measurement will be used in the Equity Tracker and Displacement Risk Mapping projects.
```{r setup, include=FALSE}
# Read in Libraries
library(magrittr)
library(stringr)
library(data.table)
library(tidycensus)
library(tidyverse)
library(openxlsx)
library(psrcelmer)
library(psrccensus)

library(sf)
library(leaflet)
library(leafem) #home button
library(psrcplot)
library(here)

library(readxl)

# reference supplemental script with supporting settings/functions
# source(here::here('GitHub/equity-tracker/data-visualization/equity-tracker-supplemental-script.R'))
source('C:/Users/mrichards/Documents/GitHub/equity-tracker/data-visualization/equity-tracker-supplemental-script.R')
```

```{r, include=FALSE}
# Helper functions
fetch_turnout <- function(zip_url, data_file){
  temp_zip <- tempfile(fileext = ".zip")
  temp_dir <- tempdir()
  
  # Download the zip file
  download.file(zip_url, temp_zip)
  
  # Extract the zip file
  unzip(temp_zip, exdir = temp_dir)
  
  # Read the Excel/CSV file
  data_file <- file.path(temp_dir, data_file)
  
  if (endsWith(data_file, ".xlsx")) {
    turnout_data <- read.xlsx(data_file)
  } else if (endsWith(data_file, ".csv")) {
    turnout_data <- read.csv(data_file)
  } else {
    print("data is a different file type")
  }
  
  # Clean up temporary files
  unlink(temp_zip)
  unlink(data_file)
  
  return(turnout_data)
}
```

::: panel-tabset
# Data 
## 1.) Elmer geography splits: precinct-tract
These data tables include the proportion of the population based on OFM (total population) between voting precincts (2024, from SOS) and census tracts (2020). There are three variations: 1) distribution based on total population; 2) distribution based on adult population, based on the 2023 5y ACS tract estimates for adult 18+ population; 3) distribution based on adult population, based on the 2020 decennial census block group estimates for adult 18+ population.
```{r}
total_sql24 <- "select *
    from stg.get_any_geography_splits(
        'voter_precinct_2024', --@data_geog_type
        'tract20',              --@planning_geog_type
        2024,                   --@ofm_estimate_year
        2024,                   --@ofm_vintage
        2018)                   --@parcel_year"
adult_sql24 <- "select *
    from general.get_any_geography_splits(
        'voter_precinct_2024', --@data_geog_type
        'tract20',              --@planning_geog_type
        2024,                   --@ofm_estimate_year
        2024,                   --@ofm_vintage
        2018)                   --@parcel_year"

adult_sql24_dec <- "select *
    from stg.get_geography_splits_scaled_decennial(
        'voter_precinct_2024', --@data_geog_type
        'tract20',              --@planning_geog_type
        2024,                   --@ofm_estimate_year
        2024,                   --@ofm_vintage
        2018)                   --@parcel_year"

df24_total <- get_query(sql=total_sql24, db_name = "Elmer")
df24_adult <- get_query(sql=adult_sql24, db_name = "Elmer")
df24_adult_dec <- get_query(sql=adult_sql24_dec, db_name = "Elmer")
```

## 2.) Secretary of State: turnout data
This data set includes the number of registered voters and the number of ballots cast by voting precinct. The voting precinct geographies change between election years. 
```{r}
voting_url <-"https://www.sos.wa.gov/sites/default/files/"
turnout_zip_url24 <- paste0(voting_url, "2025-01/2024Gen_Precinct_Results_GIS-Ready.zip")

psrc_counties <- data.frame("CountyName"=c("King","Kitsap","Pierce","Snohomish"),
                            "FIPS"=c("033","035","053","061"),
                            "CountyCode" = c("KI", "KI", "PI", "SN"))
acs_year <- 2023

# Load voter turnout data using new source and column names
# 2024 available through the SOS website
vt24 <- fetch_turnout(turnout_zip_url24, 
                      '2024Gen_Precinct_Turnout_GIS-Ready.csv')

# Clean turnout data sets - filter to PSRC region, rename columns
vt24_df <- vt24 %>% setDT() %>%                                             
  filter(str_detect(St_Code, "^KI|KP|PI|SN")) %>%               # Filter to PSRC counties
  .[, .(precinctcode = St_Code, 
        reg_voters = G24TREGVOT, ballots_cast = G24TBALCST)]

# 2024 precinct spatial data
# https://www.sos.wa.gov/elections/data-research/reports-data-and-statistics/precinct-shapefiles
folder <- "Y:/Equity Indicators/tracker-webpage-content/h-public-services/h02-voter-participation/raw-data/Statewide_Precincts_2024General"

precinct_24 <- st_read(file.path(folder, "Statewide_Precincts_2024General.shp"))
```

## 3.) ACS 5y (2023) population data
These values will be used as the denominator - dividing the ballots cast by the total population 18+.
```{r}
base_acs_data <- get_acs_recs(geography ='tract',
                              table.names = 'S0101', #subject table code
                              years = c(as.numeric(acs_year)),
                              acs.type = 'acs5')

age_18over <- base_acs_data %>%
  filter(variable=="S0101_C01_026") #Total population!!SELECTED AGE CATEGORIES!!18 years and over

# calculate the proportion of 18+ adults/total population to check if there are some tracts with more/fewer adults
age_18over_total <- base_acs_data %>%
  filter(variable=="S0101_C01_001" | #Total population
           variable=="S0101_C01_026") %>% #Total population!!SELECTED AGE CATEGORIES!!18 years and over
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable,
              values_from = estimate) %>%
  reframe(GEOID,
          pop_18=S0101_C01_026,
          pop_tot=S0101_C01_001,
          pop18_total=S0101_C01_026/S0101_C01_001)

# calculate total number for reference
sum(age_18over$estimate) #3,394,729
```

## 4.) PSRC Region spatial file
This dataset includes the census tracts within the region and provides a way to visually explore the voter turnout values. 
```{r, echo=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
arc_service <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services"

tracts20.url <- file.path(arc_service, "Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")

tracts20.lyr <- st_read(tracts20.url)
```

## 5.) OFM small area estimates (2024) population data
These values will be used as the denominator for the fourth approach - dividing the ballots cast by the total population (OFM - Small Area Demographic Estaimtes, 2024), weighted by the proportion of adults 18+/total population from ACS 5y 2023 estimates.
```{r}
# # OFM Small Area Demographic Estimates (2020-2024): https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/estimates-april-1-population-age-sex-race-and-hispanic-origin#age
# folder_loc <- here::here("data_development/voter_turnout")
# 
# ofm_raw <- read_excel(file.path(folder_loc,
#                                 "ofm_pop_sade_tract_2020_2024.xlsx"),
#                       sheet = "Total")
# 
# # clean data set
# ofm_total <- ofm_raw %>% 
#   filter(`Age Group`== "Total",
#          `Area Name` != "Washington",
#          Year=="2024") %>% 
#   select(`Area ID`, Year, Total)
# 
# # isolate PSRC tracts
# tracts20_psrc <- st_drop_geometry(tracts20.lyr) %>% 
#   select(county_name, geoid20) 
# 
# ofm_psrc <- ofm_total %>% 
#   right_join(tracts20_psrc, by=c(`Area ID`='geoid20')) %>% 
#   rename(geoid20=`Area ID`)

# OFM Small Estimates Program: https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/small-area-estimates-program
# folder_loc <- here::here("data_development/voter_turnout")
folder_loc <- "C:/Users/mrichards/Documents/GitHub/equity-tracker/data_development/voter_turnout"


ofm_raw <- read_excel(file.path(folder_loc,
                                "saep_tract20.xlsx"),
                      sheet = "Total Population",
                      skip=10) #first rows includes metadata

# clean data set
ofm_psrc <- ofm_raw %>% 
  select(`County Name`, `Census Tract Code Complete`, `Estimated Total Population 2024`) %>% 
  rename("County_Name"=`County Name`,
         "GEOID20"=`Census Tract Code Complete`,
         "population_2024"=`Estimated Total Population 2024`) %>% 
  filter(County_Name=='King' |
           County_Name=='Kitsap' |
           County_Name=='Pierce' |
           County_Name=='Snohomish')
```

# Methods
**Voter turnout = ballots cast / eligible population** 
\
The main challenge is that the raw voting data is available at voting precincts, which are unique geographies that do not align with census geographies. This requires translation between the two geographies. 
\
\
There are three approaches we are exploring to calculating the numerator (the number of ballots cast):

A) total population distribution based on OFM
B) weighting the population distribution by 2023 5y adult 18+ population estimates (TRACT)
C) weighting the population distribution by 2020 decennial census adult 18+ population estimates (BLOCK GROUP)
\

For all three approaches, the 2023 5y adult 18+ population estimates are used as the denominator (eligible population). 
\
\

There is one additional approach that will be explored after discussing with data applications team (9/30/2025). The approach to get the numerator mirrors the second approach described above, but calculating the denominator varies:

D) the denominator will be based on OFM small area total population estimates, but weighted by the proportion of adults 18+/total population (calculated from 2023 5y ACS estimates)


*DATA CAVEATS:* 

* For this exercise, it is assumed that the eligible population most closely aligns with the adult 18+ population. Although this may include individuals who are not able to legally vote (non-citizens), the alternative is using the number of registered voters (available at voting precincts), which is a subset of the adult 18+ population who are already engaged enough to have registered. While using the general 18+ population as a denominator may result in slightly lower voter turnout values (based on our working definition), using registered voters as the denominator would results in inflated voter turnout values.
* Some voters may be registered in one location, but reside in another. This could cause mismatches in the way that they are accounted for in the numerator (ballots cast) and the denominator (population). 
* There is a distinction between residential addresses (which must be in Washington State and where you are considered registered for voting purposes) and mailing addresses (which can be anywhere). Therefore, the number of registered voters refers to the number of voters who have their residential address in that precinct. This is based on correspondence with staff at the Washington Secretary of State office (8/13/2025).
* According to correspondences with staff at the Washington Secretary of State office (10/9/2025), data from smaller precincts are occasionally “masked” into a larger precinct in order to protect the privacy of voters and maintain confidentiality. It is also possible for last minuet ballots to be cast (such as by UOCAVA voters, or Military and Overseas Voters, defined by https://www.eac.gov/uocava) that must be accorded to a precinct. As a result, the voter turnout results calculated at the tract level are estimates and may reflect some margin of error. 


# Reference Maps
## Voting precints (2024) and census tracts (2020)
```{r, echo=FALSE}
# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voting precincts (2024)",
                   overlayGroups = "census tracts (2020)",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=precinct_24,
              color = "gray",
              stroke=TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "voting precincts (2024)",
              label = precinct_24$St_Code) %>%
  addPolygons(data=tracts20.lyr,
              color = "blue",
              stroke = TRUE, 
              weight = 2.5,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "census tracts (2020)",
              label = tracts20.lyr$geoid20) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

## 2024 voter participation (ballots cast/registered voters)
```{r}
# join voting data to precinct spatial layer 
join_voting24 <- precinct_24 %>% 
  right_join(vt24_df,
            by = c('St_Code'='precinctcode'))

# calculate the ballots cast/registered voters for reference numbers
voting_24 <- join_voting24 %>% 
  mutate(vote_participation=(ballots_cast/reg_voters)*100)

summary(voting_24$vote_participation)
```

While the median is quite high, there are some precincts with low voter participation (based on the registered voters).
```{r}
lowest_15 <- voting_24 %>%
  st_drop_geometry() %>% 
  arrange(vote_participation) %>% 
  slice_head(n = 15) %>% 
  select(St_Code, ballots_cast, reg_voters, vote_participation) %>% 
  print
```

This map shows the ballots cast/registered voters. The legend starts at 0% because there are 7 precincts where no ballots were cast. It is important to note that these precincts also have quite low numbers of registered voters
```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = voting_24$vote_participation)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=voting_24,
              fillColor = psrc_palette(voting_24$vote_participation),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(voting_24$vote_participation, digits=0),"%")) %>%
  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = voting_24$vote_participation,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

# A) Total population (OFM)
## Join precinct-tract to geographies to turnout data
```{r}
nrow(df24_total) #precinct-tract records: 6294
nrow(vt24_df) #voting data by unique precinct: 4608
vt24_df_0 <- vt24_df %>% 
  filter(reg_voters<1) # there are 62 precincts without data
# data_geog = precinct
# planning_geog = tract
length(unique(df24_total$data_geog)) #unique precincts: 4552
length(unique(df24_total$planning_geog)) #unique tracts: 919
```
There is a difference in the number of records - in the voting data, there are 4,608 records (precincts) while the geography split data only has 4,552 unique precincts. The difference between these two data sets is 56 precincts. There are 62 precincts in the voting data without registered voters.

### checking the records that don't line up between the two data sets
```{r}
outer <- df24_total %>% 
  anti_join(vt24_df, by = c('data_geog'='precinctcode')) 
nrow(outer) #54 records
summary(outer$percent_of_total_pop)

outer2 <- vt24_df %>% 
  anti_join(df24_total, by = c('precinctcode'='data_geog')) 
nrow(outer2) #57 records
summary(outer2$percent_of_total_pop)
```
The output table from the first anti_join() includes census tract IDs without associated precincts.


## calculate total numbers for reference 
```{r}
sum(vt24_df$reg_voters) #2,736,409
sum(vt24_df$ballots_cast) #2,166,325
```
In the 2024 voting data (by precinct), there were 2,736,409 registered voters and 2,166,325 ballots cast.  

## join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
join_df24 <- df24_total %>% 
  full_join(vt24_df, 
            by = c('data_geog'='precinctcode')) #6351 records


# simplify data - remove unnecessary fields
join_df24_simp <- join_df24 %>% 
  select(-c(percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units))

# review data
summary(join_df24_simp$reg_voters) #54 NAs
summary(join_df24_simp$ballots_cast) #54 NAs

# calculate weighted registered voters and ballots cast 
df24_calc <- join_df24_simp %>% 
  mutate(reg_voters_wt = reg_voters*percent_of_total_pop,
         ballots_cast_wt = ballots_cast*percent_of_total_pop)
```

## calculate total numbers to check 
These calculations sum the values after being weighted by the geography splits. They should match the original voting precinct data. 
```{r}
sum(df24_calc$reg_voters_wt, na.rm = TRUE) #2,736,352 ~ 57 missing voters?
sum(df24_calc$ballots_cast_wt, na.rm = TRUE) #2,166,294 ~ 31 missing ballots cast

# does rounding make a difference?
round(sum(df24_calc$reg_voters_wt, na.rm = TRUE), digits = 0) #2,736,352 ~ still 57 missing voters
round(sum(df24_calc$ballots_cast_wt, na.rm = TRUE), digits =0) #2,166,294 ~ still 31 missing ballots cast
```
After weighting, the total registered voters is 2,736,352 and the total number of ballots cast is 2,166,294. This means that there are 57 missing voters and 31 missing ballots. These go missing during the weighting step.

### checking for missing voters/ballots
```{r}
# grouping and adding by precinct 
test <- df24_calc %>% 
  # mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
  #                                  TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
  #        ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
  #                                    TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog) %>% # grouping on precinct
  summarise(#reg_voters_test = sum(reg_voters_wt), # need to round
            #ballots_cast_test = sum(ballots_cast_wt), # need to round
            reg_voters_test = round(sum(reg_voters_wt), digits = 0),
            ballots_cast_test = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop')

test2 <- df24_calc %>% 
  full_join(test) %>% 
  mutate(dif_voters = if_else(reg_voters == reg_voters_test, "Match", "No Match"),
         dif_ballots = if_else(ballots_cast == ballots_cast, "Match", "No Match"))


# check number of matching/not matching records
table(test2$dif_voters) # all matching once rounded 'reg_voters_wt' above
table(test2$dif_ballots) # all matching once rounded 'ballots_cast_wt' above

test3 <- df24_calc %>% 
  group_by(data_geog) %>% # grouping on precincts
  reframe(total_pop = sum(percent_of_total_pop))
# when viewing the data frame in a separate window, all values are '1' but when looking at df in console there are some '1' and some '1.00'

summary(test3$total_pop) #57 NAs
```
Unable to find missing values?

## calculate voter turnout by registered voters 
(ballots cast/registered voters)
\
This is not the final calculation because it is based on those who are already engaged to an extent, not the general population who are capable of voting.
```{r}
df24_calc_tract <- df24_calc %>% 
  mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
                                   TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
         ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
                                     TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog_type, planning_geog_type, 
           ofm_estimate_year, ofm_vintage, parcel_year,
           planning_geog) %>% # grouping on tract ids, retaining other fields for reference
  summarise(#reg_voters = sum(reg_voters_wt), # need to round?
            #ballots_cast = sum(ballots_cast_wt), # need to round?
            reg_voters = round(sum(reg_voters_wt), digits = 0),
            ballots_cast = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop') %>% 
  mutate(regvote_turnout = ballots_cast/reg_voters)

# calculate total numbers to check
sum(df24_calc_tract$reg_voters, na.rm=T) #2,736,352 when not rounded, 2,736,358 when rounded
sum(df24_calc_tract$ballots_cast, na.rm=T) #2,166,294 when not rounded, 2,166,281 when rounded
```

## calculate voter turnout by population 18+ 
(ballots cast/pop 18+)
\
This is closer to the final calculation because it takes into account all people who are capable of voting. This value may include non-citizens, or adults 18+ who are not legally allowed to vote, but it is more comprehensive than just considering those who are registered.
```{r}
# simplify acs table and join to voter/ballot data
age_18over_simp <- age_18over %>% 
  select(GEOID, 
         adult_18plus = estimate, 
         acs_year = year,
         acs_reliability = reliability)

# join to precinct/voter turnout data to tract 18+ population data to calculate population 18+ turnout
df24_calc_tract_adults <- df24_calc_tract %>% 
  left_join(age_18over_simp, 
            by = c('planning_geog'='GEOID')) 
  

# calculate voter turnout by population 18+
df24_tract_total <- df24_calc_tract_adults %>% 
  mutate(pop18_turnout = ballots_cast/adult_18plus,
         registered_pop18 = reg_voters/adult_18plus)

# check 
summary(df24_tract_total$pop18_turnout) #1.21 max
summary(df24_tract_total$registered_pop18) #1.39 max
```
There are some tracts with higher than 100% voter turnout and registered voters. 

### Checking tracts with over 100%
#### where # of ballots cast is greater than the number of adults 18+
```{r}
df24_tract_above1_totalpop <- df24_tract_total %>% 
  filter(pop18_turnout>1)

df24_tract_above1_totalpop

df24_tract_above1_totalpop_simp <- df24_tract_above1_totalpop %>% 
  select(planning_geog, ballots_cast, adult_18plus, pop18_turnout)
```
There are 11 tracts over 100%.

##### map of tracts >100% voter turnout
```{r}
# join full data to tract spatial file
df24_tract_total_geo <- merge(tracts20.lyr, df24_tract_total,
                              by.x="geoid20",
                              by.y="planning_geog", 
                              all.x=TRUE)

df24_tract_total_geo <- df24_tract_total_geo %>% 
  mutate(pop18_turnout=pop18_turnout*100)

# filter to tracts >100%
df24_tract_total_geo_100plus <- df24_tract_total_geo %>% 
  filter(pop18_turnout>100)
```

```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = df24_tract_total_geo$pop18_turnout)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   overlayGroups = "tracts >100%",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=df24_tract_total_geo,
              fillColor = psrc_palette(df24_tract_total_geo$pop18_turnout),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(df24_tract_total_geo$pop18_turnout, digits=0),"%")) %>%
  addPolygons(data=df24_tract_total_geo_100plus,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "tracts >100%",
              label = paste(round(df24_tract_total_geo_100plus$pop18_turnout, digits=0),"%")) %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = df24_tract_total_geo$pop18_turnout,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

#### where # of registered voters is greater than the number of adults 18+
```{r}
df24_tract_above1_totalpop_reg <- df24_tract_total %>% 
  filter(registered_pop18>1)

df24_tract_above1_totalpop_reg
```
There are 111 tracts over 100%.


We could check against the proportion of adults 18+ out of the total population, in case some tracts have more/fewer children. The geography splits are based on total population, which assumes the distribution of adults 18+ matches the distribution of the total population.
```{r}
# check distribution of 18+ 
summary(age_18over_total$pop18_total)
hist(age_18over_total$pop18_total) 
```
There are some tracts where all of the population is 18+

```{r}
# calculate weighted values for voter turnout based on population 18+ and based on registered voters
df24_tract_calculations <- df24_tract_total %>% 
  left_join(age_18over_total,
            by = c('planning_geog'='GEOID')) %>% 
  mutate(reg_voters_wt18 = reg_voters*pop18_total, #weight values calc from geog splits by eligible pop
         ballots_cast_wt18 = ballots_cast*pop18_total) %>% #weight values calc from geog splits by eligible pop
  mutate(pop18_turnout_wt18 = ballots_cast_wt18/adult_18plus,
         registered_pop18_wt18 = reg_voters_wt18/adult_18plus)

summary(df24_tract_calculations$pop18_turnout_wt18) #max: 0.961
summary(df24_tract_calculations$registered_pop18_wt18) #max: 1.05

# checking tracts where registered voters/population 18+ (weighted) >100%
df24_tract_calculations_above1 <- df24_tract_calculations %>% 
  filter(registered_pop18_wt18>1)
```
When weighted by the proportion 18+/total population, there are no tracts with over 100% voter tunrout based on ballots cast, however there are still 2 tracts that are over 100% when looking at registered voters.


# B) Adult population by tract (ACS 5y-estimates, 2023)
## Join precinct-tract to geographies to turnout data
```{r}
nrow(df24_adult) #precinct-tract records: 6294
nrow(vt24_df) #voting data by unique precinct: 4608
vt24_df_0 <- vt24_df %>% 
  filter(reg_voters<1) # there are 62 precincts without data
# data_geog = precinct
# planning_geog = tract
length(unique(df24_adult$data_geog)) #unique precincts: 4552
length(unique(df24_adult$planning_geog)) #unique tracts: 919
```
There is a difference in the number of records - in the voting data, there are 4,608 records (precincts) while the geography split data only has 4,552 unique precincts. The difference between these two data sets is 56 precincts. There are 62 precincts in the voting data without registered voters.

### checking the records that don't line up between the two data sets
```{r}
outer <- df24_total %>% 
  anti_join(vt24_df, by = c('data_geog'='precinctcode')) 
nrow(outer) #54 records
summary(outer$percent_of_total_pop)

outer2 <- vt24_df %>% 
  anti_join(df24_adult, by = c('precinctcode'='data_geog')) 
nrow(outer2) #57 records
summary(outer2$percent_of_total_pop)
```
The output table from the first anti_join() includes census tract IDs without associated precincts.


## calculate total numbers for reference 
```{r}
sum(vt24_df$reg_voters) #2,736,409
sum(vt24_df$ballots_cast) #2,166,325
```
In the 2024 voting data (by precinct), there were 2,736,409 registered voters and 2,166,325 ballots cast.  

## join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
join_df24 <- df24_adult %>% 
  full_join(vt24_df, 
            by = c('data_geog'='precinctcode')) #6351 records


# simplify data - remove unnecessary fields
join_df24_simp <- join_df24 %>% 
  select(-c(percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units))

# review data
summary(join_df24_simp$reg_voters) #54 NAs
summary(join_df24_simp$ballots_cast) #54 NAs

# calculate weighted registered voters and ballots cast 
df24_calc <- join_df24_simp %>% 
  mutate(reg_voters_wt = reg_voters*percent_of_total_pop,
         ballots_cast_wt = ballots_cast*percent_of_total_pop)
```

## calculate total numbers to check 
These calculations sum the values after being weighted by the geography splits. They should match the original voting precinct data. 
```{r}
sum(df24_calc$reg_voters_wt, na.rm = TRUE) #2,736,352 ~ 57 missing voters?
sum(df24_calc$ballots_cast_wt, na.rm = TRUE) #2,166,294 ~ 31 missing ballots cast

# does rounding make a difference?
round(sum(df24_calc$reg_voters_wt, na.rm = TRUE), digits = 0) #2,736,352 ~ still 57 missing voters
round(sum(df24_calc$ballots_cast_wt, na.rm = TRUE), digits =0) #2,166,294 ~ still 31 missing ballots cast
```
After weighting, the total registered voters is 2,736,352 and the total number of ballots cast is 2,166,294. This means that there are 57 missing voters and 31 missing ballots. These go missing during the weighting step.

### checking for missing voters/ballots
```{r}
# grouping and adding by precinct 
test <- df24_calc %>% 
  # mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
  #                                  TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
  #        ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
  #                                    TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog) %>% # grouping on precinct
  summarise(#reg_voters_test = sum(reg_voters_wt), # need to round
            #ballots_cast_test = sum(ballots_cast_wt), # need to round
            reg_voters_test = round(sum(reg_voters_wt), digits = 0),
            ballots_cast_test = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop')

test2 <- df24_calc %>% 
  full_join(test) %>% 
  mutate(dif_voters = if_else(reg_voters == reg_voters_test, "Match", "No Match"),
         dif_ballots = if_else(ballots_cast == ballots_cast, "Match", "No Match"))


# check number of matching/not matching records
table(test2$dif_voters) # all matching once rounded 'reg_voters_wt' above
table(test2$dif_ballots) # all matching once rounded 'ballots_cast_wt' above

test3 <- df24_calc %>% 
  group_by(data_geog) %>% # grouping on precincts
  reframe(total_pop = sum(percent_of_total_pop))
# when viewing the data frame in a separate window, all values are '1' but when looking at df in console there are some '1' and some '1.00'

summary(test3$total_pop) #57 NAs
```
Unable to find missing values?

## calculate voter turnout by registered voters 
(ballots cast/registered voters)
\
This is not the final calculation because it is based on those who are already engaged to an extent, not the general population who are capable of voting.
```{r}
df24_calc_tract <- df24_calc %>% 
  mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
                                   TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
         ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
                                     TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog_type, planning_geog_type, 
           ofm_estimate_year, ofm_vintage, parcel_year,
           planning_geog) %>% # grouping on tract ids, retaining other fields for reference
  summarise(#reg_voters = sum(reg_voters_wt), # need to round?
            # ballots_cast = sum(ballots_cast_wt), # need to round?
            reg_voters = round(sum(reg_voters_wt), digits = 0),
            ballots_cast = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop') %>% 
  mutate(regvote_turnout = ballots_cast/reg_voters)

# calculate total numbers to check
sum(df24_calc_tract$reg_voters, na.rm=T) #2,736,352 when not rounded, 2,736,354 when rounded
sum(df24_calc_tract$ballots_cast, na.rm=T) #2,166,294 when not rounded, 2,166,297 when rounded
```

## calculate voter turnout by population 18+ 
(ballots cast/pop 18+)
\
This is closer to the final calculation because it takes into account all people who are capable of voting. This value may include non-citizens, or adults 18+ who are not legally allowed to vote, but it is more comprehensive than just considering those who are registered.
```{r}
# simplify acs table and join to voter/ballot data
age_18over_simp <- age_18over %>% 
  select(GEOID, 
         adult_18plus = estimate, 
         acs_year = year,
         acs_reliability = reliability)

# join to precinct/voter turnout data to tract 18+ population data to calculate population 18+ turnout
df24_calc_tract_adults <- df24_calc_tract %>% 
  left_join(age_18over_simp, 
            by = c('planning_geog'='GEOID')) 
  

# calculate voter turnout by population 18+
df24_tract_adult <- df24_calc_tract_adults %>% 
  mutate(pop18_turnout = ballots_cast/adult_18plus,
         registered_pop18 = reg_voters/adult_18plus)

# check 
summary(df24_tract_adult$pop18_turnout) #1.21 max
summary(df24_tract_adult$registered_pop18) #1.39 max
```
There are some tracts with higher than 100% voter turnout and registered voters. 

### Checking tracts with over 100%
#### where # of ballots cast is greater than the number of adults 18+
```{r}
df24_tract_above1_adultpop <- df24_tract_adult %>% 
  filter(pop18_turnout>1)

df24_tract_above1_adultpop

df24_tract_above1_adultpop_simp <- df24_tract_above1_adultpop %>% 
  select(planning_geog, ballots_cast, adult_18plus, pop18_turnout)
```
There are 9 tracts over 100% - compared to 11 tracts when considering the total population.

##### map of tracts >100% voter turnout
```{r}
# join full data to tract spatial file
df24_tract_adult_geo <- merge(tracts20.lyr, df24_tract_adult,
                              by.x="geoid20",
                              by.y="planning_geog", 
                              all.x=TRUE)

df24_tract_adult_geo <- df24_tract_adult_geo %>% 
  mutate(pop18_turnout=pop18_turnout*100)

# filter to tracts >100%
df24_tract_adult_geo_100plus <- df24_tract_adult_geo %>% 
  filter(pop18_turnout>100)
```

```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = df24_tract_adult_geo$pop18_turnout)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   overlayGroups = "tracts >100%",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=df24_tract_adult_geo,
              fillColor = psrc_palette(df24_tract_adult_geo$pop18_turnout),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(df24_tract_adult_geo$pop18_turnout, digits=0),"%")) %>%
  addPolygons(data=df24_tract_adult_geo_100plus,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "tracts >100%",
              label = paste(round(df24_tract_adult_geo_100plus$pop18_turnout, digits=0),"%")) %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = df24_tract_adult_geo$pop18_turnout,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

#### where # of registered voters is greater than the number of adults 18+
```{r}
df24_tract_above1_adultpop_reg <- df24_tract_adult %>% 
  filter(registered_pop18>1)

df24_tract_above1_adultpop_reg
```
There are 109 tracts over 100% - compared to 111 tracts when considering the total population.

```{r}
# calculate weighted values for voter turnout based on population 18+ and based on registered voters
df24_tract_calculations <- df24_tract_adult %>% 
  left_join(age_18over_total,
            by = c('planning_geog'='GEOID')) %>% 
  mutate(reg_voters_wt18 = reg_voters*pop18_total, #weight values calc from geog splits by eligible pop
         ballots_cast_wt18 = ballots_cast*pop18_total) %>% #weight values calc from geog splits by eligible pop
  mutate(pop18_turnout_wt18 = ballots_cast_wt18/adult_18plus,
         registered_pop18_wt18 = reg_voters_wt18/adult_18plus)

summary(df24_tract_calculations$pop18_turnout_wt18) #max: 0.96375, compared to 0.961 when considering total population
summary(df24_tract_calculations$registered_pop18_wt18) #max: 1.05687, compared to 1.05 when considering total population

# checking tracts where registered voters/population 18+ (weighted) >100%
df24_tract_calculations_above1 <- df24_tract_calculations %>% 
  filter(registered_pop18_wt18>1)
```
When weighted by the proportion 18+/total population, there are no tracts with over 100% voter tunrout based on ballots cast, however there are still 3 tracts that are over 100% when looking at registered voters, compared to the 2 tracts when considering the total population.


### Comparing tracts
To look compare the differences in the voter turnout for all tracts (based on ballots cast) between those calculated based on total population distribution and adult 18+ distribution

#### All tracts
```{r}
compare_all <- df24_tract_total %>% 
  full_join(df24_tract_adult,
            by = c("data_geog_type", "planning_geog_type",
                  "ofm_estimate_year", "ofm_vintage",
                  "parcel_year", "planning_geog",
                  "acs_year"))

compare_all_calc <- compare_all %>% 
  mutate(dif_total_adult=pop18_turnout.x-pop18_turnout.y)
compare_all_calc_simp <- compare_all_calc %>% 
  select(planning_geog, pop18_turnout.x, pop18_turnout.y, dif_total_adult)

summary(compare_all_calc$dif_total_adult)
hist(compare_all_calc$dif_total_adult)

filter <- compare_all_calc %>% 
  filter(dif_total_adult<0)
```
This shows that the difference in voter turnout is more negative which means that when weighting by adult 18+ distribution the values are slightly higher than the values when just taking the total population distribution into account. The differences are relatively small, but 386 tracts have differences that are positive, while 418 are negative. 

#### Those above 100%
To compare the differences in the voter turnout values for the tracts over 100% (based on ballots cast) between those calculated based on total population distribution and adult 18+ distribution
```{r}
compare_above1 <- df24_tract_above1_totalpop %>% 
 full_join(df24_tract_above1_adultpop,
           by = c("data_geog_type", "planning_geog_type",
                  "ofm_estimate_year", "ofm_vintage",
                  "parcel_year", "planning_geog",
                  "acs_year", 
                  "acs_reliability")) #reliability is good for all tracts in both datasets

compare_above1_calc <- compare_above1 %>% 
  mutate(dif_total_adult=pop18_turnout.x-pop18_turnout.y)
compare_above1_calc_simp <- compare_above1_calc %>% 
  select(planning_geog, pop18_turnout.x, pop18_turnout.y, dif_total_adult)

summary(compare_above1_calc$dif_total_adult)
hist(compare_above1_calc$dif_total_adult)
```

# C) Adult population by block group (decennial census, 2020)
## Join precinct-tract to geographies to turnout data
```{r}
nrow(df24_adult_dec) #precinct-tract records: 6294
nrow(vt24_df) #voting data by unique precinct: 4608
vt24_df_0 <- vt24_df %>% 
  filter(reg_voters<1) # there are 62 precincts without data
# data_geog = precinct
# planning_geog = tract
length(unique(df24_adult_dec$data_geog)) #unique precincts: 4552
length(unique(df24_adult_dec$planning_geog)) #unique tracts: 919
```
There is a difference in the number of records - in the voting data, there are 4,608 records (precincts) while the geography split data only has 4,552 unique precincts. The difference between these two data sets is 56 precincts. There are 62 precincts in the voting data without registered voters.

### checking the records that don't line up between the two data sets
```{r}
outer <- df24_adult_dec %>% 
  anti_join(vt24_df, by = c('data_geog'='precinctcode')) 
nrow(outer) #54 records
summary(outer$percent_of_total_pop)

outer2 <- vt24_df %>% 
  anti_join(df24_adult_dec, by = c('precinctcode'='data_geog')) 
nrow(outer2) #57 records
summary(outer2$percent_of_total_pop)
```
The output table from the first anti_join() includes census tract IDs without associated precincts.


## calculate total numbers for reference 
```{r}
sum(vt24_df$reg_voters) #2,736,409
sum(vt24_df$ballots_cast) #2,166,325
```
In the 2024 voting data (by precinct), there were 2,736,409 registered voters and 2,166,325 ballots cast.  

## join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
join_df24 <- df24_adult_dec %>% 
  full_join(vt24_df, 
            by = c('data_geog'='precinctcode')) #6351 records


# simplify data - remove unnecessary fields
join_df24_simp <- join_df24 %>% 
  select(-c(percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units))

# review data
summary(join_df24_simp$reg_voters) #54 NAs
summary(join_df24_simp$ballots_cast) #54 NAs

# calculate weighted registered voters and ballots cast 
df24_calc <- join_df24_simp %>% 
  mutate(reg_voters_wt = reg_voters*percent_of_total_pop,
         ballots_cast_wt = ballots_cast*percent_of_total_pop)
```

## calculate total numbers to check 
These calculations sum the values after being weighted by the geography splits. They should match the original voting precinct data. 
```{r}
sum(df24_calc$reg_voters_wt, na.rm = TRUE) #2,736,352 ~ 57 missing voters?
sum(df24_calc$ballots_cast_wt, na.rm = TRUE) #2,166,294 ~ 31 missing ballots cast

# does rounding make a difference?
round(sum(df24_calc$reg_voters_wt, na.rm = TRUE), digits = 0) #2,736,352 ~ still 57 missing voters
round(sum(df24_calc$ballots_cast_wt, na.rm = TRUE), digits =0) #2,166,294 ~ still 31 missing ballots cast
```
After weighting, the total registered voters is 2,736,352 and the total number of ballots cast is 2,166,294. This means that there are 57 missing voters and 31 missing ballots. These go missing during the weighting step.

### checking for missing voters/ballots
```{r}
# grouping and adding by precinct 
test <- df24_calc %>% 
  # mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
  #                                  TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
  #        ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
  #                                    TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog) %>% # grouping on precinct
  summarise(#reg_voters_test = sum(reg_voters_wt), # need to round
            #ballots_cast_test = sum(ballots_cast_wt), # need to round
            reg_voters_test = round(sum(reg_voters_wt), digits = 0),
            ballots_cast_test = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop')

test2 <- df24_calc %>% 
  full_join(test) %>% 
  mutate(dif_voters = if_else(reg_voters == reg_voters_test, "Match", "No Match"),
         dif_ballots = if_else(ballots_cast == ballots_cast, "Match", "No Match"))


# check number of matching/not matching records
table(test2$dif_voters) # all matching once rounded 'reg_voters_wt' above
table(test2$dif_ballots) # all matching once rounded 'ballots_cast_wt' above

test3 <- df24_calc %>% 
  group_by(data_geog) %>% # grouping on precincts
  reframe(total_pop = sum(percent_of_total_pop))
# when viewing the data frame in a separate window, all values are '1' but when looking at df in console there are some '1' and some '1.00'

summary(test3$total_pop) #57 NAs
```
Unable to find missing values?

## calculate voter turnout by registered voters 
(ballots cast/registered voters)
\
This is not the final calculation because it is based on those who are already engaged to an extent, not the general population who are capable of voting.
```{r}
df24_calc_tract <- df24_calc %>% 
  mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
                                   TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
         ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
                                     TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog_type, planning_geog_type, 
           ofm_estimate_year, ofm_vintage, 
           planning_geog) %>% # grouping on tract ids, retaining other fields for reference
  summarise(#reg_voters = sum(reg_voters_wt), # need to round?
            # ballots_cast = sum(ballots_cast_wt), # need to round?
            reg_voters = round(sum(reg_voters_wt), digits = 0),
            ballots_cast = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop') %>% 
  mutate(regvote_turnout = ballots_cast/reg_voters)

# calculate total numbers to check
sum(df24_calc_tract$reg_voters, na.rm=T) #2,736,352 when not rounded, 2,736,354 when rounded
sum(df24_calc_tract$ballots_cast, na.rm=T) #2,166,294 when not rounded, 2,166,293 when rounded
```

## calculate voter turnout by population 18+ 
(ballots cast/pop 18+)
\
This is closer to the final calculation because it takes into account all people who are capable of voting. This value may include non-citizens, or adults 18+ who are not legally allowed to vote, but it is more comprehensive than just considering those who are registered.
```{r}
# simplify acs table and join to voter/ballot data
age_18over_simp <- age_18over %>% 
  select(GEOID, 
         adult_18plus = estimate, 
         acs_year = year,
         acs_reliability = reliability)

# join to precinct/voter turnout data to tract 18+ population data to calculate population 18+ turnout
df24_calc_tract_adults <- df24_calc_tract %>% 
  left_join(age_18over_simp, 
            by = c('planning_geog'='GEOID')) 
  

# calculate voter turnout by population 18+
df24_tract_adult_dec <- df24_calc_tract_adults %>% 
  mutate(pop18_turnout = ballots_cast/adult_18plus,
         registered_pop18 = reg_voters/adult_18plus)

# check 
summary(df24_tract_adult_dec$pop18_turnout) #1.20 max
summary(df24_tract_adult_dec$registered_pop18) #1.39 max
```
There are some tracts with higher than 100% voter turnout and registered voters. 

### Checking tracts with over 100%
#### where # of ballots cast is greater than the number of adults 18+
```{r}
df24_tract_above1_adultpop_dec <- df24_tract_adult_dec %>% 
  filter(pop18_turnout>1)

df24_tract_above1_adultpop_dec

df24_tract_above1_adultpop_dec_simp <- df24_tract_above1_adultpop_dec %>% 
  select(planning_geog, ballots_cast, adult_18plus, pop18_turnout)
```
There are 10 tracts over 100% - compared to 9 tracts over 100% when considering the adult population (by tract, 2023 5y estimates) and compared to 11 tracts when considering the total population.

##### map of tracts >100% voter turnout
```{r}
# join full data to tract spatial file
df24_tract_adult_dec_geo <- merge(tracts20.lyr, df24_tract_adult_dec,
                              by.x="geoid20",
                              by.y="planning_geog", 
                              all.x=TRUE)

df24_tract_adult_dec_geo <- df24_tract_adult_dec_geo %>% 
  mutate(pop18_turnout=pop18_turnout*100)

# filter to tracts >100%
df24_tract_adult_dec_geo_100plus <- df24_tract_adult_dec_geo %>% 
  filter(pop18_turnout>100)
```

```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = df24_tract_adult_dec_geo$pop18_turnout)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   overlayGroups = "tracts >100%",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=df24_tract_adult_dec_geo,
              fillColor = psrc_palette(df24_tract_adult_dec_geo$pop18_turnout),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(df24_tract_adult_dec_geo$pop18_turnout, digits=0),"%")) %>%
  addPolygons(data=df24_tract_adult_dec_geo_100plus,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "tracts >100%",
              label = paste(round(df24_tract_adult_dec_geo_100plus$pop18_turnout, digits=0),"%")) %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = df24_tract_adult_dec_geo$pop18_turnout,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

#### where # of registered voters is greater than the number of adults 18+
```{r}
df24_tract_above1_adultpop_reg_dec <- df24_tract_adult_dec %>% 
  filter(registered_pop18>1)

df24_tract_above1_adultpop_reg_dec
```
There are 115 tracts over 100% - compared to 109 tracts over 100% when considering the adult population (by tract, 2023 5y estimates) and compared to 111 tracts when considering the total population.

```{r}
# calculate weighted values for voter turnout based on population 18+ and based on registered voters
df24_tract_calculations <- df24_tract_adult_dec %>% 
  left_join(age_18over_total,
            by = c('planning_geog'='GEOID')) %>% 
  mutate(reg_voters_wt18 = reg_voters*pop18_total, #weight values calc from geog splits by eligible pop
         ballots_cast_wt18 = ballots_cast*pop18_total) %>% #weight values calc from geog splits by eligible pop
  mutate(pop18_turnout_wt18 = ballots_cast_wt18/adult_18plus,
         registered_pop18_wt18 = reg_voters_wt18/adult_18plus)

summary(df24_tract_calculations$pop18_turnout_wt18) #max: 0.96342, compared to 0.96375 when considering adult population by tract, compared to 0.961 when considering total population
summary(df24_tract_calculations$registered_pop18_wt18) #max: 1.05653, compared to 1.05687 when considering adult population by tract, compared to 1.05 when considering total population

# checking tracts where registered voters/population 18+ (weighted) >100%
df24_tract_calculations_above1 <- df24_tract_calculations %>% 
  filter(registered_pop18_wt18>1)
```
There are still 3 tracts that are over 100% when looking at registered voters, compared to the 2 tracts when considering the total population.


### Comparing tracts
To look compare the differences in the voter turnout for all tracts (based on ballots cast) between those calculated based on total population distribution and adult 18+ distribution

#### All tracts
```{r}
compare_all <- df24_tract_adult %>% 
  full_join(df24_tract_adult_dec,
            by = c("data_geog_type", "planning_geog_type",
                  "ofm_estimate_year", "ofm_vintage",
                  "planning_geog",
                  "acs_year"))

compare_all_calc <- compare_all %>% 
  mutate(dif_tract_bg=pop18_turnout.x-pop18_turnout.y)
compare_all_calc_simp <- compare_all_calc %>% 
  select(planning_geog, pop18_turnout.x, pop18_turnout.y, dif_tract_bg)

summary(compare_all_calc$dif_tract_bg)
hist(compare_all_calc$dif_tract_bg)

filter <- compare_all_calc %>% 
  filter(dif_tract_bg>0)
```
This shows that the difference in voter turnout is slightly more positive which means that when weighting by adult 18+ distribution (tract) the values are slightly higher than the values when just taking the adult 18+ distribution (block group, dec) into account. The differences are relatively small, but 453 tracts have differences that are positive, while 446 are negative. 

#### Those above 100%
To compare the differences in the voter turnout values for the tracts over 100% (based on ballots cast) between those calculated based on total population distribution and adult 18+ distribution
```{r}
compare_above1 <- df24_tract_above1_adultpop %>% 
 full_join(df24_tract_above1_adultpop_dec,
           by = c("data_geog_type", "planning_geog_type",
                  "ofm_estimate_year", "ofm_vintage",
                  "planning_geog",
                  "acs_year", "adult_18plus", 
                  "acs_reliability")) #reliability is good for all tracts in both datasets

compare_above1_calc <- compare_above1 %>% 
  mutate(dif_tract_bg=pop18_turnout.x-pop18_turnout.y)
compare_above1_calc_simp <- compare_above1_calc %>% 
  select(planning_geog, pop18_turnout.x, pop18_turnout.y, dif_tract_bg)

summary(compare_above1_calc$dif_tract_bg)
hist(compare_above1_calc$dif_tract_bg)
```

# D) Adult population by tract (OFM SAEP 2024; ACS 5y 2023)
## Join precinct-tract to geographies to turnout data
```{r}
nrow(df24_adult) #precinct-tract records: 6294
nrow(vt24_df) #voting data by unique precinct: 4608
vt24_df_0 <- vt24_df %>% 
  filter(reg_voters<1) # there are 62 precincts without data
# data_geog = precinct
# planning_geog = tract
length(unique(df24_adult$data_geog)) #unique precincts: 4552
length(unique(df24_adult$planning_geog)) #unique tracts: 919
```
There is a difference in the number of records - in the voting data, there are 4,608 records (precincts) while the geography split data only has 4,552 unique precincts. The difference between these two data sets is 56 precincts. There are 62 precincts in the voting data without registered voters.

## calculate total numbers for reference 
```{r}
sum(vt24_df$reg_voters) #2,736,409
sum(vt24_df$ballots_cast) #2,166,325
```
In the 2024 voting data (by precinct), there were 2,736,409 registered voters and 2,166,325 ballots cast.  

## join voter turnout data to precinct-tract splits on precinct field
Using a full_join()
```{r}
join_df24 <- df24_adult %>% 
  full_join(vt24_df, 
            by = c('data_geog'='precinctcode')) #6351 records


# simplify data - remove unnecessary fields
join_df24_simp <- join_df24 %>% 
  select(-c(percent_of_group_quarters_pop, 
            percent_of_household_pop, 
            percent_of_housing_units,
            percent_of_occupied_housing_units))

# review data
summary(join_df24_simp$reg_voters) #54 NAs
summary(join_df24_simp$ballots_cast) #54 NAs

# calculate weighted registered voters and ballots cast 
df24_calc <- join_df24_simp %>% 
  mutate(reg_voters_wt = reg_voters*percent_of_total_pop,
         ballots_cast_wt = ballots_cast*percent_of_total_pop)
```

## join tract level OFM estimates to ACS adult 18+ ratio
```{r}
tract_population <- ofm_psrc %>% 
  left_join(age_18over_total, by= c("GEOID20"="GEOID"))

tract_pop_18plus <- tract_population %>% 
  mutate(population_18plus=round(as.numeric(population_2024)*pop18_total,0))
```

## calculate voter turnout by registered voters 
(ballots cast/registered voters)
\
This is not the final calculation because it is based on those who are already engaged to an extent, not the general population who are capable of voting.
```{r}
df24_calc_tract <- df24_calc %>% 
  mutate(reg_voters_wt = case_when(is.na(reg_voters_wt)~0, # need to convert NA to 0 so they can be added
                                   TRUE ~ reg_voters_wt), # if this isn't included, when they're grouped, the NA's will cancel out the other values within the census tract
         ballots_cast_wt = case_when(is.na(ballots_cast_wt)~0,
                                     TRUE ~ ballots_cast_wt)) %>% 
  group_by(data_geog_type, planning_geog_type, 
           ofm_estimate_year, ofm_vintage, parcel_year,
           planning_geog) %>% # grouping on tract ids, retaining other fields for reference
  summarise(#reg_voters = sum(reg_voters_wt), # need to round?
            # ballots_cast = sum(ballots_cast_wt), # need to round?
            reg_voters = round(sum(reg_voters_wt), digits = 0),
            ballots_cast = round(sum(ballots_cast_wt), digits = 0),
            .groups = 'drop') %>% 
  mutate(regvote_turnout = ballots_cast/reg_voters)

# calculate total numbers to check
sum(df24_calc_tract$reg_voters, na.rm=T) #2,736,352 when not rounded, 2,736,354 when rounded
sum(df24_calc_tract$ballots_cast, na.rm=T) #2,166,294 when not rounded, 2,166,297 when rounded
```

## calculate voter turnout by population 18+ 
(ballots cast/pop 18+)
\
This is closer to the final calculation because it takes into account all people who are capable of voting. This value may include non-citizens, or adults 18+ who are not legally allowed to vote, but it is more comprehensive than just considering those who are registered. The denominator will be based on OFM population estimates (2024), weighted by the ratio of adults 18+/total population from ACS (2023). 
```{r}
# join to precinct/voter turnout data to tract 18+ population data (OFM/ACS) to calculate population 18+ turnout
df24_calc_tract_adults_ofm <- df24_calc_tract %>% 
  left_join(tract_pop_18plus, by = c('planning_geog'='GEOID20')) 
  

# calculate voter turnout by population 18+
df24_tract_adult_ofm <- df24_calc_tract_adults_ofm %>% 
  mutate(pop18_turnout = ballots_cast/population_18plus,
         registered_pop18 = reg_voters/population_18plus)

# check 
summary(df24_tract_adult_ofm$pop18_turnout) #0.99887 max
summary(df24_tract_adult_ofm$registered_pop18) #1.19 max
```
There are no tracts with higher than 100% voter turnout and registered voters. 

### Checking tracts with over 100%
#### where # of ballots cast is greater than the number of adults 18+
```{r}
df24_tract_adult_ofm_simp <- df24_tract_adult_ofm %>% 
  select(planning_geog, ballots_cast, population_18plus, pop18_turnout)

df24_tract_above1_adultpop_ofm <- df24_tract_adult_ofm %>% 
  filter(pop18_turnout>1)

df24_tract_above1_adultpop_ofm

df24_tract_above1_adultpop_ofm_simp <- df24_tract_above1_adultpop_ofm %>% 
  select(planning_geog, ballots_cast, population_18plus, pop18_turnout)
```
There aren't any tracts over 100%.

##### map of tracts by voter turnout
```{r}
# join full data to tract spatial file
df24_tract_adult_ofm_geo <- merge(tracts20.lyr, df24_tract_adult_ofm,
                              by.x="geoid20",
                              by.y="planning_geog", 
                              all.x=TRUE)

df24_tract_adult_ofm_geo <- df24_tract_adult_ofm_geo %>% 
  mutate(pop18_turnout=pop18_turnout*100)

# filter to tracts >100%
df24_tract_adult_ofm_geo_100plus <- df24_tract_adult_ofm_geo %>% 
  filter(pop18_turnout>100)
```

```{r, echo=FALSE}
# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = df24_tract_adult_ofm_geo$pop18_turnout)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = "voter participation",
                   overlayGroups = "tracts >100%",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=df24_tract_adult_ofm_geo,
              fillColor = psrc_palette(df24_tract_adult_ofm_geo$pop18_turnout),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = "voter participation",
              label = paste(round(df24_tract_adult_ofm_geo$pop18_turnout, digits=0),"%")) %>%
  addPolygons(data=df24_tract_adult_ofm_geo_100plus,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "tracts >100%",
              label = paste(round(df24_tract_adult_ofm_geo_100plus$pop18_turnout, digits=0),"%")) %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = df24_tract_adult_ofm_geo$pop18_turnout,
                       position = "bottomright",
                       title = "Voter Participation",
                       group = "voter participation",
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat(suffix = "%")) %>% 

  #set view extent
  leaflet::setView(lng=-122.257, lat=47.615, zoom=8.5) %>%
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

#### where # of registered voters is greater than the number of adults 18+
```{r}
df24_tract_above1_adultpop_reg <- df24_tract_adult %>% 
  filter(registered_pop18>1)

df24_tract_above1_adultpop_reg
```
There are 52 tracts over 100% - about half as many from the other methods.

### Comparing tracts
To look compare the differences in the voter turnout for all tracts (based on ballots cast) between those calculated based on total population distribution and adult 18+ distribution

#### All tracts
```{r}
compare_all <- df24_tract_adult %>% 
  full_join(df24_tract_adult_ofm,
            by = c("data_geog_type", "planning_geog_type",
                  "ofm_estimate_year", "ofm_vintage",
                  "parcel_year", "planning_geog"))

compare_all_calc <- compare_all %>% 
  mutate(dif_adult_ofm=pop18_turnout.x-pop18_turnout.y)
compare_all_calc_simp <- compare_all_calc %>% 
  select(planning_geog, pop18_turnout.x, pop18_turnout.y, dif_adult_ofm)

summary(compare_all_calc$dif_adult_ofm)
hist(compare_all_calc$dif_adult_ofm)

filter <- compare_all_calc %>% 
  filter(dif_adult_ofm>0)
```
This shows that the difference in voter turnout is more positive which means that when the denominator is based on OFM estimates and weighted by ACS, the values are smaller than the values when just using just ACS adult 18+ estimates into account. 597 tracts have differences that are positive, while 321 are negative. 

#### Top 10 tracts
Looking at the ten tracts with the highest turnout
```{r}
fourth <- df24_tract_adult_ofm %>% 
  arrange(desc(pop18_turnout)) %>% 
  top_n(10) %>% 
  select(planning_geog, pop18_turnout)

fourth
```

# Conclusions
## Overall
Out of the three methods explored above, the fourth approach appears to be the best. 

## Process
We tested the second approach to determine if using adult 18+ population estimates in order to weight the initial population distribution (based on OFM parcels) would improve the issue of tracts with over 100% voter turnout. This was mostly based on the assumption that the population distribution of adults 18+ does not mirror the total population distribution - there are likely some areas where there are more or fewer households with children or young adults than others. 
\
\
We tested the third approach to determine if using adult 18+ population estimates at a more granular geographic scale in order to weight the population distribution would improve the issue of tracts with over 100% voter turnout. Perhaps estimates at the block group were better at accurately reflecting the population distribution, which would have resulted in more accurate weighting/translation between the precinct and tract. 
\
\
We tested the fourth approach to determine if the accuracy of the denominator could be improved. Instead of using the 2023 5y estimates for adults 18+ population, ACS 5y estimates may be better for proportions at the tract level. Perhaps the 5y average, outdated by one year, was too low for certain areas with more recent population growth. The OFM estimates may be a better more accurate representation of population, especially as they are available for 2024. 

## Rationale
Although there are still some concerns about the final numbers because tracts with over 100% voter turnout persist, the ability to weight the population distribution by the adult 18+ population seems to be the most accurate in terms or representing eligible voters. This makes the second and third approach more attractive than the first. 

Although the geographic resolution of block groups (third approach) is more refined than tracts (second approach), the additional geographic detail did not resolve the issue of tracts with over 100% voter turnout. In addition, the ability to use population estimates most closely aligned with the election year (2023 for second approach vs. 2020 for third approach) seems to outweigh the potential benefits of additional geographic accuracy. Using block group data from the decennial census would become more challenging to justify as the distance between 2020 and the election year increases (2028). This makes the second approach more attractive than the third.

The fourth approach closely resembles the second in terms of the numerator, but calculating the denominator from OFM values seemed to yield slightly better results - there were no tracts with over 100% voter turn out. 

## Summary of results
### 1st approach
#### All tracts
```{r}
summary(df24_tract_total$pop18_turnout)
hist(df24_tract_total$pop18_turnout)
```

#### Tracts with voter turnout over 100%
```{r}
df24_tract_above1_totalpop_simp %>% arrange(desc(pop18_turnout))

summary(df24_tract_above1_totalpop_simp$pop18_turnout)
hist(df24_tract_above1_totalpop_simp$pop18_turnout)
```

### 2nd approach
#### All tracts
```{r}
summary(df24_tract_adult$pop18_turnout)
hist(df24_tract_adult$pop18_turnout)
```

#### Tracts with voter turnout over 100%
```{r}
df24_tract_above1_adultpop_simp %>% arrange(desc(pop18_turnout))

summary(df24_tract_above1_adultpop_simp$pop18_turnout)
hist(df24_tract_above1_adultpop_simp$pop18_turnout)
```

### 3rd approach
#### All tracts
```{r}
summary(df24_tract_adult_dec$pop18_turnout)
hist(df24_tract_adult_dec$pop18_turnout)
```

#### Tracts with voter turnout over 100%
```{r}
df24_tract_above1_adultpop_dec_simp %>% arrange(desc(pop18_turnout))

summary(df24_tract_above1_adultpop_dec_simp$pop18_turnout)
hist(df24_tract_above1_adultpop_dec_simp$pop18_turnout)
```

### 4rd approach
#### All tracts
```{r}
summary(df24_tract_adult_ofm$pop18_turnout)
hist(df24_tract_adult_ofm$pop18_turnout)
```

#### Tracts with voter turnout over 100%
```{r}
df24_tract_above1_adultpop_ofm_simp %>% arrange(desc(pop18_turnout))

summary(df24_tract_above1_adultpop_ofm_simp$pop18_turnout)
# hist(df24_tract_above1_adultpop_ofm_simp$pop18_turnout)
```

## Comparing tracts across four methods
Comparing the census tracts with over 100% voter turn out from methods #1-3 and the top ten tracts from method #4
```{r}
# compare to the tracts >100% from the previous 3 methods:
first <- df24_tract_above1_totalpop %>% 
  select(planning_geog, pop18_turnout) %>% 
  arrange(desc(pop18_turnout)) %>% 
  rename(first=pop18_turnout)
  
second <- df24_tract_above1_adultpop%>% 
  select(planning_geog, pop18_turnout) %>% 
  arrange(desc(pop18_turnout)) %>% 
  rename(second=pop18_turnout)

third <- df24_tract_above1_adultpop_dec%>% 
  select(planning_geog, pop18_turnout) %>% 
  arrange(desc(pop18_turnout)) %>% 
  rename(third=pop18_turnout)

fourth <- fourth %>% 
  rename(fourth=pop18_turnout)
  

all_methods <- first %>% 
  full_join(second, by="planning_geog") %>% 
  full_join(third, by="planning_geog") %>% 
  full_join(fourth, by="planning_geog")

all_methods
```