---
title: "Sidewalk Completeness on Arterials in the Region"
subtite: "Data Gen: Exploring, Cleaning, Transforming (tract data)"
author: "Grant Gibson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Sidewalk Completeness**

# Download data 
## Primary data 
Data comes from the regional_bikeped and 2020 U.S. Census tract spatial layers stored on ElmerGeo. ArcMap was used to spatially join the arterial line segments to tract polygons.
```{r, include=FALSE}
folder_path <- "Y:/Equity Indicators/tracker-webpage-content/g-transportation/g03-sidewalk-completeness/raw-data/"
load(file.path(folder_path, "ped-facility-data.rda"))
```
\
The regional bike-ped data can also be explored on the [Transportation System Visualization Tool](https://experience.arcgis.com/experience/a587d27d1c444a6e891fe1b58508622d/page/Existing-Conditions/). Data is from 2020.


## Additional data

### equity quintile data

* low, low medium, etc.) at the tract-level for each equity focus group for the years that correspond to the sidewalk completeness data (2020
```{r, include=FALSE}
# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares") %>% dplyr::filter(data_year == 2020)
table(equity_tracts$data_year)
```

### tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)

nrow(tracts20.lyr) #919
```

# Explore data
## Most Recent Data
### Map
#### Join recent data to spatial file
```{r, include=FALSE}
# join base 2020 geographic data to regional arterial data
data_tract <- merge(tracts20.lyr, tract_ped_share,
                    by.x="geoid20",
                    by.y="geoid20", 
                    all.x=TRUE)

data_tract$estimate = data_tract$ped_share
```

```{r}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - to get legend high-low with correct color order
addLegend_decreasing <-
  function (map,
            position = c("topright", "bottomright", "bottomleft",
                         "topleft"),
            pal,
            values,
            na.label = "NA",
            bins = 7,
            colors,
            opacity = 0.5,
            labels = NULL,
            labFormat = labelFormat(),
            title = NULL,
            className = "info legend",
            layerId = NULL,
            group = NULL,
            data = getMapData(map),
            decreasing = FALSE) {
    position <- match.arg(position)
    type <- "unknown"
    na.color <- NULL
    extra <- NULL
    if (!missing(pal)) {
      if (!missing(colors))
        stop("You must provide either 'pal' or 'colors' (not both)")
      if (missing(title) && inherits(values, "formula"))
        title <- deparse(values[[2]])
      values <- evalFormula(values, data)
      type <- attr(pal, "colorType", exact = TRUE)
      args <- attr(pal, "colorArgs", exact = TRUE)
      na.color <- args$na.color
      if (!is.null(na.color) &&
          col2rgb(na.color, alpha = TRUE)[[4]] ==
          0) {
        na.color <- NULL
      }
      if (type != "numeric" && !missing(bins))
        warning("'bins' is ignored because the palette type is not numeric")
      if (type == "numeric") {
        cuts <- if (length(bins) == 1)
          pretty(values, bins)
        else
          bins
        
        if (length(bins) > 2)
          if (!all(abs(diff(bins, differences = 2)) <=
                   sqrt(.Machine$double.eps)))
            stop("The vector of breaks 'bins' must be equally spaced")
        n <- length(cuts)
        r <- range(values, na.rm = TRUE)
        cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
        n <- length(cuts)
        p <- (cuts - r[1]) / (r[2] - r[1])
        extra <- list(p_1 = p[1], p_n = p[n])
        p <- c("", paste0(100 * p, "%"), "")
        if (decreasing == TRUE) {
          colors <- pal(rev(c(r[1], cuts, r[2])))
          labels <- rev(labFormat(type = "numeric", cuts))
        } else{
          colors <- pal(c(r[1], cuts, r[2]))
          labels <- rev(labFormat(type = "numeric", cuts))
        }
        colors <- paste(colors, p, sep = " ", collapse = ", ")
        
      }
      else if (type == "bin") {
        cuts <- args$bins
        n <- length(cuts)
        mids <- (cuts[-1] + cuts[-n]) / 2
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "bin", cuts))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "bin", cuts)
        }
        
      }
      else if (type == "quantile") {
        p <- args$probs
        n <- length(p)
        cuts <- quantile(values, probs = p, na.rm = TRUE)
        mids <- quantile(values, probs = (p[-1] + p[-n]) / 2,
                         na.rm = TRUE)
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "quantile", cuts, p))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "quantile", cuts, p)
        }
      }
      else if (type == "factor") {
        v <- sort(unique(na.omit(values)))
        colors <- pal(v)
        labels <- labFormat(type = "factor", v)
        if (decreasing == TRUE) {
          colors <- pal(rev(v))
          labels <- rev(labFormat(type = "factor", v))
        } else{
          colors <- pal(v)
          labels <- labFormat(type = "factor", v)
        }
      }
      else
        stop("Palette function not supported")
      if (!any(is.na(values)))
        na.color <- NULL
    }
    else {
      if (length(colors) != length(labels))
        stop("'colors' and 'labels' must be of the same length")
    }
    legend <-
      list(
        colors = I(unname(colors)),
        labels = I(unname(labels)),
        na_color = na.color,
        na_label = na.label,
        opacity = opacity,
        position = position,
        type = type,
        title = title,
        extra = extra,
        layerId = layerId,
        className = className,
        group = group
      )
    invokeMethod(map, data, "addLegend", legend)
  }

```

```{r tract data map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Sidewalk Completeness"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Missing Data
The data used to measure the share of sidewalk completeness in the region is only comprised of arterial road segments. Since not every Census tract in the region will have arterial roadways in the tract, there are six of 919 tracts that are missing data.

#### Isolate NA or missing data
```{r}
# join NA data to the shapefile to map
data_tract_na <- data_tract %>% 
  filter(is.na(estimate))

nrow(data_tract_na) # 6 tracts
```

```{r missing tract map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Sidewalk Completeness"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = var_name,
                   overlayGroups = "missing data",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%
  addPolygons(data=data_tract_na,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "missing data") %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Descriptive Statistics
The following descriptive statistics reflect the census tracts that have values - the six census tracts with NAs have been removed. These are not weighted by population numbers. 
```{r}
data_tract_nona <- data_tract %>%
  filter(!is.na(estimate))

summary(data_tract_nona$estimate)
```

```{r, include=FALSE}
distrib <- data_tract_nona$estimate

med.value <- median(distrib) #0.88
```
*Median:* `r med.value`
\
\

#### Histogram
A histogram is a visual representation of the distribution of a dataset. The y-axis shows how frequently the values on the x-axis occur in the data, while the bars group ranges of values or continuous categories on the x-axis [(source)](https://www.datacamp.com/tutorial/make-histogram-basic-r).  
```{r}
hist(distrib)
```
\

#### Boxplot
A boxplot helps to visualize a quantitative variable by displaying five common location summary (minimum, median, first and third quartiles and maximum) and any observation that was classified as a suspected outlier using the interquartile range (IQR) criterion [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
boxplot(distrib, horizontal = TRUE)
```
\

#### Outliers
The IQR criterion means that all observations above or below the first and third quartile respectively, and IQR is the difference between the third and first quartile) are considered as potential outliers by R [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
outliers <- boxplot.stats(distrib)$out
outlier_rle <- rle(outliers)
outlier_df <- data.frame(unclass(outlier_rle)) %>% 
  rename(sidewalk_comp = lengths) %>% 
  arrange(values)

outlier_df
```

There are no tracts that have outliers. 
\
\

# **MAP .rda**
Save final data set (.rda) for map
```{r, include=FALSE}
# This is the data set used to create the map of the most recent data. For this data set, there is one additional transformation that is needed before saving it for final visualization in the 'vis' script. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk. 

```

```{r, region and county numbers for map labels}
# County data has already been calculated using a 50 ft buffer - identical to the method used for tract data

county_reg_data <- county_ped_share %>% 
  dplyr::select(county_name = county, cnty_estimate = ped_share) %>% 
  dplyr::mutate(reg_estimate = region_ped_share)

# merge tract, county, and region data into one data set 
data_tract <- merge(data_tract, county_reg_data,
                    by.x="county_name",
                    by.y="county_name",
                    all.x=TRUE)
```

```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'g-transportation'
ind_dir <- 'g03-sidewalk-completeness'
file_name <- 'g03-sidewalk-completeness-map-data'

# save final data set as .rda
save(data_tract, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

# Explore data by equity quintile
## Join primary data to equity quintiles
```{r}
# Tracts by equity quintiles have already been filtered to 2020
data_equity_tract <- merge(tract_ped_share, equity_tracts,
                           by.x="geoid20",
                           by.y="geoid", 
                           all.x=TRUE)

# check data set
nrow(data_equity_tract)
```
This data set should include about twice the number of census tracts in the region - census tracts are listed twice - once as part of the region and another as part of their corresponding county.

## Transform/pivot
The data sets need to be cleaned and pivoted to a longer table so that there is only one field for the 6 different equity categories, instead of one field for each.
```{r, include=FALSE}
# clean/simplify data sets
data_equity_quintiles20 <- data_equity_tract %>% 
  dplyr::mutate(data_year=format(data_year,format="%Y"))

# pivot data set so that there is one column with quintile designation
data_equity_pivot20 <- data_equity_quintiles20 %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")
```

```{r}
# check data sets
nrow(data_equity_pivot20)
```

These data sets should include records for:

* 2 geographies (region and whichever county the tract belongs to) x # census tracts (919 for 2020)
* 6 equity focus groups


## Calculations
### Calculate by region
The shares of sidewalk completeness are available at the census tract level, but we want to calculate the regional average.
```{r}
data_equity_pivot20_calc <- data_equity_pivot20 %>% 
  filter(!is.na(quintile)) %>%
  dplyr::group_by(equity_group, quintile) %>%
  summarise(ped_share = round(sum(ped_length, na.rm = TRUE)/sum(segment_length, na.rm = TRUE), 2)) %>% 
  mutate(county="Region") %>% 
  mutate(data_year = "2020")
```

```{r}
# check data sets
nrow(data_equity_pivot20_calc)
```
These data sets should include records for:

* 1 geography: region
* 6 equity focus groups
* 5 equity quintiles 


### Calculate by equity quintile
The life expectancy values are available at the census tract level but we want to calculate the weighted values by the 6 equity/5 quintile groups, using the census bureau's tract-level population estimates (2010, 2015, 2020). 
```{r}
# calculations to get weighted sidewalk completeness share per equity group/quintile
data_equity_wt20 <- data_equity_pivot20 %>%
  filter(county != "Region") %>% # necessary because of added na columns from pivot
  dplyr::group_by(county, data_year, equity_group, quintile) %>%
  dplyr::summarise(tracts=length(geoid20),
                   ped_length = sum(ped_length, na.rm = TRUE),
                   segment_length = sum(segment_length, na.rm = TRUE)) %>%
  dplyr::mutate(ped_share = round(ped_length / segment_length, 2))

# check data set
nrow(data_equity_wt20)
```

This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 6 equity focus groups
* 5 equity quintiles
    * 2020: (subtracting) 2 missing quintiles (High: Kitsap - LEP, Pierce - LEP) 

## Combine years into one data set
```{r}
data_tract <- data_equity_wt20 %>% 
  dplyr::select(-ped_length, -segment_length, -tracts) %>% 
  rbind(data_equity_pivot20_calc) %>% 
  filter(!is.na(quintile))

# check data set
str(data_tract)
```
This data set should include records for: 

* 5 geographies: 4 PSRC counties + region
* 3 years
* 6 equity focus groups
    * 2010: (subtracting) 2 missing focus groups(disability, lep) * 5 equity quintiles * 5 geographies
    * 2015: (subtracting) 1 missing focus group (lep) * 5 equity quintiles * 5 geographies
* 5 equity quintiles

### Address "Low - Low Medium" quintile group
This code was added because of Kitsap/LEP data - there were too many tracts with 0, so we combined the bottom two quintiles into low - low medium. This change requires some additional processing so that all of the tracts (bottom 40%) in this combined quintile category are averaged and this average value is assigned to separate 'low' and 'low medium' quintile categories. 
```{r}
# nrow(data_tract) #149
# unique(data_tract$quintile) #6 quintiles because of additional "Low - Low Medium"

low_quintile <- data_tract %>% 
  filter(quintile == "Low - Low Medium") %>% 
  rename(quintile_original = quintile) %>% 
  mutate(quintile = "Low") %>% 
  filter(county != "Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications
lowmed_quintile <- data_tract %>% 
  filter(quintile == "Low - Low Medium") %>%
  rename(quintile_original = quintile) %>%
  mutate(quintile = "Low Medium") %>% 
  filter(county != "Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications

low_lowmed <- rbind(low_quintile,
                    lowmed_quintile)
# nrow(low_lowmed) #4

# gather counties that have low - medium categorization
low_lowmed_counties <- unique(low_lowmed$county) #Kitsap, Pierce
low_lowmed_years <- unique(low_lowmed$data_year) #2020
low_lowmed_equity <- unique(low_lowmed$equity_group) #lep_quintile

other_data <- data_tract %>% 
  filter(quintile!="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>% 
  mutate(quintile=quintile_original)
# nrow(other_data) #146

data_tract <- rbind(low_lowmed, #4
                    other_data) #146

nrow(data_tract) #150 - should be fewer because of duplicates (375)- low and low medium present in data set because of aggregation/calculations in previous steps - need to be removed and replaced with averaged values
# table(data_tract$quintile, data_tract$County.Name) #should not have extra in low and low medium quintile categories
# unique(data_tract$quintile) #5 quintiles

# remove rows where it is one of the "Low - Low Medium" quintiles AND the original quintile is "Low" or "Low Medium" - only want to keep the rows that are "Low"/"Low Medium" and they share a value
data_tract <- data_tract %>% 
  filter(!(county %in% low_lowmed_counties & 
           data_year %in% low_lowmed_years &
           equity_group %in% low_lowmed_equity &
           (quintile_original=="Low" | 
             quintile_original=="Low Medium")))

# nrow(data_tract) #150
# table(data_tract$quintile, data_tract$county) #should not have extra in low and low medium quintile categories
```

## Rename and factor
```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

quintile_order <- c("Low", 
                    "Low\nMedium", 
                    "Medium", 
                    "Medium\nHigh", 
                    "High")

equity_group_order <- c("People of Color", 
                        "Households with Lower Income", 
                        "People with a Disability", 
                        "Households with Limited English Proficiency", 
                        "Households with Youth <18", 
                        "Households with Older Adults 65+")

# transforming data labels ----
data_clean <- data_tract %>% 
  mutate(county_ord = factor(county, levels=county_order)) %>%
  mutate(equity_group_ord = case_when(
    equity_group=="poc_quintile"~"People of Color",
    equity_group=="disability_quintile"~"People with a Disability",
    equity_group=="lep_quintile"~"Households with Limited English Proficiency",
    equity_group=="income_quintile"~"Households with Lower Income",
    equity_group=="youth_quintile"~"Households with Youth <18",
    equity_group=="older_quintile"~"Households with Older Adults 65+")) %>%
  mutate(equity_group_ord = factor(equity_group_ord, levels = equity_group_order)) %>% 
  mutate(quintile_ord = str_wrap(quintile, width=7)) %>%
  mutate(quintile_ord = factor(quintile_ord, levels = quintile_order))

# Minor edits to the data
data_clean <- data_clean %>%
  dplyr::mutate(estimate = ped_share) #rename value variable for consistency

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county_ord, equity_group_ord, quintile_ord, data_year)
```


# **CHART .rda**
Save final data set for charts (.rda)
```{r}
# set folder structure
file_name <- 'g03-sidewalk-completeness-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>