---
title: "Access to High Capacity Transit"
subtite: "Data Gen: Exploring, Cleaning, Transforming (tract data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(psrctrends)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)
library(ggplot2)

library(summarytools) #freq
library(vtable) #summary stats
library(table1)  #nice descriptive summary table
library(scales) #number formatting
library(ggpubr) #graphing - ggarrange fx
library(forcats) #for factor re-leveling
library(plotly) #for interactive charts

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(raster)
library(ggspatial)
library(lubridate) #year formatting
library(stringr) #add leading zero's
library(reshape2) #formatting data
library(readxl)
library(gridExtra)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext
```

**Access to High Capacity Transit**

# Download data 
## Primary data 
Data comes from GTFS (General Transit Feed Specification). More information about the way that this data set is generated can be found on the Equity Tracker wiki [page](https://github.com/psrc/equity-tracker/blob/main/data_development/distance/README.md). Weighted distance to high capacity transit is available for each census tract, for each year, starting in 2011. 2011-2019 data are available at the 2010 census geographies, while 2020 onward uses 2020 census geographies.
```{r, include=FALSE}
folder_path <- "Y:/Equity Indicators/tracker-webpage-content/g-transportation/g02-hct-access/raw-data"

hct <- read.csv(file.path(folder_path,"tract_hct_distance_compiled.csv"),
                header = TRUE)
```
\
Because data is available at each year we will be able to use the data at 1y increments. 


## Additional data

### equity quintile data and 2010-20 crosswalk file 

* low, low medium, etc. at the tract-level for each equity focus group for the years that correspond to the hct access data
* crosswalk data shouldn't be necessary because the hct data is available in the corresponding year geographies
```{r, include=FALSE}
# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares")
table(equity_tracts$data_year)

# crosswalk_10_20 <- get_table(schema="census",
#                              tbl_name="v_geo_relationships_tracts")
```

### tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
tracts20.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
tracts10.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

tracts20.lyr<-st_read(tracts20.url)
tracts10.lyr<-st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

# Explore data
The equity quintile tracts are in 2010 and 2020 geographies depending on the year and include ~770 or ~920 census tracts, each of which are assigned one of the 5 quintile categories for each of the 6 equity demographic groups. 
```{r, include=FALSE}
# check number of rows in data sets
nrow(hct) #42620
# 2011-2019 -> 9 years * 720 census tracts = ~6480
# 2020-2023 -> 4 years * 920 census tracts = ~3680
# 6480+3680= ~10160 census tracts 
list(hct)
unique(hct$access_to) # 5 different types of hct
# 5*10160= ~50800
```
The hct access dataset includes 5 unique transit types, which means that each census tract for each year is going to be listed 5 times, as long as the data was available. Since there seems to be fewer rows than we would expect, there may be certain years where data for certain transit types were not available.

```{r, include=FALSE}
# to check the field names and data class
str(hct)
```

## Most Recent Data
### Map
#### Join recent data to spatial file
```{r, include=FALSE}
# filter to most recent year's data and 'all_hct' instead of the different modes
hct_2023 <- hct %>% 
  filter(analysis_year==2023,
         access_to=="all_hct")

# join base 2023 hct access data to 2020 census tracts 
data_tract <- merge(tracts20.lyr, hct_2023,
                    by.x="geoid20",
                    by.y="tract_geoid", 
                    all.x=TRUE)

data_tract <- data_tract %>% 
  dplyr::mutate(estimate=as.numeric(wt_avg_miles)) # rename value variable for consistency
```

```{r}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - to get legend high-low with correct color order
addLegend_decreasing <-
  function (map,
            position = c("topright", "bottomright", "bottomleft",
                         "topleft"),
            pal,
            values,
            na.label = "NA",
            bins = 7,
            colors,
            opacity = 0.5,
            labels = NULL,
            labFormat = labelFormat(),
            title = NULL,
            className = "info legend",
            layerId = NULL,
            group = NULL,
            data = getMapData(map),
            decreasing = FALSE) {
    position <- match.arg(position)
    type <- "unknown"
    na.color <- NULL
    extra <- NULL
    if (!missing(pal)) {
      if (!missing(colors))
        stop("You must provide either 'pal' or 'colors' (not both)")
      if (missing(title) && inherits(values, "formula"))
        title <- deparse(values[[2]])
      values <- evalFormula(values, data)
      type <- attr(pal, "colorType", exact = TRUE)
      args <- attr(pal, "colorArgs", exact = TRUE)
      na.color <- args$na.color
      if (!is.null(na.color) &&
          col2rgb(na.color, alpha = TRUE)[[4]] ==
          0) {
        na.color <- NULL
      }
      if (type != "numeric" && !missing(bins))
        warning("'bins' is ignored because the palette type is not numeric")
      if (type == "numeric") {
        cuts <- if (length(bins) == 1)
          pretty(values, bins)
        else
          bins
        
        if (length(bins) > 2)
          if (!all(abs(diff(bins, differences = 2)) <=
                   sqrt(.Machine$double.eps)))
            stop("The vector of breaks 'bins' must be equally spaced")
        n <- length(cuts)
        r <- range(values, na.rm = TRUE)
        cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
        n <- length(cuts)
        p <- (cuts - r[1]) / (r[2] - r[1])
        extra <- list(p_1 = p[1], p_n = p[n])
        p <- c("", paste0(100 * p, "%"), "")
        if (decreasing == TRUE) {
          colors <- pal(rev(c(r[1], cuts, r[2])))
          labels <- rev(labFormat(type = "numeric", cuts))
        } else{
          colors <- pal(c(r[1], cuts, r[2]))
          labels <- rev(labFormat(type = "numeric", cuts))
        }
        colors <- paste(colors, p, sep = " ", collapse = ", ")
        
      }
      else if (type == "bin") {
        cuts <- args$bins
        n <- length(cuts)
        mids <- (cuts[-1] + cuts[-n]) / 2
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "bin", cuts))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "bin", cuts)
        }
        
      }
      else if (type == "quantile") {
        p <- args$probs
        n <- length(p)
        cuts <- quantile(values, probs = p, na.rm = TRUE)
        mids <- quantile(values, probs = (p[-1] + p[-n]) / 2,
                         na.rm = TRUE)
        if (decreasing == TRUE) {
          colors <- pal(rev(mids))
          labels <- rev(labFormat(type = "quantile", cuts, p))
        } else{
          colors <- pal(mids)
          labels <- labFormat(type = "quantile", cuts, p)
        }
      }
      else if (type == "factor") {
        v <- sort(unique(na.omit(values)))
        colors <- pal(v)
        labels <- labFormat(type = "factor", v)
        if (decreasing == TRUE) {
          colors <- pal(rev(v))
          labels <- rev(labFormat(type = "factor", v))
        } else{
          colors <- pal(v)
          labels <- labFormat(type = "factor", v)
        }
      }
      else
        stop("Palette function not supported")
      if (!any(is.na(values)))
        na.color <- NULL
    }
    else {
      if (length(colors) != length(labels))
        stop("'colors' and 'labels' must be of the same length")
    }
    legend <-
      list(
        colors = I(unname(colors)),
        labels = I(unname(labels)),
        na_color = na.color,
        na_label = na.label,
        opacity = opacity,
        position = position,
        type = type,
        title = title,
        extra = extra,
        layerId = layerId,
        className = className,
        group = group
      )
    invokeMethod(map, data, "addLegend", legend)
  }

```

```{r tract data map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "HCT access"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Missing Data
Life expectancy calculations can fluctuate considerably in smaller populations or populations experiencing low or no deaths for the year(s) being calculated. Because of these issues the Life Expectancy at birth calculation for the Census Tract geographies is suppressed for Census Tracts with a population (for the 5 years combined) of <5000 or a result with a Standard Error >2 or a record of <50 deaths for the time period. The Washington Department of Health, Center for Health Statistics estimates data gathered from death certificates to be 99% complete.

#### Isolate NA or missing data
```{r}
# join NA data to the shapefile to map
data_tract_na <- data_tract %>% 
  filter(is.na(estimate))

nrow(data_tract_na) #0 tracts
```

```{r missing tract map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$estimate)

# set the variable
var_name <- "Life Expectancy"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = var_name,
                   overlayGroups = "missing data",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$estimate, digits=1)) %>%
  addPolygons(data=data_tract_na,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "missing data") %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

### Descriptive Statistics
The following descriptive statistics reflect the census tracts that have values - the `r na_rows_num` census tracts with NAs have been removed. These are not weighted by population numbers. 
```{r}
data_tract_nona <- data_tract %>% 
  filter(!is.na(estimate))

summary(data_tract_nona$estimate)
```

**When weighted by tract populations.** 
```{r, include=FALSE}
# data_tract_nona_weight <- data_tract_nona %>% 
#   mutate(tot_age = estimate*total_pop10,
#          reg_tot = sum(tot_age),
#          tot_pop = sum(total_pop10),
#          est_weight_avg = reg_tot/tot_pop)

distrib <- rep(data_tract_nona$estimate, 1)

```

#### Histogram
A histogram is a visual representation of the distribution of a dataset...The y-axis shows how frequently the values on the x-axis occur in the data, while the bars group ranges of values or continuous categories on the x-axis [(source)](https://www.datacamp.com/tutorial/make-histogram-basic-r).  
```{r}
hist(distrib)
```
\

#### Boxplot
A boxplot helps to visualize a quantitative variable by displaying five common location summary (minimum, median, first and third quartiles and maximum) and any observation that was classified as a suspected outlier using the interquartile range (IQR) criterion [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
boxplot(distrib, horizontal = TRUE)
```
\

#### Outliers
The IQR criterion means that all observations above or below the first and third quartile respectively, and IQR is the difference between the third and first quartile) are considered as potential outliers by R [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
outliers <- boxplot.stats(distrib)$out
outlier_rle <- rle(outliers)
outlier_df <- data.frame(unclass(outlier_rle)) %>% 
  rename(population = lengths,
         health_in = values) %>% 
  arrange(health_in)

outlier_df
```
\

##### *Check outliers*
The following steps are specific to the life expectancy dataset and may or may not be applicable depending on the data.

###### Identify census tracts with outliers
```{r include=FALSE}
outlier_values <- outlier_df[,2]

data_tract_nona_outliers <- data_tract_nona %>% 
  dplyr::filter(estimate %in% outlier_values)

num_outliers <- nrow(data_tract_nona_outliers) #61
```
There are `r num_outliers` tracts that have outliers. 
\
\

###### Outliers on map
```{r tract data map}
# get layer with hct stations
hct.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/high_capacity_transit/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

hct.lyr <-st_read(hct.url)

# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract_nona_outliers$estimate)

# set the variable
var_name <- "HCT access"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract_nona_outliers,
              fillColor = psrc_palette(data_tract_nona_outliers$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract_nona_outliers$estimate, digits=1)) %>%
  addPolylines(data = hct.lyr) %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract_nona_outliers$estimate,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```


# **MAP .rda**
Save final data set (.rda) for map
```{r, include=FALSE}
# This is the data set used to create the map of the most recent data. For this data set, there is one additional transformation that is needed before saving it for final visualization in the 'vis' script. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk. 

str(data_tract) # most recent data: 2023 joined to 2020 spatial file (generated above)
```

```{r, region and county numbers for map labels}
# calculating data by county 
county_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  dplyr::group_by(county_name) %>% 
  dplyr::mutate(cnty_estimate=mean(estimate, na.rm =TRUE)) %>% #calculate mean value by county
  dplyr::distinct(county_name, cnty_estimate)

# calculating data by region
reg_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  dplyr::mutate(reg_estimate=mean(estimate, na.rm =TRUE)) %>% #calculate mean value for region
  dplyr::distinct(reg_estimate)

# add region data to county data
county_reg_data <- county_tract_data %>% 
  mutate(reg_estimate=reg_tract_data$reg_estimate)

# merge tract, county, and region data into one data set 
data_tract <- merge(data_tract, county_reg_data,
                    by.x="county_name",
                    by.y="county_name",
                    all.x=TRUE)
```

```{r}
# set folder structure
base_dir <- 'Y:/Equity Indicators/tracker-webpage-content'
theme_dir <- 'g-transportation'
ind_dir <- 'g02-hct-access'
file_name <- 'g02-hct-access-map-data'

# save final data set as .rda
save(data_tract, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

# Explore data by equity quintile
## Join primary data to equity quintiles
```{r}
# this will require joining each of the years of hct access data (1-year increments) to each of the PUMS equity quintile/tract data (also available for each year, but based on 5-year time periods)
# determine which years each dataset includes
# because GTFS data is available starting in 2015, we are filtering hct data to 2015 and after (removing 2011-2014)
hct_2015 <- hct %>% 
  filter(analysis_year>2014) 

yrs_hct <- unique(hct_2015$analysis_year) #2015-2023
yrs_equity_tracts <- unique(equity_tracts$data_year) #2009-2021

# because of the mismatch in years between the two data sets we will take the common years
years_of_interest <- intersect(yrs_hct,yrs_equity_tracts) #2015-2021

# filter data sets so that they have overlapping/common years
hct_for_join <- hct %>% 
  filter(analysis_year %in% years_of_interest,
         access_to=="all_hct") # focus on all_hct, not individuals modes

# unique(hct_for_join$analysis_year) #2015-2021
# nrow(hct_for_join) #5717: ~ (5*772) + (2*919)

equity_tracts_for_join <- equity_tracts %>% 
  filter(data_year %in% years_of_interest) 

# unique(equity_tracts_for_join$data_year) #2015-2021
# nrow(equity_tracts_for_join) #11396: ~[(5*772) + (2*919)]*2 <- 2 because each census tract is listed for region and for whichever county it belongs to

# join data sets - join on year and census tract - should have same number of rows as equity_tracts_for_join, with the same number of unique values
# n_distinct(hct_for_join$wt_avg_miles) #5690

data_equity <- merge(equity_tracts_for_join, hct_for_join,
                     by.x=c("data_year","geoid"),
                     by.y=c("analysis_year","tract_geoid"), 
                     all.x=TRUE)

# n_distinct(data_equity$wt_avg_miles) #5682
```

## Transform/pivot
The data sets need to be cleaned and pivoted to a longer table so that there is only one field for the 6 different equity categories, instead of one field for each.
```{r, include=FALSE}
# pivot data set so that there is one column with quintile designation
data_equity_tracts <- data_equity %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile") %>% 
  dplyr::mutate(data_year=format(data_year,format="%Y")) %>% 
  dplyr::select(-c(X, index)) # remove unnecessary columns

# check data set
nrow(data_equity) #11396
nrow(data_equity_tracts) #68376 = ~11396*6 (6 equity groups)
n_distinct(data_equity_tracts$wt_avg_miles) #5682
```
These data sets should include records for:

* 2 geographies (region and whichever county the tract belongs to) x # census tracts (~773 for 2015-2019, ~919 for 2020-2021)
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015 - but still show up as NA)


## Calculations
### Calculate by region
The hct access distance values are available at the census tract level, but we want to calculate the regional average. Because the values are already weighted we don't need to take into account the tracts' population counts.
```{r}
data_equity_tracts_reg <- data_equity_tracts %>%
  filter(!is.na(quintile)) %>% #missing lep and disability data
  filter(county=="Region") %>% #only want census tracts to appear once, not twice
  dplyr::group_by(data_year, equity_group, quintile) %>% 
  summarise(wt_avg_miles=mean(wt_avg_miles, na.rm=TRUE),.groups='drop') %>% 
  mutate(county="Region")

# check data
table(data_equity_tracts_reg$data_year, data_equity_tracts_reg$equity_group) #5 quintiles
table(data_equity_tracts_reg$quintile, data_equity_tracts_reg$equity_group) #7 years
```
These data sets should include records for:

* 1 geography: region
* 7 data years
* 6 equity focus groups (lep missing data in 2015)
* 5 equity quintiles 


### Calculate by equity quintile
The hct access distance values are available at the census tract level, but we want to calculate the regional average. Because the values are already weighted we don't need to take into account the tracts' population counts. 
```{r}
data_equity_tracts_county <- data_equity_tracts %>%
  filter(!is.na(quintile)) %>% #missing lep and disability data
  filter(county!="Region") %>% #only want census tracts to appear once, not twice
  dplyr::group_by(county, data_year, equity_group, quintile) %>% 
  summarise(wt_avg_miles=mean(wt_avg_miles, na.rm=TRUE),.groups='drop')

# check data
table(data_equity_tracts_county$data_year, data_equity_tracts_county$equity_group) #20 - 5 quintiles, 4 counties
table(data_equity_tracts_county$quintile, data_equity_tracts_county$equity_group) #28 - 7 years, 4 counties
```
\
\
This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 7 years
* 6 equity focus groups (lep missing data in for 2015)
* 5 equity quintiles 

## Combine regional and county level data 
```{r}
data_tract <- as.data.frame(rbind(data_equity_tracts_reg,
                                  data_equity_tracts_county)) %>% 
  filter(!is.na(quintile))

# check data set
str(data_tract)
unique(data_tract$county)
table(data_tract$data_year, data_tract$county)
```
This data set should include records for: 

* 5 geographies: 4 PSRC counties + region
* 7 years
* 6 equity focus groups
* 5 equity quintiles

### Address "Low - Low Medium" quintile group
This code was added because of Kitsap/LEP data - there were too many tracts with 0, so we combined the bottom two quintiles into low - low medium. This change requires some additional processing so that all of the tracts (bottom 40%) in this combined quintile category are averaged and this average value is assigned to separate 'low' and 'low medium' quintile categories. 
```{r}
# nrow(data_tract) #378
# unique(data_tract$quintile) #6 quintiles because of additional "Low - Low Medium"

low_quintile <- data_tract %>% 
  filter(quintile=="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>% 
  mutate(quintile="Low") %>% 
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications
lowmed_quintile <- data_tract %>% 
  filter(quintile=="Low - Low Medium") %>%
  rename(quintile_original=quintile) %>%
  mutate(quintile="Low Medium") %>% 
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications

low_lowmed <- rbind(low_quintile,
                    lowmed_quintile)
# nrow(low_lowmed) #4

# gather counties that have low - medium categorization
low_lowmed_counties <- unique(low_lowmed$County.Name) #Kitsap, Pierce
low_lowmed_years <- unique(low_lowmed$data_year) #2020
low_lowmed_equity <- unique(low_lowmed$equity_group) #lep_quintile

other_data <- data_tract %>% 
  filter(quintile!="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>% 
  mutate(quintile=quintile_original)
# nrow(other_data) #375

data_tract <- rbind(low_lowmed, #4
                    other_data) #375

nrow(data_tract) #379 - should be fewer because of duplicates (375)- low and low medium present in data set because of aggregation/calculations in previous steps - need to be removed and replaced with averaged values
# table(data_tract$quintile, data_tract$County.Name) #should not have extra in low and low medium quintile categories
# unique(data_tract$quintile) #5 quintiles

# remove rows where it is one of the "Low - Low Medium" quintiles AND the original quintile is "Low" or "Low Medium" - only want to keep the rows that are "Low"/"Low Medium" and they share a value
data_tract <- data_tract %>% 
  filter(!(county %in% low_lowmed_counties & 
           data_year %in% low_lowmed_years &
           equity_group %in% low_lowmed_equity &
           (quintile_original=="Low" | 
             quintile_original=="Low Medium")))

# nrow(data_tract) #375
# table(data_tract$quintile, data_tract$County.Name) #should not have extra in low and low medium quintile categories
```

## Rename and factor
```{r}
# wrap/order labels ----
county_order <- c("Region", "King", "Kitsap", "Pierce", "Snohomish")

quintile_order <- c("Low", 
                    "Low\nMedium", 
                    "Medium", 
                    "Medium\nHigh", 
                    "High")

equity_group_order <- c("People of Color", 
                        "Households with Lower Income", 
                        "People with a Disability", 
                        "Households with Limited English Proficiency", 
                        "Households with Youth <18", 
                        "Households with Older Adults 65+")

# transforming data labels ----
data_clean <- data_tract %>% 
  mutate(county_ord = factor(county, levels=county_order)) %>%
  mutate(equity_group_ord = case_when(
    equity_group=="poc_quintile"~"People of Color",
    equity_group=="disability_quintile"~"People with a Disability",
    equity_group=="lep_quintile"~"Households with Limited English Proficiency",
    equity_group=="income_quintile"~"Households with Lower Income",
    equity_group=="youth_quintile"~"Households with Youth <18",
    equity_group=="older_quintile"~"Households with Older Adults 65+")) %>%
  mutate(equity_group_ord = factor(equity_group_ord, levels = equity_group_order)) %>% 
  mutate(quintile_ord = str_wrap(quintile, width=7)) %>%
  mutate(quintile_ord = factor(quintile_ord, levels = quintile_order))

# Minor edits to the data
data_clean <- data_clean %>%
  dplyr::mutate(estimate=as.numeric(wt_avg_miles)) #rename value variable for consistency

# Sort the data to ensure the charts work
data_clean <- data_clean %>%
  arrange(county_ord, equity_group_ord, quintile_ord, data_year)
```


# **CHART .rda**
Save final data set for charts (.rda)
```{r}
# set folder structure
file_name <- 'g02-hct-access-data'

# save final data set as .rda
save(data_clean, file = file.path(base_dir,
                                  theme_dir,
                                  ind_dir, "rda-data",
                                  paste0(file_name,'.rda')))
```

<a href="#top">Back to top of the page</a>