---
title: "Health Insurance"
subtitle: Visuals for Equity Tracker (PUMS data)
author: "Kristin Mitchell, updated by Christy Lam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'a-regional-health-collaboration',
                   ind_dir = 'a04-health-insurance',
                   map = 'a04-health-insurance-map-data',
                   chart = 'a04-health-insurance-chart-data',
                   map_name = 'a04-health-insurance-map.html',
                   column_chart_name = 'a04-health-insurance-column',
                   line_chart_name = 'a04-health-insurance-line')

# set most recent year
year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
According to the Centers for Disease Control and Prevention, “without health insurance, people are less likely to have a regular health care provider and more likely to skip routine health care. This puts them at increased risk for serious health problems.” It is helpful to know which communities have lower access to health insurance to help address health disparities.

The majority of households in all places have health insurance. Therefore, the scale for the map starts at 70% and the vertical axes for the charts have been adjusted to start at 60%.

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# move the decimal point for reg_estimate and cnty_estimate
acs_data_tract <- acs_data_tract %>%
   mutate(reg_estimate = round((reg_estimate*100), 1)) %>%
   mutate(cnty_estimate = round((cnty_estimate*100),1))
```


```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(acs_data_tract$estimate) # max is 250001, so rounding to 260000 makes sense
max_value <- 100 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
# acs_data_plus_zero<-append(acs_data_tract$estimate,0)
acs_data_plus_zero<-append(acs_data_tract$estimate, 70)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = acs_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Health Insurance <br> Coverage" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  acs_data_tract$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", 
        prettyNum((round(acs_data_tract$estimate, digits=2))), " %"), 
  paste("Tract data reliability: ", acs_data_tract$reliability),
  paste("Region value: ",
        prettyNum((round(acs_data_tract$reg_estimate, digits=2))), " %"), 
  paste("County value: ",
        prettyNum((round(acs_data_tract$cnty_estimate, digits=2))), " %"), 
  paste0("(",acs_data_tract$county_name," County)")) %>%
  lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", acs_data_tract$geoid20,
#   paste(prettyNum((round(acs_data_tract$estimate, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)


# map settings
acs_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=acs_data_tract,
              fillColor = psrc_palette(acs_data_tract$estimate),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(acs_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
acs_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
acs_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2701; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(acs_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- acs_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))

tracts_low_cov <- acs_data_tract %>% 
  arrange(estimate) |> 
  head(5)

tracts_low_cov_labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</body><br/><body>%s</bodyD>",
  tracts_low_cov$geoid20, # the code in <> makes the geoid value bold/italic
  paste("Tract value: ", 
        prettyNum((round(tracts_low_cov$estimate, digits=2))), " %"), 
  paste("Tract data reliability: ", tracts_low_cov$reliability),
  paste("Region value: ",
        prettyNum((round(tracts_low_cov$reg_estimate, digits=2))), " %"), 
  paste("County value: ",
        prettyNum((round(tracts_low_cov$cnty_estimate, digits=2))), " %"), 
  paste0("(",tracts_low_cov$county_name," County)")) %>%
  lapply(htmltools::HTML)

tracts_low_cov_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data = tracts_low_cov,
              fillColor = psrc_palette(acs_data_plus_zero),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = tracts_low_cov_labels) |> 
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

```

1. 94.4%: The region’s average health insurance coverage
2. 77.5%: The lowest average health insurance coverage at census tract level in the region
3. 13%: The regional health insurance coverage is 13 percentage points lower for households with limited English proficiency compared to households proficient in English

\

#### Insights & Analysis

* Kitsap County has the highest average for health insurance coverage (94.9%), followed by King (94.7%), Pierce (94.0%), and Snohomish (93.8%).
* There are 7 census tracts with 100% health insurance coverage: 5 in King County, and 1 each in Kitsap, and Pierce counties.
* The five census tracts with the lowest health insurance coverage, Edmonds (77.5%), Lake Stickney near Everett (77.7%), Alderton, between Sumner and Orting (78.0%), Lynnwood (78.9%), South Tacoma (79.1%).
* The national average health insurance coverage in 2022 is 92%.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year == year) #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = .50
y_max = 1
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Health Insurance by Community"
subtitle = "Percent of persons with health insurance coverage"
source = "U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")

# region/king low income households: .87	
# region/king non-low income households: .95
# .87-.95=-.08

# kitsap low income households: .90	
# kitsap non-low income households: .96

# pierce low income households: .88	
# pierce non-low income households: .95

# sno low income households: .87	
# sno non-low income households: .95

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat",
         county=="Region")

# 2022
# region lep households: .79	
# region non-lep households: .94
# .79-.94=-0.15

# King : .79
# King non : .95
# diff: .16

# Pierce : .77
# Pierce non: .94
# diff: 17


filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat",
         vulnerability=="high vulnerability")
# 2022
# region lep households: .81	
# king lep households: .81
# kitsap lep households: .81
# pierce lep households: .78
# snohomish lep households: .80

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Youth_cat")

# region youth households: .92	
# region non-youth households: .94

filter <- data_clean_column %>% 
  filter(focus_type=="POC_cat")
# 2022
# region poc: .90
# region non-poc: .96

# snohomish poc: .89
# snohomish non-poc: .95

# king poc: .91
# king non-poc: .96

# kitsap poc: .91
# kitsap non-poc: .96

# pierce poc: .90
# pierce non-poc: .95

filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# with disability has higher coverage across all counties/region

filter <- data_clean_column %>% 
  filter(focus_type=="Income_cat",
         vulnerability=="high vulnerability") %>% 
  dplyr::arrange(desc(fact_value))
```
\

#### Insights & Analysis

* Health care coverage for households with limited English proficiency in Pierce County is 17 percentage points less than households proficient in English, the largest such disparity in the region across all communities and geographies.
* The largest disparity in health care coverage between people of color and white non-Hispanic people is 6 percentage points in Snohomish County and in the region.
* Coverage for households with lower income in King County, Snohomish County, and the region are each 8 percentage points less than other households.
* Coverage is 6 percentage points less for people of color in the region than for white non-Hispanic people.


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df <- data_clean
x <- "data_year" # different from variables for facet column
fill <- "focus_attribute_ord" # different from variables for facet column
color <- "blues"
title <- "Health Insurance Trend by Community"
subtitle <- "Percent of persons with health insurance coverage, in 5-year spans between 2017 and 2022"
source <- "U.S. Census Bureau, 2008-2012, 2013-2017, 2018-2022 American Community Survey 5-Year Public Use Microdata Sample"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums - 1}
### Biggest disparities ----

middle_year <- 2017
recent_year <- 2022

# is the gap between two lines (low vulnerability vs high vulnerability) lessening or widening?
disparity_df <- data_clean |> 
  select(data_year, county, vulnerability, focus_type, focus_attribute_ord, fact_value_notrounded) 

disparity_df_wide <- disparity_df |> 
  filter(data_year != middle_year) |> 
  pivot_wider(id_cols = c(data_year, county, focus_type),
              names_from = vulnerability,
              values_from = fact_value_notrounded) |> 
  relocate(`high vulnerability`, .after = `low vulnerability`) |> 
  mutate(disparity = `low vulnerability` - `high vulnerability`) |> 
  group_by(county, focus_type) |> 
  mutate(disparity_btwn_years = disparity-lag(disparity, 1)) |>  # numerator
  mutate(disparity_share = (disparity_btwn_years/lag(disparity, 1)))

disparity_top <- disparity_df_wide |> 
  group_by(data_year, county) |> 
  slice_min(disparity_share, n = 10) |> 
  arrange(data_year, county, disparity_share)

### Deltas ----

delta_df <- data_clean |> 
  select(data_year, county, vulnerability, focus_type, focus_attribute_ord, fact_value_notrounded) |> 
  filter(data_year != middle_year) 

delta_years <- unique(delta_df$data_year)
names(delta_years) <- c("start", "end")

delta_df_wide <- delta_df |> 
  pivot_wider(id_cols = c(county, focus_type, vulnerability),
              names_from = data_year,
              values_from = fact_value_notrounded
              ) |> 
  rename(delta_years) |> 
  mutate(delta = end - start) |> 
  mutate(delta_share = delta/start)
```

```{r line facet calculations - pums - 2, include=FALSE, eval=FALSE}

filter <- disparity_df_wide |> 
  filter(county == "Region" & focus_type == "Income_cat")
# column disparity_share = -.653

# 2022 low inc: 0.87
# 2022 non low inc: 0.95
# 2022 diff: .081

# 2012 low inc: 0.68
# 2012 non low in: 0.91
# 2012 diff: .232

# ((.081 - .232)/.232)*100

filter <- disparity_df_wide |> 
  filter(county == "Region" & focus_type == "POC_cat")
# column disparity_share = -.544

filter <- data_clean |> 
  filter(county == "Snohomish" & focus_type == "LEP_cat")
```
\

#### Insights & Analysis

* From 2012 to 2022, health insurance coverage increased for all groups.
* The gap between people of color and white non-Hispanic people in the region declined by 50% between 2012 and 2022.
* Health coverage for households with lower income in the region increased by 29% between 2012 and 2022. The disparity between households with lower income versus other households in the region dropped by 67%.
* The largest disparity in health care coverage is between households with limited English proficiency and households proficient in English. Kitsap County experienced increased health coverage for households with limited English proficiency by 29 percentage points between 2012 and 2017, however between 2017 and 2022, it dropped 8 percentage points.
* Households with older adults 65+ and people with disabilities have higher health insurance coverage due to programs like Medicare and Medicaid.
* From 2012 to 2022, households with older adults 65+ in King, Pierce, and Snohomish counties experienced a three percentage point increase in health insurance coverage, while those in Kitsap County experienced a two percentage point increase. 
\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>