---
title: "Life Expectancy"
subtite: "Data Gen: Exploring, Cleaning, Transforming (tract data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcelmer",
#                          force=TRUE)
library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)

library(vtable) #summary stats
library(forcats) #for factor re-leveling

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(ggspatial)
library(lubridate) #year formatting
library(readxl)
library(summarytools) ##freq/ctable functions

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(here)
```

```{r sources}

# reference supplemental script with supporting settings/functions
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))
```

```{r variables setup}

file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'd-community-and-health',
                   ind_dir = 'd01-life-expectancy',
                   chart = 'd01-life-expectancy-chart-data',
                   map = 'd01-life-expectancy-map-data')

folder_path <- file.path(file_names$base_dir,
                         file_names$theme_dir,
                         file_names$ind_dir,
                         "raw-data")

# create list of separate CSVs - OPTIONAL depending on format of data, if in one file this step is unnecessary
raw_data <- list("Life_Expectancy_at_Birth_2010tracts_08_12.csv",
                 "Life_Expectancy_at_Birth_2010tracts_13_17.csv",
                 "Life_Expectancy_at_Birth_2020tracts_18_22.csv")

year <- 2022
years <- c(2012, 2017, 2022)
data <- 1:3 #number of datasets in raw_data list - to iterate through
```

**Life Expectancy at Birth**

# Download data 
## Primary data 
These data comes from [WTN](https://fortress.wa.gov/doh/wtn/WTNPortal/?preview=true/#!q0=8967). Separate CSVs at the census tract geography are available (manual download) for 5y time spans at 1y increments, starting in 2000. The data are available at 2010 and 2020 geographies, depending on the year. The raw data sets have been saved in the corresponding Y drive project folder - folder address included above. [Please replace this text with a description of your data and data source here to maintain a record for future updates.]
```{r}

# this function was created because the indicator data are in separate CSV files that were downloaded from the WTN. [If your data source if one csv, you can ignore/comment out this step and skip to the bottom of this code chunk - use the commented out example code.]
read_data <- function(order){
  
  # read csv files 
  csv_data <- read.csv(file.path(folder_path, raw_data[order]))
  
  # add year field (step necessary because each data set was downloaded separately and there wasn't an existing year field)
  data <- csv_data %>% 
    mutate(data_year=years[order])
}

df_raw_data <- map_dfr(data, ~read_data(order=.x))

# example code if data source is one csv -----
# df_raw_data <- read.csv(file.path(folder_path,"Cardiovascular_Disease_Mortality_06_10.csv"))
```

### Look at raw data
Use this space to explore the raw data - the values and format. Are all of the census tracts included? Are the years of interest correct?Are there data that are not necessary to keep? This step will determine how the data may need to be cleaned and transformed. 
```{r}
# print data set to see field names and field types
df_raw_data

# check data
str(df_raw_data)
# table(df_raw_data$data_year)
```

### Clean and format data
This will step will require different cleaning and formatting steps depending on the data set.
```{r}
# For the WTN data, all of the state's counties are included in addition to a row for the indicator at a state-level. 

# add year and filter to only PSRC counties
df_psrc_data <- df_raw_data %>% 
  # remove counties outside of PSRC region
  filter(County.Name == "King" |
           County.Name == "Kitsap" |
           County.Name == "Pierce" |
           County.Name == "Snohomish") %>% 
  # there are 2 census tract fields - because of the 2 different census tract vintages (2010 and 2020) 
  mutate(GEOID= case_when(is.na(Census.Tract)~Census.Tract.2020,
                          is.na(Census.Tract.2020)~Census.Tract))

# check data
str(df_psrc_data)
# there are 2,614 observations when there should be about 2,471 [776 census tracts X 2 (2012 and 2017 data) + 919 census tracts (2022)]
table(df_psrc_data$data_year)
# there are the expected number of rows for 2012 and 2017, but there are extra in the 2022 data set

test_2022 <- df_psrc_data %>% 
  filter(data_year=="2022")

length(unique(test_2022$Census.Tract.2020)) #1062 rows in 2022, when there should be 919 or 923, so about 140 additional records

table(test_2022$County.Name)
# Based on the shapefile (downloaded below, there should be King: 494, Kitsap: 59, Pierce: 193, Snohomish: 173), so there are extra census tracts in all counties. This should be taken care of when joining to the equity quintiles (for the charts) or the spatial file (for the map). But curious about the extra tracts in the WTN data set.
```

## Additional data
These additional data sets will help get the indicator data to the correct format for creating the maps and charts that are part of the Equity Tracker.They may not all be necessary depending on the indicator data set. 

### Equity quintile tract-level data 
This data set includes the quintile designation (low, low medium, etc.) for each equity focus group / year of interest / census tract - this data set is used to join to the indicator data by tract id and year. This is required for all tract-based data. 
```{r, include=FALSE}

# Elmer data sets - equity quintile tracts using psrcelmer() and crosswalk
equity_tracts <- get_table(schema="equity", tbl_name="v_tract_shares")

# to check
str(equity_tracts)
table(equity_tracts$data_year)
```

### 2010-20 crosswalk file (OPTIONAL - data dependent)
This data set includes the relationship between the 2010 and 2020 census tract ids for indicator data sets that span the two time periods. If data includes the 2010 and 2020 tract geographies, this crosswalk could be necessary to join to the indicator data.
```{r}
crosswalk_10_20 <- get_table(schema="census",
                             tbl_name="v_geo_relationships_tracts")
```

### Weighting data (OPTIONAL - data dependent)
This step is only necessary for data sets that require normalization or weighting. For example, life expectancy at the tract-level is the average number of years a newborn is expected to live by census tract. This number for each census tract would need to be weighted by the population of each census tract to better understand trends throughout the region. 
```{r, include=FALSE}

# getting census population data by tract
tracts_pop_acs <- get_acs_recs(geography ='tract', 
                           table.names = 'B01003',
                           years = years, # refers to the list of years that was set above
                           acs.type = 'acs5') 

# simplify data set
tracts_pop <- tracts_pop_acs %>% 
  select(GEOID, estimate, year) 

# check data set
table(tracts_pop$year)
```

### Geographic tract file
This spatial file is required to map the most recent year's data. It includes all of the census tracts within the PSRC region. Depending on the most recent indicator data, the 2010 census tract spatial file may be needed. 
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
arc_service <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services"

tracts20.url <- file.path(arc_service, "Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
tracts10.url <- file.path(arc_service, "Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")

tracts20.lyr <- st_read(tracts20.url)
tracts10.lyr <- st_read(tracts10.url)

nrow(tracts20.lyr) #919
nrow(tracts10.lyr) #773
```

# Create dataset
This step is optional depending on the format of the indicator. For this specific indicator, the data needs to be weighted by the population.
```{r}
# indicator data set
df_psrc_data
# weighting data set
tracts_pop

# need to combine
df_psrc_data_pop <- df_psrc_data %>% 
  inner_join(tracts_pop, join_by("GEOID"=="GEOID", "data_year"=="year"))

# check issues with extra census tracts in 2022
# test22 <- df_psrc_data_pop %>% 
#   filter(data_year==2022)
# nrow(test22)
# table(test22$County.Name)

# simplify data and calculate weighted estimates
df_psrc_data_pop_simp <- df_psrc_data_pop %>% 
  select(County.Name, Life.Expectancy, data_year, GEOID, estimate) %>% 
  # rename 'estimate' to population for clarity
  rename(population=estimate)
```

# Finalize data set
This steps is included so that the data set and variable of interest and year variable are named in a consistent way for downstream processing. 
```{r}
tract_indicator <- df_psrc_data_pop_simp %>% # insert the name of your data set
  rename(variable_value=Life.Expectancy, # insert the name of your indicator variable
         data_year=data_year, # insert the name of your year variable
         geoid=GEOID) %>% # insert the name of your census tract or geography variable
  mutate(variable_value=as.numeric(variable_value)) # ensure the indicator variable is numeric so it can be mapped/visualized

```

Use this space to make sure the final data set is formatted correctly (includes GEOID, year, variable of interest) with the consistent variable names. Depending on the number of years that are available and the geography vintage (2010 or 2020 census tracts), the number of rows will vary, but there should be rows for each census tract and each year. 
```{r, include=FALSE}
nrow(tract_indicator)

table(tract_indicator$data_year)

# to check the field names and data class
str(tract_indicator)
```

# Explore most recent data
## Join recent data to spatial file
```{r, include=FALSE}
# refine data to most recent year
tract_indicator_now <- tract_indicator %>% 
  filter(data_year==year)

# join to spatial file
data_tract <- merge(tracts20.lyr, tract_indicator_now,
                    by.x="geoid20",
                    by.y="geoid", 
                    all.x=TRUE)
```

## Visualize map
```{r tract data map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$variable_value)

# set the variable
var_name <- "Life Expectancy"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$variable_value),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$variable_value, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$variable_value,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

## Missing Data
This step is dependent on the availability of your data set. If you have NAs you can use this step to visualize where in the region the NAs are occurring. If you do not have any NAs, you can skip this section and move on to the 'Descriptive Statistics' section.

### Isolate NA or missing data
Life expectancy calculations can fluctuate considerably in smaller populations or populations experiencing low or no deaths for the year(s) being calculated. Because of these issues the Life Expectancy at birth calculation for the Census Tract geographies is suppressed for Census Tracts with a population (for the 5 years combined) of <5000 or a result with a Standard Error >2 or a record of <50 deaths for the time period. The Washington Department of Health, Center for Health Statistics estimates data gathered from death certificates to be 99% complete.
```{r}
# join NA data to the shapefile to map
data_tract_na <- data_tract %>% 
  filter(is.na(variable_value))

nrow(data_tract_na) #5 tracts
```

```{r missing tract map}
# set map extent
map.lat<- 47.615
map.lon<- -122.257
map.zoom<- 8.5

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = data_tract$variable_value)

# set the variable
var_name <- "Life Expectancy"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  leaflet::addLayersControl(baseGroups = var_name,
                   overlayGroups = "missing data",
                   options = layersControlOptions(collapsed = FALSE)) %>%
  
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$variable_value),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(data_tract$variable_value, digits=1)) %>%
  addPolygons(data=data_tract_na,
              color = "red",
              stroke = TRUE, 
              weight = 3,
              smoothFactor = 0.5,
              fillOpacity = 0,
              group = "missing data") %>% 

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = data_tract$variable_value,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

## Descriptive Statistics
The following descriptive statistics reflect the census tracts that have values - the census tracts with NAs have been removed. 
```{r}
data_tract_nona <- data_tract %>% 
  filter(!is.na(variable_value))

# The summary function treats each tract value as equal - doesn't account for any weighting
summary(data_tract_nona$variable_value)
```

**When weighted by tract populations.** 
```{r, include=FALSE}
data_tract_nona_weight <- data_tract_nona %>% 
  mutate(tot_age = variable_value*population,
         reg_tot = sum(tot_age),
         tot_pop = sum(population),
         est_weight_avg = reg_tot/tot_pop)

distrib <- rep(data_tract_nona$variable_value, data_tract_nona$population)

avg.value <- data_tract_nona_weight$est_weight_avg[1] #80.43404
med.value <- median(distrib) #80.16
```
\

### Histogram
A histogram is a visual representation of the distribution of a dataset...The y-axis shows how frequently the values on the x-axis occur in the data, while the bars group ranges of values or continuous categories on the x-axis [(source)](https://www.datacamp.com/tutorial/make-histogram-basic-r).  
```{r}
hist(distrib)
```
\

### Boxplot
A boxplot helps to visualize a quantitative variable by displaying five common location summary (minimum, median, first and third quartiles and maximum) and any observation that was classified as a suspected outlier using the interquartile range (IQR) criterion [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
boxplot(distrib, horizontal = TRUE)
```
\

### Outliers
The IQR criterion means that all observations above or below the first and third quartile respectively, and IQR is the difference between the third and first quartile) are considered as potential outliers by R [(source)](https://statsandr.com/blog/outliers-detection-in-r/).  
```{r}
outliers <- boxplot.stats(distrib)$out
outlier_rle <- rle(outliers)
outlier_df <- data.frame(unclass(outlier_rle)) %>% 
  rename(population = lengths,
         health_in = values) %>% 
  arrange(health_in)

outlier_df
```
\

#### *Check outliers* (OPTIONAL)
The following steps are specific to the life expectancy data set and may or may not be applicable depending on the data.

##### Identify census tracts with outliers
```{r include=FALSE}
outlier_values <- outlier_df[,2]

data_tract_nona_weight_outliers <- data_tract_nona_weight %>% 
  dplyr::filter(variable_value %in% outlier_values)

num_outliers <- nrow(data_tract_nona_weight_outliers) #21
```
\

##### Outliers by Displacement Risk score (OPTIONAL)
This check uses the displacement risk data, which is mapped to 2010 census tracts. This will help determine if there is a connection between displacement risk and your indicator. This step may not apply to your data set. 
```{r displacement risk data, inlcude=FALSE}
# Connecting to Portal for displacement tract geographies ----
displacement.url <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services/Displacement_Risk_Data/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"

displacement.lyr <-st_read(displacement.url)
# head(displacement.lyr)
# plot(displacement.lyr$geometry)
# class(displacement.lyr$geoid10)

displacement_simp <- displacement.lyr %>%
  dplyr::select(geoid10,
                risk_score) %>% 
  as_tibble() %>% 
  dplyr::select(-geometry)
```

```{r, include=FALSE}
data_tract_nona_weight_outliers <- data_tract_nona_weight_outliers %<>% 
  dplyr::mutate(geoid=trimws(geoid20))

# join outlier tract datas to spatial dispalcement tract data to isolate tracts with outlier data
data_tract_risk_outliers <- merge(x=displacement_simp, 
                                      y=data_tract_nona_weight_outliers,
                                      by.x="geoid10",
                                      by.y="geoid",
                                      all.y=TRUE)
head(data_tract_risk_outliers)
# class(data_tract_risk_outliers)
# plot(data_tract_risk_outliers$geometry)
nrow(data_tract_risk_outliers) #21
pop_outliers <- as.numeric(sum(data_tract_risk_outliers$population, na.rm = T)) #76,772
# There are 76,772 people (2022 ACS 5y population) in these outlier tracts.
```

\
\
```{r}
# compare the two measures
plot(data_tract_risk_outliers$risk_score, 
     data_tract_risk_outliers$variable_value)
```

```{r}
# this will be different for each indicator - the code and the text included below...
# based on the plot in the last chunk, this code is meant to isolate outlier(s) - in this case it filters for the outliers with high risk scores and high life expectancy

# outlier #1
data_tract20_outliers_1 <- data_tract_risk_outliers %>% 
  filter(risk_score > 50)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_1

# outlier #2
data_tract20_outliers_2 <- data_tract_risk_outliers %>% 
  filter(risk_score < 17)

# code to look at the scores and census tracts that are outliers
data_tract20_outliers_2

# To locate the census tracts on the map, its easy to use the Data Portal (https://psrc-psregcncl.hub.arcgis.com/datasets/census-tracts-2020/explore?location=47.506129%2C-121.980700%2C9.23) - just filter on the geoid20 field. 
```


# **MAP .rda**
The map .rda is the data set used to create the map of the most recent data. For this data set, there is one additional transformation that is needed before saving it for final visualization in the 'vis' script. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk.
```{r}
# workspace for edits and reformatting before saving as .rda
str(data_tract)
```

## Region and County numbers
Because the measure (life expectancy) does not account for the population of the tract, it hasn't been normalized and needs to be weighted in order to aggregate to the region/county level - these values will be referred to in the code and visualized on the map for regional context. 

Depending on the indicator, weighting by population may not be necessary, but region and county numbers should be calculated, if possible. 
```{r, region and county numbers for map labels}
# calculating data by county 
county_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  dplyr::select(County.Name, variable_value, population) %>% #to simplify table to necessary fields
  dplyr::group_by(County.Name) %>% 
  dplyr::mutate(cnty_estimate=(sum(variable_value*population, na.rm =TRUE))/(sum(population, na.rm=TRUE))) %>% #calculate weighted value by county
  dplyr::distinct(County.Name, cnty_estimate)

# calculating data by region
reg_tract_data <- data_tract %>% 
  st_drop_geometry() %>%  #to remove spatial component
  dplyr::select(variable_value, population) %>% #to simplify table to necessary fields
  dplyr::mutate(reg_estimate=(sum(variable_value*population, na.rm =TRUE))/(sum(population, na.rm=TRUE))) %>% #calculate weighted value for region
  dplyr::distinct(reg_estimate)

# add region data to county data
county_reg_data <- county_tract_data %>% 
  mutate(reg_estimate=reg_tract_data$reg_estimate)

# merge tract, county, and region data into one data set 
data_tract <- merge(data_tract, county_reg_data,
                    by="County.Name",
                    all.x=TRUE)
```

## Save final data set (.rda) for map
```{r}
# save final data set as .rda
save(data_tract, file = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir, 
                                  "update",
                                  "rda-data",
                                  paste0(file_names$map,'.rda')))
```

# Explore data by equity quintile
## Join primary data to equity quintiles.
```{r}
tract_indicator
equity_tracts

# join indicator data to equity quintiles 
data_equity_tracts <- merge(tract_indicator, equity_tracts,
                            by=c("geoid", "data_year"),
                            all.x=TRUE)

# check data set
nrow(data_equity_tracts) #4938
table(data_equity_tracts$data_year)
```
This data set should include rows for each census tracts twice (once as part of the region and another as part of their corresponding county) for each year.


## Transform/pivot
The data set need to be cleaned and pivoted to a longer table so that there is only one field for the 6 different equity categories, instead of one field for each.
```{r, include=FALSE}
str(data_equity_tracts)

# clean/simplify data sets
data_equity_quintile <- data_equity_tracts %>% 
  dplyr::mutate(data_year=format(data_year,format="%Y"))

# pivot data set so that there is one column with quintile designation
data_equity_pivot <- data_equity_quintile %>% 
  pivot_longer(cols = poc_quintile:lep_quintile,
               names_to = "equity_group",
               values_to = "quintile")

# check data sets
nrow(data_equity_pivot)
str(data_equity_pivot)
```

These data sets should include records for:

* 2 geographies (region and whichever county the tract belongs to) x # census tracts (~773 for 2010, ~919 for 2020)
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015 - but still show up as NA)


## Calculations
### Calculate by region
The data are available at the census tract level, but we want to calculate the regional average. For life expectancy we also want to weight the values using the ACS population estimates. 

The step of calculating by region is important for all data sets because the region is one of the visualization options (radio buttons), along with each county. The weighting by population is optional depending on the data set. 
```{r}
data_equity_wt_region <- data_equity_pivot %>% 
  filter(!is.na(quintile)) %>% #missing lep and disability data
  dplyr::group_by(equity_group, quintile, data_year) %>%
  summarise(variable_value_wt=(sum(variable_value*population, na.rm =TRUE))/(sum(population, na.rm=TRUE))) %>% 
  mutate(County.Name="Region")

# check data set
nrow(data_equity_wt_region)
```

These data sets should include records for:

* 1 geography: region
* 6 equity focus groups (disability and lep are not available in for 2010, lep not available in 2015)
* 5 equity quintiles 
* 3 years (this is dependent on the indicator data set)


### Calculate by equity quintile
The data are available at the census tract level but we want to calculate the weighted values by the 6 equity/5 quintile groups. For life expectancy we also want to weight the values using the ACS population estimates. 

The step of calculating by county/equity group/quintile is important for all data sets. The weighting by population is optional depending on the data set. 
```{r}

# calculations to get weighted average life expectancy
data_equity_wt_county <- data_equity_pivot %>% 
  filter(!is.na(quintile)) %>% # necessary because of added na columns from pivot
  mutate(total_risk_tract=variable_value*population) %>% 
  group_by(County.Name, data_year, equity_group, quintile) %>%
  summarise(tracts=length(geoid),
            total_variable=(sum(total_risk_tract, na.rm = TRUE)),
            total_pop=(sum(population, na.rm = TRUE)),
            .groups = "drop") %>% 
  mutate(variable_value_wt=total_variable/total_pop)

# check data set
nrow(data_equity_wt_county)
table(data_equity_wt_county$data_year)
```

This data sets should include records for:

* 4 geographies: 4 PSRC counties
* 6 equity focus groups (disability is not available in for 2010, but is for 2012; lep is not available in 2012)
* 5 equity quintiles 
* 3 years


## Combine region and county data
```{r}
data_tract <- as.data.frame(rbind(data_equity_wt_region,
                                  data_equity_wt_county)) %>% 
  dplyr::select(-total_variable, -total_pop, -tracts) %>% 
  filter(!is.na(quintile)) %>% 
  rename(variable_value=variable_value_wt,
         county=County.Name)

# check data set
str(data_tract)
```
This data set should include records for: 

* 5 geographies: 4 PSRC counties + region
* 3 years
* ~6 equity focus groups (depends on the years for disability an lep)
* 5 equity quintiles 

### Address "Low - Low Medium" quintile group
This code was added because of Kitsap/LEP data - there were too many tracts with 0, so we combined the bottom two quintiles into low - low medium. This change requires some additional processing so that all of the tracts (bottom 40%) in this combined quintile category are averaged and this average value is assigned to separate 'low' and 'low medium' quintile categories. 
```{r}
nrow(data_tract) #429
# unique(data_tract$quintile) #6 quintiles because of additional "Low - Low Medium"

low_quintile <- data_tract %>%
  filter(quintile=="Low - Low Medium") %>% 
  rename(quintile_original=quintile) %>%
  mutate(quintile="Low") %>%  # quintile label that matches rest of data
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications
lowmed_quintile <- data_tract %>% 
  filter(quintile=="Low - Low Medium") %>%
  rename(quintile_original=quintile) %>%
  mutate(quintile="Low Medium") %>% # quintile label that matches rest of data
  filter(county!="Region") # need to remove region from this because there should be enough tracts when all the tracts are combined to have separate low and low medium classifications

low_lowmed <- rbind(low_quintile,
                    lowmed_quintile)
# nrow(low_lowmed) #4

# gather counties that have low - medium categorization
low_lowmed_counties <- unique(low_lowmed$county) #Kitsap
low_lowmed_years <- unique(low_lowmed$data_year) #2017 and 2022
low_lowmed_equity <- unique(low_lowmed$equity_group) #lep_quintile

other_data <- data_tract %>% 
  filter(quintile!="Low - Low Medium") %>% 
  mutate(quintile_original=quintile)
# nrow(other_data) #425

data_tract <- rbind(low_lowmed, #4
                    other_data) #425

nrow(data_tract) #429 - should be fewer because of duplicates - low and low medium present in data set because of aggregation/calculations in previous steps - need to be removed and replaced with averaged values
table(data_tract$quintile_original, data_tract$county) # should not have extra in low and low medium quintile categories
# unique(data_tract$quintile) #5 quintiles

# remove rows where it is one of the "Low - Low Medium" quintiles AND the original quintile is "Low" or "Low Medium" - only want to keep the rows that are "Low"/"Low Medium" and they share a value
data_full <- data_tract %>% 
  filter(!(county %in% low_lowmed_counties & 
           data_year %in% low_lowmed_years &
           equity_group %in% low_lowmed_equity &
           (quintile_original=="Low" | 
             quintile_original=="Low Medium")))

# nrow(data_tract) #425
# table(data_tract$quintile, data_tract$county) #should not have extra in low and low medium quintile categories
```

## Rename and factor
```{r}
# wrap/order labels ----
# variable ordering and label function in equity-tracker-supplemental-script.R
data_clean <- transform_data_labels_tract(data_full)
```


# **CHART .rda**
Save final data set for charts (.rda)
```{r}
# save final data set as .rda
save(data_clean, file = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir, 
                                  "update",
                                  "rda-data",
                                  paste0(file_names$chart,'.rda')))
```

<a href="#top">Back to top of the page</a>