---
title: "Life Expectancy"
subtitle: "Visuals for Equity Tracker (tract data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'd-community-and-health',
                   ind_dir = 'd01-life-expectancy',
                   map = 'd01-life-expectancy-map-data',
                   chart = 'd01-life-expectancy-chart-data',
                   map_name = 'd01-life-expectancy-map.html',
                   column_chart_name = 'd01-life-expectancy-column',
                   line_chart_name = 'd01-life-expectancy-line')

# set most recent year
year <- "2022"
years <- c(2012, 2017, 2022)
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time. 
\
\
**If the indicator is available through a tract-level data set.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-tract-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Life expectancy provides a way to understand a population's overall health. Differences between groups of people can highlight differing access to healthcare and public services, adequate nutrition and living conditions, all of which impact a population's mortality rate.
[Please include the following for a general, layperson audience: 1-2 sentences explaining what your indicator is/measures, what it says about people's life outcomes; 1 sentence describing why it is useful/relevant in the indicator's theme]

## 1. Map of most recent data

### Create Visual
```{r map .rda}

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

To map data in this form, there should be a value corresponding to each census tract. Depending on the year or source of the data, this could be about 700 rows for data at the 2010 census tract resolution, or about 900 rows for data at the 2020 census tract resolution. 
```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# check data set
str(data_tract)
```


```{r tract data map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(data_tract$variable_value) # max is 95.39, so rounding to 100 makes sense
max_value <- 100 # this value will be the max value in the legend

# set up palettes
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
# tract_data_plus_zero<-append(data_tract$variable_value,0)
# 

# psrc_palette <- leaflet::colorNumeric(palette=psrc_purple_plus,
#                                       domain = data_tract$variable_value) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(data_tract$variable_value,na.rm = TRUE),max_value, NA)) # this is the value field

# set the variable
var_name <- "Life Expectancy"

labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>", data_tract$geoid20, # the code in <> makes the geoid value italic 
  paste("Tract value: ", prettyNum((round(data_tract$variable_value, digits=1))), " years"),
  paste("Region value: ", prettyNum((round(data_tract$reg_estimate, digits=1))), " years"),
  paste("County value: ", prettyNum((round(data_tract$cnty_estimate, digits=1))), 
        paste0("(",data_tract$county_name," County)"))) %>% 
  lapply(htmltools::HTML)

# if the label required a prefix, for example a $, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20, # the code in <> makes the geoid value italic
#   paste("$",
#         prettyNum((round(data_tract$variable_value, digits=-2)),
#                   big.mark = ","))) %>%
#   lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20,
#   paste(prettyNum((round(data_tract$variable_value, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$variable_value), #tract_data_plus_zero
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              # label = round(data_tract$variable_value, digits=1),
              label = labels) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(data_tract$variable_value,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```
Source(s): Washington State Department of Health, 2018-2022 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate tract map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(tract_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - tract, include=FALSE, eval=FALSE}
# simplify data set
data <- data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))
# 3 highest: 53053070316 (Auburn) Pierce (95.39), 53033004400 (Ravenna/U-District/Roosevelt) King (92.14), 53033032309 (Redmond) King (91.25)

sort <- data %>% 
  dplyr::arrange(estimate)
```

1. 78.8 years: The region's average life expectancy
2. 80.9 years: The region's median life expectancy
3. 28.1 years: The difference in the number of years between those living in the census tracts with the highest and lowest life expectancy

\

#### Insights & Analysis

* Snohomish County has the highest life expectancy (80.7 years), followed by King (79.4 years), Kitsap (78.9 years), and Pierce (75.7 years)
* The census tracts with the highest life expectancy are in Auburn, Pierce County (95.4 years); Ravenna/U-District neighborhoods (Seattle), King County (92.1 years); Redmond, King County (91.3 years)
* The census tracts with the lowest life expectancy are in the Lincoln International District (Tacoma), Pierce County (67.3 years); Lakewood, Pierce County (69.7 years); Tulalip Reservation, Snohomish County (70.1 years)

\
\

## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r load rda for chart, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - tract}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (y_min, y_max, dec, esttype, title, subtitle, source)
df = data_clean_column
geo = "county"
x = "quintile_ord"
y = "variable_value"
facet = "equity_group_ord"
fill = "quintile_ord"
y_min = 0 
y_max = 95 # will be based on indicator data set
dec = 1
esttype = "number"
color = "oranges"
num_colors = 5
color_rev = FALSE
title = "Life Expectancy by Community"
subtitle = "Number of years a newborn can be expected to live based on current conditions"
source = "Washington State Department of Health, 2016-2020 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, 2016-2020 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\d01-life-expectancy\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - tract, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(county=="Region",
         equity_group=="poc_quintile")

# region low quintile poc households: 81.19030	
# region high quintile poc households: 76.67422
# 81.19030-76.67422=4.51608

filter <- data_clean_column %>% 
  filter(quintile=="High",
         equity_group=="income_quintile")

# region high quintile low income households: 71.45515	
# pierce high quintile low income households: 63.50990
# 71.45515-63.50990=7.94525

filter <- data_clean_column %>% 
  filter(equity_group=="lep_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for lep: 79.04127-77.92853=1.11274
# king dif bt high/low quintiles for lep: 80.32595-77.63977=2.68618
# pierce dif bt high/low quintiles for lep: 78.46118-70.37597=8.08521
# snohomish dif bt high/low quintiles for lep: 80.35905-80.50632=-0.14727

filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for income: 81.98596-71.45515=10.53081
# king dif bt high/low quintiles for income: 81.94740-73.03439=8.91301
# kitsap dif bt high/low quintiles for income: 85.48813-69.90245=15.58568
# pierce dif bt high/low quintiles for income: 80.93128-63.50990=17.42138
# snohomish dif bt high/low quintiles for income: 82.09946-78.50544=3.59402

filter <- data_clean_column %>% 
  filter(county!="Region",
         equity_group=="youth_quintile",
         quintile=="Low"|
           quintile=="High")
# king dif bt high/low quintiles for income: 79.29009-76.35762=2.93247
# kitsap dif bt high/low quintiles for income: 59.86058-82.22206=-22.36148
# pierce dif bt high/low quintiles for income: 66.70910-75.97165=-9.26255
# snohomish dif bt high/low quintiles for income: 81.56795-80.16112=1.40683
```

1. 76.7 years: The regional life expectancy of people living in census tracts with the highest concentration of people of color, compared to 81.2 in census tracts with the lowest concentrations of people of color
2. 7.9 years: The life expectancy of those living census tracts with the highest concentration of low income households is lowest in Pierce County - a difference of 7.9 years when compared to the regional average
3. 7.0 years: Those living in census tracts with the highest concentration of limited English proficient households have a shorter life expectancy in Pierce County by 7.0 years, when compared to the regional average 

\

#### Insights & Analysis

* The smallest difference in life expectancy between census tracts with the highest and lowest concentrations of households with lower income is in Snohomish County (3.6 years), while the largest difference is in Pierce County (17.4 years)
* The life expectancy of those living in census tracts with the highest concentration of limited English proficient households is lower in all the counties except for Snohomish County  
* There is no noticeable relationship between life expectancy and living in census tracts with high and low concentrations of people with disabilities
* When comparing the life expectancy of those living in census tracts with the highest and lowest concentrations of households with children under 18 years old, life expectancy is slightly higher in King County (2.9 years) and Snohomish County (1.4 years) and much lower in Kitsap County (-22.4 years) and Pierce County (-9.3 years) 

\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r insert nulls for line chart, include=FALSE}

# insert null records ----
standard_base_year <- 2012 # earliest data year

## main table
df_base <- data_clean %>% 
  mutate(data_year_num = as.numeric(data_year))

## set-up columns to identify missing years by the unique categories/geography
d <- df_base %>% 
  group_by(county, equity_group, quintile, equity_group_ord, quintile_ord) %>% 
  summarise(avail_years = list(unique(as.numeric(data_year))),
            num_unique_years = length(unique(data_year)),
            start_year = as.numeric(min(unique(data_year))),
            end_year = as.numeric(max(unique(data_year))),
            .groups = "drop") %>% 
  mutate(base_year = standard_base_year)

## create list-column, a vector of missing years in a cell
d_calc <- d %>% 
  rowwise() %>% 
  mutate(missing_years = ifelse(start_year == base_year, NA, list(seq(base_year, (start_year-1))))) 

## create list-column, a vector of missing years in a cell
d_calc <- d %>% 
  rowwise() %>% 
  mutate(missing_years = list(setdiff(seq(base_year, end_year), avail_years)))

## unpack so each missing year is a row of its own
d_unnest <- d_calc %>% 
  unnest(missing_years) 

## create table with null records for chart
df_nulls <- d_unnest %>% 
  dplyr::select(county, equity_group, quintile, equity_group_ord, quintile_ord, data_year_num = missing_years) %>% 
  mutate(wt_Avg_Per = NA, estimate = NA, data_year = as.character(data_year_num))

## assemble main table  
df <- bind_rows(df_base, df_nulls) %>% 
  arrange(county, equity_group_ord, quintile_ord, data_year_num)

# filter for relevant years
df2 <- df %>% 
  filter(data_year_num %in% years)
```

```{r line facet - tract, include=FALSE}
# Create Facet Line Chart

# set variables
df = dplyr::filter(df2, data_year %in% years)
x = "data_year" # different from variables for facet column
fill = "quintile_ord" # different from variables for facet column
y_min = 50
y_max = 95
color = "blues"
title = "Life Expectancy Trend by Community"
subtitle = "Number of years a newborn can be expected to live based on current conditions, in 5-year spans between 2012 and 2022"
source = "Washington State Department of Health, 2008-2012, 2012-2017, 2018-2022 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, 2008-2012, 2012-2017, 2018-2022 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - tract, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2010" |
           data_year=="2020")

# 2010 region low quintile poc households: 80.54142	
# 2010 region high quintile poc households: 76.64042
# 80.54142-76.64042=3.901
# 2020 region low quintile poc households: 81.19030
# 2020 region high quintile poc households: 76.67422
# 81.19030-76.67422=4.51608
# ((4.51608-3.901)/(3.901))*100: 15.8% change

filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2010" |
           data_year=="2020")

# 2010 region low quintile income: 75.48994	
# 2010 region high quintile income: 72.81326
# 75.48994-72.81326=2.67668
# 2020 region low quintile income: 81.98596
# 2020 region high quintile income: 71.45515
# 81.98596-71.45515=10.53081
# ((10.53081-2.67668)/(2.67668))*100: 293.4281% change

filter <- data_clean %>% 
  filter(county=="Snohomish",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High")

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2010" |
           data_year=="2020")

# 2020 King dif low/high income: 81.94740-73.03439=8.91301
# 2010 King dif low/high income: 74.67479-75.27990=-0.60511
# 2020 Kitsap dif low/high income: 85.48813-69.90245=15.58568
# 2010 Kitsap dif low/high income: 78.56102-64.58841=13.97261
# 2020 Pierce dif low/high income: 80.93128-63.50990=17.42138
# 2010 Pierce dif low/high income: 76.74249-66.53859=10.2039
# 2020 Snohomish dif low/high income: 82.09946-78.50544=3.59402
# 2010 Snohomish dif low/high income: 76.01877-77.99171=-1.97294

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="lep_quintile",
         quintile=="High")

```

1. 3.9 years: The regional life expectancy gap between census tracts with the highest and lowest concentrations of people of color in 2010, a 15.8% smaller gap than in 2020
2. 293%: The increase in the regional life expectancy gap between census tracts with the highest and lowest concentrations of households below 200% of the poverty line between 2010 and 2020
3. 71.5 years: The 2020 regional life expectancy in tracts with the highest concentrations of households below 200% of the poverty line, 7 years fewer than the regional average (78.8) and 9 years fewer than the regional median (80.9)

\

#### Insights & Analysis

* Snohomish County is the only county in the region where the life expectancy of the census tracts with the highest concentration of people of color exceeds the life expectancy of the census tracts with the lowest concentration of people of color, a trend consistent over the past decade
* The life expectancy gaps between census tracts with the highest and lowest concentration of households below 200% of the poverty line increased in all counties between 2010 and 2020, with the largest gap change in Pierce County (7.2 years) and the smallest gap change in Kitsap County (1.6 years)
* The 2020 life expectancy of those living in the census tracts with the highest concentrations of households with limited English proficiency is shortest in Pierce County (70.4 years), compared to King County () and Snohomish County (80.5 years)

\
\

# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>