---
title: "Life Expectancy"
subtitle: "Visuals for Equity Tracker (tract data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'd-community-and-health',
                   ind_dir = 'd01-life-expectancy',
                   map = 'd01-life-expectancy-map-data',
                   chart = 'd01-life-expectancy-chart-data',
                   map_name = 'd01-life-expectancy-map.html',
                   column_chart_name = 'd01-life-expectancy-column',
                   line_chart_name = 'd01-life-expectancy-line')

# set most recent year
year <- "2022"
years <- c(2012, 2017, 2022)

# set past version year - this will set organize archived files in Y drive
past_year <- "2022"
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time. 
\
\
**If the indicator is available through a tract-level data set.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-tract-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
Life expectancy provides a way to understand a population's overall health. Differences between groups of people can highlight differing access to healthcare and public services, adequate nutrition and living conditions, all of which impact a population's mortality rate.
[Please include the following for a general, layperson audience: 1-2 sentences explaining what your indicator is/measures, what it says about people's life outcomes; 1 sentence describing why it is useful/relevant in the indicator's theme]

## 1. Map of most recent data

### Create Visual
```{r map .rda}

# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

To map data in this form, there should be a value corresponding to each census tract. Depending on the year or source of the data, this could be about 700 rows for data at the 2010 census tract resolution, or about 900 rows for data at the 2020 census tract resolution. 
```{r data format - map, include=FALSE}
# This space is for any data transformations that may be needed before the final visualizations are generated. This code may or may not be necessary depending on the way the data were formatted in the data-gen script and saved in the rda.

# check data set
str(data_tract)
```


```{r tract data map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(data_tract$variable_value) # max is 95.39, so rounding to 100 makes sense
max_value <- 100 # this value will be the max value in the legend

# set up palettes
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
# tract_data_plus_zero<-append(data_tract$variable_value,0)
# 

# psrc_palette <- leaflet::colorNumeric(palette=psrc_purple_plus,
#                                       domain = data_tract$variable_value) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(data_tract$variable_value,na.rm = TRUE),max_value, NA)) # this is the value field

# set the variable
var_name <- "Life Expectancy"

labels <- sprintf(
  "Census tract: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>", data_tract$geoid20, # the code in <> makes the geoid value italic 
  paste("Tract value: ", prettyNum((round(data_tract$variable_value, digits=1))), " years"),
  paste("Region value: ", prettyNum((round(data_tract$reg_estimate, digits=1))), " years"),
  paste("County value: ", prettyNum((round(data_tract$cnty_estimate, digits=1))), 
        paste0("(",data_tract$county_name," County)"))) %>% 
  lapply(htmltools::HTML)

# if the label required a prefix, for example a $, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20, # the code in <> makes the geoid value italic
#   paste("$",
#         prettyNum((round(data_tract$variable_value, digits=-2)),
#                   big.mark = ","))) %>%
#   lapply(htmltools::HTML)

# if the label required a suffix instead of a prefix, for example a %, the code would look like this:
# labels <- sprintf(
#   "Census tract: <em>%s</em><br/><strong>%s</strong>", data_tract$geoid20,
#   paste(prettyNum((round(data_tract$variable_value, digits=2)), #the digits would probably be 2 for %
#               big.mark = ","), #this wouldn't be necessary to include for %s    
#         '%')) %>%
#   lapply(htmltools::HTML)

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=data_tract,
              fillColor = psrc_palette(data_tract$variable_value), #tract_data_plus_zero
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              # label = round(data_tract$variable_value, digits=1),
              label = labels) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(data_tract$variable_value,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```
Source(s): Washington State Department of Health, 2018-2022 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate tract map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(tract_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

#### Data call outs
```{r map calculations - tract, include=FALSE, eval=FALSE}
# simplify data set
data <- data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(variable_value))
# 3 highest: 53053072801 (DuPont) Pierce (99.40), 53033007101 (Uptown Seattle) King (96.63), 53033022803 (Redmond) King (95.82)

sort <- data %>% 
  dplyr::arrange(variable_value)
# lowest by county: 53053061900 (South End, Tacoma) Pierce (68.43), 53033005303 (UW) King (68.51), 53033031101 (Auburn, south of SR18) King (71.65), 53061940002 (Tulalip Reservation) Snohomish (69.97), 53035081100 (Charleston, Bremerton) kitsap (73.16)

# difference between highest and lowest
99.40-68.43=30.97
```

1. 80 years: The region's average life expectancy
2. 78.6 years: The regional life expectancy of communities with high concentrations of people of color, compared to 80.8 in communities with low concentrations, a difference of 2.2 years
3. 30 years: The difference in the number of years between those living in the census tracts with the highest and lowest life expectancy

\

#### Insights & Analysis

* King County has the highest life expectancy (80.8 years), followed by Kitsap (80.6 years), Snohomish (79.8 years), and Pierce (78.2 years)
* Communities with the highest life expectancy are in DuPont, Pierce County (99.4 years); the Uptown neighborhood in Seattle, King County (96.6 years); Redmond, King County (95.8 years)
* Among the communities with the lowest life expectancy are the South End neighborhood in Tacoma, Pierce County (68.4 years); Auburn, King County (71.7 years); Tulalip Reservation, Snohomish County (70.0 years); the Charleston neighborhood in Bremerton, Kitsap County (73.2 years)
* There is a 2.6 year difference between life expectancy in King and Pierce counties, counties with the highest and lowest countywide averages, respectively.

\
\

## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r load rda for chart, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - tract}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year==year) #filter on most recent year

# set variables - adjust as needed (y_min, y_max, dec, esttype, title, subtitle, source)
df = data_clean_column
geo = "county"
x = "quintile_ord"
y = "variable_value"
facet = "equity_group_ord"
fill = "quintile_ord"
y_min = 50
y_max = 90 # will be based on indicator data set
dec = 1
esttype = "number"
color = "oranges"
num_colors = 5
color_rev = FALSE
title = "Life Expectancy by Community"
subtitle = "Number of years a newborn can be expected to live based on current conditions"
source = "Washington State Department of Health, 2016-2020 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, 2016-2020 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\d01-life-expectancy\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\d-regional-health-collaboration\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - tract, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information

# data call out (#2) from above 
filter <- data_clean_column %>% 
  filter(county=="Region",
         equity_group=="poc_quintile")

# region low quintile poc households: 80.75294	
# region high quintile poc households: 78.58048
# 80.75294-78.58048=2.17246

filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High",
         county!="Region")

# king low-high quintile low income households: 82.30115-	77.81201=4.48914
# kitsap low-high quintile low income households: 83.36568-78.39460=4.97108
# pierce low-high quintile low income households: 81.40780-75.06289=6.34491
# snohomish low-high quintile low income households: 81.37299-78.03709=3.3359

filter <- data_clean_column %>% 
  filter(equity_group=="lep_quintile",
         quintile=="Low"|
           quintile=="High",
         county!="Region")

# king dif bt high/low quintiles for lep: 82.61406-79.57669=3.03737
# kitsap dif bt high/low quintiles for lep: 80.75554-79.68579=1.06975
# pierce dif bt high/low quintiles for lep: 79.19723-76.27642=2.92081
# snohomish dif bt high/low quintiles for lep: 79.25054-78.99658=0.25396


filter <- data_clean_column %>% 
  filter(equity_group=="disability_quintile",
         quintile=="Low"|
           quintile=="High")
# region dif bt low/high quintiles for disability: 81.42043-77.88053=3.5399

filter <- data_clean_column %>% 
  filter(county!="Region",
         equity_group=="youth_quintile",
         quintile=="Low"|
           quintile=="High")
# king dif bt high/low quintiles for income: 80.94906-79.57012=1.37894
# kitsap dif bt high/low quintiles for income: 81.29869-80.12047=1.17822
# pierce dif bt high/low quintiles for income: 78.64674-78.50646=0.14028
# snohomish dif bt high/low quintiles for income: 81.13082-79.70847=1.42235


filter <- data_clean_column %>% 
  filter(equity_group=="income_quintile",
         quintile=="Low"|
           quintile=="High")

# region dif bt high/low quintiles for income: 81.99937-77.20026=4.79911
# king dif bt high/low quintiles for income: 82.30115-77.81201=4.48914
# kitsap dif bt high/low quintiles for income: 83.36568-78.39460=4.97108
# pierce dif bt high/low quintiles for income: 81.40780-75.06289=6.34491
# snohomish dif bt high/low quintiles for income: 81.37299-78.03709=3.3359
```
\

#### Insights & Analysis

* The smallest difference in life expectancy between communities with high and low concentrations of households with lower income is in Snohomish County (3.3 years), while the largest difference is in Pierce County (6.3 years).
* The life expectancy of communities with high concentrations of households with limited English proficiency is lower in all counties - ranging from Pierce (76.3 years), with the lowest life expectancy, to Kitsap (79.7 years), with the highest life expectancy.
* Regionally, communities with higher concentrations of people with disabilities have lower life expectancies than those without disabilities by 3.5 years.
* Communities with high concentrations of households with youth have life expectancies 1.4 years longer than households without youth in 3 of the counties: King, Kitsap, and Snohomish. The difference in Pierce County is almost negligible (0.1 year).
* The life expectancy of communities with high concentrations of households with lower income is shortest in Pierce County at 75.1 years- 2.1 fewer years than the regional average for communities with high concentrations of households with lower income (77.2 years).

\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r insert nulls for line chart, include=FALSE}

# insert null records ----
standard_base_year <- 2012 # earliest data year

## main table
df_base <- data_clean %>% 
  mutate(data_year_num = as.numeric(data_year))

## set-up columns to identify missing years by the unique categories/geography
d <- df_base %>% 
  group_by(county, equity_group, quintile, equity_group_ord, quintile_ord) %>% 
  summarise(avail_years = list(unique(as.numeric(data_year))),
            num_unique_years = length(unique(data_year)),
            start_year = as.numeric(min(unique(data_year))),
            end_year = as.numeric(max(unique(data_year))),
            .groups = "drop") %>% 
  mutate(base_year = standard_base_year)

## create list-column, a vector of missing years in a cell
d_calc <- d %>% 
  rowwise() %>% 
  mutate(missing_years = ifelse(start_year == base_year, NA, list(seq(base_year, (start_year-1))))) 

## create list-column, a vector of missing years in a cell
d_calc <- d %>% 
  rowwise() %>% 
  mutate(missing_years = list(setdiff(seq(base_year, end_year), avail_years)))

## unpack so each missing year is a row of its own
d_unnest <- d_calc %>% 
  unnest(missing_years) 

## create table with null records for chart
df_nulls <- d_unnest %>% 
  dplyr::select(county, equity_group, quintile, equity_group_ord, quintile_ord, data_year_num = missing_years) %>% 
  mutate(wt_Avg_Per = NA, estimate = NA, data_year = as.character(data_year_num))

## assemble main table  
df <- bind_rows(df_base, df_nulls) %>% 
  arrange(county, equity_group_ord, quintile_ord, data_year_num)

# filter for relevant years
df2 <- df %>% 
  filter(data_year_num %in% years)
```

```{r line facet - tract, include=FALSE}
# Create Facet Line Chart

# set variables
df = dplyr::filter(df2, data_year %in% years)
x = "data_year" # different from variables for facet column
fill = "quintile_ord" # different from variables for facet column
y_min = 50
y_max = 90
color = "blues"
title = "Life Expectancy Trend by Community"
subtitle = "Number of years a newborn can be expected to live based on current conditions, in 5-year spans between 2012 and 2022"
source = "Washington State Department of Health, 2008-2012, 2012-2017, 2018-2022 Washington Tracking Network, Life Expectancy at Birth; U.S. Census Bureau, 2008-2012, 2012-2017, 2018-2022 American Community Survey 5-Year Estimates, Tables B02001, C17002, B22010, B11005, B11007, C16002"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - tract, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\a-regional-health-collaboration\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - tract, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - tract, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2012" |
           data_year=="2022")

# 2010 region low quintile poc households: 80.54142	
# 2010 region high quintile poc households: 76.64042
# 80.54142-76.64042=3.901
# 2020 region low quintile poc households: 81.19030
# 2020 region high quintile poc households: 76.67422
# 81.19030-76.67422=4.51608
# ((4.51608-3.901)/(3.901))*100: 15.8% change

filter <- data_clean %>% 
  filter(county=="Snohomish",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2012" |
           data_year=="2022")

# 2012 king low-high quintile income: 79.07988-74.11040=4.96948	
# 2022 king low-high quintile income: 82.30115-77.81201=4.48914
# 4.96948-4.48914=0.48034
# 2012 kitsap low-high quintile income: 78.34657-63.55216=14.79441
# 2022 kitsap low-high quintile income: 83.36568-78.39460=4.97108
# 14.79441-4.97108=9.82333
# 2012 pierce low-high quintile income: 79.83607-68.12831=11.70776
# 2022 pierce low-high quintile income: 81.40780-75.06289=6.34491
# 11.70776-6.34491=5.36286
# 2012 snohomish low-high quintile income: 79.68945-78.47776=1.21169
# 2022 snohomish low-high quintile income: 81.37299-78.03709=3.3359
# 1.21169-3.3359=-2.12421

filter <- data_clean %>% 
  filter(county!="Region",
         equity_group=="lep_quintile", 
         quintile=="High",
         data_year=="2022")


filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="poc_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2012" |
           data_year=="2022")
# 2012 low-high poc: 80.25792-77.01544=3.24248
# 2022 low-high poc: 80.75294-78.58048=2.17246
# ((2.17246-3.24248)/3.24248)*100=-33%

filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="income_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2012" |
           data_year=="2022")

# 2012 low-high low income: 79.28106-72.72990=6.55116
# 2022 low-high low income: 81.99937-77.20026=4.79911
# ((4.79911-6.55116)/6.55116)*100=-26.74412%

filter <- data_clean %>% 
  filter(county=="Region",
         equity_group=="disability_quintile",
         quintile=="Low" | 
           quintile=="High",
         data_year=="2012" |
           data_year=="2022")

# 2012 low-high dis: 76.94577-77.51709=-0.57132
# 2022 low-high dis: 81.42043-77.88053=3.5399
# ((3.5399--0.57132)/0.57132)*100=719.6%
```
\

#### Insights & Analysis

* Snohomish County is the only county in the region where the life expectancy of communities with high concentrations of people of color is similar to or slightly higher than the life expectancy of communities with low concentrations of people of color, a trend consistent over most of the past decade.
* The life expectancy gaps between areas with high and low concentrations of households with lower income decreased in King (-0.5 years), Kitsap (-9.8 years), and Pierce (-5.4 years) counties between 2012 and 2022. In Snohomish County the gap increased from 1.2 years in 2012 to 3.3 years in 2022.
* The 2022 life expectancy of communities with high concentrations of households with limited English proficiency is shortest in Pierce County (76.3 years), compared to Kitsap (79.7 years), King (79.6 years), and Snohomish (79 years).
* The regional life expectancy gap is 2.2 years between communities with high and low concentrations of people of color in 2022, a 33% smaller gap than in 2012.
* The regional life expectancy gap decreased 27% between areas with high and low concentrations of households with lower income between 2012 and 2022.
*  In 2022, communities with high concentrations of individuals with a disability had a lower life expectancy (77.9 years) than communities with low concentrations of individuals without a disability (81.4 years). 

\
\

# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. *Always run this chunk of code. If this is the first time you are generating visuals for an indicator, comment out the 'update' part of the 'y.drive.folder.update' variable because this is the first version that is being created.*
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update' #comment out if this is the first time you are generating visuals
                                   )

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r to archive, include=FALSE, eval=FALSE}
# add previous year folder to archive folder - to retain and organize all archived files
dir.create(file.path(file_names$base_dir,
                     file_names$theme_dir,
                     file_names$ind_dir,'archive', past_year))

# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive', past_year)

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
This step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall). *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. *Only run this chunk of code if updating visuals for indicator - if there is already existing visuals on a webpage. Don't run if this is the first time visuals are being created for indicator.*
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage. *Always run this chunk of code.*
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>