---
title: "Kindergarten Readiness" 
subtitle: "Data Gen: Exploring, Cleaning, Transforming (PUMS/OSPI data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

# Organize workspace
```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r library setup, include=FALSE}
# install.packages("devtools")
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)

library(tidyverse)
library(psrcelmer)
library(psrccensus)
library(psrcplot)
library(rlang) #required for psrccensus
library(emmeans) #required for rlang
library(magrittr)

library(vtable) #summary stats
library(forcats) #for factor re-leveling

library(odbc) #connect to ElmerGeo
library(DBI) #connect to ElmerGeo
library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html
library(ggspatial)
library(lubridate) #year formatting
library(readxl)

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(here)
```

```{r sources}
# https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend - this code helps to set up the map legend so that it is arranged high-low with correct color order

# reference supplemental script with supporting settings/functions
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))
```

```{r variables setup}

file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'f-economy',
                   ind_dir = 'f02-kindergarten-readiness',
                   chart = 'f02-kindergarten-readiness-chart-data',
                   map = 'f02-kindergarten-readiness-map-data')
```

**Median Income**

# Download data 
## Primary data 
The data set referenced in this script is generated from the American Community Survey (ACS) and the Washington Office of Superintendent of Public Instruction [(OSPI)](https://www.k12.wa.us/data-reporting/data-portal). These data sets provide data at the **person- or household-level**, with the ability to look at the different indicators by the FOUR equity demographic groups of interest. 
```{r, include=FALSE}
# DATA IS AVAILABLE FOR EVERY SCHOOL YEAR, BACK TO 2011

pums_ospi_year <- "2022"
years <- c("2011", "2012", "2013", "2014",
           "2015", "2016", "2017", "2018",
           "2019", #"2020", #2020 data not available from OSPI because of COVID
           "2021", "2022")

# PUMS/OSPI data set
pums_ospi_elmer <- get_table(schema = "equity", tbl_name = "v_tracker_indicators")
```

```{r, include=FALSE}
# PUMS/OSPI data set
pums_ospi_elmer <- get_table(schema="equity",
                             tbl_name="v_tracker_indicators")
```
\
This data set was compiled from PUMS data.
\
Looking at the fields in the data set:
```{r}
# explore the data
head(pums_ospi_elmer)

# view the equity demographic groups, different indicators, and data years
pums_ospi_cols <- c('focus_type', 'indicator_type', 'data_year') 
walk(pums_ospi_cols, ~print(unique(pums_ospi_elmer[[.x]])))
```

```{r indicator specific information, include=FALSE}
# set indicator information
indicator_measure <- 'Kindergarten readiness'
indicator_title <- 'Kindergarten readiness'
```


## 1. Explore data
*In this section we make sure that the data set has the correct dimensions - year, geography, equity focus group, etc.*  
```{r}
# clean data set
data_full <- pums_ospi_elmer %>% 
  # distinct(indicator_fact_id, .keep_all = TRUE) %>% #this may be necessary depending on indicator
  dplyr::filter(indicator_type==indicator_measure) %>% 
  dplyr::filter(focus_type!="Total") %>% 
  filter(focus_attribute!="Total") %>%
  filter(indicator_attribute!="Total") %>%
  dplyr::mutate(data_year_yr=format(data_year,format="%Y"))
```

### Data fields
#### *consistent base data*

* There should be 5 geographies - the 4 counties and the Region. 
* There should be 4 equity focus categories - POC, income, disability, and LEP
    + 2 sub-groups in each of the 4 equity focus group (e.g. people of color, non-people of color)

*THE WARNING MESSAGES DON'T APPLY BECAUSE 'YOUTH' AND 'OLDER ADULTS' AREN'T INCLUDED IN THIS DATASET. ALSO, THERE ARE DATA FOR EVERY SCHOOL YEAR.*
```{r, include=FALSE}
# geographies
if(length(unique(data_full$county)) < 4){
  warning("There is a missing geography. Please check.")
} else if (length(unique(data_full$county)) < 5) {
  warning("There is a missing geography. Please check.")
} else {
  message("All geographies are included: King, Kitsap, Pierce, Snohomish, and Region.")
}

# equity focus groups
if(length(unique(data_full$focus_type)) < 5){
  warning("There are missing equity groups. Please check.")
} else if (length(unique(data_full$focus_type)) < 6) {
  warning("There is a missing equity group. Please check.")
} else {
  message("All 6 equity groups are included.")
}

# equity focus attributes
if(length(unique(data_full$focus_attribute)) < 11){
  warning("There are missing equity sub-groups. Please check.")
} else if (length(unique(data_full$focus_attribute)) < 12) {
  warning("There is a missing equity sub-groups. Please check.")
} else {
  message("All 12 equity sub-groups are included.")
}
```

#### *indicator-specific data*
These fields will vary by indicator:

* Type of metric - this will determine how the data are visualized *(est ="percent" or "currency" or "number")*
* Number of years (5-year span) - this can vary depending on data availability
* Number of indicator-specific categories - this can vary depending on the indicator of interest, ranging from N/A (*median income*) to multiple levels (*crowding, housing cost burden*)
```{r}
# metric
data_type <- unique(data_full$fact_type) #median
message("This indicator's data type/format: ", data_type)

# years
num_yr <- length(unique(data_full$data_year)) #3
range <- list(paste(as.character(unique(data_full$data_year)),collapse=", ",sep=""))
message("There are ", num_yr, " available years: ", range)

# indicator attributes
num_indatt <- length(unique(data_full$indicator_attribute)) #1
attributes_list <- list(paste(unique(data_full$indicator_attribute), collapse=", ", sep=""))
if(length(unique(data_full$indicator_attribute)) > 1){
  warning(paste("There are",num_indatt, "indicator attributes: ",
              attributes_list, ". Please choose 1."))
} else {
  message("There is only one indicator attribute: ", attributes_list,
              ". No need to filter.")
}
```

```{r}
# space to filter if necessary
# data_full <- data_full %>%
#   filter(indicator_attribute=="Yes")
```

*THE WARNING MESSAGES DON'T APPLY BECAUSE 'YOUTH' AND 'OLDER ADULTS' AREN'T INCLUDED IN THIS DATASET. ALSO, THERE ARE DATA FOR EVERY SCHOOL YEAR.*
```{r}
# calculate expected number of rows
num_row <- num_yr*5*6*2 #720
message("There should be data for each year (",num_yr,"), geography (5), equity focus group (6), and equity sub-group (2) = ",num_row ," rows")

if(nrow(data_full) < num_row){
  warning(paste("There are", nrow(data_full), "rows, which is fewer than expected. Please check."))
} else if (nrow(data_full )> num_row){
  warning(paste("There are", nrow(data_full), "rows, which is more than expected. Please check and filter as necessary."))
} else {
  message(paste("There are the expected number of rows in the dataset:", nrow(data_full)))
}
```

```{r check data elements}
# extra detail: if there are any missing elements, can run this code
check_data_fields <- function(column) {
  l <- length(unique(column))
  v <- unique(column)
  return(list(length = l, values = v))
}

data_fields <- c("num_county" = "county", # geographies
                 "num_group" = "focus_type", # equity focus groups
                 "metric" = "fact_type", 
                 "num_yr" = "data_year", 
                 "num_sub_group" = "focus_attribute") #equity sub-focus groups

data_fields_summary <- map(data_fields, ~check_data_fields(data_full[[.x]]))

data_fields_summary


# calculate expected number of rows
num_row <- data_fields_summary |> 
  map(~pluck(.x, "length")) |> # isolate nums
  discard_at("metric") |> # remove metric element
  reduce(`*`) * 2 # multiply all numbers and by 2 for subgroups

#180
```

#### *checking for missing data*
*THE WARNING MESSAGES DON'T APPLY BECAUSE 'YOUTH' AND 'OLDER ADULTS' AREN'T INCLUDED IN THIS DATASET. ALSO, THERE ARE DATA FOR EVERY SCHOOL YEAR.*
##### Year / geography
```{r}
# Function in equity-tracker-supplemental-script.R
num_yr_geo <- check_missing_data(vars = c("num_yr", "num_county"),
                                 multiply_by_subgroups = FALSE)
#15
```

```{r}
table(data_full$data_year,
      data_full$county)
```


##### Year / equity focus group
```{r}
num_yr_grp <- check_missing_data(vars = c("num_yr", "num_group"),
                                 multiply_by_subgroups = FALSE)
#18
```

```{r}
table(data_full$data_year,
      data_full$focus_type)
```


##### Year / equity focus sub-group
```{r}
num_yr_subgrp <- check_missing_data(vars = c("num_yr", "num_sub_group"),
                                    multiply_by_subgroups = FALSE)
#96
```

```{r}
table(data_full$data_year,
      data_full$focus_attribute)
```
\
\

### Numeric data
To check for *0*s and *NULL*s
```{r}
# checking for NULLs
summary(data_full$fact_value)
```
\
\

To look at distribution of all data - not the most useful visual, but provides a sense of the range of values at a high level in one plot. 
\
```{r}
qplot(data_year_yr, fact_value, 
      colour = county, shape = focus_type, 
      data = data_full, 
      main = indicator_title) + theme(axis.title.x = element_blank(),
                                               axis.title.y = element_blank())
```
\
This table includes a lot of information about the data set and helps to show the different levels of each field. It provides another way to check if data are available for all counties and all years, or where there may be gaps in the data set. 
```{r}
data_refined <- data_full |> 
  select(data_year_yr, county, vulnerability, focus_type, focus_attribute,
        indicator_type, indicator_attribute, fact_value)

sumtable(data_refined,
         add.median = TRUE,
         group = "county",
         out = "return")
```

```{r}
# wrap/order labels ----
# variable ordering and label function in equity-tracker-supplemental-script.R
data_clean <- transform_data_labels(data_full)
```

```{r}
# indicator specific re-ordering - this step is unnecessary for SNAP and likely for most indicators since we are focused on one aspect of the variable
# data_clean$indicator_attribute <- factor(data_clean$indicator_attribute,
#                                     levels = c("Less than a Bachelor's degree",
#                                                "Bachelor's degree or higher"))
```
\

### Data labels, shares
These charts were generated to ensure the labels across years are consistent/make sense. There had been an issue with misassigned labels because tidycensus::pums_variables, i.e. the only digital data dictionary available to associate labels with codes, exists only from 2017 forward. Most variables have had consistent codes, but in cases where the codes have shifted over time, using the 2017 lookup winds up mischaracterizing categories.
\
\
These charts also help to confirm that the shares add up to 100% - only relevant when *indicator_attribute* has more than one category. The *indicator_attribute* for median household income is NA.
\
\
*The colors of the charts may not be consistent between the years depending on missing data.* 
\
```{r}
check_region <- function(df, year) {
  df |> 
    filter(county == "Region") |> 
    filter(data_year == year) |> 
    arrange(focus_attribute_ord)
}

check_region_facet <- function(df, year) {
  static_facet_column_chart(t = df,
                            x = "data_year", 
                            y = "fact_value",
                            fill = "vulnerability", 
                            facet = "focus_type_ord",
                            est = "percent",
                            scales = "fixed",
                            color = "psrc_pairs", 
                            ncol = 3,
                            title = paste0(indicator_title, ": Region (", year, ")"),
                            subtitle = paste(str_to_title(str_replace_all(indicator_measure, "_", " ")), "by Equity Focus Group"),
                            source = "U.S. Census Bureau, American Community Survey (ACS) 5-Year Public Use Microdata Sample (PUMS)")
}

# the years below reflect the years that are available from unique(data_year)
just_region_dfs <- map(years, ~check_region(df = data_clean, 
                                            year = .x)) |> 
  set_names(paste0('r', years))

# check est type - make sure it corresponds to your indicator (percent, currency, or number)
check_region_facets <- map2(just_region_dfs, 
                            years, 
                            ~check_region_facet(df = .x, year = .y))

walk(check_region_facets, ~print(.x))

```

## 2. Visually explore data
### Scatter plots
*In this section we start to explore the data visually - distribution by the different dimensions within the data set. These plots are helpful to check for outliers and get a higher level understanding of the data in one visual, before slicing the data by geography and equity focus group in the following sections.* 
\
\
The following code will need to be adjusted to fit the fields specific to the data indicator. For educational attainment, we focus on those with a Bachelor's degree or higher. The following code establishes the data frame that the rest of the analysis uses. If there are fewer than 2 *indicator attributes*, this section can be skipped/commented out, but the code will need to be adjusted throughout. 
\
\

#### By *indicator_attribute*
This section isn't relevant for this specific indicator because there aren't unique indicator attributes. 

#### By Year
```{r}
# separating by year ----

data_clean_dfs <- map(years, ~data_clean |> filter(data_year == .x))
```

```{r}
# plot data, adjust y limits based on data set
data_clean_plot <- function(df, year) {
  qplot(focus_type_ord, 
        fact_value, 
        colour = county, 
        shape = vulnerability, 
        data = df, 
        main = paste0(indicator_title, ": ", year)) + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()#,
          # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
          ) + 
    ylim(0, (max(data_clean$fact_value)*1.5))
}

data_clean_plots <- map2(data_clean_dfs, years, ~data_clean_plot(.x, .y))

walk(data_clean_plots, ~print(.x))
```


# **CHART .rda**
```{r, include=FALSE}
head(data_clean) # make sure this is the correct data set

# make any additional edits 
data_clean <- data_clean %>% 
  mutate(fact_value_notrounded=fact_value) %>% #create column to retain full median income values
  mutate(fact_value=round(fact_value, digits=2)) #round median income value to hundreds and set retain consistent name - if you are not working with $ you will need to change this to the correct number of digits you would like to round to (for example, if using decimals, you may want to set digits=2 or if you using whole numbers you may want to set digits=0)

data_clean #check to make sure the rounding worked for your indicator - if not will need to go back to generate data set in chunk around line 245
```
Save final data set (.rda) for charts
```{r}
# save final data set as .rda
save(data_clean, file = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir, 
                                  "update",
                                  "rda-data",
                                  paste0(file_names$chart,'.rda')))
```


## Create data set for map .rda 
### Additional data
#### Tract spatial file
for mapping most recent data
```{r census tract spatial data - for tract data, include=FALSE}
# Connecting to ElmerGeo for census geographies through Portal----
arc_service <- "https://services6.arcgis.com/GWxg6t7KXELn1thE/arcgis/rest/services"

# tracts20.url <- file.path(arc_service, "Census_Tracts_2020/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
# tracts10.url <- file.path(arc_service, "Census_Tracts_2010/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
# 
# tracts20.lyr <- st_read(tracts20.url)
# tracts10.lyr <- st_read(tracts10.url)
# 
# nrow(tracts20.lyr) #919
# nrow(tracts10.lyr) #773

# school district spatial file (without water)
sch_districts.url <- file.path(arc_service,
                               "School_Districts/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")
sch_districts <- st_read(sch_districts.url)
```

#### Corresponding ACS table
This step is not required because the data is available by school district, which is a unique geographic boundary

#### Most recent OSPI data by school district
Will need to download the corresponding report card from OSPI and save to Y drive indicator/raw-data folder.
https://ospi.k12.wa.us/data-reporting/data-portal
```{r}
# most recent KG readiness data by school district
kgreadiness_schdistrict <- read.csv(file.path(file_names$base_dir,
                                              file_names$theme_dir,
                                              file_names$ind_dir, "raw-data",
                                              'Report_Card_WaKids_2022-23_School_Year.csv'))

# edit data to more usable form
kg_schdistrict <- kgreadiness_schdistrict %>% 
  dplyr::filter(OrganizationLevel=="District",
                StudentGroupType=="All",
                Domain=="NumberofDomains", # removes Skykomish
                MeasureValue=="6") %>%
  mutate(Percent_num=Percent*100) %>% 
  arrange(DistrictName)
```

### Joining data
```{r}
# merge to school districts spatial file to ospi data
ospi_data_tract <- merge(sch_districts, kg_schdistrict,
                         by.x = "organization",
                         by.y = "OrganizationId",
                         all.x=TRUE)

# aggregating data by county
county_data <- ospi_data_tract %>%
  sf::st_drop_geometry() %>% 
  dplyr::select(county, Numerator, Denominator) %>% 
  dplyr::group_by(county) %>% 
  dplyr::summarise(cnty_num=sum(Numerator, na.rm=TRUE),
                   cnty_denom=sum(Denominator, na.rm=TRUE),
                   cnty_percent_num=(cnty_num/cnty_denom)*100) %>% #to distinguish between tract and county values
  dplyr::select(-cnty_num, -cnty_denom)

# aggregating data by region
reg_data <- ospi_data_tract %>%
  sf::st_drop_geometry() %>% 
  dplyr::select(OrganizationLevel, Numerator, Denominator) %>% 
  dplyr::group_by(OrganizationLevel) %>% 
  dplyr::summarise(reg_num=sum(Numerator, na.rm=TRUE),
                   reg_denom=sum(Denominator, na.rm=TRUE),
                   reg_percent_num=(reg_num/reg_denom)*100) %>% #to distinguish between county, tract, and region values
  na.omit() #remove NA row
 

# add region to county data
county_reg_data <- county_data %>% 
  mutate(reg_percent_num=reg_data$reg_percent_num) #add column with region value

# merge tract, county, and region data into one data set 
ospi_data_tract <- merge(ospi_data_tract, county_reg_data,
                         by.x="county",
                         by.y="county",
                         all.x=TRUE)
```

```{r}
# check NAs
summary(ospi_data_tract$Percent) #2 NAs

ospi_data_tract_na <- ospi_data_tract %>% 
  filter(is.na(Percent))

# Index data are suppressed 
# Skykomish data are suppressed 
```

### Explore map/acs data
```{r explore map}
# map settings and legend function in equity-tracker-supplemental-script.R

# set up palettes
psrc_palette <- leaflet::colorNumeric(palette=psrc_colors$purples_inc,
                                      domain = ospi_data_tract$Percent_num)

# set the variable
var_name <- "Kindergarten <br> Readiness"

# map settings
tract_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=ospi_data_tract,
              fillColor = psrc_palette(ospi_data_tract$Percent_num),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = round(ospi_data_tract$Percent_num, digits=1)) %>%

  # legends
  addLegend_decreasing(pal = psrc_palette,
                       values = ospi_data_tract$Percent_num,
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       decreasing = TRUE,
                       labFormat = labelFormat()) %>% 
  
  #set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)   
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fix the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML
tract_map %<>% htmlwidgets::prependContent(html_fix)

# print map
tract_map
```

# **MAP .rda**
Save final data set (.rda) for map
```{r}
# This is the data set used to create the map of the most recent data. If there are any edits they should be made before saving it as an .rda. If not, you can skip this code chunk.

# save final data set as .rda
save(ospi_data_tract, file = file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       "update",
                                       "rda-data",
                                       paste0(file_names$map,'.rda')))
```

<a href="#top">Back to top of the page</a>