---
title: "Kindergarten Readiness"
subtitle: "Visuals for Equity Tracker (OSPI data)"
author: "Mary Richards"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
---

```{r rmarkdown setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting

# to reference chart functions in Github
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r library setup, include=FALSE}
# devtools::install_github("psrc/psrcplot",
#                          force=TRUE)
library(psrcplot)
library(tidyverse)
library(magrittr)

library(sf)
library(leaflet)
library(leafem) #home button
library(htmlwidgets) #save visuals as html

install_psrc_fonts()
library(showtext) #trying to fix PSRC font issues
library(sysfonts) #required for showtext
library(showtextdb) #required for showtext

library(echarts4r)
library(webshot2) #static image creation
library(here)
```

```{r, choose output type}
# output_type <- "word"
output_type <- "html"
```

```{r variables setup}
file_names <- list(base_dir = 'Y:/Equity Indicators/tracker-webpage-content',
                   theme_dir = 'f-economy',
                   ind_dir = 'f02-kindergarten-readiness',
                   map = 'f02-kindergarten-readiness-map-data',
                   chart = 'f02-kindergarten-readiness-chart-data',
                   map_name = 'f02-kindergarten-readiness-map.html',
                   column_chart_name = 'f02-kindergarten-readiness-column',
                   line_chart_name = 'f02-kindergarten-readiness-line')
```

# 3 visuals for webpage
This code will help produce the three visuals that are going to be a part of each equity tracker indicator webpage: regional map (tract level) of most recent data, chart of the most recent data, chart of trends over time.
\
\
**If the indicator is a PUMS/OPSI indicator that can be accessed through Elmer.** Getting the data to a workable version may require some data transformation. To explore, clean, transform, and generate a final data set, please use the *data-gen-pums-template*. This script will generate an .rda for the map and an .rda for the charts. These data sets will be loaded in before the data visualization code.  

## Indicator Explanation
[Please include the following for a general, layperson audience: 1-2 sentences explaining what your indicator is/measures, what it says about people's life outcomes; 1 sentence describing why it is useful/relevant in the indicator's theme]

## 1. Map of most recent data
To map data in this form, it requires accessing data at the regional/tract level from ACS since the Elmer data set is already aggregated to equity group/quintile. 

### Create Visual
```{r map .rda}
# load map .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$map,'.rda')))
``` 

```{r acs map}
# reference addLegend_decreasing function
source(here::here('data-visualization/equity-tracker-supplemental-script.R'))

# check maximum value of data
# summary(ospi_data_tract$Percent_num)
max_value <- 90 # this value will be the max value in the legend

# setting color palette
psrc_purple_plus<-c("#FFFFFF", "#FFFFFF", "#F6CEFC", psrc_colors$purples_inc)
ospi_data_plus_zero<-append(ospi_data_tract$Percent_num,0)

psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = ospi_data_plus_zero) # this is the value field
psrc_palette <- leaflet::colorNumeric(palette = psrc_purple_plus,
                                      domain = c(min(ospi_data_plus_zero,na.rm = TRUE),max_value, NA)) # this is the value field

# setting the legend/label variables
var_name <- "Kindergarten <br> Readiness" # the <br> could be adjusted depending on the length/spacing of the name
labels <- sprintf(
  "School District: <em>%s</em><br/><strong>%s</strong><br/><body>%s</body><br/><body>%s</bodyD>",
  ospi_data_tract$lea_name, # the code in <> makes the district italic
  paste("School district value: ", prettyNum(round(ospi_data_tract$Percent_num, digits=0)),'%'),
  paste("Region value: ", prettyNum((round(ospi_data_tract$reg_percent_num, digits=0))), '%'),
  paste("County value: ", prettyNum(round(ospi_data_tract$cnty_percent_num, digits=0)), '%', 
        paste0("(", ospi_data_tract$county," County)"))) %>%
  lapply(htmltools::HTML)

# map settings
ospi_map <- leaflet() %>%
  leaflet::addMapPane(name = "polygons", zIndex = 410) %>%
  leaflet::addMapPane(name = "maplabels", zIndex = 500) %>% # higher zIndex rendered on top
  leaflet::addProviderTiles("CartoDB.VoyagerNoLabels") %>%
  leaflet::addProviderTiles("CartoDB.VoyagerOnlyLabels",
                            options = leaflet::leafletOptions(pane = "maplabels"),
                            group = "Labels") %>%
  addPolygons(data=ospi_data_tract,
              fillColor = psrc_palette(ospi_data_plus_zero),
              stroke=FALSE, 
              smoothFactor = 0.2,
              fillOpacity = 0.7,
              group = var_name,
              label = labels) %>% 

  # legend
  addLegend_decreasing(pal = psrc_palette,
                       values = c(min(ospi_data_plus_zero,na.rm = TRUE),max_value, NA),
                       position = "bottomright",
                       title = var_name,
                       group = var_name,
                       opacity = 0.7,
                       labFormat = labelFormat(suffix = "%"),
                       decreasing = TRUE) %>% #to get legend high-low with correct color order
  
  # set view extent
  leaflet::setView(lng=map.lon, lat=map.lat, zoom=map.zoom) %>% 
  addEasyButton(easyButton(
    icon = htmltools::span(class = "globe", htmltools::HTML("&#127758;")),  #&#127760; (another emoji option) #"fa-globe", (font awesome icon no longer works because of the conversion to Poppins font below)  
    title ="Region",
    onClick=JS("function(btn, map){map.setView([47.615,-122.257],8.5); }")))

# fixing the legend NA placement (https://github.com/rstudio/leaflet/issues/615)
css_fix <- "div.info.legend.leaflet-control br {clear: both;} html * {font-family: Poppins !important;}" # CSS to correct spacing and font family
html_fix <- htmltools::tags$style(type = "text/css", css_fix) # Convert CSS to HTML
ospi_map %<>% htmlwidgets::prependContent(html_fix) # Insert into leaflet HTML code

# printing map
ospi_map
```
Source(s): U.S. Census Bureau, 2018-2022 American Community Survey 5-Year Estimates, Table S2201; U.S. Census Bureau, Geography Division 2020 TIGER/Line Shapefiles
```{r generate acs map html, include=FALSE}
# to save html visual to folder, you must create an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
  
# interactive webpage output process
htmlwidgets::saveWidget(ospi_map,
                        file=file.path(file_names$base_dir,
                                       file_names$theme_dir,
                                       file_names$ind_dir, 
                                       'update/webpage-html-outputs', 
                                       file_names$map_name))
```
\
\

```{r map calculations - pums, include=FALSE, eval=FALSE}
# simplify data set
data <- ospi_data_tract %>%
  st_drop_geometry()

# calculate call out information
sort <- data %>% 
  dplyr::arrange(desc(estimate))
# Highest percentage of SNAP participation is 53053940006 (east Tacoma where two low income housing complexes are - 49.4%), 53053061601 (downtown Tacoma - 44.1%), 53033029004 (southeast Des Moines - 39.3%), 53033029604 (southeast Kent - 35.8%), 53053063301 (east Tacoma just south of 53053940006 - 35.0%)

# national average
# acs table: S2201, https://data.census.gov/table/ACSST5Y2022.S2201?q=S2201

sort <- data %>% 
  dplyr::arrange(estimate)

filter <- data %>% 
  filter(estimate==0)
# There are 38 census tracts with a 0% participation. 
table(filter$county_name)
# 30 tracts in King County, 2 in Kitsap, 2 in Pierce, and 4 in Snohomish 
```

1. 53%: Regional share of students ready for kindergarten
2. 40%: Percentage point difference between the school district with the most kindergarten ready students and the one with the least
3. 47%: Students with disabilities in the region are less likely to be prepared for kindergarten compared to students without disabilities, a gap which has increased by 47% over time

\

#### Insights & Analysis

* King County has the highest share of students who are kindergarten ready (59%), followed by Kitsap (52%), Pierce (47%), and Snohomish (46%).
* Of the 37,800 kindergarten students in the four-county region, almost half of them are in King County (49%), one-quarter are in Pierce (25%), one-fifth are in Snohomish (20%), and the remaining are in Kitsap (6%).
* Snoqualmie Valley School District (King County) has the highest share of students ready for kindergarten (81%).
* Six school districts straddle two counties:
    * Three are entirely in the PSRC region: Auburn (King/Pierce), Fife (King/Pierce), and Northshore (King/Snohomish).
    * Three extend outside the region: Darrington (Skagit/Snohomish), Eatonville (Lewis/Pierce), and Stanwood-Camano (Island/Snohomish).
* Index (Snohomish County) and Skykomish (King County) school districts are the only two with suppressed data because of small sample sizes.

\
\


## 2. Facet of most recent data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r data setup, include=FALSE}
# load visualization functions for charts
source(here::here('data-visualization/equity-tracker-chart-functions.R'))

# load chart .rda
load(file=file.path(file_names$base_dir,
                    file_names$theme_dir,
                    file_names$ind_dir,
                    "update",
                    "rda-data",
                    paste0(file_names$chart,'.rda')))
```

```{r column facet - pums}
# Create Facet Column Chart
data_clean_column <- data_clean %>% 
  dplyr::filter(data_year=="2023") #filter on most recent year

# set variables - adjust as needed (esp. y_min, y_max, dec, esttype, etc.)
df = data_clean_column
geo = "county"
x = "focus_attribute_ord"
y = "fact_value"
facet = "focus_type_ord"
fill = "focus_attribute_ord"
y_min = 0
y_max = 1
dec = 0
esttype = "percent"
color = "oranges"
num_colors = 2
color_rev = TRUE
title = "Kindergarten Readiness by Community"
subtitle = "Percent of students demonstrating readiness in 6 of 6 domains*"
source = "Washington Office of Superintendent of Public Instruction, Report Card WaKids 2022-23 School Year, Washington Kindergarten Inventory of Developing Skills <br> *Social-Emotional, Physical, Language, Cognitive, Literacy, and Math"

column_chart <- equity_tracker_column_facet(df = df,
                                            geo = geo,
                                            x = x,
                                            y = y,
                                            facet = facet,
                                            title = title,
                                            y_min = y_min,
                                            y_max = y_max,
                                            dec = dec,
                                            esttype = esttype,
                                            color = color,
                                            num_colors = num_colors,
                                            color_rev = color_rev)
```

```{r output final column chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Column Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle),
                  output_file = paste0(file_names$column_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,'update/webpage-html-outputs'))

# Column Chart - png for use in static file
# need to create a 'static-images' sub-folder in theme sub-folder if not already existing
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$column_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$column_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r column chart output - pums, echo=FALSE}

if (output_type == "html") {

  column_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$column_chart_name, '.png')))
  
}

# column_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r column facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean_column %>% 
  filter(focus_type=="Disability_cat")
# Region: 0.21 vs. 0.06
# with disability has higher coverage across all counties/region

filter <- data_clean_column %>% 
  filter(focus_type=="LEP_cat")
# King lep households: .22	
# King non-lep households: .07

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="Income_cat")
# region low income households: .32	
# region non-low income households: .04
# .32-.04=.28

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="LEP_cat")
# region non-lep:0.09
# region lep: 0.21
# 0.21/0.09=1.857143

filter <- data_clean_column %>% 
  filter(county=="Region",
         focus_type=="POC_cat")
# region non-poc:0.07
# region poc: 0.13
# 0.13/0.07=2.3333
```
\

#### Insights & Analysis

* The share of kindergarten ready students from households with lower income in the region is 37%, 24 percentage points lower than other households.
* The share of kindergarten ready students in the region for those with a disability is 59% lower than those without.
* In 2022, fewer students of color were kindergarten ready than white non-Hispanic students, a difference of 11 percentage points.
* The smallest percentage point difference in kindergarten readiness between students of color and white non-Hispanic students is in Kitsap County (3%), while the largest difference is in King County (16%); Pierce (11%) and Snohomish (9%) fall in the middle.
* For students in households with lower income, kindergarten readiness is lowest in Snohomish County (32%), followed by Pierce (37%), King (38%), and Kitsap (48%).
* The largest percentage point difference in kindergarten readiness between students with a disability and those without is in King County (37%), while the smallest difference is in Kitsap and Pierce counties (28%).


\
\


## 3. Facet of trend data
### Create Visual
<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>

```{r line facet - pums, include=FALSE}
# Create Facet Line Chart

# set variables
df = data_clean
x = "data_year" # different from variables for facet column
fill = "focus_attribute_ord" # different from variables for facet column
color = "blues"
title = "Kindergarten Readiness Trend by Community"
subtitle = "Percent of students demonstrating readiness in 6 of 6 domains*, in 1-year spans between 2010 and 2022"
source = "Washington Office of Superintendent of Public Instruction, Report Card WaKids 2010-23 School Year, Washington Kindergarten Inventory of Developing Skills <br> Limited data available for students in Kitsap County <br> No data available for school year 2020-21 <br> *Social-Emotional, Physical, Language, Cognitive, Literacy, and Math"

line_chart <- equity_tracker_line_facet(df = df,
                                        geo = geo,
                                        x = x,
                                        y = y,
                                        fill = fill,
                                        facet = facet,
                                        y_min = y_min,
                                        y_max = y_max,
                                        dec = dec,
                                        esttype = esttype,
                                        color = color,
                                        num_colors = num_colors,
                                        color_rev = color_rev,
                                        width = '420px',
                                        height = '380px')
```

```{r output final line chart - pums, include=FALSE}
# to save html visual to folder, there must be an 'webpage-html-outputs' folder within each indicator sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\f01-median-income\webpage-html-outputs
# to save png visuals to folder, there must be a 'static-images' folder within each thematic sub-folder: Y:\Equity Indicators\tracker-webpage-content\f-economy\static-images

# Line Chart - html for interactive
rmarkdown::render(here("data-visualization/equity-tracker-chart-creator.RMD"), 
                  params = list(df = df,
                                geo = geo,
                                x = x,
                                y = y,
                                fill = fill,
                                facet = facet,
                                y_min = y_min,
                                y_max = y_max,
                                dec = dec,
                                esttype = esttype,
                                color = color,
                                num_colors = num_colors,
                                color_rev = color_rev,
                                title = title,
                                # source = source,
                                subtitle = subtitle,
                                width = '420px',
                                height = '380px',
                                charttype = "line"),
                  output_file = paste0(file_names$line_chart_name, '.html'),
                  output_dir = file.path(file_names$base_dir,
                                         file_names$theme_dir,
                                         file_names$ind_dir,
                                         'update/webpage-html-outputs'))

# Line Chart - png for use in static file
webshot2::webshot(url = file.path(file_names$base_dir,
                                  file_names$theme_dir,
                                  file_names$ind_dir,
                                  'update/webpage-html-outputs', 
                                  paste0(file_names$line_chart_name, '.html')), 
                  file = file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   'static-images', 
                                   paste0(file_names$line_chart_name, '.png')), 
                  vwidth = 1920, vheight = 1080)
```

<p style="font-family: Poppins; font-size:14pt; font-weight:bold">
`r title`
</p>
<p style="font-family: Poppins; font-size:12pt; font-weight:normal">
`r subtitle`
</p>
```{r line chart - pums, echo=FALSE}
if (output_type == "html") {

  line_chart  
  
} else {
  
  knitr::include_graphics(file.path(file_names$base_dir,
                                    file_names$theme_dir,
                                    'static-images', 
                                    paste0(file_names$line_chart_name, '.png')))
  
}
# line_chart
```
<p style="font-family: Poppins; font-size:10pt; font-style:italic">
`r source`
</p> 

#### Data call outs
```{r line facet calculations - pums, include=FALSE, eval=FALSE}
# calculate call out information
filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# king 2012: 0.17-0.07=0.10
# king 2022: 0.12-0.06=0.06
# kitsap 2012: 0.19-0.10=0.09
# kitsap 2022:0.14-0.08=0.06
# pierce 2012: 0.19-0.10=0.09
# pierce 2022:0.17-0.10=0.07
# snohomish 2012: 0.16-0.10=0.06
# snohomish 2022: 0.11-0.08=0.03

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="LEP_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# king 2022-2012: 0.22-0.24=-0.02
# kitsap 2022-2012: 0.13-0.18=-0.05
# pierce 2022-2012: 0.23-0.21=0.02
# snohomish 2022-2012: 0.18-0.26=-0.08

filter <- data_clean %>%
  filter(county!="Region",
         focus_type=="Disability_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# King 2022-2012: 0.21-0.22=-0.01
# Kitsap 2022-2012: 0.20-0.20=0
# Pierce 2022-2012: 0.22-0.22=0
# Snohomish: 0.19-0.23=-0.04

filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year!="2017")
# King 2012 poc-non poc: 0.17-0.07=0.10
# King 2022: 0.12-0.06=0.06
# King: ((0.06-0.10)/(0.10))*100: -40% change
# Kitsap 2012: 0.19-0.10=0.09
# Kitsap 2022: 0.14-0.08=0.06
# Kitsap: ((0.06-0.09)/(0.09))*100: -33% change
# Pierce 2012:0.19-0.10=0.09
# Pierce 2022: 0.17-0.10=0.07
# Pierce: ((0.07-0.09)/(0.09))*100: -22% change
# Snohomish 2012: 0.16-0.10=0.06
# Snohomish 2022: 0.11-0.08=0.03 
# Snohomish: ((0.03-0.06)/(0.06))*100: -50% change


filter <- data_clean %>%
  filter(county == "Region",
         focus_type=="Income_cat",
         vulnerability=="high vulnerability",
         data_year!="2017")
# 2012: 0.35
# 2022: 0.32


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="POC_cat",
         data_year=="2016" |
           data_year=="2021")


filter <- data_clean %>%
  filter(county=="Snohomish",
         focus_type=="Income_cat",
         vulnerability=="low vulnerability",
         data_year=="2016" |
           data_year=="2021")

```
\

#### Insights & Analysis

* The share of students prepared for kindergarten decreased over the past decade for all groups in Pierce and Snohomish counties, and increased for all groups in King County.
* Students of color in the region are less likely to be prepared for kindergarten compared to white non-Hispanic students, a gap of 10 percentage points and a decrease of almost half (48%) over the past decade.
* The gap in kindergarten readiness in the region between students from households with lower income and those from other households has remained relatively consistent in the last decade, decreasing from 27 percentage points to 24 percentage points in 2022.
* In King County, the difference in kindergarten readiness between students with a disability and those without almost doubled over the past decade (+95%), while the difference decreased in Snohomish County (-9%) and in Pierce County (-10%). Kitsap County didn’t report data for students with disabilities until 2015 and in the past half decade, the difference has decreased slightly (-7%).


\
\


# Transfer files
## Copy files from Github > Y drive/update folder
This step will transfer all of the Rmd output files (html and docx) to the network for review. It will keep the Rmd files within GitHub so that code is kept in a central place. 
```{r to y drive, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(getwd())
y.drive.folder.update <- file.path(file_names$base_dir,
                                   file_names$theme_dir,
                                   file_names$ind_dir,
                                   'update')

# isolate files in Github
list.of.files <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$")

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder.update,
          overwrite = TRUE,
          recursive = TRUE)

# keep only .Rmd files in Github (cleaner)
files.in.dir <- list.files(current_folder, 
                            full.names = T)
files.to.keep <- list.files(current_folder, 
                            full.names = T,
                            pattern = "\\.Rmd$")
files.to.remove <- list(files.in.dir[!(files.in.dir %in% grep(paste(files.to.keep, collapse = "|"),
                                                              files.in.dir, value=TRUE))])
do.call(unlink, files.to.remove)
```

## Copy files from Y drive/indicator folder > Y drive/indicator/archive folder
This step will transfer the previous data and files to the archive folder. This step is meant to retain the older versions in case they are needed for reference. 
```{r to archive, include=FALSE, eval=FALSE}
# set folder structure
y.drive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir)
archive.folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,'archive')

# isolate old files/directories
list.of.files <- list.files(y.drive.folder, 
                            full.names = T,
                            pattern = "\\.(html|docx)$|rda-data|webpage-html-outputs")

# copy the files to the archive folder
file.copy(from=list.of.files,
          to=archive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

### Delete old files from indicator folder
This step will clean the indicator folder to make room for the new versions.
```{r clean indicator folder, include=FALSE, eval=FALSE}
# identify html and docx files to remove
list.of.files <- list.files(y.drive.folder,  
                            full.names = T,
                            pattern = "\\.(html|docx)$",
                            recursive=F)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# # identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
# map.files.dir <- list.dirs(file.path(y.drive.folder, 'webpage-html-outputs'),
#                            full.names = T,
#                            recursive=F)
# 
# # clean the files from the update folder
# dir.to.remove <- list(map.files.dir)
# unlink(dir.to.remove, recursive=T)
```

## Copy new files from Y drive/update folder > Y drive/indicator folder
Once the data and planning reviews are complete, this step will move all of the updated data/files to the general indicator folder. They should be moved from the update (draft staging directory) to the parent folder so that the htmls can be copied to the webpage folder (outside the firewall).
```{r update to indicator folder, include=FALSE, eval=FALSE}
# isolate old files/directories
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T)

# copy the files to the new folder
file.copy(from=list.of.files,
          to=y.drive.folder,
          overwrite = TRUE,
          recursive = TRUE)
```

## Clear Y drive/update folder
This step will help keep the folders organized and ready for the next update. 
```{r clear update, include=FALSE, eval=FALSE}
# clean the map files from the indicator/webpage-html-outputs folders
# identify files/data to remove
list.of.files <- list.files(y.drive.folder.update, 
                            full.names = T,
                            pattern = "\\.(html|docx|rda)$",
                            recursive=T)
# clean the files from the update folder
files.to.remove <- list(list.of.files)
do.call(unlink, files.to.remove)

# the webpage-html-output and rda files will get overwritten, but the map_files folder is still in the webpage-html-outputs folder

# identify map_files sub-directory, wanted to keep the rda-data and webpage-html-outputs folders
map.files.dir <- list.dirs(file.path(y.drive.folder.update, 'webpage-html-outputs'),
                           full.names = T,
                           recursive=F)

# clean the files from the update folder
dir.to.remove <- list(map.files.dir)
unlink(dir.to.remove, recursive=T)
```

## Copy files from Y drive/indicator folder > website folder
This step copies the htmls for the webpage (3 visuals) from the network to the folder outside the firewall- this 'external' folder connects directly to the webpage.
```{r to web dev folder, include=FALSE, eval=FALSE}
# set folder structure
current_folder <- file.path(file_names$base_dir,
                            file_names$theme_dir,
                            file_names$ind_dir,
                            'webpage-html-outputs')

web.base.folder <- '//WEB/website_data/equity-tracker-webpages'
new.folder <- file.path(web.base.folder,
                        file_names$theme_dir)

list.of.files <- list.files(current_folder, 
                            full.names = T)
# list.of.files

# copy the files to the new folder
file.copy(list.of.files,
          new.folder,
          overwrite = T)
```

<a href="#top">Back to top of the page</a>